### 2012-12-06 mcmc.tex
initial commit

\documentclass[12pt,twoside,pdftex]{article}

\usepackage{styles/dar_endnotes}

\newcommand{\this}{Introduction to Markov chain Monte Carlo}
\include{styles/dar}

\begin{document}

\thispagestyle{plain}\raggedbottom
\section*{Data analysis recipes:\ \this\footnotemark}

\footnotetext{%
    The notes begin on page~\pageref{note:first}, including the
    license\note{\label{note:first}
        Copyright 2010 by the authors. This work is licensed under a
        \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
            Creative Commons Attribution-NonCommercial-NoDerivs 3.0 Unported
            License}.}
    and the acknowledgements\note{%
        We would like to thank\ldots
    }.
}

\noindent
Dan~Foreman-Mackey\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,%
       New York University}
\\[1ex]
David~W.~Hogg\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,%
       New York University}\\
\affil{Max-Planck-Institut f\"ur Astronomie, Heidelberg}

\begin{abstract}
    WTF is MCMC?
\end{abstract}

\section{Statistical error bars}

% \begin{figure}[htbp]
% \exampleplot{ex17}
% \caption{Partial solution to \problemname~\ref{prob:bayesintrinsic}:
% The marginalized posterior probability distribution for the intrinsic
% variance.}\label{fig:bayesintrinsic}
% \end{figure}

\clearpage
\markright{Notes}\theendnotes

\clearpage
\begin{thebibliography}{}\markright{References}

\bibitem[Bovy, Hogg, \& Roweis(2009)]{bovy}
  Bovy,~J., Hogg,~D.~W., \& Roweis, S.~T., 2009,
  Extreme deconvolution: inferring complete distribution functions from noisy, heterogeneous, and incomplete observations,
  \arxiv{0905.2979} [stat.ME]

\bibitem[Foreman-Mackey et al.(2012)]{emcee} Foreman-Mackey,
    D., Hogg, D.~W., Lang, D., \& Goodman, J.\ 2012, \arxiv{1202.3665}

\end{thebibliography}

\end{document}

### 2013-01-26 mcmc.tex
title and acks

@@
 
-\newcommand{\this}{Introduction to Markov chain Monte Carlo}
+\newcommand{\this}{Using Markov Chain Monte Carlo}
 \include{styles/dar}
@@
 \thispagestyle{plain}\raggedbottom
-\section*{Data analysis recipes:\ \this\footnotemark}
+\section*{Data analysis recipes:\\ \this\footnotemark}
 
@@
     and the acknowledgements\note{%
-        We would like to thank\ldots
-    }.
+        We would like to thank
+          Jo Bovy (IAS),
+          Brendon Brewer (Auckland),
+          Jonathan Goodman (NYU),
+          Dustin Lang (CMU),
+          \ldots
+        for valuable advice and comments.}.
 }

### 2013-01-26 mcmc.tex
outline and references

@@
 
-\section{Statistical error bars}
+\section{When do you need MCMC?}
+
+...Search, optimization, and sampling in inference.
+
+\section{What is MCMC?}
+
+\section{Metropolis--Hastings, stretch move, and more}
+
+...M-H, emcee, and references to competitor methods like nested and hamiltionian etc.
+
+\section{Results, error bars, and figures}
+
+...using the output to make figures, get numbers with error bars, and show results
+
+\section{Tuning}
+
+...tuning and Gibbs sampling and the like
+
+\section{Convergence}
+
+...how do you know you have run long enough?
+
+\section{Initialization and burn-in}
+
+\section{Advice and discussion}
+
+...Don'ts and no-nos and breaking Markov-ness.
 
@@
 \begin{thebibliography}{}\markright{References}
-
-\bibitem[Bovy, Hogg, \& Roweis(2009)]{bovy}
-  Bovy,~J., Hogg,~D.~W., \& Roweis, S.~T., 2009,
-  Extreme deconvolution: inferring complete distribution functions from noisy, heterogeneous, and incomplete observations,
-  \arxiv{0905.2979} [stat.ME]
-
-\bibitem[Foreman-Mackey et al.(2012)]{emcee} Foreman-Mackey,
-    D., Hogg, D.~W., Lang, D., \& Goodman, J.\ 2012, \arxiv{1202.3665}
-
+\bibitem[Foreman-Mackey et al.(2012)]{emcee}
+  Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2012,
+  \project{emcee}: The MCMC Hammer,
+  \arxiv{1202.3665}
 \end{thebibliography}

### 2013-01-26 mcmc.tex
typography
more outline

@@
 \documentclass[12pt,twoside,pdftex]{article}
-
 \usepackage{styles/dar_endnotes}
@@
           Jonathan Goodman (NYU),
+          Fengji Hou (NYU),
           Dustin Lang (CMU),
@@
 
-\section{What is MCMC?}
+\section{What is M--H MCMC?}
 
-\section{Metropolis--Hastings, stretch move, and more}
+...The algorithm
 
-...M-H, emcee, and references to competitor methods like nested and hamiltionian etc.
+...Why does it work (briefly)?
+
+...What is a typical proposal distribution?
+
+\section{stretch move and other variants}
+
+...How does the stretch move work?
+
+...Why is the stretch move useful?
+
+...What, briefly, is nested sampling, and why is it good?
+
+...What, briefly, is hamiltonian, and why is it good?
 
@@
 
-...using the output to make figures, get numbers with error bars, and show results
+...So you have a chain output, what do you do with it?
+
+...What is the number and error bar?  What is an upper limit?
+
+...What figures should you make to check the chain?
+
+\section{Convergence}
+
+...How do you know you have run long enough?  Heuristically...
+
+...Autocorrelation time
 
@@
 
-...tuning and Gibbs sampling and the like
+...How do you make it run fast?
 
-\section{Convergence}
+...Acceptance ratio and ESJD and etc.
 
-...how do you know you have run long enough?
+...Gibbs sampling and interior marginalizations
 
@@
 
+...Precede with optimization
+
+...Start from different locations
+
+...Dealing with multi-modal problems
+
 \section{Advice and discussion}
 
-...Don'ts and no-nos and breaking Markov-ness.
+...Write and debug your own M--H sampler and then download a real
+package \emph{afterwards}.
+
+...breaking Markov-ness.
+
+...Initialization and burn-in advice we haven't addressed earlier?
+
+...The Bayes integral / evidence integral?
+
+...Posterior predictive checks?
 

### 2013-01-26 mcmc.tex
abstract; and some outline tweaks

@@
 \begin{abstract}
-    WTF is MCMC?
+Markov Chain Monte Carlo (MCMC) methods for sampling probability
+distribution functions, plus abundant computation, have transformed
+the sciences.  Here we give a fast overview of basic MCMC methods and
+then turn to practical advice for their use in real inference
+problems.  We give advice on method choice, tuning for performance,
+initialization and burn-in, judging convergence, and use of the chain
+output to make results with error bars. [Insert some generally useful
+  point here!]
 \end{abstract}
@@
 
-...Search, optimization, and sampling in inference.
+...Search, optimization, and sampling in inference.  MCMC does only
+one of these things well!
 
@@
 ...What is a typical proposal distribution?
+
+...You can't sample the likelihood!  You can only sample a posterior
+pdf!
 
@@
 
+...Finally, to repeat: You can't sample the likelihood!  You can only
+sample a posterior pdf!
+
 % \begin{figure}[htbp]

### 2013-01-26 mcmc.tex
tweaks

@@
     license\note{\label{note:first}
-        Copyright 2010 by the authors. This work is licensed under a
+        Copyright 2013 by the authors. This work is licensed under a
         \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
@@
           Jonathan Goodman (NYU),
-          Fengji Hou (NYU),
-          Dustin Lang (CMU),
-          \ldots
+          Fengji Hou (NYU), and
+          Dustin Lang (CMU)
         for valuable advice and comments.}.

### 2013-01-28 mcmc.tex
introduction section

@@
+% to-do
+% -----
+% - write zeroth draft
+% - write first draft
+% - send to friendlies for comments
+%   - Brewer
+%   - Lang
+%   - Rix
+%   - Bovy
+% - post on arXiv
+
+% style notes
+% -----------
+% - [LaTeX] newline after every full stop for proper git diffing
+% - [LaTeX] eqnarray not equation
+% - pdf not PDF
+
 \documentclass[12pt,twoside,pdftex]{article}
@@
 \include{styles/dar}
+
+\newcommand{\data}{D}
+\newcommand{\pars}{\theta}
 
@@
 distribution functions, plus abundant computation, have transformed
-the sciences.  Here we give a fast overview of basic MCMC methods and
-then turn to practical advice for their use in real inference
-problems.  We give advice on method choice, tuning for performance,
+the sciences.
+Here we give a fast overview of basic MCMC methods and then turn to
+practical advice for their use in real inference problems.
+We give advice on method choice, tuning for performance,
 initialization and burn-in, judging convergence, and use of the chain
-output to make results with error bars. [Insert some generally useful
-  point here!]
+output to make results with error bars.
+[Insert some generally useful point here!]
 \end{abstract}
@@
 
-...Search, optimization, and sampling in inference.  MCMC does only
-one of these things well!
+Markov Chain Monte Carlo (MCMC) methods are methods for sampling
+probability distribution functions (pdfs).
+They don't require that you have a full analytic description of the
+properly normalized pdf for sampling to proceed; they only require
+that you be able to compute ratios of the pdf at pairs of locations.
+These properties make MCMC methods ideal for sampling \emph{posterior
+  pdfs} in probabilistic inferences:
+In probabilistic inferences, the posterior pdf $p(\pars\given\data)$,
+or pdf for the parameters $\pars$ given the data $\data$, is
+constructed from the likelihood $p(\data\given\pars)$, or pdf for the
+data given the parameters, and the prior pdf $p(\pars)$ for the
+parameters by what's often known as ``Bayes rule'',
+\begin{eqnarray}
+p(\pars\given\data) &=& \frac{1}{Z}\,p(\data\given\pars)\,p(\pars)
+\quad .
+\end{eqnarray}
+In these contexts, the constant $Z$ is sometimes written as
+$p(\data)$, is sometimes called the ``evidence'' or the ``Bayes
+integral'', and is usually \emph{extremely hard to
+  calculate}.\note{Say words about why we call it $Z$ not $p(\data)$
+  and why it is extremely hard to calculate.}
+That is, you often know the function $p(\pars\given\data)$ up to a
+constant factor; you can compute ratios of the pdf at pairs of points,
+but not the precise value at any individual point.
+
+In addition to this normalization-insensitive property of MCMC, in its
+simplest forms it can be run without computing any derivatives or
+integrals of the function, and (as we will show below) in its simplest
+forms it is \emph{extremely easy to implement}.
+For all these reasons, MCMC is ideal for sampling posterior pdfs in
+the real situations in which scientists find themselves.
+
+Say you are in this situation.
+You have a huge blob of data $\data$ (think of this as a vector or
+list or heterogeneous collection of observations\note{In general in
+  this \documentname\ we will treat any variable as being capable of
+  containing not just a value but some huge arbitrarily structured set
+  of values in matrix, vector, or more complex form.}).
+You also have a model sophisticated enough---a probabilistic,
+generative model, if you will\note{For our definition of a
+  \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
+setting of a huge blob of parameters $\pars$\note{Again, the parameter
+  blob $\pars$ should also be thought of as an immense, arbitrarily
+  structured set or list of parameters, possibly even with no fixed
+  dimension!}, you can compute a pdf for data (or likelihood\note{We
+  bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
+  series.}) $p(\data\given\pars)$.
+And say you also can write down some kind of informative or
+uninformative prior pdf $p(\pars)$ for the parameter blob.
+If all these things are true, then---even if you can't compute
+anything else---in principle a trivial-to-implement MCMC can give you
+a fair sampling of the posterior pdf.
+That is, you can run MCMC (for a very long time\note{A very, very long
+  time; we will return to that below.}) and you will be left with a
+set of $K$ parameter-blob settings $\pars_k$ such that the full set
+$\setofall{\pars_k}_{k=1}^K$ constitutes a fair sampling from the
+posterior pdf $p(\pars\given\data$).
+
+All that said, and adhering to the traditions of the \project{Data
+  Analysis Recipes} project\note{Every chapter in the \project{Data
+    Analysis Recipes} series begins with a rant in which we argue that
+  most uses of the methods in question are not appropriate!}, we are
+compelled to note that MCMC is in fact \emph{over-used}.
+Because MCMC provably (under assumptions\note{Hint at assumptions
+  here.}, some of which will be discussed below) samples the full
+posterior pdf in all of parameter space, many investigators use MCMC
+because it will sample \emph{all} of the parameter space
+($\pars$-space).
+That is, they are using MCMC because they want to
+\emph{search the parameter space for good models}.
+Because MCMC samples the parameter representatively, it spends most of
+its time near very good models; models that (within the confines of
+the prior pdf) do a good job of explaining the data.
+For this reason, many investigators are using MCMC because it
+effectively \emph{optimizes the posterior pdf}.
+
+Both of these reasons for using MCMC---that it is a parameter-space
+search algorithm, and that it is a simple-to-code effective
+optimizer---are \emph{not good reasons}.  MCMC is first and foremost a
+\emph{sampler}.  If you are trying to find the optimum of the
+likelihood or the posterior pdf, you should use an \emph{optimzer},
+not a sampler.  If you want to make sure you search all of parameter
+space, you should use a \emph{search algorithm}, not a sampler.  MCMC
+is good at one thing, and one thing only: Sampling ill-normalized
+pdfs.
 

### 2013-01-28 mcmc.tex
Small comments on abstract

@@
 Markov Chain Monte Carlo (MCMC) methods for sampling probability
-distribution functions, plus abundant computation, have transformed
+distribution functions, plus abundant computational resources,
+have transformed
 the sciences.
@@
 initialization and burn-in, judging convergence, and use of the chain
-output to make results with error bars.
+output to make results {\it with error bars}.
+% Giving MCMC results with error bars is uncommon, but cool. I thought it
+% deserved an italic
 [Insert some generally useful point here!]
@@
 probability distribution functions (pdfs).
+% pdf usually stands for probability density function. You are using non-standard
+% terminology here.
 They don't require that you have a full analytic description of the

### 2013-01-28 mcmc.tex
More comments, prefaced by my initials
Names for evidence

@@
 output to make results {\it with error bars}.
+% ******* BJB
 % Giving MCMC results with error bars is uncommon, but cool. I thought it
@@
 Markov Chain Monte Carlo (MCMC) methods are methods for sampling
-probability distribution functions (pdfs).
+probability distribution functions (pdfs). These pdfs may be either probability
+mass functions on a discrete space, or probability densities on a continuous
+space.
+% ******* BJB
 % pdf usually stands for probability density function. You are using non-standard
-% terminology here.
+% terminology here. My words above might be too technical given the aim of your
+% document.
 They don't require that you have a full analytic description of the
@@
   pdfs} in probabilistic inferences:
+% ******** BJB
+% I know you don't like it, but LOTS of people use and understand the word
+% Bayesian.
 In probabilistic inferences, the posterior pdf $p(\pars\given\data)$,
@@
 \end{eqnarray}
-In these contexts, the constant $Z$ is sometimes written as
-$p(\data)$, is sometimes called the ``evidence'' or the ``Bayes
-integral'', and is usually \emph{extremely hard to
+% ********* BJB
+% Unfortunately there are many more names for the evidence. I prefer
+% marginal likelihood myself.
+In these contexts, the constant $Z$, sometimes written as
+$p(\data)$, is known by the names ``evidence'', ``marginal likelihood''
+the ``Bayes integral'' and ``prior predictive probability'', and is usually
+\emph{extremely hard to
   calculate}.\note{Say words about why we call it $Z$ not $p(\data)$

### 2013-01-28 mcmc.tex
Rant
More ranty comments

@@
   series.}) $p(\data\given\pars)$.
+% ************* BJB
+% <Wild Subjective Bayesian>
+% The "generative model" is a probabilistic description of your *prior
+% beliefs about what the data are going to be*. </Wild Subjective Bayesian>
 And say you also can write down some kind of informative or
-uninformative prior pdf $p(\pars)$ for the parameter blob.
+vague prior pdf $p(\pars)$ for the parameter blob.
+% ************* BJB
+% Uninformative should be banished! The correct word is vague :). Attempts
+% at finding formal uninformative priors are misguided IMHO.
 If all these things are true, then---even if you can't compute
@@
 a fair sampling of the posterior pdf.
-That is, you can run MCMC (for a very long time\note{A very, very long
-  time; we will return to that below.}) and you will be left with a
+That is, you can run MCMC (for a very long time\note{Sometimes a very, very long
+  time; we will return to that below.})
+% ************* BJB
+% It's not always super slow, so added the word "sometimes"
+and you will be left with a
 set of $K$ parameter-blob settings $\pars_k$ such that the full set

### 2013-01-31 mcmc.tex
More comments
Done for now
revising Brewer edits

@@
 % - send to friendlies for comments
-%   - Brewer
 %   - Lang
@@
 initialization and burn-in, judging convergence, and use of the chain
-output to make results {\it with error bars}.
-% ******* BJB
-% Giving MCMC results with error bars is uncommon, but cool. I thought it
-% deserved an italic
+output to produce parameter estimates with associated uncertainties.
 [Insert some generally useful point here!]
@@
 Markov Chain Monte Carlo (MCMC) methods are methods for sampling
-probability distribution functions (pdfs). These pdfs may be either probability
-mass functions on a discrete space, or probability densities on a continuous
-space.
-% ******* BJB
-% pdf usually stands for probability density function. You are using non-standard
-% terminology here. My words above might be too technical given the aim of your
-% document.
-They don't require that you have a full analytic description of the
+probability distribution functions or probability density functions (pdfs).
+These pdfs may be either probability mass functions on a discrete
+space, or probability densities on a continuous space, though we will
+concentrate on the latter in this \documentname.
+MCMC methods don't require that you have a full analytic description of the
 properly normalized pdf for sampling to proceed; they only require
 that you be able to compute ratios of the pdf at pairs of locations.
-These properties make MCMC methods ideal for sampling \emph{posterior
+This makes MCMC methods ideal for sampling \emph{posterior
   pdfs} in probabilistic inferences:
-% ******** BJB
-% I know you don't like it, but LOTS of people use and understand the word
-% Bayesian.
-In probabilistic inferences, the posterior pdf $p(\pars\given\data)$,
+
+In a probabilistic inference, the posterior pdf $p(\pars\given\data)$,
 or pdf for the parameters $\pars$ given the data $\data$, is
@@
 \end{eqnarray}
-% ********* BJB
-% Unfortunately there are many more names for the evidence. I prefer
-% marginal likelihood myself.
 In these contexts, the constant $Z$, sometimes written as
@@
 vague prior pdf $p(\pars)$ for the parameter blob.
-% ************* BJB
-% Uninformative should be banished! The correct word is vague :). Attempts
-% at finding formal uninformative priors are misguided IMHO.
 If all these things are true, then---even if you can't compute
@@
 a fair sampling of the posterior pdf.
-That is, you can run MCMC (for a very long time\note{Sometimes a very, very long
-  time; we will return to that below.})
+That is, you can run MCMC (for a very long time\note{Sometimes a very,
+  very long time; we will return to that below.})  and you will be
+left with a set of $K$ parameter-blob settings $\pars_k$ such that the
+full set $\setofall{\pars_k}_{k=1}^K$ constitutes a fair sampling from
+the posterior pdf $p(\pars\given\data$).
 % ************* BJB
-% It's not always super slow, so added the word "sometimes"
-and you will be left with a
-set of $K$ parameter-blob settings $\pars_k$ such that the full set
-$\setofall{\pars_k}_{k=1}^K$ constitutes a fair sampling from the
-posterior pdf $p(\pars\given\data$).
+% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
+% refer the interested reader to the paper "Monte Carlo is Fundamentally
+% Unsound" by Tony O'Hagan.
 
@@
 For this reason, many investigators are using MCMC because it
-effectively \emph{optimizes the posterior pdf}.
+effectively \emph{optimizes the posterior pdf}, or, for certain
+choices of prior pdf, \emph{optimizes the likelihood}.\note{Of course
+  a committed Bayesian would argue that any time you are optimizing a
+  posterior pdf or optimizing a likelihood, you \emph{should} be
+  sampling a posterior pdf.  That is, for some, the fact that when
+  someone ``wants'' to optimize, it is actually useful that they
+  choose the ``wrong'' tool, because that ``wrong'' tool gives them
+  back something far more useful than the output of any optimizer!}
 
@@
 space, you should use a \emph{search algorithm}, not a sampler.  MCMC
-is good at one thing, and one thing only: Sampling ill-normalized
-pdfs.
+is good at one thing, and one thing only: Sampling ill-normalized (or
+otherwise hard to sample) pdfs.
 
@@
 pdf!
+% *********** BJB
+% YES!!!! Although you can sample the likelihood if your samples are in
+% data space. :)
+
+% *********** BJB
+% I suggest partitioning the discussion of different methods into the following
+% two categories:
+%
+%  i) Methods for getting around the target distribution
+% ii) Methods that change the target distribution to something
+% other than the posterior pdf, to help the sampling somehow
+%
+% i) Would include things like M-H, the stretch move,
+% Hamiltonian, slice sampling (maybe skip the latter)
+% ii) would include things like annealing and Nested Sampling
 
@@
 
-...What is the number and error bar?  What is an upper limit?
+...What should you report as the inferred parameter value and
+associated uncertainty (error bar)?  What is an upper limit?
 
@@
 
+% BJB
+% Please suggest mixtures of scales for M-H proposals.
+% You can mimic the properties of slice sampling but with
+% much easier code.
+
 ...Gibbs sampling and interior marginalizations
@@
 ...Precede with optimization
+% BJB
+% If applicable. Bad idea in high dimension/underconstrained
+% problem.
 
 ...Start from different locations
+% BJB
+% What different locations?
+% Drawn from the prior can be useful. Then you can
+% look at the chains and see which ones "made it" and
+% which ones "got stuck" and delete the latter
 
 ...Dealing with multi-modal problems
+% Starting from different locations is good
+% for DIAGNOSING a problem as multimodal but it's
+% terrible for quantifying the modes. Explain why,
+% it's a good teaching point.
 
@@
 package \emph{afterwards}.
+% ***** BJB
+% I AGREE WHOLEHEARTEDLY. That's how to learn.
 
 ...breaking Markov-ness.
+% ***** BJB
+% "Diminishing adaptations" is an idea that is quite easy
+% to explain. There are other ways to break Markov-ness but
+% this probably covers enough of them for an intro document
+% ***** Hogg
+% I was going to say you should NEVER break Markov-ness
+% unless you REALLY understand what you are doing.
 

### 2013-01-31 mcmc.tex
small tweaks and notes to self

@@
 Dan~Foreman-Mackey\\
-\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,%
-       New York University}
-\\[1ex]
+\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
+       New York University}\\[1ex]
 David~W.~Hogg\\
-\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,%
+\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
        New York University}\\
@@
 distribution functions, plus abundant computational resources,
-have transformed
-the sciences.
+have transformed the sciences.
 Here we give a fast overview of basic MCMC methods and then turn to
@@
   \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
-setting of a huge blob of parameters $\pars$\note{Again, the parameter
+setting of a huge blob of parameters\note{Again, the parameter
   blob $\pars$ should also be thought of as an immense, arbitrarily
   structured set or list of parameters, possibly even with no fixed
-  dimension!}, you can compute a pdf for data (or likelihood\note{We
+  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We
   bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
@@
 % beliefs about what the data are going to be*. </Wild Subjective Bayesian>
+% ************* Hogg
+% That comment should go in the ``What is a Model?'' Chapter.
 And say you also can write down some kind of informative or

### 2013-01-31 mcmc.tex
are you converged?

@@
 
-...What figures should you make to check the chain?
+...What do you do in high-dimensional spaces?  Note the issue
+that---even in a unimodal posterior PDF---the most probable model
+might not be at the set of parameter values that are reported if you
+compute the parameter values to report in one-dimensional projections
+of the posterior pdf.  This requires a figure to demonstrate.  What to
+do here?
+
+...What figures should you make to display the chain?
+
+...If you are really good, you will pass forward a sampling.  How big
+should it be?  Refer to Mackay here.
 
@@
 
-...How do you know you have run long enough?  Heuristically...
+A key question for an MCMC operator---\emph{the} key question in some
+sense---is \emph{how long to run} to be sure of having reliable
+results.
+It is disappointing and annoying to many that there is no extremely
+simple and reliable answer to this question.
 
-...Autocorrelation time
+The reason that there is no simple answer is that you can't really
+ever know that you have sampled the full posterior pdf, and the reason
+for \emph{that} is that if you \emph{could} know that, you would also
+be able to solve the famously difficult discrete optimization
+problem.\note{explain this remark and cite something here}
+Another way to put it is that you might have two modes---two separated
+regions of substantial total probability---in the posterior pdf that
+are both important but separated by a large region of low probability.
+If you are using a simple sampler, you could sample one of these modes
+very well but the sampler will take effectively infinite time to find
+the other mode.
+Any test of convergence would be passed, but in fact the sampling
+would be incomplete and not representative.
+That is, knowing that you have a complete and representative sampling
+is nearly impossible.
+
+Heuristically, however, you have sampled long enough when you can see
+that the (or each) walker has traversed the high-probability parts of
+the parameter space many times in the length of the chain.
+Or, equivalently, you have sampled long enough when the first half of
+the chain shows very much the same posterior pdf morphology as the
+second half of the chain, or indeed as any substantial subset.
+
+...What plot should you make?  Make plots that demonstrate a non-converged and a converged chain.
+
+...Autocorrelation time is the premier tool.  It is not a panacea, and why?
 

### 2013-02-06 mcmc.tex
start of results reporting

@@
 % - pdf not PDF
+% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
 
@@
 % That comment should go in the ``What is a Model?'' Chapter.
-And say you also can write down some kind of informative or
+Furthermore, say you also can write down some kind of informative or
 vague prior pdf $p(\pars)$ for the parameter blob.
@@
 
-...So you have a chain output, what do you do with it?
+For a committed probabilistic scientist---frequentist or
+bayesian---there is no ``answer'', there are only probability density
+functions.  Any time in a paper or in an email or in a conversation
+we say what ``the answer is''---even if we say it with an error
+bar---we have departed the probabilistic program.  There are many
+principled ways to depart the program; in another \documentname, we
+discuss the economic explanation of how we can make hard decisions in
+the context of probabilistic reasoning\note{CITE DECISION DOCUMENT}.
+But without going into the economic model, it is fair to say that a
+substantial problem with being a committed probabilist is that when we
+\emph{publish}, we are not permitted (not now, at least) to publish a
+\emph{probability distribution over pubications}; we have to publish
+\emph{one single, deterministic text}; we have to \emph{decide} what
+to write in the title, the abstract, the tables, figures, and results
+section.  Given a MCMC sampling of the posterior pdf, what do we
+report as our \emph{results}?
 

### 2013-02-11 mcmc.tex
notes to self
outline notes

@@
 
+\section{What is a sampling?}
+
+...What is a sampling, heurstically?..
+
+...What is the expectation integral in the sampling approximation?  Why is that a good idea?..
+
+...How is the sampling of the marginalized pdf trivially constructed from the full sampling?..
+
+...Some guidance of where this comes back up below..
+
 \section{What is M--H MCMC?}
@@
 ...What should you report as the inferred parameter value and
-associated uncertainty (error bar)?  What is an upper limit?
+associated uncertainty (error bar)?  What is an upper limit?  Do this
+all for ONE DIMENSION.
+
+...Now for multiple dimensions...return to the marginalization point
+from above...
 

### 2013-02-12 mcmc.tex
sampling approximation

@@
 
-...What is a sampling, heurstically?..
+...What is a sampling, heurstically?.. Along the way, what is a pdf?..
 
-...What is the expectation integral in the sampling approximation?  Why is that a good idea?..
+If there is a pdf $p(x)$ for some quantity $x$, which can be a scalar
+or a multi-element vector or list or blob, we can define
+\emph{expectation values} $E[x]$ for $x$ or for any quantity that can
+be expressed as a function $q(x)$ of $x$:
+\begin{eqnarray}
+E[x] &\equiv& \int x\,p(x)\,\dd x
+\\
+E[q] &\equiv& \int q(x)\,p(x)\,\dd x
+\quad ,
+\end{eqnarray}
+where the integrals are implicitly definite integrals over the domain
+of $x$ (all the parts of $x$ space in which $p(x)$ is finite) and the
+integrals are multi-dimensional if $x$ is multi-dimensional.  These
+expectation values are the mean values of $x$ and $q(x)$ under the
+pdf.  A good sampling---really the definition of a good
+sampling---makes the sampling approximation to these integrals
+accurate.  With a good sampling $\setofall{x_k}_{k=1}^K$ the integrals
+get replaced with sums over samples $x_k$:
+\begin{eqnarray}
+E[x] &\approx& \frac{1}{K}\,\sum_{k=1}^K x_k
+\\
+E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
+\quad .
+\end{eqnarray}
+That is, a sampling is good when any expectation value of interest is
+accurately computed via the sampling approximation.  Your conditions
+on your sampling will depend on the expectations you want to compute,
+and the accuracies you need.
 

### 2013-02-15 mcmc.tex
outline notes
outline points

@@
 
-...Some guidance of where this comes back up below..
+...Is there any way to reconstruct $p(x)$ from the samples
+$\setofall{x_k}$?  In short, no, but in many cases of interest, sort-of!..
+
+...You can't sample the likelihood!  You can only sample a posterior
+pdf!.. why not?.. Give a sense of the difficulties you get into if you
+fail here; like discussion with Sven Kreiss on 2013-02-14...
 
@@
 
-...You can't sample the likelihood!  You can only sample a posterior
-pdf!
 % *********** BJB

### 2013-03-24 mcmc.tex
outline notes
nothing
what is a pdf, what is a sampling, badly normalized functions

@@
 left with a set of $K$ parameter-blob settings $\pars_k$ such that the
-full set $\setofall{\pars_k}_{k=1}^K$ constitutes a fair sampling from
+full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
 the posterior pdf $p(\pars\given\data$).
@@
 
-...What is a sampling, heurstically?.. Along the way, what is a pdf?..
+Heuristically, a sampling $\setof{x_k}_{k=1}^K$ from some pdf $p(x)$
+is a set of $K$ values $x_k$ that are independent draws from the pdf.
+Not being mathematicians, we don't know the precise definition of a
+pdf\note{Seriously.  We don't.}, but for our purposes here a pdf is
+any single-valued (scalar) function that is non-negative everywhere
+(the entire domain of $x$), and obeys a normalization condition
+\begin{eqnarray}
+0 &\leq& p(x) \quad \mbox{for all $x$}
+\\
+1 &=& \int p(x)\,\dd x
+\quad ,
+\end{eqnarray}
+where, implicitly, the integral is over the full domain of $x$.
+Importantly, although $p(x)$ is single-valued, $x$ can be a scalar or
+a multi-element vector or list or blob; it can be arbitrarily large
+and complicated.
+In data-analysis contexts, $x$ will often be the full blob of free
+parameters in the model.
 
-If there is a pdf $p(x)$ for some quantity $x$, which can be a scalar
-or a multi-element vector or list or blob, we can define
-\emph{expectation values} $E[x]$ for $x$ or for any quantity that can
-be expressed as a function $q(x)$ of $x$:
+Given this pdf $p(x)$, we can define \emph{expectation values} $E[x]$
+for $x$ or for any quantity that can be expressed as a function $q(x)$
+of $x$:
 \begin{eqnarray}
@@
 sampling---makes the sampling approximation to these integrals
-accurate.  With a good sampling $\setofall{x_k}_{k=1}^K$ the integrals
+accurate.  With a good sampling $\setof{x_k}_{k=1}^K$ the integrals
 get replaced with sums over samples $x_k$:
@@
 That is, a sampling is good when any expectation value of interest is
-accurately computed via the sampling approximation.  Your conditions
-on your sampling will depend on the expectations you want to compute,
-and the accuracies you need.
+accurately computed via the sampling approximation.  The word
+``accurately'' here translates into some kinds of theorems about
+limiting behavior; the general idea is that the sampling approximation
+becomes exact as $K$ goes to infinity.  The size $K$ of the sampling
+you need \emph{in practice} will depend on the expectations you want
+to compute, and the accuracies you need.
+
+As we noted above, in the context of MCMC, we are often sampling or
+using some \emph{badly normalized} function $f(x)$.
+This function is just the pdf $p(x)$ multiplied by some unknown and
+hard-to-compute scalar.
+In this case, for our purposes, the conditions on $f(x)$ are that it
+be non-negative everywhere and have \emph{finite} integral $Z$
+\begin{eqnarray}
+0 &\leq& f(x) \quad \mbox{for all $x$}
+\\
+Z &=& \int f(x)\,\dd x
+\quad .
+\end{eqnarray}
+
+When the sampling $\setof{x_k}_{k=1}^K$ is of one of these badly
+normalized functions $f(z)$---as it usually will be---the
+sampling-approximation expectation values are the expectation values
+under the properly-normalized corresponding pdf, even though you might
+never \emph{learn that normalization}.
+When we run MCMC sampling on $f(x)$, a function which differs from a
+pdf $p(x)$ by some unknown normalization constant, then the sampling
+permits computation of the following kinds of quantities:
+\begin{eqnarray}
+E[q] &\equiv& \frac{\int q(x)\,f(x)\,\dd x}{\int f(x)\,\dd x}
+\\
+E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
+\quad .
+\end{eqnarray}
+That is, the sampling can be constructed (as we will show below) from
+$f(x)$ directly, and it permits you to compute expectation values
+without ever requiring you to integrate either the numerator or the
+denominator integrals, both of which are generally intractable.\note{As
+  we will
+  see, the intractability comes from various directions, but one is
+  that the dimension of $x$ gets large in most realistic situations.
+  Another is that the support of $p(x)$ tends to be, in normal
+  inference situations, \emph{far smaller} than the full domain of
+  $x$.
+  That is, the pdf is at or very close to zero over all but some tiny
+  and very hard-to-find part of the space.}
 
@@
 ...Is there any way to reconstruct $p(x)$ from the samples
-$\setofall{x_k}$?  In short, no, but in many cases of interest, sort-of!..
+$\setof{x_k}$?  In short, no, but in many cases of interest, sort-of!..
 
-...You can't sample the likelihood!  You can only sample a posterior
-pdf!.. why not?.. Give a sense of the difficulties you get into if you
-fail here; like discussion with Sven Kreiss on 2013-02-14...
+...You can't sample a likelihood!  You can only sample a posterior
+pdf!..
+why not?..
+Give a sense of the difficulties you get into if you fail here; like
+discussion with Sven Kreiss on 2013-02-14...
 
@@
 ...What is a typical proposal distribution?
-
-% *********** BJB
-% YES!!!! Although you can sample the likelihood if your samples are in
-% data space. :)
 
@@
 
-...Autocorrelation time is the premier tool.  It is not a panacea, and why?
+...Autocorrelation time is the premier tool...
+How is Autocorrelation time defined?..
+How is Autocorrelation time measured?..
+Why is it is not a panacea?..
 
@@
 
-...How do you make it run fast?
+...How do you make it run fast?..
 
-...Acceptance ratio and ESJD and etc.
+...Acceptance ratio...
+
+...ESJD and etc...
 
@@
 
-...Gibbs sampling and interior marginalizations
+...Gibbs sampling and interior marginalizations...
 

### 2013-03-24 mcmc.tex
tweaks and line breaking

@@
 \end{eqnarray}
-where the integrals are implicitly definite integrals over the domain
+where again the integrals are implicitly definite integrals over the
+entire domain
 of $x$ (all the parts of $x$ space in which $p(x)$ is finite) and the
-integrals are multi-dimensional if $x$ is multi-dimensional.  These
-expectation values are the mean values of $x$ and $q(x)$ under the
-pdf.  A good sampling---really the definition of a good
+integrals are multi-dimensional if $x$ is multi-dimensional.
+These expectation values are the mean values of $x$ and $q(x)$ under
+the pdf.  A good sampling---really the definition of a good
 sampling---makes the sampling approximation to these integrals
-accurate.  With a good sampling $\setof{x_k}_{k=1}^K$ the integrals
-get replaced with sums over samples $x_k$:
+accurate.
+With a good sampling $\setof{x_k}_{k=1}^K$ the integrals get replaced
+with sums over samples $x_k$:
 \begin{eqnarray}
@@
 That is, a sampling is good when any expectation value of interest is
-accurately computed via the sampling approximation.  The word
-``accurately'' here translates into some kinds of theorems about
-limiting behavior; the general idea is that the sampling approximation
-becomes exact as $K$ goes to infinity.  The size $K$ of the sampling
-you need \emph{in practice} will depend on the expectations you want
-to compute, and the accuracies you need.
+accurately computed via the sampling approximation.
+The word ``accurately'' here translates into some kinds of theorems
+about limiting behavior; the general idea is that the sampling
+approximation becomes exact as $K$ goes to infinity.
+The size $K$ of the sampling you need \emph{in practice} will depend
+on the expectations you want to compute, and the accuracies you need.
 

### 2013-03-24 mcmc.tex
more heuristics
typo

@@
 is a set of $K$ values $x_k$ that are independent draws from the pdf.
+Heuristically, if an enormous (large $K$) sampling could be displayed
+in a finely binned $x$-space histogram, the histogram would look---up
+to a total normalization---just like the original function $p(x)$.
+
 Not being mathematicians, we don't know the precise definition of a
@@
 When the sampling $\setof{x_k}_{k=1}^K$ is of one of these badly
-normalized functions $f(z)$---as it usually will be---the
+normalized functions $f(x)$---as it usually will be---the
 sampling-approximation expectation values are the expectation values

### 2013-03-24 mcmc.tex
tweaks, notes to selves re other methods

@@
 $f(x)$ directly, and it permits you to compute expectation values
-without ever requiring you to integrate either the numerator or the
-denominator integrals, both of which are generally intractable.\note{As
-  we will
+without ever requiring you to integrate either the numerator integral
+or the denominator integral, both of which are generally
+intractable.\note{As we will
   see, the intractability comes from various directions, but one is
@@
 
-% *********** BJB
+...Why everyone reading this document should stop here and JUST FREAKIN' CODE THIS UP.
+
+% *********** BJB talking:
 % I suggest partitioning the discussion of different methods into the following
@@
 
-...How does the stretch move work?
+%%%% Hogg talking:
+% DFM: In this section, I think we should just summarize the other
+% methods, not describe them in detail.  Then we can cite the relevant
+% papers for the enthusiastic.
 
-...Why is the stretch move useful?
+...What, briefly, is the stretch move, and why is it useful?
 
@@
 ...What, briefly, is hamiltonian, and why is it good?
+
+...What, briefly, is reversible-jump, and why is it useful?
 

### 2013-05-16 mcmc.tex
note to self
notes to self

@@
 
+...Return here to the proposal distribution, which exists for most methods...
+
 ...How do you make it run fast?..
@@
 
+...Priors: If you describe your priors in one parameterization but
+then perform sampling in a different space, it is easy to make
+mistakes.  For example: Flat prior in amplitude and phase, sample in
+$A\,\cos\phi$ and $A\,\sin\phi$.  There are many ways to do this
+wrong.
+
 ...Initialization and burn-in advice we haven't addressed earlier?

### 2013-07-03 mcmc.tex
notes to self
give the no-sample-likelihood it's own section

@@
 
+\section{Likelihoods and priors}
+
 ...You can't sample a likelihood!  You can only sample a posterior
@@
 discussion with Sven Kreiss on 2013-02-14...
+
+...It is a good idea to have proper priors.
+It is a good test of your priors that you can sample with a ``null'' likelihood function.
+You can often have your priors wrong despite having beautiful code.
+
+...You must describe your priors in the same parameterization that you ``step'' in.
+That is, if you change your parameters, you \emph{must} change your priors correspondingly.
+Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
 
@@
 
-...Priors: If you describe your priors in one parameterization but
+...Priors: Once again:
+If you describe your priors in one parameterization but
 then perform sampling in a different space, it is easy to make
@@
 wrong.
+
+...Posteriors: On the other hand, you can do anything you like (within
+reason) to the output samples and be sampling the posterior in that
+transformed quantity.
 

### 2013-07-03 mcmc.tex
section headings
sample the posterior; don't sample the likelihood

@@
 
-...You can't sample a likelihood!  You can only sample a posterior
-pdf!..
-why not?..
-Give a sense of the difficulties you get into if you fail here; like
+MCMC is used to obtain samples $x_k$ from PDF $p(x)$,
+  or a badly normalized all-positive function $f(x)$
+  that is different from the PDF by an unknown factor $Z$.
+In the context of data analysis,
+  MCMC is usually being used to obtain samples
+  that are \emph{parameter values} for a probabilistic model of some data.
+
+That is, for the purposes of this \documentname, we will imagine that
+  there is some (possibly large block of) data $D$,
+  and there are parameters $\theta$ of a \emph{likelihood function}
+  $p(D\given\theta)$.%
+\note{...something here about the definition of a model, for our purposes...}
+  The likelihood function is used to compute the probability (density) at any
+  particular observed data set, given any parameter setting.%
+\note{...strictly this is only correctly called the ``likelihood'' when the data are treated as fixed...}
+In standard probabilistic inference,
+  the likelihood is used to update beliefs about the parameters.%
+\note{...something here about us being Bayesian;
+   if you aren't Bayesian, why are you considering MCMC for your data analysis?..
+   probabilities are representations of beliefs not frequencies...
+   beliefs can be thought of as subjective or ``official''...
+   etc...}
+In particular, if you started with prior probabilistic beliefs
+  represented by a prior PDF $p(\theta)$,
+  your posterior beliefs can be represented by a posterior PDF
+  $p(\theta\given D)$, created via Bayes's rule:
+\begin{eqnarray}
+p(\theta\given D) = \frac{1}{Z}\,p(D\given\theta)\,p(\theta)
+\quad,
+\end{eqnarray}
+  where $Z$ is a normalization constant.
+MCMC is used in the context of data analysis
+  (and hence the context of this \documentname)
+  to sample posterior PDFs.
+
+Importantly, \emph{you can't use MCMC to draw samples $\theta_k$ from the likelihood $p(D\given\theta)$!}
+Why not?
+Because $p(D\given\theta)$ is a PDF for the \emph{data} $D$ not a PDF for the \emph{parameters} $\theta$.
+Many times data analysts believe they are sampling the likelihood,
+  when in fact they are sampling a posterior PDF but with implicit priors.
+
+...Give a sense of the difficulties you get into if you fail here; like
 discussion with Sven Kreiss on 2013-02-14...
+
+...Explain how your code \emph{should look} with a prior function and a likelihood function...
 
@@
 
-\section{stretch move and other variants}
+\section{Ensembles, Hamiltonians, and other variants}
 

### 2013-07-07 mcmc.tex
bug
moved likelihood sampling rant to later, removed repitition

--- 2013-07-03 mcmc.tex
@@
 
-\section{Likelihoods and priors}
-
-...You can't sample a likelihood!  You can only sample a posterior
-pdf!..
-why not?..
-Give a sense of the difficulties you get into if you fail here; like
-discussion with Sven Kreiss on 2013-02-14...
-
-...It is a good idea to have proper priors.
-It is a good test of your priors that you can sample with a ``null'' likelihood function.
-You can often have your priors wrong despite having beautiful code.
-
-...You must describe your priors in the same parameterization that you ``step'' in.
-That is, if you change your parameters, you \emph{must} change your priors correspondingly.
-Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
-
 \section{What is M--H MCMC?}
@@
 
+\section{Likelihoods and priors}
+
+MCMC is used to obtain samples $x_k$ from pdf $p(x)$,
+  or a badly normalized all-positive function $f(x)$
+  that is different from the pdf by an unknown factor $Z$.
+In the context of data analysis,
+  MCMC is usually being used to obtain samples
+  that are \emph{parameter values} for a probabilistic model of some data.
+That is, MCMC is sampling a pdf \emph{for the parameters}.
+If it is returning parameter values, then it is not sampling a pdf for the data.
+That is, MCMC can only sample a posterior (or prior) pdf for the parameters.
+\emph{It cannot sample a likelihood.}%
+\note{Well, technically, MCMC \emph{can} be used to sample a likelihood function,
+  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
+  The purist might say that even this is not sampling a likelihood function,
+  because you should only call it a ``likelihood function'' in contexts in which you are
+  treating the data as fixed (at the true data values) and the parameters as variable.}
+
+Despite all this, in many cases,
+  data analysts believe they are sampling the likelihood.
+This is because (we presume) they have put a likeilihood function
+  (or log-likelihood function)
+  in as the input to the MCMC code, where the probability function
+  (or log-probability function) should go.
+Then is it the likelihood that is being sampled?
+No, not really;
+it is a posterior probability that is directly proportional to the likelihood function.
+That is, it is a posterior probability for some implicit (and improper) ``flat'' priors.
+
+...Give a sense of the difficulties you get into if you fail here; like
+discussion with Sven Kreiss on 2013-02-14...
+
+...Explain how your code \emph{should look} with a prior function and a likelihood function...
+
+...It is a good idea to have proper priors.
+It is a good test of your priors that you can sample with a ``null'' likelihood function.
+You can often have your priors wrong despite having beautiful code.
+
+...You must describe your priors in the same parameterization that you ``step'' in.
+That is, if you change your parameters, you \emph{must} change your priors correspondingly.
+Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
+
 \section{Advice and discussion}

### 2013-07-07 mcmc.tex
notes to self

@@
   data analysts believe they are sampling the likelihood.
-This is because (we presume) they have put a likeilihood function
+This is because (we presume) they have put a likelihood function
   (or log-likelihood function)
@@
 
+...Simple problems where the outputs are analytic.  Functional testing.
+
 ...breaking Markov-ness.
@@
 wrong.
+
+...Sample the prior (likelihood set to unity) to check priors and parameterization (another functional test).
 

### 2013-07-11 mcmc.tex
troubleshooting section notes

@@
 
+\section{Troubleshooting}
+
+...Run on standard distributons.
+
+...Set likelihood to 1 and check that you sample your prior pdf.
+
+...Low acceptance fraction?  Reducing step size \emph{must} increase fraction.
+
+...Appearance of good and bad chains (show examples).
+
+...Different initializations leading to different posterior samplings?
+
+...Jammed ensemble sampler?
+
+...Degeneracies and duplications in parameter space; reparameterizations.
+
 \section{Advice and discussion}
@@
 
+...How to deal with forbidden regions in prior, or sharp prior boundaries?
+
+...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.
+
 ...Initialization and burn-in advice we haven't addressed earlier?

### 2013-07-11 mcmc.tex
troubleshooting text

--- 2013-07-07 mcmc.tex
@@
 
+\section{Troubleshooting}
+
+\paragraph{functional testing:}
+...Run on standard distributons.
+
+...Set likelihood to 1 and check that you sample your prior pdf.
+
+\paragraph{low acceptance fraction:}
+If you find that you are getting low acceptance fraction
+  (in standard M-H MCMC or other varieties for which there is such a concept)
+  you can simply reduce the sizes of the steps you consider to increase the fraction that are accepted.
+This means decreasing the variance of the proposal distribution,
+  or whatever parameter or parameters control the width of that distribution.
+If reducing the variance of the proposal distribution does \emph{not} increase the acceptance fraction,
+  then \emph{you have a bug}.
+You must find that bug by auditing your code or else returning to the functional tests listed above.
+
+Similarly for high acceptance fraction:
+If you are finding a high acceptance fraction,
+  then increasing the variance of the proposal distributon \emph{must} decrease the acceptance fraction.
+If it does \emph{not}, then \emph{you have a bug}.
+
+\paragraph{sticky chains:}
+...Appearance of good and bad chains (show examples).
+
+\paragraph{initialization-dependence:}
+...Different initializations leading to different posterior samplings?
+
+...Jammed ensemble sampler?
+
+\paragraph{parameterization problems:}
+...Degeneracies and duplications in parameter space; reparameterizations.
+
 \section{Advice and discussion}

### 2013-07-12 mcmc.tex
convergence

@@
 That is, knowing that you have a complete and representative sampling
-is nearly impossible.
+is effectively impossible.
 
-Heuristically, however, you have sampled long enough when you can see
+In this context,
+  it is \emph{simply not fair} to require an MCMC run to have fully and completely
+  sampled the posterior pdf,
+  at least not in any provable sense.
+The question of whether a sampling is converged to a representative sampling
+  of the posterior pdf is actually outside the domain of science,
+  because the domain of science is the domain of questions answerable by scientists.
+That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
+  despite many statements in the literature to the contrary.%
+  \note{Give example statements... Mention cases in which you \emph{might} be able to claim this...}
+The upshot of this is that we don't and can't require any kind of absolute convergence;
+  indeed, it would be impossible even to test for it.
+
+In this pragmatic situation---the Real World, as it is called---we have to rely on heuristics.
+Heuristically, you have sampled long enough when you can see
 that the (or each) walker has traversed the high-probability parts of

### 2013-07-30 mcmc.tex
something old
notes to self

@@
 
+...Why is it that the sampling is only really useful for computing integrals.  More on this below...
+
 ...Is there any way to reconstruct $p(x)$ from the samples
@@
 report as our \emph{results}?
+
+...Reminder that a sampling is good for computing integrals (see above).
+Therefore we should endeavor to use outputs that are based on integrals computed with the sampling.
+This includes expectations and quantiles, but not optima...
 
@@
 
-...What plot should you make?  Make plots that demonstrate a non-converged and a converged chain.
-
-...Autocorrelation time is the premier tool...
+Making this heuristic condition more precise...
+Autocorrelation time is the premier tool...
 How is Autocorrelation time defined?..
@@
 Why is it is not a panacea?..
+
+...What plots should you make?
+Make plots that demonstrate a non-converged and a converged chain...
 
@@
 
+\paragraph{model checking:}
+...Beyond the scope of this document...
+Something about cross-validation and posterior predictive checks...
+Something about plotting models (drawn from posterior) on data, where appropriate...
+How could you possibly know that your sampling really represents your \emph{real} posterior PDF?..
+
 \section{Advice and discussion}

### 2013-09-10 mcmc.tex
emph

@@
 under the properly-normalized corresponding pdf, even though you might
-never \emph{learn that normalization}.
+\emph{never learn that normalization}.
 When we run MCMC sampling on $f(x)$, a function which differs from a

### 2013-09-10 mcmc.tex
integrals and histograms

@@
 
+The ``correctness'' of a sampling is defined (above) in terms of its use in
+  performing integrals or approximate computation of integrals.
+In a deep sense, the \emph{only} thing a sampling is good for is
+  computing integrals.
+There are many uses of the MCMC sampling, some of them good and some of them bad.
+Most of the good or sensible uses will somehow involve integration.
+
+For example, one magical property of a sampling (in a $d$-dimensional space)
+  is that a \emph{histogram} (a $d$-dimensional histogram) of the samples
+  (divided, if you like, by the number of samples in each bin and the bin width\note{something about units and normalization})
+  looks very much like the pdf from which the samples were drawn.
+This is a way to ``reconstruct'' the pdf from the sampling:
+Make a histogram from the samples.
+Even in this case, the sampling is being used to do \emph{integrals};
+  the (possibly odd) idea is that the approximate or effective value of the pdf in each bin
+  of the histogram is an average over the bin.
+That average is obtained by performing an integral.
+
 ...How is the sampling of the marginalized pdf trivially constructed from the full sampling?..
-
-...Why is it that the sampling is only really useful for computing integrals.  More on this below...
-
-...Is there any way to reconstruct $p(x)$ from the samples
-$\setof{x_k}$?  In short, no, but in many cases of interest, sort-of!..
 

### 2013-09-13 mcmc.tex
marginalization

@@
 
+% in the next paragraph I adopt $d$ to be the dimensionality of the space
+
 For example, one magical property of a sampling (in a $d$-dimensional space)
@@
 
-...How is the sampling of the marginalized pdf trivially constructed from the full sampling?..
+Finally---and perhaps most importantly---another magical propert of a sampling
+  (in a $d$-dimensional space)
+  is that if some of your $d$ dimensions are totally uninteresting
+  (nuisance parameters, if you will)
+  and some of your $d$ dimensions are of great interest,
+  the sampling in the full $d$-space is trivially converted into a sampling in the subspace of interest:
+You just drop from each $x_k$ vector (or blob or list) the dimensions of no interest!
+That is, the projection of the sampling to the subspace of interest produces
+  a sampling of the marginalized pdf, marginalizing (or projecting) out the nuisance parameters.
+That is extremely important for inference,
+  where there are always parameters with very different levels of importance to the scientific conclusions.
+This point will return again below, either explicitly or implicitly. % does it?
 

### 2013-09-13 mcmc.tex
note

@@
 That is, the projection of the sampling to the subspace of interest produces
-  a sampling of the marginalized pdf, marginalizing (or projecting) out the nuisance parameters.
+  a sampling of the marginalized pdf, marginalizing (or projecting) out the nuisance parameters.%
+  \note{This point about marginalization is, once again, a use of the sampling to perform an \emph{integral};
+        the marginalized pdf is obtained from the full pdf by an integration.
+        The sampling performs this integration automatically.}
 That is extremely important for inference,

### 2013-09-18 mcmc.tex
measurements and outputs

@@
 
+\section{Ensembles, Hamiltonians, and other variants}
+
 % *********** BJB talking:
@@
 
-\section{Ensembles, Hamiltonians, and other variants}
-
 %%%% Hogg talking:
@@
 
+...What, briefly, is importance sampling?
+
 ...What, briefly, is nested sampling, and why is it good?
+
+...What, briefly, is tempering?
 
@@
 bayesian---there is no ``answer'', there are only probability density
-functions.  Any time in a paper or in an email or in a conversation
+functions (pdfs).  Any time in a paper or in an email or in a conversation
 we say what ``the answer is''---even if we say it with an error
@@
 the context of probabilistic reasoning\note{CITE DECISION DOCUMENT}.
-But without going into the economic model, it is fair to say that a
+But without going into that ``economic'' model, it is fair to say that a
 substantial problem with being a committed probabilist is that when we
@@
 \emph{probability distribution over pubications}; we have to publish
-\emph{one single, deterministic text}; we have to \emph{decide} what
+\emph{one single, deterministic text}; we have to make a decision about what
 to write in the title, the abstract, the tables, figures, and results
@@
 
-...Reminder that a sampling is good for computing integrals (see above).
-Therefore we should endeavor to use outputs that are based on integrals computed with the sampling.
-This includes expectations and quantiles, but not optima...
+One thing we can keep in mind to guide us in this is what we said above,
+  which is that samplings are good for doing \emph{integrals}.
+We should endeavor to use as our ``results'' outputs from the MCMC
+  that are based on integrals computed with the sampling.
+This includes expectations, medians, quantiles, one-dimensional histograms,
+  and multi-dimensional histograms.
+This does \emph{not} include the ``best'' sample or a mode or optimum.
+The latter things are not necessarily illegitimate outputs of the MCMC,
+  but they do not make best use of the fact that the sampling is a tool for integration,
+  and we do not recommend them.
 
-...What should you report as the inferred parameter value and
-associated uncertainty (error bar)?  What is an upper limit?  Do this
-all for ONE DIMENSION.
+Imagine that you are in the simplest case:
+You have a model with a small number of parameters (say three-ish),
+  and only one of them is of great interest.
+What are your options for the ``measurement'' of this parameter?
+The only simple integral-based options are the posterior mean
+  or posterior median value for the parameter.
+\\HOGG CHOOSE NOTATION AND GIVE FORMULAE HERE.
+
+What are your best options for the ``uncertainty'' or ``limits'' on this parameter?
+The only simple integral-based options are posterior quantiles.
+That is, the ``one-sigma'' error bar can be the size of the central (or smallest%
+  \note{HOGG shizzle about choosing quantiles}) interval of the parameter that contains 68 percent of the posterior samples.
+\\HOGG GIVE FORMULA HERE.\\
+The ``two-sigma'' error bar would be the same but for the 95-percent interval.
+
+One amusing thing is that if you choose the 68-percent interval sensibly,
+  but use as the ``measurement'' the posterior mean,
+  you can get pathological situations in which the posterior mean measurement
+  is actually \emph{outside} the one-sigma confidence interval.
+For this reason (and others), we usually recommend as a default behavior%
+  ---in the one-dimensional case---%
+  to choose the \emph{median} of sampling as the measurement value,
+  the 16-percent quantile as the lower error bar,
+  and the 84-percent quantile as the upper error bar.
+This has pathologies in higher dimensions (as we are about to see),
+  but is pretty safe for one-dimensional answers.
+
+The conservative scientist would show not just 68-percent error bars,
+  but also 95-percent
+  (to help readers visualize the skew of the posterior pdf.
+The very conservative scientist would also show a histogram
+  of the posterior samples,
+  with the relvant quantiles (2.5, 16, 50, 84, and 97.5~percent) indicated.
+
+As we mentioned above, scientists often like to report the ``best sample'';
+  that is, the sample with the highest posterior pdf value.
+This is not usually a good idea.
+For one, MCMC is a sampling algorithm, not an optimization algorithm;
+  there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.
+For two, the closeness of the best sample to the posterior mode is a strong function of sample size.
+For three, if you just want an optimum, use an optimizer!\note{HOGG, something?}
+That is, we \emph{strongly advise against} reporting,
+  as the measurement or among the results,
+  the parameter value or values for the best sample.
 

### 2013-12-01 mcmc.tex
stuff about best sample and MAP

@@
 \newcommand{\pars}{\theta}
+\newcommand{\best}{{\mathrm{(best)}}}
+\newcommand{\better}{{\mathrm{(better)}}}
 
@@
 For one, MCMC is a sampling algorithm, not an optimization algorithm;
-  there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.
-For two, the closeness of the best sample to the posterior mode is a strong function of sample size.
-For three, if you just want an optimum, use an optimizer!\note{HOGG, something?}
+  there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.%
+  \note{The sampler samples the pdf fairly.
+    The tallest peak in the posterior is not guaranteed to be---%
+    and in general is not---%
+    also the peak with the greatest posterior ``mass''.
+    That is, the samples in a fair sampling will not necessarily come
+    predominantly from the \emph{tallest} peak in the posterior pdf.
+    They will come from the peak with the greatest total integral
+    of the likelihood over the prior.
+    These issues might sound ``academic'',
+    but when the number of parameters gets large (greater than, say, 10),
+    many normal intuitions one might have (if any) about what is reasonably likely in a pdf
+    are regularly violated.}
+For two, the closeness of the best sample to the posterior mode is a strong function of sample size,
+  and with very bad scaling.%
+  \note{If you have a $K$-point sampling in a $d$-dimensional parameter space,
+    there is some best sample $\pars_k^{\best}$.  Now imagine that you want to make
+    that best sample 10 times better---that is, you want to find a point $\pars^{\better}$
+    that is ten times closer to the true optimum.  How many more samples do you need to take?
+    In the limit that $K$ is large, on average, you have to take $10^d$ more samples!
+    Most sensible optimizers are far, far better than this.}
+For three, if you just want an optimum, use an optimizer!%
+  \note{HOGG, something?}
 That is, we \emph{strongly advise against} reporting,
-  as the measurement or among the results,
+  as ``the'' measurement or ``the result'',
   the parameter value or values for the best sample.
+Giving some posterior samples is a good idea (see below).
+If you feel drawn to give the best sample,
+  instead give the parameter values found by optimizing the posterior PDF,
+  starting at the best sample as an initialization
+  (and sacrifice your probabilism\note{something}).
+This optimal-posterior parameter value is called the
+  ``maximum \foreign{a posterior}'' or ``MAP'' value for the parameters;
+  it is like a maximum-likelihood value but regularized by the prior.
+It is an estimator with some good (and some bad) properties, but it is not the point of MCMC,
+  and therefore outside the scope of this \documentname.
 

### 2013-12-01 mcmc.tex
typos

@@
   but also 95-percent
-  (to help readers visualize the skew of the posterior pdf.
+  (to help readers visualize the skew of the posterior pdf).
 The very conservative scientist would also show a histogram
@@
 This optimal-posterior parameter value is called the
-  ``maximum \foreign{a posterior}'' or ``MAP'' value for the parameters;
+  ``maximum \foreign{a posteriori}'' or ``MAP'' value for the parameters;
   it is like a maximum-likelihood value but regularized by the prior.

### 2013-12-04 mcmc.tex
tuning section; some notes to self

@@
 %   - Bovy
+%   - Brewer
 % - post on arXiv
@@
 
-...If you are really good, you will pass forward a sampling.  How big
-should it be?  Refer to Mackay here.
+...If you are really good, you will pass forward a sampling.
+...How big should it be? Refer to Mackay here.
+...Usefulness of a big sampling for hierarchical inferences; concept of interim prior?
 
@@
 ...What plots should you make?
-Make plots that demonstrate a non-converged and a converged chain...
+...Plots showing individual walkers going through the space.
+...(now show plots that demonstrate a non-converged and a converged chain)...
 
 \section{Tuning}
-
-...Return here to the proposal distribution, which exists for most methods...
-
-...How do you make it run fast?..
-
-...Acceptance ratio...
-
-...ESJD and etc...
 
@@
 
-...Gibbs sampling and interior marginalizations...
+Most MCMC methods
+  (though not all\note{We are promoters of an ensemble method that has a
+  nearly-fixed proposal distribution and requires no tuning (CITE EMCEE).})
+  make use of something like a ``proposal distribution''
+  which determines what kinds of steps the walker can take
+  as it random-walks through the parameter space.
+The user generally has a lot of control over what this proposal distribution might be,
+  and how to choose the parameters.
+For example, in a $d$-dimensional parameter space, %do we consistently use d here?
+  a zero-mean but anisotropic Gaussian (normal distribution)
+  is often used to draw offsets for the walker to take from point to point.
+Even within this choice (which is of one function among many possibilities),
+  there are $d\,(d+1)/2$ parameters in the $d\times d$ symmetric, positive definite covariance matrix
+  to set ``by hand''.
+How to choose this function and set these parameters?
+
+The key idea here is that if the proposal distribution is too narrow---%
+  it proposes steps too small---%
+  almost all steps will be accepted (REFERENCE ALGORITHM HERE),
+  but it will take a long time to move anywhere because of timidity.
+If the propsoal distribution is too wide---%
+  it proposes steps too large---%
+  the moves will cover parameter space easily,
+  but almost no steps will be accepted;
+  it will tend to jump to much lower probability regions.
+There is a Goldilocks step size (proposal distribution root-variance) that is ``just right''.
+In one dimension this might be easy to find,
+  but, as we say, in large numbers of dimensions,
+  there is a lot of freedom in choosing the parameters of the distribution.
+
+In a deep sense, the only scalar that makes sense to optimize,
+  when choosing proposal distribution (or, loosely speaking, step size),
+  is the autocorrelation time, described above.
+The optimal step size is the step size that makes for the shortest autocorrelation time.
+This is easy to state, but hard to use in practice;
+  being a second-order statistic of the MCMC chain,
+  the autocorrelation time is a hard thing to measure without a lot of data,
+  so it is hard to quickly estimate it and adjust,
+  much less put it into some kind of optimization loop.
+We usually use proxies for this of various kinds.
+
+The simplest proxy for autocorrelation time optimization is the \emph{acceptance fraction}.
+If you are accepting almost all proposed steps, your step sizes are too small on average.
+If you are accepting almost none, your step sizes are too large.
+The Goldilocks value is around a half,
+  with a strong argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}.
+In the burn-in phase (discussed below) of an MCMC run,
+  it makes sense to track the acceptance ratio,
+  and adjust the proposal distribution variance as you get acceptance ratios
+  that are far from the Goldilocks ratio.
+This process can be automated easily;
+  such automation is part of many projects that use MCMC.
+
+A very common---%
+  and very useful---%
+  kind of proposal distribution is one that cycles through parameters,
+  taking a random step in just one parameter at a time.
+Aside from involving some persistence of state,
+  this kind of proposal distribution can be valuable,
+  in part because it reduces the (almost impossible) $d$-dimensional tuning problem
+  to $d$ one-dimensional problems:
+In this form of proposal distrubution,
+  it is possible to track a separate acceptance fraction for every parameter.
+Code can be built that uses the burn-in phase to tune all $d$ proposal variances
+  such that each of them, individually,
+  obtains acceptance at the same Goldilocks ratio.
+
+One very important note to make here,
+  because it is relevant to tuning,
+  is that tuning can \emph{only} take place during the burn-in phase;
+  you cannot tune \emph{while} you run your final MCMC run.
+Why not?
+Because tuning the proposal distribution based on the past history of the chain
+  violates the ``Markov'' property
+  that each step depends \emph{only} on the state at the previous step.
+Violating the Markov property can be very bad;
+  when you violate it,
+  you lose all the provable properties of MCMC
+  on which all our righteous power is based.
+
+Another proxy for autocorrelation time is the
+  Expected Squared Jump Distance.\note{(CITE GELMAN)}
+This is the mean squared distance the walker moves, per step.
+It is maximized when the acceptance ratio is reasonable
+  and the step size is large;
+  it is large when the exploration of the space is fast.
+This proxy is easy to measure and use for tuning,
+  and more directly related to autocorrelation time than the acceptance ratio.
+We recommend using it for tuning, though we have never used it ourselves.
+Like the acceptance fraction, it can also be used to tune
+  in the case that the proposal distribution loops over parameters;
+  once again, the user gets (in this case) $d$ one-dimensional tunings.
+
+When sampling really gets very slow,
+  there are various tricks and tips to work through.
+One is to see if any of the parameters has a conditional posterior
+  probability that is guaranteed to be Gaussian
+  (this is true for some linear parameters in least-square problems
+  when the priors are Gaussian, for example).
+These dimensions can often be sampled \emph{exactly},
+  outside of the MCCM framework;
+  the algorithm looks like:
+Take an MCMC step in the hard parameters,
+  then take an exact step in the easy parameters,
+  repeat.
+Another related trick is to analytically marginalize,
+  rather than sample,
+  easy nuisance parameters.
+The huge bag of possible tricks is so large,
+  it goes way beyond the scope of this introductory \documentname.
+In our experience, it is valuable to make friends with at least one statistician,
+  one applied mathematician, and one computer scientist.
+Between them, they ought to span the literature.
 

### 2013-12-11 mcmc.tex
paragraph about initialization

@@
 That is, the ``one-sigma'' error bar can be the size of the central (or smallest%
-  \note{HOGG shizzle about choosing quantiles}) interval of the parameter that contains 68 percent of the posterior samples.
+  \note{HOGG shizzle about choosing quantiles}) interval of the parameter
+  that contains 68 percent of the posterior samples.
 \\HOGG GIVE FORMULA HERE.\\
@@
 
-...Precede with optimization
-% BJB
-% If applicable. Bad idea in high dimension/underconstrained
-% problem.
+Just as most (though not all) MCMC methods require a choice of proposal distribution,
+  most (though not all) require a choice about initialization:
+The walker needs (or walkers need) to be started somewhere in the parameter space.
+In many use cases for MCMC,
+  the investigator wants to use it to obtain uncertainty information about
+  or propagate uncertainty into parameter estimates.
+In these cases, it makes sense to initialize the walker or walkers at sensible parameter estimates,
+  found by optimizing a likelihood or a posterior PDF in advance of sampling.
+In extremely high dimensions (large numbers of parameters), this can be a bad idea,
+  since the optimal parameters are not necessarily near typical posterior samples,
+  but in low dimensions (few to tens) this is often sensible.
+
+In other use cases,
+  the investigator wants to use MCMC to search all of parameter space
+  for good models, and also sample.
+In this case, it makes sense to be more catholic with initialization...
+
+...but note that MCMC is not the ideal search algorithm for any parameter space.
 

### 2013-12-12 mcmc.tex
functional testing

@@
 In extremely high dimensions (large numbers of parameters), this can be a bad idea,
-  since the optimal parameters are not necessarily near typical posterior samples,
+  since the optimal parameters are not necessarily near typical posterior samples\note{Make the counterintuitive point.},
   but in low dimensions (few to tens) this is often sensible.
@@
 
+When issues arise with MCMC sampling,
+  they are sometimes difficult to diagnose:
+Is it the MCMC code, the priors, the likelihood function,
+  the initialization, the tuning, or something else?
+
 \paragraph{functional testing:}
-...Run on standard distributons.
+In addition to a full set of unit tests for every part of your code
+  (yeah right!\note{If you aren't unit testing, you are probably making some very big mistakes.}),
+  we recommend building some end-to-end functional tests that permit you to perform MCMC
+  runs with output that must meet certain expectations.
+The simplest kind of functional test is to sample a known distribution,
+  and check that the sampling has the moments you expect (to within tolerances).
+For example, you can sample a $d$-dimensional Gaussian distribution with known non-trivial covariance,
+  and check that the empirical covariance comes out as expected.
+This is an easy test to fail, so a success builds confidence in any MCMC code.
+
+In the most extreme form of this test,
+  for a Gaussian in $d$ dimensions,
+  build the covariance matrix input to the Gaussian distribution
+  by taking the sum of the outer product of $d+1$ randomly generated vectors.
+Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
+Set up the MCMC sampler with the Gaussian function with this covariance matrix
+  as the probability function (or the log of it as the log-probability function).
+Sample for many autocorrelation times,
+  or until you are confident you are getting good samples.
+Check that the empirical $d$-dimensional covariance of the samples
+  has the same (or similar) eigenvectors and eigenvlues to the input matrix.
+Also check that the individual one-dimensional samplings
+  (along each axis or any projection in the $d$-space)
+  look Gaussian and have nearly vanishing skew and kurtosis.
+It takes some thinking and some calculating to see ``how well'' these things all ought to match
+  expectations, but one thing that can be said with confidence is that agreement should improve
+  as the sampling proceeds.
+If it doesn't, it is likely that there is something broken about the sampler itself.
 
 ...Set likelihood to 1 and check that you sample your prior pdf.
+...Checks properness of priors.
+...Checks for think-os about parameterization.
+...Make triangle plots.
 

### 2013-12-14 mcmc.tex
toubleshooting and advice

@@
 % -----
-% - write zeroth draft
-% - write first draft
+% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
+% - need to mention case where there is an MC integral INSIDE the likelihood function
+% - need to mention testing the likelihood function
 % - send to friendlies for comments
@@
 
+One note to make here
+  is that you must \emph{propose}
+  in the same coordinates (parameterization or transformation of parameters)
+  as that in which your priors are \emph{specified},
+  or else multiply in a (possibly nasty) Jacobian.
+That is, if your prior is ``flat in $x$''
+  but you find it easier to run your sampler by proposing steps in $\ln x$,
+  if you don't modify your acceptance fraction by the Jacobian,
+  your \emph{real} prior won't be flat in $x$.
+These think-os can get subtle;
+  the best way to avoid them
+  is to write your code and your likelihood function
+  and your prior pdf function and your proposal distribution
+  all in precisely the same parameterization.
+We will return to this point below in our comments on functional testing.
+
 ...Why everyone reading this document should stop here and JUST FREAKIN' CODE THIS UP.
@@
 
-One very important note to make here,
-  because it is relevant to tuning,
+One note to make here---%
+  because it is relevant to tuning---%
   is that tuning can \emph{only} take place during the burn-in phase;
@@
 In addition to a full set of unit tests for every part of your code
-  (yeah right!\note{If you aren't unit testing, you are probably making some very big mistakes.}),
+  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
   we recommend building some end-to-end functional tests that permit you to perform MCMC
@@
 
-...Set likelihood to 1 and check that you sample your prior pdf.
-...Checks properness of priors.
-...Checks for think-os about parameterization.
-...Make triangle plots.
+Another important functional test is to \emph{sample the prior pdf}.
+This tests the sampler,
+  tests that your priors really are what you think they are
+  (and there are so many ways for them \emph{not} to be\note{HOGG}),
+  and tests that your prior pdfs are proper
+  (functions that can be normalized).
+This test is most effective as a true end-to-end test of your model
+  if you structure your code such that the log probability function \code{lnprob}
+  input to the MCMC scheme
+  is set up as a sum of a log likelihood function \code{lnlikelihood}
+  and a prior function \code{lnprior}.
+The sampler can be switched to sampling the prior by a simple replacement
+  of the \code{lnlikelihood} function
+  with a trivial function that always returns zero.
+
+MCMC should run well in this flat-likelihood case (priors are usually easy to sample)
+  and the output sample histograms should look like what you expect
+  given the prior pdfs you coded.
+If the walkers run off to positive or negative infinity,
+  and nothing seems to converge,
+  your priors are probably not proper.
+If the priors don't look as you expect,
+  you either have a bug in your code,
+  or else a think-o about how your priors ought to look.
+Either way, diagnosis in order.
+
+...Make triangle plots.  Great value!
+
+Finally, if you pass the above tests but things are still incomprehensible,
+  it is often a problem with the likelihood function.
+Things to check include the following:
+When you re-call the likelihood function with the same parameters,
+  do you get \emph{exactly} the same answer?
+If you don't, your likelihood function effectively depends on some random numbers;
+  WE DISCUSS THIS WHERE?
+When you call the likelihood function on a grid in each parameter
+  (with the others fixed, say, to reasonable values),
+  is the likelihood function a smooth function of the parameter,
+  or at least as smooth as you expect?
+Things like numerical integration,
+  truncated expansions,
+  or adaptive approximations
+  inside likelihood functions
+  can make the function noisy or jagged at small scales in parameter space.
+These problems hurt optimizers more than samplers,
+  but they usually point to code issues.
 

### 2013-12-30 mcmc.tex
diy advice

@@
 
-...Why everyone reading this document should stop here and JUST FREAKIN' CODE THIS UP.
+As we mention below,
+  there are many MCMC methods more advanced than M--H.
+However, they all share characteristics with M--H,
+  such that it is very valuable to understand M--H well before using anything more advanced.
+Furthermore, a scientist new to MCMC benefits enormously from
+  building, tuning, and using her or his own MCMC software.
+One piece of advice we give then,
+  to the new user of MCMC,
+  is to code up, tune, and use a M--H MCMC sampler for a scientific project.
+A huge amount is learned in doing this,
+  and it is very often the case that the home-built M--H MCMC does everything needed;
+  you often don't need any more advanced tool.
+Only after you have concluded that your home-built M--H MCMC sampler
+  is not suited to your project
+  (or not developed properly into a properly versatile software package)
+  should you download and start to use any professionally developed alternative.
+That is, even if---in the end---%
+  you want to leave the sampling code to the experts and run an industrial-strength code,
+  it is still valuable to build your own,
+  given the simplicity of the algorithm,
+  and given the intuition you gain by doing it (at least once) yourself.
 

### 2013-12-30 mcmc.tex
tweak

@@
   there are many MCMC methods more advanced than M--H.
-However, they all share characteristics with M--H,
+However, they all share characteristics with M--H
+  (initialization, tuning, judging convergence),
   such that it is very valuable to understand M--H well before using anything more advanced.

### 2013-12-30 mcmc.tex
advice

@@
 
-\section{Troubleshooting}
+\section{Troubleshooting and advice}
 
@@
 
-...Make triangle plots.  Great value!
+\paragraph{visualization:}
+...Make triangle plots.  Great value!...  Show example here.
 
-Finally, if you pass the above tests but things are still incomprehensible,
-  it is often a problem with the likelihood function.
-Things to check include the following:
-When you re-call the likelihood function with the same parameters,
+\paragraph{likelihood issues:}
+Sometimes there can be problems with the likelihood function itself.
+For example,
+  when you re-call the likelihood function with the same parameters,
   do you get \emph{exactly} the same answer?
 If you don't, your likelihood function effectively depends on some random numbers;
-  WE DISCUSS THIS WHERE?
+  maybe it includes within it an integral performed by Monte Carlo method?
+In general, you want to change your likelihood function such that,
+  even if it includes a numerical or stochastic integral,
+  it is designed such that it produces \emph{precisely} the same
+  value when it returns to the same point in parameter space.
+One way to make this happen is to choose the random numbers
+  (those used for any internal integration inside the likelihood function)
+  in advance and re-use them on every likelihood call.
+This provides the same integral according to the same approximation,
+  but both speeds up the code (no re-generation of random numbers)
+  and also makes the likelihood function single valued.
+Even better is to replace internal Monte Carlo integrals with
+  deterministic numerical (or even better analytic) integrals.
+
+If things are giving trouble and you suspect the likelihood function,
+  another important test is to visualize slices through parameter space.
 When you call the likelihood function on a grid in each parameter
   (with the others fixed, say, to reasonable values),
-  is the likelihood function a smooth function of the parameter,
-  or at least as smooth as you expect?
+  you should see a likelihood value that is a smooth function of each parameter,
+  or at least as smooth as you expect.
 Things like numerical integration,
@@
   but they usually point to code issues.
+In general, plotting the likelihood function as a function of parameter along
+  slices in parameter space often reveals bugs or think-os.
+
+SHOW EXAMPLES?
 
@@
 \paragraph{sticky chains:}
-...Appearance of good and bad chains (show examples).
+If you are concerned about acceptance fraction or convergence,
+  it makes sense to plot parameter values as a function of ``step number'' or iteration number.
+That is, plot the ordered chain in each parameter dimension.
+These plots will have long horizontal patches if the chain is getting stuck;
+  a converged chain will show the walker traversing the parameter space in every dimension fully many times
+  over the length of the run.
+Not a few times; many times.
+If you are using an ensemble sampler\note{CITE EMCEE},
+  make plots showing all $M$ walkers, each a different color.
+A converged run for an ensemble sampler will show every single walker traversing the parameter space
+  in every dimension many times.
+
+SHOW EXAMPLES OF GOOD AND BAD?
 
 \paragraph{initialization-dependence:}
-...Different initializations leading to different posterior samplings?
+If you initialize your MCMC runs in different places in parameter space,
+  do you get the same (or very similar) final parameter samplings,
+  in mean and variance?
+If you don't,
+  then your posterior PDF is badly multi-modal;
+  trivial MCMC methods probably won't fix your problem.
+Not to put too fine a point on it:
+If your results are initialization-dependent,
+  then your MCMC runs \emph{are not converged}.
 
-...Jammed ensemble sampler?
+You either have to go to a method that tries to explore the posterior
+  more liberally, such as nested sampling or simulated tempering,
+  or else split your ``model'' up into several sub-models,
+  each of which contains different posterior modes.
+In general, this problem is hard to fix;
+  as we mentioned above, it is provably impossible
+  to explore all of parameter space if the posterior PDF is
+  complex in morphology.
+We don't have very useful advice here, except to
+  spend time learning about the multiple modes in the posterior PDF
+  and what each ``means'' or corresponds to,
+  and to do your best to express in your results the multiple optima.
+One of the miracles of MCMC is that if you \emph{can} fairly sample
+  as complex posterior PDF,
+  the fraction of samples in each mode
+  tends to the relative total integrated posterior probability inside each mode,
+  as the sampling converges.
+
+Ensemble samplers react to multiple modes
+  by obtaining very low acceptance ratio,
+  because the samplers in the different modes are not easily able to ``help one another''
+  move efficiently.
+Again, if the acceptance ratio is low and there appear to be multiple modes,
+  it is likely that the sampling will also be initialization dependent,
+  and it is very unlikely that your sampling is going to be converged.
 
 \paragraph{parameterization problems:}
-...Degeneracies and duplications in parameter space; reparameterizations.
+It helps to think carefully about how to parameterize.
+For example, if there is an angle $\phi$ in your problem,
+  and your prior requires it to be between $0$ and $2\pi$,
+  and the posterior mode is very near $0$,
+  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
+  makes an intrinsically unimodal problem multimodal!
+In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
+  to two vector components $A\,\cos\phi$ and $A\,\sin\phi$.
+Of course this re-parameterization changes significantly the \emph{form} of the prior PDF
+  (the transformation brings in a Jacobian);
+  it has to be adjusted with care.
+
+DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?
+
+%% ...Priors: Once again:
+%% If you describe your priors in one parameterization but
+%% then perform sampling in a different space, it is easy to make
+%% mistakes.  For example: Flat prior in amplitude and phase, sample in
+%% $A\,\cos\phi$ and $A\,\sin\phi$.  There are many ways to do this
+%% wrong.
+
+For another example, sometimes there is almost an exact degeneracy between two parameters,
+  say $a$ and $b$.  Then it makes sense to switch to parameters $a+b$ and $a-b$,
+  or some other linear combinations where correlations or near-degeneracies are likely to be reduced.
+Again, such transformations require also prior transformations, which must be made with care.
+There are some affine invariant samplers\note{CITE EMCEE AGAIN HERE} that are invariant to such transformations,
+  but if you aren't using such a sampler,
+  it is worth getting out ahead of these problems.
+They can be found by performing an exploratory sampling,
+  making a triangle plot as described above,
+  and looking for narrow diagonal lines in the two-dimentional scatter plots.
 
@@
 
-\section{Advice and discussion}
-
-...Write and debug your own M--H sampler and then download a real
-package \emph{afterwards}.
-% ***** BJB
-% I AGREE WHOLEHEARTEDLY. That's how to learn.
-
-...Simple problems where the outputs are analytic.  Functional testing.
-
-...breaking Markov-ness.
-% ***** BJB
-% "Diminishing adaptations" is an idea that is quite easy
-% to explain. There are other ways to break Markov-ness but
-% this probably covers enough of them for an intro document
-% ***** Hogg
-% I was going to say you should NEVER break Markov-ness
-% unless you REALLY understand what you are doing.
-
-...Priors: Once again:
-If you describe your priors in one parameterization but
-then perform sampling in a different space, it is easy to make
-mistakes.  For example: Flat prior in amplitude and phase, sample in
-$A\,\cos\phi$ and $A\,\sin\phi$.  There are many ways to do this
-wrong.
-
-...Sample the prior (likelihood set to unity) to check priors and parameterization (another functional test).
-
-...Posteriors: On the other hand, you can do anything you like (within
-reason) to the output samples and be sampling the posterior in that
-transformed quantity.
-
-...How to deal with forbidden regions in prior, or sharp prior boundaries?
-
-...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.
-
-...Initialization and burn-in advice we haven't addressed earlier?
-
 ...The Bayes integral / evidence integral?
@@
 
-...Finally, to repeat: You can't sample the likelihood!  You can only
-sample a posterior pdf!
+%% HOGG DELETED / COMMENTED OUT THE BELOW
+%% - make sure everything below is discussed somewhere above and then delete it below.
 
-% \begin{figure}[htbp]
-% \exampleplot{ex17}
-% \caption{Partial solution to \problemname~\ref{prob:bayesintrinsic}:
-% The marginalized posterior probability distribution for the intrinsic
-% variance.}\label{fig:bayesintrinsic}
-% \end{figure}
+%% \section{Advice and discussion}
+
+%% ...breaking Markov-ness.
+%% % ***** BJB
+%% % "Diminishing adaptations" is an idea that is quite easy
+%% % to explain. There are other ways to break Markov-ness but
+%% % this probably covers enough of them for an intro document
+%% % ***** Hogg
+%% % I was going to say you should NEVER break Markov-ness
+%% % unless you REALLY understand what you are doing.
+
+%% ...Sample the prior (likelihood set to unity) to check priors and parameterization (another functional test).
+
+%% ...Posteriors: On the other hand, you can do anything you like (within
+%% reason) to the output samples and be sampling the posterior in that
+%% transformed quantity.
+
+%% ...How to deal with forbidden regions in prior, or sharp prior boundaries?
+
+%% ...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.
+
+%% ...Initialization and burn-in advice we haven't addressed earlier?
+
+%% ...Finally, to repeat: You can't sample the likelihood!  You can only
+%% sample a posterior pdf!
+
+%% % \begin{figure}[htbp]
+%% % \exampleplot{ex17}
+%% % \caption{Partial solution to \problemname~\ref{prob:bayesintrinsic}:
+%% % The marginalized posterior probability distribution for the intrinsic
+%% % variance.}\label{fig:bayesintrinsic}
+%% % \end{figure}
 

### 2013-12-30 mcmc.tex
estimators

@@
 % That comment should go in the ``What is a Model?'' Chapter.
+% (Or a \note{} right here.)
 Furthermore, say you also can write down some kind of informative or
@@
   or posterior median value for the parameter.
-\\HOGG CHOOSE NOTATION AND GIVE FORMULAE HERE.
+
+For example,
+  imagine that by MCMC you have generated
+  $K$ samples $\pars_k$ of a parameter vector $\pars$.
+Now you have a scalar function $q(\theta)$ which takes the parameter vector
+  and returns a scalar value%
+  \note{For the median-of-sampling value, $q(\theta)$ needs to return a scalar,
+  but technically, for the mean-of-sampling, $q(\theta)$ can be a vector or something high-dimensional.}
+  which could be as simple as a single component of the parameter vector
+  (one parameter from the list),
+  or something more complex.
+The mean-of-sampling value $\mean{q}$ is just
+\begin{eqnarray}
+\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K q(\theta_k)
+\quad ,
+\end{eqnarray}
+  and the median-of-sampling value is just the $[K/2]$th value of $q(\theta_k)$
+  when the $q(\theta_k)$ have been ordered (sorted).
+These are both produced by integrals;
+  the first is the value returned by the expectation integral estimate,
+  the second is the value past which half of the integrated pdf lies.
 
@@
   that contains 68 percent of the posterior samples.
-\\HOGG GIVE FORMULA HERE.\\
 The ``two-sigma'' error bar would be the same but for the 95-percent interval.
+Again, these limits can be estimated%
+  \note{One amusing thing about samplings,
+  which are so beloved of us probabilistic (Bayesian) reasoners,
+  is that anything we \emph{do} with a sampling,
+  like estimate an integral or a quantile,
+  is just an old-school frequentist estimator.
+  A frequentist estimator of a Bayesian quantity, to be sure, but a frequentist estimator nonetheless.}
+  by ordering the $q(\theta_k)$ values,
+  and finding the $\theta$ values such that some fraction (0.68 or 0.95)y
+  of the samples lie between them.
+Again, to form these limits, $q(\theta)$ must be a scalar function (so sorting is well defined).
 
@@
   to choose the \emph{median} of sampling as the measurement value,
-  the 16-percent quantile as the lower error bar,
-  and the 84-percent quantile as the upper error bar.
+  the 16-percent quantile as the lower one-sigma error bar,
+  and the 84-percent quantile as the upper one-sigma error bar.
 This has pathologies in higher dimensions (as we are about to see),
@@
 
-Another proxy for autocorrelation time is the
+Another proxy for autocorrelation time useful for tuning is the
   Expected Squared Jump Distance.\note{(CITE GELMAN)}
@@
   because you should only call it a ``likelihood function'' in contexts in which you are
-  treating the data as fixed (at the true data values) and the parameters as variable.}
+  treating the data as fixed (at the true data values) and the parameters as variable.
+  In this context, the likelihood function is \emph{not} a pdf for the parameters,
+  so it can't be sampled.}
 
@@
 In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
-  to two vector components $A\,\cos\phi$ and $A\,\sin\phi$.
+  to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
 Of course this re-parameterization changes significantly the \emph{form} of the prior PDF
@@
 DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?
+Choose a $p(A)$ and a $p(\phi) = [2\pi]^{-1}$,
+   and then transform to $p(a, b)$.
 
@@
 
-%% \section{Advice and discussion}
-
 %% ...breaking Markov-ness.
@@
 
-%% ...Sample the prior (likelihood set to unity) to check priors and parameterization (another functional test).
-
-%% ...Posteriors: On the other hand, you can do anything you like (within
+%% ...Posteriors:  You can do anything you like (within
 %% reason) to the output samples and be sampling the posterior in that

### 2014-12-26 mcmc.tex
typo and note to self

@@
 
-Finally---and perhaps most importantly---another magical propert of a sampling
+Finally---and perhaps most importantly---another magical property of a sampling
   (in a $d$-dimensional space)
@@
 ...What is a typical proposal distribution?
+
+...How to avoid violating detailed balance (without deeply understanding it)?
 

### 2014-12-26 mcmc.tex
notes and content

@@
+% This file is part of the MCMC project,
+%   which is part of the Data Analysis Recipes project
+% Copyright 2013, 2014 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).
+
 % to-do
@@
 %   - Brewer
+%   - Hill
 % - post on arXiv
@@
 % - is it ``a MCMC sampling'' or ``an MCMC sampling''?
+% - \note{} right after word if you are endnoting a phrase,
+%   but after the full stop if endnoting the sentence or paragraph.
 
@@
     license\note{\label{note:first}
-        Copyright 2013 by the authors. This work is licensed under a
+        Copyright 2013, 2014 by the authors. This work is licensed under a
         \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
@@
 \affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
-       New York University}\\[1ex]
+       New~York~University}\\[1ex]
 David~W.~Hogg\\
 \affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
-       New York University}\\
-\affil{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
+       New~York~University}\\
+\affil{Center~for~Data~Science,
+       New~York~University}\\
+\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}
 
@@
 Not being mathematicians, we don't know the precise definition of a
-pdf\note{Seriously.  We don't.}, but for our purposes here a pdf is
+pdf\note{Seriously.  We don't.  It has something to do with measure theory.}, but for our purposes here a pdf is
 any single-valued (scalar) function that is non-negative everywhere
@@
 0 &\leq& p(x) \quad \mbox{for all $x$}
-\\
+\\ \label{eq:normp}
 1 &=& \int p(x)\,\dd x
@@
 parameters in the model.
+Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
+has as many dimensions as there are elements or entries or components of $x$.
+Also, if there are elements or entries or components of $x$ that are discrete
+(that is, take on only integer values or equivalent),
+then along those dimensions the integral becomes a discrete sum.
+This latter is a detail to which we return below.
 
@@
 
-\section{What is M--H MCMC?}
+\section{What is Metropolis--Hastings MCMC?}
+
+The simplest algorithm for MCMC is the Metropolis--Hastings algorithm.\note{%
+  HOGG put here some short discussion of the name of the algorithm and the original reference.}
+It is so simple, that we recommend that any reader of this document
+who has not previously implemented the algorithm take a break at the end
+of this \sectionname\ and implement it forthwith, in a short piece of computer code,
+in the context of some simple problem.\note{%
+  HOGG put reference to fitting a line here and point inside to the simple problem suggestion.}
 
@@
 
+HOGG:  How to deal with discrete or integer or binary parameters?
+
 \paragraph{model checking:}

### 2014-12-26 mcmc.tex
algo

@@
 %   but after the full stop if endnoting the sentence or paragraph.
+% - do we work in $x$ or $\pars$?  I am confused...
 
@@
 
-The simplest algorithm for MCMC is the Metropolis--Hastings algorithm.\note{%
+The simplest algorithm for MCMC is the Metropolis--Hastings algorithm (M--H MCMC).\note{%
   HOGG put here some short discussion of the name of the algorithm and the original reference.}
-It is so simple, that we recommend that any reader of this document
+It is so simple that we recommend that any reader of this document
   who has not previously implemented the algorithm take a break at the end
@@
 
-...The algorithm
+The M--H MCMC algorithm requires two inputs.
+The first is a handle to the function $f(\pars)$ that is the function to be sampled,
+  such that the algorithm can evaluate $f(\pars)$ for any value of the parameters $\pars$.
+In data-analysis contexts,
+  this function would be the prior $p(\pars)$ times the likelihood $p(\data\given\pars)$
+  evaluated at the observed data $\data$.
+The second input is a proposal function $p(\pars'\given\pars)$,
+  such that the algorithm can draw a new position $\pars'$ in the parameter space
+  given an ``old'' position $\pars$.
+This second function must meet some symmetry requirements we discuss further below.
+It permits us to random-walk around the parameter space in a fair way.
+
+The M--H MCMC algorithm is the following:
+We have generated some set of samples, the most recent of which is $\pars_k$
+To generate the next sample $\pars_{k+1}$ do the following:
+\begin{itemize}
+\item Draw a proposal $\pars'$ from the proposal disribution $p(\pars'\given\pars_k)$.
+\item Draw a random number $0<r<1$ from the uniform distribution.
+\item If $p(\pars') / p(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
+      otherwise $\pars_{k+1} \leftarrow \pars_k$.
+\end{itemize}
+That is, at each step, either a new proposed position in the parameter
+  space gets accepted or else the prior position in the parameter space
+  gets repeated.
+The algorithm can be iterated a large number $K$ of times to produce $K$ samples.
+
+...The algorithm is a biased random walk; nearby samples are covariant...etc.
 
@@
 ...How to avoid violating detailed balance (without deeply understanding it)?
+
+...Cycling through parameters??  Mention here?
 
@@
 
-\section{Ensembles, Hamiltonians, and other variants}
+\section{Gibbs, Ensembles, Hamiltonians, and other variants}
 
@@
   $K$ samples $\pars_k$ of a parameter vector $\pars$.
-Now you have a scalar function $q(\theta)$ which takes the parameter vector
+Now you have a scalar function $q(\pars)$ which takes the parameter vector
   and returns a scalar value%
-  \note{For the median-of-sampling value, $q(\theta)$ needs to return a scalar,
-  but technically, for the mean-of-sampling, $q(\theta)$ can be a vector or something high-dimensional.}
+  \note{For the median-of-sampling value, $q(\pars)$ needs to return a scalar,
+  but technically, for the mean-of-sampling, $q(\pars)$ can be a vector or something high-dimensional.}
   which could be as simple as a single component of the parameter vector
@@
 \begin{eqnarray}
-\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K q(\theta_k)
+\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
 \quad ,
 \end{eqnarray}
-  and the median-of-sampling value is just the $[K/2]$th value of $q(\theta_k)$
-  when the $q(\theta_k)$ have been ordered (sorted).
+  and the median-of-sampling value is just the $[K/2]$th value of $q(\pars_k)$
+  when the $q(\pars_k)$ have been ordered (sorted).
 These are both produced by integrals;
@@
   A frequentist estimator of a Bayesian quantity, to be sure, but a frequentist estimator nonetheless.}
-  by ordering the $q(\theta_k)$ values,
-  and finding the $\theta$ values such that some fraction (0.68 or 0.95)y
+  by ordering the $q(\pars_k)$ values,
+  and finding the $\pars$ values such that some fraction (0.68 or 0.95)y
   of the samples lie between them.
-Again, to form these limits, $q(\theta)$ must be a scalar function (so sorting is well defined).
+Again, to form these limits, $q(\pars)$ must be a scalar function (so sorting is well defined).
 

### 2014-12-26 mcmc.tex
tweak

@@
 That is, at each step, either a new proposed position in the parameter
-  space gets accepted or else the prior position in the parameter space
-  gets repeated.
+  space gets accepted into the list of samples or else the prior sample
+  in the parameter space gets repeated.
 The algorithm can be iterated a large number $K$ of times to produce $K$ samples.

### 2014-12-26 mcmc.tex
nearly finished M-H section

--- 2014-12-26 mcmc.tex
@@
 % - [LaTeX] newline after every full stop for proper git diffing
-% - [LaTeX] eqnarray not equation
+% - [LaTeX] \eqnarray not \equation
 % - pdf not PDF
@@
 
-Finally---and perhaps most importantly---another magical property of a sampling
+Integrals are also involved in finding the mean, median, and any quantiles
+  of a pdf.
+They are not involved in finding the \emph{mode} of a pdf.
+For this reason (and others), in what follows, when we talk about what to report
+  about the outcome of your MCMC sampling,
+  we will advise in favor of mean, median, and quantiles,
+  and we will advise against modes.
+
+Finally---and perhaps most importantly---a magical property of a sampling
   (in a $d$-dimensional space)
@@
   evaluated at the observed data $\data$.
-The second input is a proposal function $p(\pars'\given\pars)$,
+The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
   such that the algorithm can draw a new position $\pars'$ in the parameter space
   given an ``old'' position $\pars$.
-This second function must meet some symmetry requirements we discuss further below.
+This second function must meet a symmetry requirement (detailed
+  balance) we discuss further below.
 It permits us to random-walk around the parameter space in a fair way.
 
-The M--H MCMC algorithm is the following:
+The algorithm is the following:
 We have generated some set of samples, the most recent of which is $\pars_k$
@@
 \begin{itemize}
-\item Draw a proposal $\pars'$ from the proposal disribution $p(\pars'\given\pars_k)$.
+\item Draw a proposal $\pars'$ from the proposal pdf $q(\pars'\given\pars_k)$.
 \item Draw a random number $0<r<1$ from the uniform distribution.
-\item If $p(\pars') / p(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
+\item If $f(\pars') / f(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
       otherwise $\pars_{k+1} \leftarrow \pars_k$.
@@
 That is, at each step, either a new proposed position in the parameter
-  space gets accepted or else the prior position in the parameter space
-  gets repeated.
+  space gets accepted into the list of samples or else the previous sample
+  in the parameter space gets repeated.
 The algorithm can be iterated a large number $K$ of times to produce $K$ samples.
 
-...The algorithm is a biased random walk; nearby samples are covariant...etc.
-
-...Why does it work (briefly)?
-
-...What is a typical proposal distribution?
-
-...How to avoid violating detailed balance (without deeply understanding it)?
-
-...Cycling through parameters??  Mention here?
+This algorithm---and indeed any MCMC algorithm---produces a
+  \emph{biased random walk} through parameter space.
+It is a random walk for the same reason that it is ``Markov'':
+The step it makes to position $\pars_{k+1}$ depends only on the state
+  of the sampler at position $\pars_k$ (and no previous state).
+It is a biased random walk, biased by the acceptance algorithm
+  involving the ratios of function values; this acceptance rule biases
+  the random walk such that the amount of time spent in the neighborhood
+  of location $\pars$ is proportional to $f(\pars)$.
+
+The principal user-settable knob in M--H MCMC is the proposal
+  pdf $q(\pars'\given\pars)$.
+A typical choice is a multi-variate Gaussian distribution for $\pars'$
+  centered on $\pars$ with some simple (diagonal, perhaps) variance
+  tensor.
+We will discuss the choice and tuning of this proposal distribution
+  below.
+
+Importantly---for the algorithm given above to work correctly---the
+  proposal pdf must satisfy a ``detailed balance'' condition; it
+  must have the property that
+\begin{eqnarray}
+q(\pars'\given\pars) &=& q(\pars\given\pars')
+\quad ;
+\end{eqnarray}
+that is, it must be just as easy to go one way in the parameter space
+  as the other.
+You can break this property, if you like, and then adjust the
+  acceptance condition accordingly, but we \emph{do not recommend this}
+  except under the supervision of a trained professional.\note{Actually,
+  our own favorite method, \project{emcee}, has a proposal pdf that does
+  violate detailed balance in this way and has a compensated
+  acceptance ratio.  Read the paper [HOGG CITE] for more details.}
+If the detailed balance condition frightens you (as it frightens us),
+  then just stick with pdfs that are symmetric in $\pars'$ and $\pars$,
+  like the Gaussian, or a centered uniform distribution and forget about
+  it.
+
+In what follows, when we discuss tuning of the proposal pdf, we will
+  often cycle through the parameter blob components and propose changes
+  in only one dimension or element or component at a time.
+That is, at step $k+1$ you might only propose a one-dimensional move
+  in the $i$th dimension, not in all $d$ dimensions of the parameter
+  space.
+That won't change anything significant; but if you are implementing
+  this right now you might want to do that.
+It will help us tune the proposal pdf and diagnose problems.
+
+...Why does M--H MCMC work (briefly)? ...Or put this in a note on an earlier paragraph?
 

### 2014-12-26 mcmc.tex
independence
agreement

@@
   we will advise in favor of mean, median, and quantiles,
-  and we will advise against modes.
+  and we will advise against mode.
 
@@
   of location $\pars$ is proportional to $f(\pars)$.
+Because of this local (Markov) property, the nearby samples are not
+  independent; the algorithm only produces fair samples in the limit of
+  arbitrary run time, and two samples are only independent when they are
+  sufficiently separated in the chain (more on this below).
 

### 2014-12-26 mcmc.tex
methods content

@@
 % - do we work in $x$ or $\pars$?  I am confused...
+% - should we use the word ``walker'' anywhere, everywhere, or never?
 
@@
 
-\thispagestyle{plain}\raggedbottom
+\thispagestyle{plain}\sloppy\sloppypar\raggedbottom
 \section*{Data analysis recipes:\\ \this\footnotemark}
@@
 
-\section{Gibbs, Ensembles, Hamiltonians, and other variants}
+\section{Gibbs, ensemble, Hamiltonian, and so on}
+
+MCMC methods are proven to be correct ``in the limit''; that is, when
+an infinite number of samples have been taken.
+When the function $f(\pars)$ has challenging properties, the results
+can approach this limit very slowly (or not at all\note{HOGG: discuss
+  situations in which MCMC will not work---related to support.}).
+For this reason, in the finite duration of a scientific project, most
+users of MCMC have one of two (related) problems:
+Either it is taking too long---they have take too many steps or
+execute too many calls of the function $f(\pars)$---or it is not
+exploring the full parameter space---there are local optima (or bad
+local geometry around those optima) ``trapping'' the algorithm.
+
+There is no general, problem-independent solution for either of these
+problems.
+Indeed, there are many different strategies that work well or badly
+for different functions $f(\pars)$.
+In this \sectionname, we will try to guide the reader towards methods
+that might help in different circumstances.
+We will not give a full description of all the MCMC algorithms out
+there, but just provide some field notes and pointers to references.
+New MCMC methods are being developed all the time, so a reader with a
+serious problem would do well to consult with the applied-mathematics
+or statistics literature.\note{HOGG: Your best interface to these
+  literatures is via a colleague.  Break down those walls!}
+The different methods to which we refer here are (in general) very
+different in their difficulty of implementation.
+We won't help with that either, except to say that there are more and
+more open-source or distributed packages every year.
+We will try to name some current implementations in the notes, but any
+list compiled today will be incomplete tomorrow.
+The spaces of MCMC methods and MCMC method implementations are both
+growing rapidly, in part because MCMC has become such a core
+technology in the empirical sciences.
 
@@
 
-...What, briefly, is the stretch move, and why is it useful?
+...Gibbs!  Every parameter is differently difficult to sample in?
+
+...What, briefly, is the stretch move, and why is it useful?  Every parameter treated identically---almost opposite to gibbs.
 

### 2014-12-26 mcmc.tex
methods content

@@
 
+[SHOULD THIS SECTION MOVE TO AFTER THE DEFINITION OF AUTOCORRELATION TIME AND DISCUSSION OF CONVERGENCE?]
+
 MCMC methods are proven to be correct ``in the limit''; that is, when
@@
 users of MCMC have one of two (related) problems:
-Either it is taking too long---they have take too many steps or
-execute too many calls of the function $f(\pars)$---or it is not
+Either \textsl{(1)}~it is taking too long---it is taking too many steps or
+executing too many calls of the function $f(\pars)$ to get independent samples---or \textsl{(2)}~it is not
 exploring the full parameter space---there are local optima (or bad
@@
 or statistics literature.\note{HOGG: Your best interface to these
-  literatures is via a colleague.  Break down those walls!}
+  literatures is a colleague in another department.  Break down those walls!}
 The different methods to which we refer here are (in general) very
@@
 technology in the empirical sciences.
+
+...Note a good general reference here?  MacKay?
+
+\paragraph{Ensembles:}
+As we discuss more below, one of the critical challenges in the use of
+MCMC is tuning the proposal pdf $q(\pars'\given\pars)$.
+In problems in which all of the parameters (all of the components or entries of $\pars$) are somehow ``equivalent'',
+or the parameters can be seen as components of a vector in a vector
+space, there are \emph{ensemble methods} that make use of not just one
+random-walker but instead many walkers to automatically generate a properly tuned
+proposal distribution.
+These methods make use of the distribution of a set of independent
+walkers in the parameter space ($\pars$-space) to gauge the typical step-sizes and
+directions at which new proposals should be made, obviating or greatly
+reducing the tuning requirements.
+We are co-authors on one of these methods, \project{emcee}\note{HOGG:
+  ref emcee}, which is popular in the astrophysical sciences.
+Because ensemble methods are updating independent MCMC chains, they
+are often easy to parallelize to make use of multiple cores.
+
+\paragraph{Gibbs:}
+In many problems, the parameters---the components or entries of the
+blob $\pars$---are not equivalent at all.
+Some are global parameters, which touch or have an effect on every
+sub-part of the data blob $\data$, while others are local parameteres,
+which only touch a small subset of the data.
+Or, in another use case, some parameters are linear parameters
+that---at fixed values of the other parameters---will have a Gaussian
+likelihood or posterior pdf.
+In these problems, there are some directions in parameter space that
+are hard to move in (they require complete re-calculation of the full
+non-linear function $f(\pars)$, and other directions that are very
+easy to move in (they require only partial re-calculations or the
+sampling could even be analytic and exact).
+For problems with these structures, Gibbs samplers are ideal, and
+there are good implementations.\note{HOGG cite JAGS}
+
+A trivial kind of Gibbs sampling was mentioned above in the
+discussion of M--H MCMC:
+Even when running vanilla M--H MCMC, it is often useful to update only
+one parameter at a time; that is, do only axis-aligned moves in the
+$\pars$-space, cycling through axes as you go from step to step.
+This helps with tuning and diagnosis, but it also permits a good
+implementation to capitalize on simpler calculations for the simpler
+parameter directions.
+
+Gibbs sampling can be very good for exploiting multiprocessing:
+Global parameter updates might require a serial re-calculation of the
+full likelihood, but local parameter updates can be done in parallel,
+with each local parameter working on its local bit of data all at the
+same time.
+This kind of structure is common in hierarchical inference, where
+local parameters touch only individual objects in some population,
+say, and global parameters are parameters \emph{of} that population as
+a whole.
+Parallelization is built in to some Gibbs implementations.\note{HOGG}
+
+\paragraph{Hamiltonian:}
+[HOGG: THIS HAS SOME OTHER NAME; WHAT?]
+In many cases of interest, the function $f(\pars)$ can be analytically
+differentiated, such that it is possible to quickly execute a gradient
+$\dd f/\dd\pars$ evaluation.
+Indeed, in the modern world of computing, there are even
+auto-differentiation systems that deliver automatically code that
+produces the gradient of any function you can write.\note{HOGG cite
+  autodiff systems}
+There is a class of MCMC methods that make use of gradient information
+to effectively inform the proposal distribution and thus speed sampling.
+... HOGG more here!
 
@@
 
-...Gibbs!  Every parameter is differently difficult to sample in?
-
-...What, briefly, is the stretch move, and why is it useful?  Every parameter treated identically---almost opposite to gibbs.
-
-...What, briefly, is importance sampling?
-
-...What, briefly, is nested sampling, and why is it good?
-
-...What, briefly, is tempering?
-
-...What, briefly, is hamiltonian, and why is it good?
-
-...What, briefly, is reversible-jump, and why is it useful?
+\paragraph{Importance:}
+
+\paragraph{Tempering:}
+
+\paragraph{Nested:}
+
+\paragraph{Reversible-jump:}
+Finally, there are some problems that \emph{cannot be done} without
+additional proposal technology...
 

### 2014-12-27 mcmc.tex
variance

@@
 What are your best options for the ``uncertainty'' or ``limits'' on this parameter?
-The only simple integral-based options are posterior quantiles.
-That is, the ``one-sigma'' error bar can be the size of the central (or smallest%
+The only simple integral-based options are either variances or else posterior quantiles.
+We usually use quantiles.
+That is, the ``one-sigma'' error bar can be taken to be the half-size of the central (or smallest%
   \note{HOGG shizzle about choosing quantiles}) interval of the parameter
@@
   by ordering the $q(\pars_k)$ values,
-  and finding the $\pars$ values such that some fraction (0.68 or 0.95)y
+  and finding the $\pars$ values such that some fraction (0.68 or 0.95)
   of the samples lie between them.

### 2015-01-02 mcmc.tex
moving comment
notes to self

@@
 
+% *********** BJB talking:
+% I suggest partitioning the discussion of different methods into the following
+% two categories:
+%
+%  i) Methods for getting around the target distribution
+% ii) Methods that change the target distribution to something
+% other than the posterior pdf, to help the sampling somehow
+%
+% i) Would include things like M-H, the stretch move,
+% Hamiltonian, slice sampling (maybe skip the latter)
+% ii) would include things like annealing and Nested Sampling
+
+%%%% Hogg talking:
+% DFM: In this section, I think we should just summarize the other
+% methods, not describe them in detail.  Then we can cite the relevant
+% papers for the enthusiastic.
+
 \paragraph{Ensembles:}
@@
 \paragraph{Hamiltonian:}
-[HOGG: THIS HAS SOME OTHER NAME; WHAT?]
+[HOGG, DFM: THIS HAS SOME OTHER NAME; WHAT?]
 In many cases of interest, the function $f(\pars)$ can be analytically
@@
 to effectively inform the proposal distribution and thus speed sampling.
+
 ... HOGG more here!
-
-% *********** BJB talking:
-% I suggest partitioning the discussion of different methods into the following
-% two categories:
-%
-%  i) Methods for getting around the target distribution
-% ii) Methods that change the target distribution to something
-% other than the posterior pdf, to help the sampling somehow
-%
-% i) Would include things like M-H, the stretch move,
-% Hamiltonian, slice sampling (maybe skip the latter)
-% ii) would include things like annealing and Nested Sampling
-
-%%%% Hogg talking:
-% DFM: In this section, I think we should just summarize the other
-% methods, not describe them in detail.  Then we can cite the relevant
-% papers for the enthusiastic.
 
@@
 
+% HOGG:  Is the following a repeat from previously and/or does it mesh with what's earlier?
+
 A very common---%

### 2015-01-02 mcmc.tex
presenting results in high dimensions

@@
 %   which is part of the Data Analysis Recipes project
-% Copyright 2013, 2014 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).
+% Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).
 
@@
 
-...Now for multiple dimensions...return to the marginalization point
-from above...
-
-...What do you do in high-dimensional spaces?  Note the issue
-that---even in a unimodal posterior PDF---the most probable model
-might not be at the set of parameter values that are reported if you
-compute the parameter values to report in one-dimensional projections
-of the posterior pdf.  This requires a figure to demonstrate.  What to
-do here?
-
-...What figures should you make to display the chain?
-
-...If you are really good, you will pass forward a sampling.
-...How big should it be? Refer to Mackay here.
-...Usefulness of a big sampling for hierarchical inferences; concept of interim prior?
+The parameter space $\pars$ is usually multi-dimensional, and more
+than one of those parameters is of interest.
+In this case, the mean or median of sampling can be produced for each
+dimension.
+Because the sampling projected onto any dimension is a marginalization
+of the posterior pdf, any such one-dimensional mean or median is the
+mean or median of the marginalized posterior pdf.
+
+There is one oddity to note here, as it catches many investigators by
+surprise:
+Even if the model is good and the posterior pdf is unimodal and well-behaved,
+the median (or even mean) of the posterior pdf for each of parameters,
+when taken together as a ``best-fit'' parameter vector $\theta^\ast$,
+will not itself necessarily be a good fit to the data!
+If there is substantial ``curvature'' to the pdf in the parameter
+space, the mean or median of sampling does not necessarily lie at a
+high-probability location of parameter space.
+This may seem counterintuitive, but it is easy to see in the case of a
+``banana-shaped'' posterior pdf.
+This all relates to the fact that MCMC is not an optimizer, it is a
+sampler.
+It is important, when reporting the output of an MCMC run by giving
+means or medians of the posterior PDF to remind the reder or user of
+that output that it does not necessarily represent (collectively) a
+good fit to the data; these outputs only give information that is
+useful one parameter at a time.
+
+In principle the only output from the sampling that safely gives both
+probabilistic information about the result of the inference and also
+good-fitting models is a few---randomly chosen---example samples.
+We strongly recommend this\note{HOGG:  And we do this WHERE? Citation!};
+in particular this is a better thing to do
+than to give the ``best sample'', which isn't even guaranteed to be
+near the bulk of the posterior pdf or the samples therefrom.
+
+If you \emph{do} return posterior samples to your audience, how many
+should you return?
+There are only heuristic answers to this, but a guideline is that, if
+you only care about a single (one-dimensional) parameter, you don't
+need many samples; a dozen suffices to give you a reasonable posterior
+pdf mean and variance.\note{HOGG cite Mackay.}
+If you care about $K$ parameters or dimensions of the parameter space,
+in general the need for samples grows exponentially (or perhaps even
+factorially!) with the number of dimensions.\note{HOGG cite something.}
+So if you have a ten-dimensional space, expect to be publishing your
+samples in an electronic table on the web!
+If you \emph{do} publish sufficient numbers of posterior samples, you
+might find users who want to make serious use of them.\note{HOGG cite
+  our importance-sampling papers with some explanation here.}
+
+Of course figures are always better than tables, at least in the
+printed form of scientific publications.
+In a $D$-dimensional parameter space, we recommend plotting all $D$
+one-dimensional sample histograms, on a pretty fine binning, and all
+$D$-choose-2 two-dimensional histograms (scatter plots) to show the
+low-level covariances and non-linearities.
+If you are clever, all these plots can be arranged into a lovely and
+informative triangle.\note{HOGG cite triangle.py and put an example in
+  the note!}
+These two-dimensional plots do not in any sense contain all the
+information in the sampling!  However, they are remarkable for
+locating expected and unexpected parameter relationships, and often
+invaluable for suggesting re-parameterizations and transformations
+that simplify your problem.
 

### 2015-01-02 mcmc.tex
notes to self

@@
 \paragraph{Importance:}
+... HOGG
 
 \paragraph{Tempering:}
+... HOGG
 
 \paragraph{Nested:}
+... HOGG
 
@@
 
-Making this heuristic condition more precise...
-Autocorrelation time is the premier tool...
-How is Autocorrelation time defined?..
-How is Autocorrelation time measured?..
-Why is it is not a panacea?..
+...Making this heuristic condition more precise
+...Autocorrelation time is the premier tool
+...How is Autocorrelation time defined?
+...How is Autocorrelation time measured?
+...Why is it is not a panacea? ...estimators ...getting stuck in one mode
 
@@
 \section{Tuning}
-
-% BJB
-% Please suggest mixtures of scales for M-H proposals.
-% You can mimic the properties of slice sampling but with
-% much easier code.
 
@@
 
+% BJB
+% Please suggest mixtures of scales for M-H proposals.
+% You can mimic the properties of slice sampling but with
+% much easier code.
+
 In a deep sense, the only scalar that makes sense to optimize,
@@
 
-The simplest proxy for autocorrelation time optimization is the \emph{acceptance fraction}.
+The simplest heuristic proxy statistic for tuning is the \emph{acceptance fraction}.
 If you are accepting almost all proposed steps, your step sizes are too small on average.
 If you are accepting almost none, your step sizes are too large.
-The Goldilocks value is around a half,
-  with a strong argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}.
+The Goldilocks value is around a quarter,
+  with an argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}
+  (though you could never tune it precisely enough to warrant that third digit of accuracy).
 In the burn-in phase (discussed below) of an MCMC run,

### 2015-01-03 mcmc.tex
tweaks
importance, nested, and f() is posterior

@@
 % -----
+% - perhaps re-order to m-h, convergence, reporting results, other methods ??
 % - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
@@
 % - [LaTeX] newline after every full stop for proper git diffing
-% - [LaTeX] \eqnarray not \equation
+% - [LaTeX] eqnarray not equation
 % - pdf not PDF
@@
   where there are always parameters with very different levels of importance to the scientific conclusions.
-This point will return again below, either explicitly or implicitly. % does it?
+This point will return again below, either explicitly or implicitly. % HOGG: does it?
+
+Although the discussion in this \documentname\ is general, the most
+common use of MCMC sampling (for us, anyway) is in probabilistic
+inference.
+For this reason, we will often refer to the function $f(\pars)$
+colloquially as ``the posterior pdf''\note{We will also sometimes
+  refer to expectations under $f(\pars)$ as posterior means, and
+  medians as median-of-posterior values, and so on.} even though it is
+implicitly ill-normalized and might not be a posterior pdf in any
+sense.
+We will also occasionally assume---just because it is true in
+inference---that the function $f(\theta)$ is the product of two
+functions, one called ``the prior pdf'' and one called ``the
+likelihood''.
+Again, this usage is colloquial and is only strictly correct in
+inference contexts with proper inputs.
+That the function $f(\theta)$ can be thought of as a product of a
+prior pdf and a likelihood is only necessary for what follows in
+the context of advanced sampling techniques like tempering or
+nested sampling, both mentioned briefly below.
 
@@
 
+\paragraph{Slice:}
+... DFM, can you help here?
+
 \paragraph{Importance:}
-... HOGG
+Sometimes it is possible to sample efficiently from a distribution
+that is close to the target function $f(\pars)$ you care about but
+hard to sample efficiently with MCMC.
+An example is a problem that has a simple (Gaussian, say) likelihood
+function times a non-trivial prior:
+It is easy to sample from a Gaussian likelihood times a Gaussian
+prior, even without MCMC, but it might be hard to do the same sampling
+under a more complex prior.
+Importance sampling is a direct sampling method (not really an MCMC
+method at all):
+You take a sampling from the approximate function (which by assumption
+was easy) and then you re-weight the sampling (or do a subsequent
+rejection of samples) using the ratio of the true function to the
+approximate function as a weight or probability.
+We use this quite a bit in our hierarchical inferences.\note{HOGG:
+  Citations}
+
+The only significant technical details for importance sampling are the
+following:
+The approximate function must have the same (or greater) support as
+the true function, and the method degrades in efficiency as the two
+functions diverge.
+If the ratio of the true function to the approximate function is to be
+used as a probability for rejection of samples, then the true function
+must be greater than or equal to the approximate function everywhere!
 
@@
 \paragraph{Nested:}
-... HOGG
+Similar to tempering are a class of methods called ``nested''
+samplings, which also smoothly change the target distribution.
+In the nested case, instead of increasing the temperature, the
+samplings are of a censored version of the prior, censored by the
+value of the likelihoood function.
 
@@
 For a committed probabilistic scientist---frequentist or
-bayesian---there is no ``answer'', there are only probability density
+Bayesian---there is no ``answer'', there are only probability density
 functions (pdfs).  Any time in a paper or in an email or in a conversation

### 2015-01-04 mcmc.tex
methods

@@
 \paragraph{Tempering:}
-... HOGG
+A huge problem for MCMC samplers---which make local moves based on
+local positioning in the parameter space---is that there can be
+multiple modes in the function $f(\pars)$, separated by valleys of
+zero or near-zero density.
+A vanilla MCMC method cannot easily cross such valleys with local
+moves.
+The idea behind simulated tempering and related methods is to ``smooth
+out'' the function $f(\pars)$ or reduce its dynamic range, to make the
+peaks less tall and the valleys less deep, to permit the MCMC random
+walk to cross the valleys.
+The standard method is to introduce a ``temperature'' variable and,
+while sampling, take the likelihood function to the power of the
+inverse temperature (or, if working in the logarithm, multiply the log
+likelihood by the inverse temperature).
+When the temperature is high, this reduces the influence of the
+likelihood function relative to the prior pdf, and (assuming that the
+prior pdf is easy to sample) makes movement in the parameter space
+easier.
+At each step, when the temperature has moved away from unity, the
+distribution being sampled is not exactly $f(\pars)$, but the results
+of the sampling are only kept for those samples taken at unit
+temperature; these in the end make a fair sampling.
 
 \paragraph{Nested:}
-Similar to tempering are a class of methods called ``nested''
-samplings, which also smoothly change the target distribution.
+Similar to tempering-like methods are a class of methods called ``nested''
+samplers, which also smoothly change the target distribution away from $f(\pars)$.
 In the nested case, instead of increasing the temperature, the
@@
 value of the likelihoood function.
+When the likelihood censoring is strong, only the most high-likelihood
+parts of the prior get sampled; when the likelihood censoring is weak,
+almost all of the prior gets sampled.
+The idea behind nested sampling---like tempering---is that it is
+designed to be a good method for exploring the full parameter space or
+searching for all of the modes of the posterior pdf.
+It is also designed such that it produces not just a sampling but also
+an estimate of the integral $Z$ of the likelihood times the prior
+(the Bayes factor or fully marginalized likelihood).
 
 \paragraph{Reversible-jump:}
-Finally, there are some problems that \emph{cannot be done} without
-additional proposal technology...
+Finally, there are some extreme problems in which it is not just hard
+to sample in the parameter space, but the parameter space itself is
+not of fixed dimension.
+For example, if you are modeling an astronomical image with a
+collection of stars, and you don't know how many stars to use; in this
+case the number of stars itself is a parameter, so the number of
+parameters is itself a function of the parameters.\note{HOGG Cite
+  Brewer et al}
+In these cases you can only use certain kinds of samplers (though M--H
+MCMC is one you can choose), and, in addition, you need to make
+special kinds of proposals that permit the system to jump from one
+parameter space to another that has different dimensionality.
+In these cases, the concept of detailed balance becomes non-trivial,
+but there are criteria for creating reversible-jump proposals between
+the parameter spaces.
+These methods are best built and operated under the supervision of a
+trained professional.
 

### 2015-01-10 mcmc.tex
notes to self; tweaks

@@
 % -----
+% - mention JAGS and STAN in methods section
 % - perhaps re-order to m-h, convergence, reporting results, other methods ??
@@
 
-The only significant technical details for importance sampling are the
+Two of the significant technical details for importance sampling are the
 following:
@@
 (the Bayes factor or fully marginalized likelihood).
+
+\paragraph{Data augmentation:}
+...HOGG?
 

### 2015-01-10 mcmc.tex
slice sampling, reordering

@@
 We will also occasionally assume---just because it is true in
-inference---that the function $f(\theta)$ is the product of two
+inference---that the function $f(\pars)$ is the product of two
 functions, one called ``the prior pdf'' and one called ``the
@@
 inference contexts with proper inputs.
-That the function $f(\theta)$ can be thought of as a product of a
+That the function $f(\pars)$ can be thought of as a product of a
 prior pdf and a likelihood is only necessary for what follows in
@@
 
-As we mention below,
+As we discuss briefly below in \sectionname~\ref{sec:methods},
   there are many MCMC methods more advanced than M--H.
@@
 
-\section{Gibbs, ensemble, Hamiltonian, and so on}
-
-[SHOULD THIS SECTION MOVE TO AFTER THE DEFINITION OF AUTOCORRELATION TIME AND DISCUSSION OF CONVERGENCE?]
+\section{Convergence}
+
+A key question for an MCMC operator---\emph{the} key question in some
+sense---is \emph{how long to run} to be sure of having reliable
+results.
+It is disappointing and annoying to many that there is no extremely
+simple and reliable answer to this question.
+
+The reason that there is no simple answer is that you can't really
+ever know that you have sampled the full posterior pdf, and the reason
+for \emph{that} is that if you \emph{could} know that, you would also
+be able to solve the famously difficult discrete optimization
+problem.\note{explain this remark and cite something here}
+Another way to put it is that you might have two modes---two separated
+regions of substantial total probability---in the posterior pdf that
+are both important but separated by a large region of low probability.
+If you are using a simple sampler, you could sample one of these modes
+very well but the sampler will take effectively infinite time to find
+the other mode.
+Any test of convergence would be passed, but in fact the sampling
+would be incomplete and not representative.
+That is, knowing that you have a complete and representative sampling
+is effectively impossible.
+
+In this context,
+  it is \emph{simply not fair} to require an MCMC run to have fully and completely
+  sampled the posterior pdf,
+  at least not in any provable sense.
+The question of whether a sampling is converged to a representative sampling
+  of the posterior pdf is actually outside the domain of science,
+  because the domain of science is the domain of questions answerable by scientists.
+That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
+  despite many statements in the literature to the contrary.%
+  \note{Give example statements... Mention cases in which you \emph{might} be able to claim this...}
+The upshot of this is that we don't and can't require any kind of absolute convergence;
+  indeed, it would be impossible even to test for it.
+
+In this pragmatic situation---the Real World, as it is called---we have to rely on heuristics.
+Heuristically, you have sampled long enough when you can see
+that the (or each) walker has traversed the high-probability parts of
+the parameter space many times in the length of the chain.
+Or, equivalently, you have sampled long enough when the first half of
+the chain shows very much the same posterior pdf morphology as the
+second half of the chain, or indeed as any substantial subset.
+
+...Making this heuristic condition more precise
+...Autocorrelation time is the premier tool
+...How is Autocorrelation time defined?
+...How is Autocorrelation time measured?
+...Why is it is not a panacea? ...estimators ...getting stuck in one mode
+
+...What plots should you make?
+...Plots showing individual walkers going through the space.
+...(now show plots that demonstrate a non-converged and a converged chain)...
+
+\section{Gibbs, ensemble, Hamiltonian, and so on}\label{sec:methods}
 
@@
 
-\paragraph{Slice:}
-... DFM, can you help here?
-
 \paragraph{Importance:}
@@
 must be greater than or equal to the approximate function everywhere!
+
+\paragraph{Slice:}
+... When is slice sampling a good idea?  DFM, can you help here?
+
+Slice sampling makes use of the point that (in one dimension) a fair
+sampling of $f(\pars)$ is a projection onto the $\pars$ axis of a
+uniform sampling of the two-dimensional region under the
+two-dimensional plot of $f(\pars)$ \vs\ $\pars$.
+The $D$-dimensional generalization is that a fair sampling of
+$f(\pars)$ is a projection onto the $\pars$ $D$-dimensional
+hypersurface of $\pars$ of a uniform sampling of the $D+1$-dimensional
+hypervolume under the hypersurface $f(\pars)$ in a hypothetical plot
+of $f(\pars)$ over the whole parameter space.
+Thinking in terms of the two-dimensional case (for simplicity), slice
+samplers make both ``horizontal'' moves in the $\pars$ direction at
+fixed $f(\pars)$ and also ``vertical'' moves in the $f(\pars)$
+direction.
+Because the sampling in the higher-dimensional space is uniform, it is
+easy to make moves to very different parts of parameter space, or
+different peaks or modes in the function $f(\pars)$.
+In our own research, we use a variant of slice sampling for problems
+involving samplings governed by Gaussian Processes.\note{DFM: Cite us
+  and Murray here.}
 
@@
 the median (or even mean) of the posterior pdf for each of parameters,
-when taken together as a ``best-fit'' parameter vector $\theta^\ast$,
+when taken together as a ``best-fit'' parameter vector $\pars^\ast$,
 will not itself necessarily be a good fit to the data!
@@
 that simplify your problem.
-
-\section{Convergence}
-
-A key question for an MCMC operator---\emph{the} key question in some
-sense---is \emph{how long to run} to be sure of having reliable
-results.
-It is disappointing and annoying to many that there is no extremely
-simple and reliable answer to this question.
-
-The reason that there is no simple answer is that you can't really
-ever know that you have sampled the full posterior pdf, and the reason
-for \emph{that} is that if you \emph{could} know that, you would also
-be able to solve the famously difficult discrete optimization
-problem.\note{explain this remark and cite something here}
-Another way to put it is that you might have two modes---two separated
-regions of substantial total probability---in the posterior pdf that
-are both important but separated by a large region of low probability.
-If you are using a simple sampler, you could sample one of these modes
-very well but the sampler will take effectively infinite time to find
-the other mode.
-Any test of convergence would be passed, but in fact the sampling
-would be incomplete and not representative.
-That is, knowing that you have a complete and representative sampling
-is effectively impossible.
-
-In this context,
-  it is \emph{simply not fair} to require an MCMC run to have fully and completely
-  sampled the posterior pdf,
-  at least not in any provable sense.
-The question of whether a sampling is converged to a representative sampling
-  of the posterior pdf is actually outside the domain of science,
-  because the domain of science is the domain of questions answerable by scientists.
-That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
-  despite many statements in the literature to the contrary.%
-  \note{Give example statements... Mention cases in which you \emph{might} be able to claim this...}
-The upshot of this is that we don't and can't require any kind of absolute convergence;
-  indeed, it would be impossible even to test for it.
-
-In this pragmatic situation---the Real World, as it is called---we have to rely on heuristics.
-Heuristically, you have sampled long enough when you can see
-that the (or each) walker has traversed the high-probability parts of
-the parameter space many times in the length of the chain.
-Or, equivalently, you have sampled long enough when the first half of
-the chain shows very much the same posterior pdf morphology as the
-second half of the chain, or indeed as any substantial subset.
-
-...Making this heuristic condition more precise
-...Autocorrelation time is the premier tool
-...How is Autocorrelation time defined?
-...How is Autocorrelation time measured?
-...Why is it is not a panacea? ...estimators ...getting stuck in one mode
-
-...What plots should you make?
-...Plots showing individual walkers going through the space.
-...(now show plots that demonstrate a non-converged and a converged chain)...
 

### 2015-01-10 mcmc.tex
cross-refs
note re Z

@@
 p(\pars\given\data) &=& \frac{1}{Z}\,p(\data\given\pars)\,p(\pars)
-\quad .
+\label{eq:bayes}\quad .
 \end{eqnarray}
@@
 \emph{extremely hard to
-  calculate}.\note{Say words about why we call it $Z$ not $p(\data)$
-  and why it is extremely hard to calculate.}
+  calculate}.\note{Some would call
+  the factor $Z$ in \equationname~(\ref{eq:bayes}) something like
+  ``$p(\data)$'' or ``the marginal likelihood'' or the 
+  probability of the data.
+  Of course it is only the probability of the data within the model
+  space under consideration!
+  We discuss this object and these issues [MORE ELSEWHERE; CITE].
+  The factor $Z$ is often hugely difficult to compute, because the
+  likelihood (or the prior) can have extremely complex structure,
+  with multiple arbitrarily compact modes, arbitrarily positioned
+  in the (presumably high dimensional) parameter space $\pars$.
+  Elsewhere, we discuss the computation of this object (cite Fengji)
+  and---perhaps more importantly---why you probably never really
+  \emph{want} to compute it.
+  The computation of $Z$ is outside the scope of this \documentname,
+  but suffice it to say:
+  If the computation of $Z$ is in your travel plans, you will need to
+  think about implementing some non-trivial integrating or sampling
+  technology; we will give some brief description of options in
+  \sectionname~\ref{sec:methods}.}
 That is, you often know the function $p(\pars\given\data)$ up to a
@@
   this right now you might want to do that.
-It will help us tune the proposal pdf and diagnose problems.
+It will help us tune the proposal pdf (\sectionname~\ref{sec:tuning})
+and diagnose problems (\sectionname~\ref{sec:troubleshooting}).
 
@@
 
-\section{Tuning}
+\section{Tuning}\label{sec:tuning}
 
@@
 
-\section{Troubleshooting and advice}
+\section{Troubleshooting and advice}\label{sec:troubleshooting}
 

### 2015-01-10 mcmc.tex
what is a model

@@
   space under consideration!
-  We discuss this object and these issues [MORE ELSEWHERE; CITE].
+  We discuss this object and these issues elsewhere (HOGG CITE).
   The factor $Z$ is often hugely difficult to compute, because the
@@
   in the (presumably high dimensional) parameter space $\pars$.
-  Elsewhere, we discuss the computation of this object (cite Fengji)
+  Elsewhere, we discuss the computation of this object (HOGG cite Fengji)
   and---perhaps more importantly---why you probably never really
-  \emph{want} to compute it.
+  \emph{want} to compute it (HOGG CITE).
   The computation of $Z$ is outside the scope of this \documentname,
@@
 You also have a model sophisticated enough---a probabilistic,
-generative model, if you will\note{For our definition of a
-  \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
+generative model, if you will\note{For more discussion
+  of our definition of a \emph{model}, see [HOGG SOME OTHER NOTE] in
+  this series.
+  Briefly, a model is a likelihood function (a pdf for the data given
+  model parameters), and a prior pdf over the parameters.
+}---that, given a
 setting of a huge blob of parameters\note{Again, the parameter
@@
   series.}) $p(\data\given\pars)$.
-% ************* BJB
-% <Wild Subjective Bayesian>
-% The "generative model" is a probabilistic description of your *prior
-% beliefs about what the data are going to be*. </Wild Subjective Bayesian>
-% ************* Hogg
-% That comment should go in the ``What is a Model?'' Chapter.
-% (Or a \note{} right here.)
 Furthermore, say you also can write down some kind of informative or

### 2015-01-10 mcmc.tex
notes

@@
 % -----------
+% - These documents are ``Chapters'', no?
 % - [LaTeX] newline after every full stop for proper git diffing
@@
   this series.
-  Briefly, a model is a likelihood function (a pdf for the data given
+  Briefly, a ``model'' for us is a likelihood function (a pdf for the data given
   model parameters), and a prior pdf over the parameters.
+  Because, under this definition, the model can always generate (by
+  sampling, say) parameters and parameters can generate (again by
+  sampling, say) data, the model is effectively (or actually, if you
+  are a true subjective Bayesian) a probability distribution (pdf)
+  over all possible data.
 }---that, given a
@@
 a fair sampling of the posterior pdf.
-That is, you can run MCMC (for a very long time\note{Sometimes a very,
-  very long time; we will return to that below.})  and you will be
-left with a set of $K$ parameter-blob settings $\pars_k$ such that the
+That is, you can run MCMC
+(for a very long time---see \sectionname~\ref{sec:convergence} for how long)
+and you will be left with a set of $K$ parameter-blob settings $\pars_k$ such that the
 full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
@@
 All that said, and adhering to the traditions of the \project{Data
-  Analysis Recipes} project\note{Every chapter in the \project{Data
+  Analysis Recipes} project\note{Every \documentname\ in the \project{Data
     Analysis Recipes} series begins with a rant in which we argue that
@@
 
-\section{Convergence}
+\section{Convergence}\label{sec:convergence}
 

### 2015-01-10 mcmc.tex
autocorrelation

@@
 
-...Making this heuristic condition more precise
-...Autocorrelation time is the premier tool
-...How is Autocorrelation time defined?
-...How is Autocorrelation time measured?
+The above heuristics can be made more precise (in terms of the amount
+of deviation one expects between the means and variances, say, of two
+disjoint subsets of the chain.
+The premier tool for making this heuristic precise, however, is to
+look at the \emph{autocorrelation time} of the chain.
+In general, when one has a sequence $(\pars_1, \pars_2, \theta_3, \cdots)$
+generated by a Markov Process in the $\pars$ space, nearby points in
+the sequence will be similar, but sufficiently distant points will not
+``know about'' one another;
+the autocorrelation function measures this.
+
+The autocorrelation function is defined by
+\begin{equation}
+C(\Delta) &\equiv& \frac{E[(\pars_t - \mu)\,(\pars_{t+\Delta} - \mu)]}{E[(\pars_t)^2]
+\\
+\mu &=& E[\pars_t]
+\quad ;
+\end{equation}
+that is, it is the covariance of points separated by $\Delta$, taken
+relative to the variance.
+By construction
+\begin{equation}
+C(0) &=& 0
+\\
+\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
+\quad ,
+\end{equation}
+and by definition, the autocorrelation ``time'' $\tau$ is
+\begin{equation}
+\tau = \int C(\Delta)\,\dd\Delta
+\quad .
+\end{equation}
+Heuristically, the autocorrelation time is the number of samples you
+need to take before you get an independent sample.
+More precisely, the autocorrelation time is the factor by which you
+have to reduce the number of samples when you compute the variance on
+a sampling-based expectation value estimate.\note{HOGG Explain more fully.}
+
+Of course, the autocorrelation-time definition does not by itself provide
+an \emph{estimator} for the time.
+
 ...Why is it is not a panacea? ...estimators ...getting stuck in one mode

### 2015-01-11 mcmc.tex
autocorrelation function content

@@
 The autocorrelation function is defined by
-\begin{equation}
-C(\Delta) &\equiv& \frac{E[(\pars_t - \mu)\,(\pars_{t+\Delta} - \mu)]}{E[(\pars_t)^2]
-\\
-\mu &=& E[\pars_t]
+\begin{eqnarray}
+C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
+\label{eq:autocorrelationfn}\\
+\mu &=& E[\pars]
 \quad ;
-\end{equation}
+\end{eqnarray}
 that is, it is the covariance of points separated by $\Delta$, taken
 relative to the variance.
+In the above equation, the assumption is that the samples $\pars_k$ are
+given in the order that they were generated by the MCMC algorithm.
 By construction
-\begin{equation}
-C(0) &=& 0
+\begin{eqnarray}
+C(0) &=& 1
 \\
@@
 \quad ,
-\end{equation}
+\end{eqnarray}
 and by definition, the autocorrelation ``time'' $\tau$ is
-\begin{equation}
-\tau = \int C(\Delta)\,\dd\Delta
-\quad .
-\end{equation}
+\begin{eqnarray}
+\tau = \sum_{\Delta = 0}^{\infty} C(\Delta)
+\label{eq:autocorrelationtime}\quad .
+\end{eqnarray}
 Heuristically, the autocorrelation time is the number of samples you
@@
 have to reduce the number of samples when you compute the variance on
-a sampling-based expectation value estimate.\note{HOGG Explain more fully.}
-
-Of course, the autocorrelation-time definition does not by itself provide
-an \emph{estimator} for the time.
-
-...Why is it is not a panacea? ...estimators ...getting stuck in one mode
-
-...What plots should you make?
-...Plots showing individual walkers going through the space.
-...(now show plots that demonstrate a non-converged and a converged chain)...
+a sampling-based expectation value estimate.\note{HOGG Explain more fully.  Maybe also cite Fengji.}
+It is slightly off-topic here, but related to this, if you have a long
+chain of length $K$ and with autocorrelation time $\tau$ and you thin
+the chain down to every $\tau$th sample (instead of using all $K$
+samples), you will get just as good sampling-based estimates as if you
+had used all $K$ samples.
+
+A sampler with a better autocorrelation time is just better; you have
+to do fewer $f(\pars)$ calls per independent sample, and you have to
+run less time to get accurate sampling-based integral estimates.
+A sampler that takes an independent sample every time would have an
+autocorrelation time of unity, which is the best possible value; this
+optimal sampling is only possible for problems where the sampling is
+analytic (for example if the posterior is perfectly Gaussian with
+known mean and variance).
+In general, the best sampler will be different for different problems,
+and we will (below) tune the samplers we have to do the best on the
+problems we have; this tuning will also be problem-specific.
+There are many heuristic bases on which different samplers might be
+compared or tuned, but fundamentally it is lower autocorrelation time
+that separates good samplers from bad ones and is the ultimate basis
+on which we compare performance.\note{HOGG: Bash Dunkley and anyone
+  else who has used something else to compare.}
+
+Of course, the autocorrelation-time definition
+(\ref{eq:autocorrelationtime}) does not by itself provide an
+\emph{estimator} for the time.
+Usually, we just take the naive estimator---the ratio of empirical
+expectations implied by \equationname~(\ref{eq:autocorrelationfn}),
+and then perform a naive sum to get the autocorrelation time.
+However, almost all empirical estimators of the autocorrelation time
+will end up being under-estimates (biased low), because the finite
+length of the chain makes any subset seem more representative than it
+truly is; indeed this is a sub-problem of the problem that convergence
+can never be known empirically with good confidence.
+
+Finally one last point about convergence:
+Since all convergence tests are fundamentally heuristic, it is useful
+to just make the heuristic visualization of the samples $\pars_k$ as a
+function of $k$, in the order they were generated.
+The chain is likely to be converged only if the random-walk process
+crossed the domain of $\pars$ fully many times during the MCMC run.
+Often some parameter directions are much worse than others; it is
+worth looking at the chain in all parameter directions.
 

### 2015-01-12 mcmc.tex
reordering of the sections, some guide at the beginning

@@
 
+In what follows, we are going to provide a kind of ``user manual'' or
+advice document or folklore capture regarding the use of MCMC for
+data analysis.
+This will not be a detailed description of multiple MCMC methods
+(indeed, we will only explain one method in detail), and it will not
+be about the mathematical properties or structure of the method or
+methods.
+It will be about how to use MCMC, including diagnosis and
+trouble-shooting.
+
+The first couple of \sectionname s will describe what a sampling is, and how
+the simplest MCMC method, the Metropolis-Hastings algorithm, can
+provide one.
+The next few \sectionname s will provide ideas about how to
+initialize, tune and operate MCMC methods for good performance.
+The last few \sectionname s will provide advice for making decisions
+among the myriad MCMC methods implementations, how to implement a good
+likelihood function and prior pdf function for inference, and how to
+trouble-shoot standard kinds of problems that arise in operating MCMC
+methods on real problems.
+We will leave parenthetical and philosophical matters to the end-notes,
+which appear after the main text.
+
 \section{What is a sampling?}
@@
 
-\section{Gibbs, ensemble, Hamiltonian, and so on}\label{sec:methods}
-
-MCMC methods are proven to be correct ``in the limit''; that is, when
-an infinite number of samples have been taken.
-When the function $f(\pars)$ has challenging properties, the results
-can approach this limit very slowly (or not at all\note{HOGG: discuss
-  situations in which MCMC will not work---related to support.}).
-For this reason, in the finite duration of a scientific project, most
-users of MCMC have one of two (related) problems:
-Either \textsl{(1)}~it is taking too long---it is taking too many steps or
-executing too many calls of the function $f(\pars)$ to get independent samples---or \textsl{(2)}~it is not
-exploring the full parameter space---there are local optima (or bad
-local geometry around those optima) ``trapping'' the algorithm.
-
-There is no general, problem-independent solution for either of these
-problems.
-Indeed, there are many different strategies that work well or badly
-for different functions $f(\pars)$.
-In this \sectionname, we will try to guide the reader towards methods
-that might help in different circumstances.
-We will not give a full description of all the MCMC algorithms out
-there, but just provide some field notes and pointers to references.
-New MCMC methods are being developed all the time, so a reader with a
-serious problem would do well to consult with the applied-mathematics
-or statistics literature.\note{HOGG: Your best interface to these
-  literatures is a colleague in another department.  Break down those walls!}
-The different methods to which we refer here are (in general) very
-different in their difficulty of implementation.
-We won't help with that either, except to say that there are more and
-more open-source or distributed packages every year.
-We will try to name some current implementations in the notes, but any
-list compiled today will be incomplete tomorrow.
-The spaces of MCMC methods and MCMC method implementations are both
-growing rapidly, in part because MCMC has become such a core
-technology in the empirical sciences.
-
-...Note a good general reference here?  MacKay?
-
-% *********** BJB talking:
-% I suggest partitioning the discussion of different methods into the following
-% two categories:
-%
-%  i) Methods for getting around the target distribution
-% ii) Methods that change the target distribution to something
-% other than the posterior pdf, to help the sampling somehow
-%
-% i) Would include things like M-H, the stretch move,
-% Hamiltonian, slice sampling (maybe skip the latter)
-% ii) would include things like annealing and Nested Sampling
-
-%%%% Hogg talking:
-% DFM: In this section, I think we should just summarize the other
-% methods, not describe them in detail.  Then we can cite the relevant
-% papers for the enthusiastic.
-
-\paragraph{Ensembles:}
-As we discuss more below, one of the critical challenges in the use of
-MCMC is tuning the proposal pdf $q(\pars'\given\pars)$.
-In problems in which all of the parameters (all of the components or entries of $\pars$) are somehow ``equivalent'',
-or the parameters can be seen as components of a vector in a vector
-space, there are \emph{ensemble methods} that make use of not just one
-random-walker but instead many walkers to automatically generate a properly tuned
-proposal distribution.
-These methods make use of the distribution of a set of independent
-walkers in the parameter space ($\pars$-space) to gauge the typical step-sizes and
-directions at which new proposals should be made, obviating or greatly
-reducing the tuning requirements.
-We are co-authors on one of these methods, \project{emcee}\note{HOGG:
-  ref emcee}, which is popular in the astrophysical sciences.
-Because ensemble methods are updating independent MCMC chains, they
-are often easy to parallelize to make use of multiple cores.
-
-\paragraph{Gibbs:}
-In many problems, the parameters---the components or entries of the
-blob $\pars$---are not equivalent at all.
-Some are global parameters, which touch or have an effect on every
-sub-part of the data blob $\data$, while others are local parameteres,
-which only touch a small subset of the data.
-Or, in another use case, some parameters are linear parameters
-that---at fixed values of the other parameters---will have a Gaussian
-likelihood or posterior pdf.
-In these problems, there are some directions in parameter space that
-are hard to move in (they require complete re-calculation of the full
-non-linear function $f(\pars)$, and other directions that are very
-easy to move in (they require only partial re-calculations or the
-sampling could even be analytic and exact).
-For problems with these structures, Gibbs samplers are ideal, and
-there are good implementations.\note{HOGG cite JAGS}
-
-A trivial kind of Gibbs sampling was mentioned above in the
-discussion of M--H MCMC:
-Even when running vanilla M--H MCMC, it is often useful to update only
-one parameter at a time; that is, do only axis-aligned moves in the
-$\pars$-space, cycling through axes as you go from step to step.
-This helps with tuning and diagnosis, but it also permits a good
-implementation to capitalize on simpler calculations for the simpler
-parameter directions.
-
-Gibbs sampling can be very good for exploiting multiprocessing:
-Global parameter updates might require a serial re-calculation of the
-full likelihood, but local parameter updates can be done in parallel,
-with each local parameter working on its local bit of data all at the
-same time.
-This kind of structure is common in hierarchical inference, where
-local parameters touch only individual objects in some population,
-say, and global parameters are parameters \emph{of} that population as
-a whole.
-Parallelization is built in to some Gibbs implementations.\note{HOGG}
-
-\paragraph{Hamiltonian:}
-[HOGG, DFM: THIS HAS SOME OTHER NAME; WHAT?]
-In many cases of interest, the function $f(\pars)$ can be analytically
-differentiated, such that it is possible to quickly execute a gradient
-$\dd f/\dd\pars$ evaluation.
-Indeed, in the modern world of computing, there are even
-auto-differentiation systems that deliver automatically code that
-produces the gradient of any function you can write.\note{HOGG cite
-  autodiff systems}
-There is a class of MCMC methods that make use of gradient information
-to effectively inform the proposal distribution and thus speed sampling.
-
-... HOGG more here!
-
-\paragraph{Importance:}
-Sometimes it is possible to sample efficiently from a distribution
-that is close to the target function $f(\pars)$ you care about but
-hard to sample efficiently with MCMC.
-An example is a problem that has a simple (Gaussian, say) likelihood
-function times a non-trivial prior:
-It is easy to sample from a Gaussian likelihood times a Gaussian
-prior, even without MCMC, but it might be hard to do the same sampling
-under a more complex prior.
-Importance sampling is a direct sampling method (not really an MCMC
-method at all):
-You take a sampling from the approximate function (which by assumption
-was easy) and then you re-weight the sampling (or do a subsequent
-rejection of samples) using the ratio of the true function to the
-approximate function as a weight or probability.
-We use this quite a bit in our hierarchical inferences.\note{HOGG:
-  Citations}
-
-Two of the significant technical details for importance sampling are the
-following:
-The approximate function must have the same (or greater) support as
-the true function, and the method degrades in efficiency as the two
-functions diverge.
-If the ratio of the true function to the approximate function is to be
-used as a probability for rejection of samples, then the true function
-must be greater than or equal to the approximate function everywhere!
-
-\paragraph{Slice:}
-... When is slice sampling a good idea?  DFM, can you help here?
-
-Slice sampling makes use of the point that (in one dimension) a fair
-sampling of $f(\pars)$ is a projection onto the $\pars$ axis of a
-uniform sampling of the two-dimensional region under the
-two-dimensional plot of $f(\pars)$ \vs\ $\pars$.
-The $D$-dimensional generalization is that a fair sampling of
-$f(\pars)$ is a projection onto the $\pars$ $D$-dimensional
-hypersurface of $\pars$ of a uniform sampling of the $D+1$-dimensional
-hypervolume under the hypersurface $f(\pars)$ in a hypothetical plot
-of $f(\pars)$ over the whole parameter space.
-Thinking in terms of the two-dimensional case (for simplicity), slice
-samplers make both ``horizontal'' moves in the $\pars$ direction at
-fixed $f(\pars)$ and also ``vertical'' moves in the $f(\pars)$
-direction.
-Because the sampling in the higher-dimensional space is uniform, it is
-easy to make moves to very different parts of parameter space, or
-different peaks or modes in the function $f(\pars)$.
-In our own research, we use a variant of slice sampling for problems
-involving samplings governed by Gaussian Processes.\note{DFM: Cite us
-  and Murray here.}
-
-\paragraph{Tempering:}
-A huge problem for MCMC samplers---which make local moves based on
-local positioning in the parameter space---is that there can be
-multiple modes in the function $f(\pars)$, separated by valleys of
-zero or near-zero density.
-A vanilla MCMC method cannot easily cross such valleys with local
-moves.
-The idea behind simulated tempering and related methods is to ``smooth
-out'' the function $f(\pars)$ or reduce its dynamic range, to make the
-peaks less tall and the valleys less deep, to permit the MCMC random
-walk to cross the valleys.
-The standard method is to introduce a ``temperature'' variable and,
-while sampling, take the likelihood function to the power of the
-inverse temperature (or, if working in the logarithm, multiply the log
-likelihood by the inverse temperature).
-When the temperature is high, this reduces the influence of the
-likelihood function relative to the prior pdf, and (assuming that the
-prior pdf is easy to sample) makes movement in the parameter space
-easier.
-At each step, when the temperature has moved away from unity, the
-distribution being sampled is not exactly $f(\pars)$, but the results
-of the sampling are only kept for those samples taken at unit
-temperature; these in the end make a fair sampling.
-
-\paragraph{Nested:}
-Similar to tempering-like methods are a class of methods called ``nested''
-samplers, which also smoothly change the target distribution away from $f(\pars)$.
-In the nested case, instead of increasing the temperature, the
-samplings are of a censored version of the prior, censored by the
-value of the likelihoood function.
-When the likelihood censoring is strong, only the most high-likelihood
-parts of the prior get sampled; when the likelihood censoring is weak,
-almost all of the prior gets sampled.
-The idea behind nested sampling---like tempering---is that it is
-designed to be a good method for exploring the full parameter space or
-searching for all of the modes of the posterior pdf.
-It is also designed such that it produces not just a sampling but also
-an estimate of the integral $Z$ of the likelihood times the prior
-(the Bayes factor or fully marginalized likelihood).
-
-\paragraph{Data augmentation:}
-...HOGG?
-
-\paragraph{Reversible-jump:}
-Finally, there are some extreme problems in which it is not just hard
-to sample in the parameter space, but the parameter space itself is
-not of fixed dimension.
-For example, if you are modeling an astronomical image with a
-collection of stars, and you don't know how many stars to use; in this
-case the number of stars itself is a parameter, so the number of
-parameters is itself a function of the parameters.\note{HOGG Cite
-  Brewer et al}
-In these cases you can only use certain kinds of samplers (though M--H
-MCMC is one you can choose), and, in addition, you need to make
-special kinds of proposals that permit the system to jump from one
-parameter space to another that has different dimensionality.
-In these cases, the concept of detailed balance becomes non-trivial,
-but there are criteria for creating reversible-jump proposals between
-the parameter spaces.
-These methods are best built and operated under the supervision of a
-trained professional.
+\section{Tuning}\label{sec:tuning}
+
+Most MCMC methods
+  (though not all\note{We are promoters of an ensemble method that has a
+  nearly-fixed proposal distribution and requires no tuning (CITE EMCEE).})
+  make use of something like a ``proposal distribution''
+  which determines what kinds of steps the walker can take
+  as it random-walks through the parameter space.
+The user generally has a lot of control over what this proposal distribution might be,
+  and how to choose the parameters.
+For example, in a $d$-dimensional parameter space, %do we consistently use d here?
+  a zero-mean but anisotropic Gaussian (normal distribution)
+  is often used to draw offsets for the walker to take from point to point.
+Even within this choice (which is of one function among many possibilities),
+  there are $d\,(d+1)/2$ parameters in the $d\times d$ symmetric, positive definite covariance matrix
+  to set ``by hand''.
+How to choose this function and set these parameters?
+
+The key idea here is that if the proposal distribution is too narrow---%
+  it proposes steps too small---%
+  almost all steps will be accepted (REFERENCE ALGORITHM HERE),
+  but it will take a long time to move anywhere because of timidity.
+If the propsoal distribution is too wide---%
+  it proposes steps too large---%
+  the moves will cover parameter space easily,
+  but almost no steps will be accepted;
+  it will tend to jump to much lower probability regions.
+There is a Goldilocks step size (proposal distribution root-variance) that is ``just right''.
+In one dimension this might be easy to find,
+  but, as we say, in large numbers of dimensions,
+  there is a lot of freedom in choosing the parameters of the distribution.
+
+% BJB
+% Please suggest mixtures of scales for M-H proposals.
+% You can mimic the properties of slice sampling but with
+% much easier code.
+
+In a deep sense, the only scalar that makes sense to optimize,
+  when choosing proposal distribution (or, loosely speaking, step size),
+  is the autocorrelation time, described above.
+The optimal step size is the step size that makes for the shortest autocorrelation time.
+This is easy to state, but hard to use in practice;
+  being a second-order statistic of the MCMC chain,
+  the autocorrelation time is a hard thing to measure without a lot of data,
+  so it is hard to quickly estimate it and adjust,
+  much less put it into some kind of optimization loop.
+We usually use proxies for this of various kinds.
+
+The simplest heuristic proxy statistic for tuning is the \emph{acceptance fraction}.
+If you are accepting almost all proposed steps, your step sizes are too small on average.
+If you are accepting almost none, your step sizes are too large.
+The Goldilocks value is around a quarter,
+  with an argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}
+  (though you could never tune it precisely enough to warrant that third digit of accuracy).
+In the burn-in phase (discussed below) of an MCMC run,
+  it makes sense to track the acceptance ratio,
+  and adjust the proposal distribution variance as you get acceptance ratios
+  that are far from the Goldilocks ratio.
+This process can be automated easily;
+  such automation is part of many projects that use MCMC.
+
+% HOGG:  Is the following a repeat from previously and/or does it mesh with what's earlier?
+
+A very common---%
+  and very useful---%
+  kind of proposal distribution is one that cycles through parameters,
+  taking a random step in just one parameter at a time.
+Aside from involving some persistence of state,
+  this kind of proposal distribution can be valuable,
+  in part because it reduces the (almost impossible) $d$-dimensional tuning problem
+  to $d$ one-dimensional problems:
+In this form of proposal distrubution,
+  it is possible to track a separate acceptance fraction for every parameter.
+Code can be built that uses the burn-in phase to tune all $d$ proposal variances
+  such that each of them, individually,
+  obtains acceptance at the same Goldilocks ratio.
+
+One note to make here---%
+  because it is relevant to tuning---%
+  is that tuning can \emph{only} take place during the burn-in phase;
+  you cannot tune \emph{while} you run your final MCMC run.
+Why not?
+Because tuning the proposal distribution based on the past history of the chain
+  violates the ``Markov'' property
+  that each step depends \emph{only} on the state at the previous step.
+Violating the Markov property can be very bad;
+  when you violate it,
+  you lose all the provable properties of MCMC
+  on which all our righteous power is based.
+
+Another proxy for autocorrelation time useful for tuning is the
+  Expected Squared Jump Distance.\note{(CITE GELMAN)}
+This is the mean squared distance the walker moves, per step.
+It is maximized when the acceptance ratio is reasonable
+  and the step size is large;
+  it is large when the exploration of the space is fast.
+This proxy is easy to measure and use for tuning,
+  and more directly related to autocorrelation time than the acceptance ratio.
+We recommend using it for tuning, though we have never used it ourselves.
+Like the acceptance fraction, it can also be used to tune
+  in the case that the proposal distribution loops over parameters;
+  once again, the user gets (in this case) $d$ one-dimensional tunings.
+
+When sampling really gets very slow,
+  there are various tricks and tips to work through.
+One is to see if any of the parameters has a conditional posterior
+  probability that is guaranteed to be Gaussian
+  (this is true for some linear parameters in least-square problems
+  when the priors are Gaussian, for example).
+These dimensions can often be sampled \emph{exactly},
+  outside of the MCCM framework;
+  the algorithm looks like:
+Take an MCMC step in the hard parameters,
+  then take an exact step in the easy parameters,
+  repeat.
+Another related trick is to analytically marginalize,
+  rather than sample,
+  easy nuisance parameters.
+The huge bag of possible tricks is so large,
+  it goes way beyond the scope of this introductory \documentname.
+In our experience, it is valuable to make friends with at least one statistician,
+  one applied mathematician, and one computer scientist.
+Between them, they ought to span the literature.
+
+\section{Initialization and burn-in}
+
+Just as most (though not all) MCMC methods require a choice of proposal distribution,
+  most (though not all) require a choice about initialization:
+The walker needs (or walkers need) to be started somewhere in the parameter space.
+In many use cases for MCMC,
+  the investigator wants to use it to obtain uncertainty information about
+  or propagate uncertainty into parameter estimates.
+In these cases, it makes sense to initialize the walker or walkers at sensible parameter estimates,
+  found by optimizing a likelihood or a posterior PDF in advance of sampling.
+In extremely high dimensions (large numbers of parameters), this can be a bad idea,
+  since the optimal parameters are not necessarily near typical posterior samples\note{Make the counterintuitive point.},
+  but in low dimensions (few to tens) this is often sensible.
+
+In other use cases,
+  the investigator wants to use MCMC to search all of parameter space
+  for good models, and also sample.
+In this case, it makes sense to be more catholic with initialization...
+
+...but note that MCMC is not the ideal search algorithm for any parameter space.
+
+...Start from different locations
+% BJB
+% What different locations?
+% Drawn from the prior can be useful. Then you can
+% look at the chains and see which ones "made it" and
+% which ones "got stuck" and delete the latter
+
+...Dealing with multi-modal problems
+% Starting from different locations is good
+% for DIAGNOSING a problem as multimodal but it's
+% terrible for quantifying the modes. Explain why,
+% it's a good teaching point.
 
@@
 
-\section{Tuning}\label{sec:tuning}
-
-Most MCMC methods
-  (though not all\note{We are promoters of an ensemble method that has a
-  nearly-fixed proposal distribution and requires no tuning (CITE EMCEE).})
-  make use of something like a ``proposal distribution''
-  which determines what kinds of steps the walker can take
-  as it random-walks through the parameter space.
-The user generally has a lot of control over what this proposal distribution might be,
-  and how to choose the parameters.
-For example, in a $d$-dimensional parameter space, %do we consistently use d here?
-  a zero-mean but anisotropic Gaussian (normal distribution)
-  is often used to draw offsets for the walker to take from point to point.
-Even within this choice (which is of one function among many possibilities),
-  there are $d\,(d+1)/2$ parameters in the $d\times d$ symmetric, positive definite covariance matrix
-  to set ``by hand''.
-How to choose this function and set these parameters?
-
-The key idea here is that if the proposal distribution is too narrow---%
-  it proposes steps too small---%
-  almost all steps will be accepted (REFERENCE ALGORITHM HERE),
-  but it will take a long time to move anywhere because of timidity.
-If the propsoal distribution is too wide---%
-  it proposes steps too large---%
-  the moves will cover parameter space easily,
-  but almost no steps will be accepted;
-  it will tend to jump to much lower probability regions.
-There is a Goldilocks step size (proposal distribution root-variance) that is ``just right''.
-In one dimension this might be easy to find,
-  but, as we say, in large numbers of dimensions,
-  there is a lot of freedom in choosing the parameters of the distribution.
-
-% BJB
-% Please suggest mixtures of scales for M-H proposals.
-% You can mimic the properties of slice sampling but with
-% much easier code.
-
-In a deep sense, the only scalar that makes sense to optimize,
-  when choosing proposal distribution (or, loosely speaking, step size),
-  is the autocorrelation time, described above.
-The optimal step size is the step size that makes for the shortest autocorrelation time.
-This is easy to state, but hard to use in practice;
-  being a second-order statistic of the MCMC chain,
-  the autocorrelation time is a hard thing to measure without a lot of data,
-  so it is hard to quickly estimate it and adjust,
-  much less put it into some kind of optimization loop.
-We usually use proxies for this of various kinds.
-
-The simplest heuristic proxy statistic for tuning is the \emph{acceptance fraction}.
-If you are accepting almost all proposed steps, your step sizes are too small on average.
-If you are accepting almost none, your step sizes are too large.
-The Goldilocks value is around a quarter,
-  with an argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}
-  (though you could never tune it precisely enough to warrant that third digit of accuracy).
-In the burn-in phase (discussed below) of an MCMC run,
-  it makes sense to track the acceptance ratio,
-  and adjust the proposal distribution variance as you get acceptance ratios
-  that are far from the Goldilocks ratio.
-This process can be automated easily;
-  such automation is part of many projects that use MCMC.
-
-% HOGG:  Is the following a repeat from previously and/or does it mesh with what's earlier?
-
-A very common---%
-  and very useful---%
-  kind of proposal distribution is one that cycles through parameters,
-  taking a random step in just one parameter at a time.
-Aside from involving some persistence of state,
-  this kind of proposal distribution can be valuable,
-  in part because it reduces the (almost impossible) $d$-dimensional tuning problem
-  to $d$ one-dimensional problems:
-In this form of proposal distrubution,
-  it is possible to track a separate acceptance fraction for every parameter.
-Code can be built that uses the burn-in phase to tune all $d$ proposal variances
-  such that each of them, individually,
-  obtains acceptance at the same Goldilocks ratio.
-
-One note to make here---%
-  because it is relevant to tuning---%
-  is that tuning can \emph{only} take place during the burn-in phase;
-  you cannot tune \emph{while} you run your final MCMC run.
-Why not?
-Because tuning the proposal distribution based on the past history of the chain
-  violates the ``Markov'' property
-  that each step depends \emph{only} on the state at the previous step.
-Violating the Markov property can be very bad;
-  when you violate it,
-  you lose all the provable properties of MCMC
-  on which all our righteous power is based.
-
-Another proxy for autocorrelation time useful for tuning is the
-  Expected Squared Jump Distance.\note{(CITE GELMAN)}
-This is the mean squared distance the walker moves, per step.
-It is maximized when the acceptance ratio is reasonable
-  and the step size is large;
-  it is large when the exploration of the space is fast.
-This proxy is easy to measure and use for tuning,
-  and more directly related to autocorrelation time than the acceptance ratio.
-We recommend using it for tuning, though we have never used it ourselves.
-Like the acceptance fraction, it can also be used to tune
-  in the case that the proposal distribution loops over parameters;
-  once again, the user gets (in this case) $d$ one-dimensional tunings.
-
-When sampling really gets very slow,
-  there are various tricks and tips to work through.
-One is to see if any of the parameters has a conditional posterior
-  probability that is guaranteed to be Gaussian
-  (this is true for some linear parameters in least-square problems
-  when the priors are Gaussian, for example).
-These dimensions can often be sampled \emph{exactly},
-  outside of the MCCM framework;
-  the algorithm looks like:
-Take an MCMC step in the hard parameters,
-  then take an exact step in the easy parameters,
-  repeat.
-Another related trick is to analytically marginalize,
-  rather than sample,
-  easy nuisance parameters.
-The huge bag of possible tricks is so large,
-  it goes way beyond the scope of this introductory \documentname.
-In our experience, it is valuable to make friends with at least one statistician,
-  one applied mathematician, and one computer scientist.
-Between them, they ought to span the literature.
-
-\section{Initialization and burn-in}
-
-Just as most (though not all) MCMC methods require a choice of proposal distribution,
-  most (though not all) require a choice about initialization:
-The walker needs (or walkers need) to be started somewhere in the parameter space.
-In many use cases for MCMC,
-  the investigator wants to use it to obtain uncertainty information about
-  or propagate uncertainty into parameter estimates.
-In these cases, it makes sense to initialize the walker or walkers at sensible parameter estimates,
-  found by optimizing a likelihood or a posterior PDF in advance of sampling.
-In extremely high dimensions (large numbers of parameters), this can be a bad idea,
-  since the optimal parameters are not necessarily near typical posterior samples\note{Make the counterintuitive point.},
-  but in low dimensions (few to tens) this is often sensible.
-
-In other use cases,
-  the investigator wants to use MCMC to search all of parameter space
-  for good models, and also sample.
-In this case, it makes sense to be more catholic with initialization...
-
-...but note that MCMC is not the ideal search algorithm for any parameter space.
-
-...Start from different locations
-% BJB
-% What different locations?
-% Drawn from the prior can be useful. Then you can
-% look at the chains and see which ones "made it" and
-% which ones "got stuck" and delete the latter
-
-...Dealing with multi-modal problems
-% Starting from different locations is good
-% for DIAGNOSING a problem as multimodal but it's
-% terrible for quantifying the modes. Explain why,
-% it's a good teaching point.
+\section{Gibbs, ensemble, Hamiltonian, and so on}\label{sec:methods}
+
+As we have discussed above, MCMC methods are proven to be correct ``in the limit''; that is, when
+an infinite number of samples have been taken.
+When the function $f(\pars)$ has challenging properties, the results
+can approach this limit very slowly (or not at all\note{HOGG: discuss
+  situations in which MCMC will not work---related to support.}).
+For this reason, in the finite duration of a scientific project, most
+users of MCMC have one of two (related) problems:
+Either \textsl{(1)}~it is taking too long---it is taking too many steps or
+executing too many calls of the function $f(\pars)$ to get independent samples---or \textsl{(2)}~it is not
+exploring the full parameter space---there are local optima (or bad
+local geometry around those optima) ``trapping'' the algorithm.
+
+There is no general, problem-independent solution for either of these
+problems.
+Indeed, there are many different strategies that work well or badly
+for different functions $f(\pars)$.
+In this \sectionname, we will try to guide the reader towards methods
+that might help in different circumstances.
+We will not give a full description of all the MCMC algorithms out
+there, but just provide some field notes and pointers to references.
+New MCMC methods are being developed all the time, so a reader with a
+serious problem would do well to consult with the applied-mathematics
+or statistics literature.\note{HOGG: Your best interface to these
+  literatures is a colleague in another department.  Break down those walls!}
+The different methods to which we refer here are (in general) very
+different in their difficulty of implementation.
+We won't help with that either, except to say that there are more and
+more open-source or distributed packages every year.
+We will try to name some current implementations in the notes, but any
+list compiled today will be incomplete tomorrow.
+The spaces of MCMC methods and MCMC method implementations are both
+growing rapidly, in part because MCMC has become such a core
+technology in the empirical sciences.
+
+...Note a good general reference here?  MacKay?
+
+% *********** BJB talking:
+% I suggest partitioning the discussion of different methods into the following
+% two categories:
+%
+%  i) Methods for getting around the target distribution
+% ii) Methods that change the target distribution to something
+% other than the posterior pdf, to help the sampling somehow
+%
+% i) Would include things like M-H, the stretch move,
+% Hamiltonian, slice sampling (maybe skip the latter)
+% ii) would include things like annealing and Nested Sampling
+
+%%%% Hogg talking:
+% DFM: In this section, I think we should just summarize the other
+% methods, not describe them in detail.  Then we can cite the relevant
+% papers for the enthusiastic.
+
+\paragraph{Ensembles:}
+As we discuss more below, one of the critical challenges in the use of
+MCMC is tuning the proposal pdf $q(\pars'\given\pars)$.
+In problems in which all of the parameters (all of the components or entries of $\pars$) are somehow ``equivalent'',
+or the parameters can be seen as components of a vector in a vector
+space, there are \emph{ensemble methods} that make use of not just one
+random-walker but instead many walkers to automatically generate a properly tuned
+proposal distribution.
+These methods make use of the distribution of a set of independent
+walkers in the parameter space ($\pars$-space) to gauge the typical step-sizes and
+directions at which new proposals should be made, obviating or greatly
+reducing the tuning requirements.
+We are co-authors on one of these methods, \project{emcee}\note{HOGG:
+  ref emcee}, which is popular in the astrophysical sciences.
+Because ensemble methods are updating independent MCMC chains, they
+are often easy to parallelize to make use of multiple cores.
+
+\paragraph{Gibbs:}
+In many problems, the parameters---the components or entries of the
+blob $\pars$---are not equivalent at all.
+Some are global parameters, which touch or have an effect on every
+sub-part of the data blob $\data$, while others are local parameteres,
+which only touch a small subset of the data.
+Or, in another use case, some parameters are linear parameters
+that---at fixed values of the other parameters---will have a Gaussian
+likelihood or posterior pdf.
+In these problems, there are some directions in parameter space that
+are hard to move in (they require complete re-calculation of the full
+non-linear function $f(\pars)$, and other directions that are very
+easy to move in (they require only partial re-calculations or the
+sampling could even be analytic and exact).
+For problems with these structures, Gibbs samplers are ideal, and
+there are good implementations.\note{HOGG cite JAGS}
+
+A trivial kind of Gibbs sampling was mentioned above in the
+discussion of M--H MCMC:
+Even when running vanilla M--H MCMC, it is often useful to update only
+one parameter at a time; that is, do only axis-aligned moves in the
+$\pars$-space, cycling through axes as you go from step to step.
+This helps with tuning and diagnosis, but it also permits a good
+implementation to capitalize on simpler calculations for the simpler
+parameter directions.
+
+Gibbs sampling can be very good for exploiting multiprocessing:
+Global parameter updates might require a serial re-calculation of the
+full likelihood, but local parameter updates can be done in parallel,
+with each local parameter working on its local bit of data all at the
+same time.
+This kind of structure is common in hierarchical inference, where
+local parameters touch only individual objects in some population,
+say, and global parameters are parameters \emph{of} that population as
+a whole.
+Parallelization is built in to some Gibbs implementations.\note{HOGG}
+
+\paragraph{Hamiltonian:}
+[HOGG, DFM: THIS HAS SOME OTHER NAME; WHAT?]
+In many cases of interest, the function $f(\pars)$ can be analytically
+differentiated, such that it is possible to quickly execute a gradient
+$\dd f/\dd\pars$ evaluation.
+Indeed, in the modern world of computing, there are even
+auto-differentiation systems that deliver automatically code that
+produces the gradient of any function you can write.\note{HOGG cite
+  autodiff systems}
+There is a class of MCMC methods that make use of gradient information
+to effectively inform the proposal distribution and thus speed sampling.
+
+... HOGG more here!
+
+\paragraph{Importance:}
+Sometimes it is possible to sample efficiently from a distribution
+that is close to the target function $f(\pars)$ you care about but
+hard to sample efficiently with MCMC.
+An example is a problem that has a simple (Gaussian, say) likelihood
+function times a non-trivial prior:
+It is easy to sample from a Gaussian likelihood times a Gaussian
+prior, even without MCMC, but it might be hard to do the same sampling
+under a more complex prior.
+Importance sampling is a direct sampling method (not really an MCMC
+method at all):
+You take a sampling from the approximate function (which by assumption
+was easy) and then you re-weight the sampling (or do a subsequent
+rejection of samples) using the ratio of the true function to the
+approximate function as a weight or probability.
+We use this quite a bit in our hierarchical inferences.\note{HOGG:
+  Citations}
+
+Two of the significant technical details for importance sampling are the
+following:
+The approximate function must have the same (or greater) support as
+the true function, and the method degrades in efficiency as the two
+functions diverge.
+If the ratio of the true function to the approximate function is to be
+used as a probability for rejection of samples, then the true function
+must be greater than or equal to the approximate function everywhere!
+
+\paragraph{Slice:}
+... When is slice sampling a good idea?  DFM, can you help here?
+
+Slice sampling makes use of the point that (in one dimension) a fair
+sampling of $f(\pars)$ is a projection onto the $\pars$ axis of a
+uniform sampling of the two-dimensional region under the
+two-dimensional plot of $f(\pars)$ \vs\ $\pars$.
+The $D$-dimensional generalization is that a fair sampling of
+$f(\pars)$ is a projection onto the $\pars$ $D$-dimensional
+hypersurface of $\pars$ of a uniform sampling of the $D+1$-dimensional
+hypervolume under the hypersurface $f(\pars)$ in a hypothetical plot
+of $f(\pars)$ over the whole parameter space.
+Thinking in terms of the two-dimensional case (for simplicity), slice
+samplers make both ``horizontal'' moves in the $\pars$ direction at
+fixed $f(\pars)$ and also ``vertical'' moves in the $f(\pars)$
+direction.
+Because the sampling in the higher-dimensional space is uniform, it is
+easy to make moves to very different parts of parameter space, or
+different peaks or modes in the function $f(\pars)$.
+In our own research, we use a variant of slice sampling for problems
+involving samplings governed by Gaussian Processes.\note{DFM: Cite us
+  and Murray here.}
+
+\paragraph{Tempering:}
+A huge problem for MCMC samplers---which make local moves based on
+local positioning in the parameter space---is that there can be
+multiple modes in the function $f(\pars)$, separated by valleys of
+zero or near-zero density.
+A vanilla MCMC method cannot easily cross such valleys with local
+moves.
+The idea behind simulated tempering and related methods is to ``smooth
+out'' the function $f(\pars)$ or reduce its dynamic range, to make the
+peaks less tall and the valleys less deep, to permit the MCMC random
+walk to cross the valleys.
+The standard method is to introduce a ``temperature'' variable and,
+while sampling, take the likelihood function to the power of the
+inverse temperature (or, if working in the logarithm, multiply the log
+likelihood by the inverse temperature).
+When the temperature is high, this reduces the influence of the
+likelihood function relative to the prior pdf, and (assuming that the
+prior pdf is easy to sample) makes movement in the parameter space
+easier.
+At each step, when the temperature has moved away from unity, the
+distribution being sampled is not exactly $f(\pars)$, but the results
+of the sampling are only kept for those samples taken at unit
+temperature; these in the end make a fair sampling.
+
+\paragraph{Nested:}
+Similar to tempering-like methods are a class of methods called ``nested''
+samplers, which also smoothly change the target distribution away from $f(\pars)$.
+In the nested case, instead of increasing the temperature, the
+samplings are of a censored version of the prior, censored by the
+value of the likelihoood function.
+When the likelihood censoring is strong, only the most high-likelihood
+parts of the prior get sampled; when the likelihood censoring is weak,
+almost all of the prior gets sampled.
+The idea behind nested sampling---like tempering---is that it is
+designed to be a good method for exploring the full parameter space or
+searching for all of the modes of the posterior pdf.
+It is also designed such that it produces not just a sampling but also
+an estimate of the integral $Z$ of the likelihood times the prior
+(the Bayes factor or fully marginalized likelihood).
+
+\paragraph{Data augmentation:}
+...HOGG?
+
+\paragraph{Reversible-jump:}
+Finally, there are some extreme problems in which it is not just hard
+to sample in the parameter space, but the parameter space itself is
+not of fixed dimension.
+For example, if you are modeling an astronomical image with a
+collection of stars, and you don't know how many stars to use; in this
+case the number of stars itself is a parameter, so the number of
+parameters is itself a function of the parameters.\note{HOGG Cite
+  Brewer et al}
+In these cases you can only use certain kinds of samplers (though M--H
+MCMC is one you can choose), and, in addition, you need to make
+special kinds of proposals that permit the system to jump from one
+parameter space to another that has different dimensionality.
+In these cases, the concept of detailed balance becomes non-trivial,
+but there are criteria for creating reversible-jump proposals between
+the parameter spaces.
+These methods are best built and operated under the supervision of a
+trained professional.
 

### 2015-01-12 mcmc.tex
note to self

@@
   given the simplicity of the algorithm,
-  and given the intuition you gain by doing it (at least once) yourself.
+  and given the intuition you gain by doing it (at least once) yourself.\note{HOGG Refer again to the DAR on fitting a line for an example project.}
 

### 2015-01-23 mcmc.tex
notetoself
typesetting code; advice, notes to self, notation

@@
 % This file is part of the MCMC project,
-%   which is part of the Data Analysis Recipes project
+%   which is (in turn) part of the Data Analysis Recipes project.
 % Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).
@@
 %   but after the full stop if endnoting the sentence or paragraph.
-% - do we work in $x$ or $\pars$?  I am confused...
+% - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
 % - should we use the word ``walker'' anywhere, everywhere, or never?
@@
 
-Heuristically, a sampling $\setof{x_k}_{k=1}^K$ from some pdf $p(x)$
-is a set of $K$ values $x_k$ that are independent draws from the pdf.
+Heuristically, a sampling $\setof{\pars_k}_{k=1}^K$ from some pdf $p(\pars)$
+is a set of $K$ values $\pars_k$ that are independent draws from the pdf.
 Heuristically, if an enormous (large $K$) sampling could be displayed
-in a finely binned $x$-space histogram, the histogram would look---up
-to a total normalization---just like the original function $p(x)$.
+in a finely binned $\pars$-space histogram, the histogram would look---up
+to a total normalization---just like the original function $p(\pars)$.
 
@@
 any single-valued (scalar) function that is non-negative everywhere
-(the entire domain of $x$), and obeys a normalization condition
+(the entire domain of $\pars$), and obeys a normalization condition
 \begin{eqnarray}
-0 &\leq& p(x) \quad \mbox{for all $x$}
+0 &\leq& p(\pars) \quad \mbox{for all $\pars$}
 \\ \label{eq:normp}
-1 &=& \int p(x)\,\dd x
+1 &=& \int p(\pars)\,\dd\pars
 \quad ,
 \end{eqnarray}
-where, implicitly, the integral is over the full domain of $x$.
-Importantly, although $p(x)$ is single-valued, $x$ can be a scalar or
+where, implicitly, the integral is over the full domain of $\pars$.
+Importantly, although $p(\pars)$ is single-valued, $\pars$ can be a scalar or
 a multi-element vector or list or blob; it can be arbitrarily large
 and complicated.
-In data-analysis contexts, $x$ will often be the full blob of free
+In data-analysis contexts, $\pars$ will often be the full blob of free
 parameters in the model.
 Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
-has as many dimensions as there are elements or entries or components of $x$.
-Also, if there are elements or entries or components of $x$ that are discrete
+has as many dimensions as there are elements or entries or components of $\pars$.
+Also, if there are elements or entries or components of $\pars$ that are discrete
 (that is, take on only integer values or equivalent),
@@
 
-Given this pdf $p(x)$, we can define \emph{expectation values} $E[x]$
-for $x$ or for any quantity that can be expressed as a function $q(x)$
-of $x$:
+Given this pdf $p(\pars)$, we can define \emph{expectation values} $E[\pars]$
+for $\pars$ or for any quantity that can be expressed as a function $q(\pars)$
+of $\pars$:
 \begin{eqnarray}
-E[x] &\equiv& \int x\,p(x)\,\dd x
+E[\pars] &\equiv& \int \pars\,p(\pars)\,\dd \pars
 \\
-E[q] &\equiv& \int q(x)\,p(x)\,\dd x
+E[q] &\equiv& \int q(\pars)\,p(\pars)\,\dd \pars
 \quad ,
@@
 entire domain
-of $x$ (all the parts of $x$ space in which $p(x)$ is finite) and the
-integrals are multi-dimensional if $x$ is multi-dimensional.
-These expectation values are the mean values of $x$ and $q(x)$ under
+of $\pars$ (all the parts of $\pars$ space in which $p(\pars)$ is finite) and the
+integrals are multi-dimensional if $\pars$ is multi-dimensional.
+These expectation values are the mean values of $\pars$ and $q(\pars)$ under
 the pdf.  A good sampling---really the definition of a good
@@
 accurate.
-With a good sampling $\setof{x_k}_{k=1}^K$ the integrals get replaced
-with sums over samples $x_k$:
+With a good sampling $\setof{\pars_k}_{k=1}^K$ the integrals get replaced
+with sums over samples $\pars_k$:
 \begin{eqnarray}
-E[x] &\approx& \frac{1}{K}\,\sum_{k=1}^K x_k
+E[\pars] &\approx& \frac{1}{K}\,\sum_{k=1}^K \pars_k
 \\
-E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
+E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
 \quad .
@@
 As we noted above, in the context of MCMC, we are often sampling or
-using some \emph{badly normalized} function $f(x)$.
-This function is just the pdf $p(x)$ multiplied by some unknown and
+using some \emph{badly normalized} function $f(\pars)$.
+This function is just the pdf $p(\pars)$ multiplied by some unknown and
 hard-to-compute scalar.
-In this case, for our purposes, the conditions on $f(x)$ are that it
+In this case, for our purposes, the conditions on $f(\pars)$ are that it
 be non-negative everywhere and have \emph{finite} integral $Z$
 \begin{eqnarray}
-0 &\leq& f(x) \quad \mbox{for all $x$}
+0 &\leq& f(\pars) \quad \mbox{for all $\pars$}
 \\
-Z &=& \int f(x)\,\dd x
+Z &=& \int f(\pars)\,\dd \pars
 \quad .
@@
 
-When the sampling $\setof{x_k}_{k=1}^K$ is of one of these badly
-normalized functions $f(x)$---as it usually will be---the
+When the sampling $\setof{\pars_k}_{k=1}^K$ is of one of these badly
+normalized functions $f(\pars)$---as it usually will be---the
 sampling-approximation expectation values are the expectation values
@@
 \emph{never learn that normalization}.
-When we run MCMC sampling on $f(x)$, a function which differs from a
-pdf $p(x)$ by some unknown normalization constant, then the sampling
+When we run MCMC sampling on $f(\pars)$, a function which differs from a
+pdf $p(\pars)$ by some unknown normalization constant, then the sampling
 permits computation of the following kinds of quantities:
 \begin{eqnarray}
-E[q] &\equiv& \frac{\int q(x)\,f(x)\,\dd x}{\int f(x)\,\dd x}
+E[q] &\equiv& \frac{\int q(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
 \\
-E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
+E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
 \quad .
@@
 That is, the sampling can be constructed (as we will show below) from
-$f(x)$ directly, and it permits you to compute expectation values
+$f(\pars)$ directly, and it permits you to compute expectation values
 without ever requiring you to integrate either the numerator integral
@@
   see, the intractability comes from various directions, but one is
-  that the dimension of $x$ gets large in most realistic situations.
-  Another is that the support of $p(x)$ tends to be, in normal
+  that the dimension of $\pars$ gets large in most realistic situations.
+  Another is that the support of $p(\pars)$ tends to be, in normal
   inference situations, \emph{far smaller} than the full domain of
-  $x$.
+  $\pars$.
   That is, the pdf is at or very close to zero over all but some tiny
@@
 
-% in the next paragraph I adopt $d$ to be the dimensionality of the space
-
-For example, one magical property of a sampling (in a $d$-dimensional space)
-  is that a \emph{histogram} (a $d$-dimensional histogram) of the samples
+% in the next paragraph I adopt $D$ to be the dimensionality of the space
+
+For example, one magical property of a sampling (in a $D$-dimensional space)
+  is that a \emph{histogram} (a $D$-dimensional histogram) of the samples
   (divided, if you like, by the number of samples in each bin and the bin width\note{something about units and normalization})
@@
 Finally---and perhaps most importantly---a magical property of a sampling
-  (in a $d$-dimensional space)
-  is that if some of your $d$ dimensions are totally uninteresting
+  (in a $D$-dimensional space)
+  is that if some of your $D$ dimensions are totally uninteresting
   (nuisance parameters, if you will)
-  and some of your $d$ dimensions are of great interest,
-  the sampling in the full $d$-space is trivially converted into a sampling in the subspace of interest:
-You just drop from each $x_k$ vector (or blob or list) the dimensions of no interest!
+  and some of your $D$ dimensions are of great interest,
+  the sampling in the full $D$-space is trivially converted into a sampling in the subspace of interest:
+You just drop from each $\pars_k$ vector (or blob or list) the dimensions of no interest!
 That is, the projection of the sampling to the subspace of interest produces
@@
 That is, at step $k+1$ you might only propose a one-dimensional move
-  in the $i$th dimension, not in all $d$ dimensions of the parameter
+  in the $i$th dimension, not in all $D$ dimensions of the parameter
   space.
@@
   or else multiply in a (possibly nasty) Jacobian.
-That is, if your prior is ``flat in $x$''
-  but you find it easier to run your sampler by proposing steps in $\ln x$,
+That is, if your prior is ``flat in $\pars$''
+  but you find it easier to run your sampler by proposing steps in $\ln \pars$,
   if you don't modify your acceptance fraction by the Jacobian,
-  your \emph{real} prior won't be flat in $x$.
+  your \emph{real} prior won't be flat in $\pars$.
 These think-os can get subtle;
@@
   and how to choose the parameters.
-For example, in a $d$-dimensional parameter space, %do we consistently use d here?
+For example, in a $D$-dimensional parameter space, %do we consistently use d here?
   a zero-mean but anisotropic Gaussian (normal distribution)
@@
 Even within this choice (which is of one function among many possibilities),
-  there are $d\,(d+1)/2$ parameters in the $d\times d$ symmetric, positive definite covariance matrix
+  there are $D\,(D+1)/2$ parameters in the $D\times D$ symmetric, positive definite covariance matrix
   to set ``by hand''.
@@
   this kind of proposal distribution can be valuable,
-  in part because it reduces the (almost impossible) $d$-dimensional tuning problem
-  to $d$ one-dimensional problems:
+  in part because it reduces the (almost impossible) $D$-dimensional tuning problem
+  to $D$ one-dimensional problems:
 In this form of proposal distrubution,
   it is possible to track a separate acceptance fraction for every parameter.
-Code can be built that uses the burn-in phase to tune all $d$ proposal variances
+Code can be built that uses the burn-in phase to tune all $D$ proposal variances
   such that each of them, individually,
@@
   in the case that the proposal distribution loops over parameters;
-  once again, the user gets (in this case) $d$ one-dimensional tunings.
+  once again, the user gets (in this case) $D$ one-dimensional tunings.
 
@@
   and with very bad scaling.%
-  \note{If you have a $K$-point sampling in a $d$-dimensional parameter space,
+  \note{If you have a $K$-point sampling in a $D$-dimensional parameter space,
     there is some best sample $\pars_k^{\best}$.  Now imagine that you want to make
@@
     that is ten times closer to the true optimum.  How many more samples do you need to take?
-    In the limit that $K$ is large, on average, you have to take $10^d$ more samples!
+    In the limit that $K$ is large, on average, you have to take $10^D$ more samples!
     Most sensible optimizers are far, far better than this.}
@@
 
-MCMC is used to obtain samples $x_k$ from pdf $p(x)$,
-  or a badly normalized all-positive function $f(x)$
+MCMC is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
+  or a badly normalized all-positive function $f(\pars)$
   that is different from the pdf by an unknown factor $Z$.
 In the context of data analysis,
-  MCMC is usually being used to obtain samples
-  that are \emph{parameter values} for a probabilistic model of some data.
+  MCMC is usually being used to obtain samples $\pars_k$
+  that are \emph{parameter values} $\pars$ for a probabilistic model of some data.
 That is, MCMC is sampling a pdf \emph{for the parameters}.
@@
 
-...Give a sense of the difficulties you get into if you fail here; like
-discussion with Sven Kreiss on 2013-02-14...
-
-...Explain how your code \emph{should look} with a prior function and a likelihood function...
+% Hogg say:  I have no idea what the comment here (below) means, even though I wrote it!
+% ---
+% Give a sense of the difficulties you get into if you fail here; like
+% discussion with Sven Kreiss on 2013-02-14...
+% ---
+
+Ideally, if you are using MCMC for inference, your code should input
+to the MCMC function or routine a probability function which is a
+product of a prior times a likelihood.
+In terms of implementation, it is usually advisable to work in the
+logarithm of the density function, so the function input to the MCMC code
+would be called something like \code{ln_f()}.
+This function \code{ln_f()} internally would compute and return the
+sum of a log prior pdf \code{ln_prior()} and a log likelihood function
+\code{ln_likelihood()}.
+(When working in logarithmic probability, in M--H-like MCMC methods,
+the accept--reject step would change from a comparison of a random
+deviate with a ratio of probabilities to comparison of the log of a
+random deviate with a difference of log probabilities.)
+
+If you are using ``flat'' (improper) priors, the \code{ln_prior()}
+function can just return a zero no matter what the parameters.
+If it is flat with bounds (that is, proper), the \code{ln_prior()}
+should check the bounds and return \code{-Inf} values if the parameter
+vector is out of bounds.
+The pseudo-code for the \code{ln_f()} function should look something
+like this:
+\begin{verbatim}
+def ln_f(pars, data):
+    x = ln_prior(pars)
+    if is_Inf(x):
+        return -Inf
+    return x + ln_likelihood(data, pars)
+\end{verbatim}
+This pseudo-code ensures that you only compute the likelihood when the
+parameters are within the prior bounds; it assumes implicitly that the
+prior pdf is easier to compute than the likelihood.
+If you find yourself in the opposite regime, adjust accordingly.
 
@@
   and check that the sampling has the moments you expect (to within tolerances).
-For example, you can sample a $d$-dimensional Gaussian distribution with known non-trivial covariance,
+For example, you can sample a $D$-dimensional Gaussian distribution with known non-trivial covariance,
   and check that the empirical covariance comes out as expected.
@@
 In the most extreme form of this test,
-  for a Gaussian in $d$ dimensions,
+  for a Gaussian in $D$ dimensions,
   build the covariance matrix input to the Gaussian distribution
-  by taking the sum of the outer product of $d+1$ randomly generated vectors.
+  by taking the sum of the outer product of $D+1$ randomly generated vectors.
 Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
@@
   or until you are confident you are getting good samples.
-Check that the empirical $d$-dimensional covariance of the samples
+Check that the empirical $D$-dimensional covariance of the samples
   has the same (or similar) eigenvectors and eigenvlues to the input matrix.
 Also check that the individual one-dimensional samplings
-  (along each axis or any projection in the $d$-space)
+  (along each axis or any projection in the $D$-space)
   look Gaussian and have nearly vanishing skew and kurtosis.

### 2015-01-26 mcmc.tex
tweaks
copyright

@@
     license\note{\label{note:first}
-        Copyright 2013, 2014 by the authors. This work is licensed under a
+        Copyright 2013, 2014, 2015 by the authors. This work is licensed under a
         \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
@@
     x = ln_prior(pars)
-    if is_Inf(x):
+    if not is_finite(x):
         return -Inf
@@
 \paragraph{visualization:}
-...Make triangle plots.  Great value!...  Show example here.
+...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?
 

### 2015-01-26 mcmc.tex
negative infinities

@@
 If you find yourself in the opposite regime, adjust accordingly.
+Of course the above pseudo-code presumes that when you perform the
+accept--reject step of the MCMC method, the programming language handles
+properly the \code{-Inf} values.\footnote{If you are working in a language
+that doesn't have an \code{Inf} or doesn't evaluate comparisons correctly
+when the \code{Inf} appears, you might have to write some case code, and
+have your \code{ln_f} function return a value \emph{and} some kind of flag
+which indicates ``zero probability'' or $f(\pars)=0$.}
 
@@
 
-SHOW EXAMPLES?
+...SHOW EXAMPLES?
 
@@
 
+\paragraph{bad initialization:}
+If you initialize your sampler in a disallowed region of parameter
+space---that is, a part of the space with $f(\pars)=0$ (so your
+\code{ln_f()} function returns \code{-Inf})---it might have a very
+hard time random-walking out of that location.
+At initialization time, it is important to test that the
+initialization is permitted.
+For ensemble methods, it is important to test that \emph{all} the
+walkers in the ensemble are initialized to permitted values.
+
 \paragraph{parameterization problems:}
@@
 
-HOGG:  How to deal with discrete or integer or binary parameters?
+HOGG: ...How to deal with discrete or integer or binary parameters?
 
@@
 
-%% ...How to deal with forbidden regions in prior, or sharp prior boundaries?
-
 %% ...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.

### 2015-08-17 mcmc.tex
notes to self
tiny tweaks

@@
 % - mention JAGS and STAN in methods section
-% - perhaps re-order to m-h, convergence, reporting results, other methods ??
+% - when we say ``below'' also give a Section Number.
 % - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
@@
 \emph{never learn that normalization}.
-When we run MCMC sampling on $f(\pars)$, a function which differs from a
+When we run MCMC sampling on $f(\pars)$, a function that differs from a
 pdf $p(\pars)$ by some unknown normalization constant, then the sampling
@@
   space gets accepted into the list of samples or else the previous sample
-  in the parameter space gets repeated.
+  in the parameter space \emph{gets repeated}.
 The algorithm can be iterated a large number $K$ of times to produce $K$ samples.
@@
 It is important, when reporting the output of an MCMC run by giving
-means or medians of the posterior PDF to remind the reder or user of
+means or medians of the posterior PDF to remind the reader or user of
 that output that it does not necessarily represent (collectively) a
@@
 parameter directions.
+
+%% DWH:  Note here that this requires writing the f(\pars) code to capitalize on the axis-aligned moves!!
 
@@
 ...The Bayes integral / evidence integral?
+It is both the biggest strength and biggest weakness that vanilla MCMC methods can't give you this integral.
+There are tricks, and then there are MCMC methods.
+All that said, you never really want to calculate this integral!
 

### 2015-10-04 mcmc.tex
author affils, small text tweaks

@@
 %   which is (in turn) part of the Data Analysis Recipes project.
-% Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).
+% Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).
 
@@
 % -----
+% - deal with all mentions of ``DFM'' or ``HOGG''
+% - mention Gelman--Rubin statistic
 % - mention JAGS and STAN in methods section
@@
 % - send to friendlies for comments
+%   - Gelman
 %   - Lang
@@
 %   - Brewer
-%   - Hill
 % - post on arXiv
@@
 Dan~Foreman-Mackey\\
+\affil{Department~of~Astronomy, University~of~Washington}\\
 \affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
@@
 David~W.~Hogg\\
+\affil{Simons~Center~for~Data~Analysis,
+       Simons~Foundation}\\
 \affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
@@
 These expectation values are the mean values of $\pars$ and $q(\pars)$ under
-the pdf.  A good sampling---really the definition of a good
+the pdf.  A good sampling---and really this is the definition of a good
 sampling---makes the sampling approximation to these integrals
@@
 
-\section{What is Metropolis--Hastings MCMC?}
+\section{Metropolis--Hastings MCMC}
 
@@
 \end{eqnarray}
+%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING?
 Heuristically, the autocorrelation time is the number of samples you
@@
 
-A sampler with a better autocorrelation time is just better; you have
+A sampler with a smaller autocorrelation time is just better; you have
 to do fewer $f(\pars)$ calls per independent sample, and you have to
@@
 
-%% DWH:  Note here that this requires writing the f(\pars) code to capitalize on the axis-aligned moves!!
+%% HOGG:  Note here that this requires writing the f(\pars) code to capitalize on the axis-aligned moves!!
 
@@
 
-% Hogg say:  I have no idea what the comment here (below) means, even though I wrote it!
+% HOGG say:  I have no idea what the comment here (below) means, even though I wrote it!
 % ---

### 2015-10-04 mcmc.tex
tweaks and comments on comments

@@
 % - deal with all mentions of ``DFM'' or ``HOGG''
+% - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
 % - mention Gelman--Rubin statistic
@@
 % - need to mention testing the likelihood function
+% - would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
 % - send to friendlies for comments
@@
   the factor $Z$ in \equationname~(\ref{eq:bayes}) something like
-  ``$p(\data)$'' or ``the marginal likelihood'' or the 
+  ``$p(\data)$'' or ``the fully marginalized likelihood'' or the 
   probability of the data.
@@
   space under consideration!
-  We discuss this object and these issues elsewhere (HOGG CITE).
+  We (intend to) discuss this object and these issues elsewhere
+  (in various things in preparation).
   The factor $Z$ is often hugely difficult to compute, because the
@@
   containing not just a value but some huge arbitrarily structured set
-  of values in matrix, vector, or more complex form.}).
+  of values in matrix, vector, or an ordered list of any set of such objects,
+  or perhaps some even more complex form.}).
 You also have a model sophisticated enough---a probabilistic,
 generative model, if you will\note{For more discussion
-  of our definition of a \emph{model}, see [HOGG SOME OTHER NOTE] in
+  of our definition of a \emph{model}, see [some other as-yet-unwritten \documentname] in
   this series.
@@
   are a true subjective Bayesian) a probability distribution (pdf)
-  over all possible data.
+  over all possible data.%
 }---that, given a
@@
   structured set or list of parameters, possibly even with no fixed
-  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We
-  bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
-  series.}) $p(\data\given\pars)$.
+  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We will
+  bash the terminology ``likelihood'' in another, as-yet-unwritten, \documentname\ in this
+  series.  Techically, this thing $p(\data\given\pars)$
+  is only properly a likelihood function when
+  we are thinking of the data $\data$ as being fixed, and the parameters $\pars$
+  as being permitted
+  to vary.}) $p(\data\given\pars)$.
 Furthermore, say you also can write down some kind of informative or
@@
     that is ten times closer to the true optimum.  How many more samples do you need to take?
-    In the limit that $K$ is large, on average, you have to take $10^D$ more samples!
-    Most sensible optimizers are far, far better than this.}
+    In the limit that $K$ is large, on average, you have to take a factor of
+    $10^D$ more samples!
+    Any reasonable optimizer is far, far better than this (in terms of compute time).}
 For three, if you just want an optimum, use an optimizer!%
@@
 If you are clever, all these plots can be arranged into a lovely and
-informative triangle.\note{HOGG cite triangle.py and put an example in
+informative triangle.\note{HOGG cite corner.py and put an example in
   the note!}
@@
 
-... HOGG more here!
+... HOGG more here!  For example, does this work better in large
+numbers of dimensions?  If so, we should say so.
 

### 2015-11-08 mcmc.tex
refs

@@
 % -----
-% - deal with all mentions of ``DFM'' or ``HOGG''
+% - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``...''
+% - Should we cite the Brooks book or just chapters within it?  I think the latter!
 % - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
@@
 \begin{thebibliography}{}\markright{References}
-\bibitem[Foreman-Mackey et al.(2012)]{emcee}
+\bibitem[Foreman-Mackey \etal(2012)]{emcee}
   Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2012,
   \project{emcee}: The MCMC Hammer,
-  \arxiv{1202.3665}
+  \arxiv{1202.3665}.
+\bibitem[Brooks \etal(2011)]{mcmchandbook}
+  Brooks,~S., Gelman~A., Jones~G., \& Meng,~X.-L., eds.\ 2011,
+  \textit{Handbook of Markov Chain Monte Carlo},
+  Chapman and Hall~/~CRC Press
+  \isbn{978-1-4200-7941-8}.
 \end{thebibliography}

### 2015-11-08 mcmc.tex
cite fengji

@@
   in the (presumably high dimensional) parameter space $\pars$.
-  Elsewhere, we discuss the computation of this object (HOGG cite Fengji)
-  and---perhaps more importantly---why you probably never really
+  Elsewhere, we discuss the computation of this object (\citealt{fengji}),
+  and so have many others before us.
+  Perhaps more importantly, we also have views about why you probably never really
   \emph{want} to compute it (HOGG CITE).
@@
 \begin{thebibliography}{}\markright{References}
-\bibitem[Foreman-Mackey \etal(2012)]{emcee}
-  Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2012,
-  \project{emcee}: The MCMC Hammer,
-  \arxiv{1202.3665}.
 \bibitem[Brooks \etal(2011)]{mcmchandbook}
@@
   \isbn{978-1-4200-7941-8}.
+\bibitem[Foreman-Mackey \etal(2012)]{emcee}
+  Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2012,
+  \project{emcee}: The MCMC Hammer,
+  (also \arxiv{1202.3665}).
+\bibitem[Hou \etal(2014)]{fengji}
+  Hou,~F., Goodman,~J., \& Hogg,~D.~W., 2014,
+  The probabilities of orbital-companion models for stellar radial velocity data,
+  \arxiv{1401.6128}.
 \end{thebibliography}

### 2015-11-29 mcmc.tex
formatting tweaks
model checking pwned

@@
 % - Should we cite the Brooks book or just chapters within it?  I think the latter!
+% - Should we make an \acronym{} environment and use it religiously?  I think yes.
 % - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
@@
 
-\newcommand{\this}{Using Markov Chain Monte Carlo}
+\newcommand{\chaptertitle}{Using Markov Chain Monte Carlo}
 \include{styles/dar}
@@
 
-\begin{document}
-
-\thispagestyle{plain}\sloppy\sloppypar\raggedbottom
-\section*{Data analysis recipes:\\ \this\footnotemark}
+\begin{document}\sloppy\sloppypar\raggedbottom\thispagestyle{plain}%
+%
+\section*{Data analysis recipes:\\ \chaptertitle\footnotemark}
 
@@
 \paragraph{model checking:}
-...Beyond the scope of this document...
-Something about cross-validation and posterior predictive checks...
-Something about plotting models (drawn from posterior) on data, where appropriate...
-How could you possibly know that your sampling really represents your \emph{real} posterior PDF?..
-
-...The Bayes integral / evidence integral?
-It is both the biggest strength and biggest weakness that vanilla MCMC methods can't give you this integral.
-There are tricks, and then there are MCMC methods.
-All that said, you never really want to calculate this integral!
-
-...Posterior predictive checks?
+\emph{All models are wrong, but some are useful.}
+That's not a quote from Box, but it's close.\note{The real quotation
+  is a parenthetical sentence ``Remember that all models are wrong;
+  the practical question is how wrong do they have to be to not be
+  useful.'' (\citealt{box}, p.~74).  It is amazing how rarely this is
+  ever quoted correctly; or maybe there is another source for this
+  quotation?}
+The point (for our purposes) is that if you check your model hard
+enough---that is, you take enough data---you will certainly rule it
+out.\note{Physicists sometimes forget this, and believe that various
+  physical theories, such as quantum electrodynamics or the
+  cold-dark-matter cosmological model, are strictly true.  First of
+  all, there is no way to know that for sure, and second of all, even
+  if they \emph{are} true, they don't precisely explain any
+  observation, which is also affected by various kinds of auxilliary
+  effects and noises for which there is (and can be, for deep reasons,
+  discussed elsewhere by us; CITE) no extremely precise model.  So in
+  detail, any model of any specific observation must be wrong, even in
+  the (exceedingly unlikely) event that the fundamental model is
+  correct.}
+``Model checking'' is an enormous subject that goes way beyond the
+scope of this \documentname, for which reason we won't say much about
+it here except to make two points:
+
+The first---and most important---point about model checking is that
+you always learn a huge amount by checking the model.
+What this entails is looking at the distribution of residuals or the
+quality of the predictions of the model when compared to the data.
+There are methods (like the chi-squared statistic and the Bayesian
+evidence integral) that check the absolute quality of the model in a
+scalar way.
+These are usually uninformative, because (as we noted), all models are
+wrong, and eventually the data will be good enough to return a bad
+statistic here.\note{Sometimes Bayesians like to complain that the
+  chi-squared statistic is a measure of the size of your data!  That's
+  true when your model is wrong (that is, always).}
+
+Much more informative are methods (like plotting residuals of the data
+away from the model in the space of the data) that look at what parts
+of the data the model ``explains well'' and what parts of the data the
+model ``explains badly''.
+These kinds of experiments usually reveal inadequacies of the model in
+a rich way, or confirm that certain assumptions are bad.
+Cross-validation\note{CITE SOMETHING} is a framework for constructing
+valuable tests of this form, as is posterior predictive checks\note{CITE SOMETHING}.
+Importantly, you want to find ways to compare your model to your data
+\emph{in the space of the data}.
+Don't just look at the parameter space of your model!
+It is the data that really exist, especially when all models are wrong!
+
+There are model checking methods (like varying the prior and
+re-running MCMC) that test the sensitivity of results to assumptions.
+Not only are these very valuable for testing the model, they are
+essential for writing any paper about your results: The reader always
+wants to know the sensitivity of results to assumptions.\note{And be
+  sure not to forget that there are (at least) as many assumptions
+  encoded in your likelihood function as there are in your prior
+  pdfs.}
+Once again, it is important to find ways to display the results of
+such tests in the space of the data.
+All we will say here about these kinds of rich model tests is that you
+should do them, and report the results.
+
+The second---and less important---point about model checking is that
+the ur-Bayesian thing to be doing is computing a Bayesian evidence
+integral.
+However, it is usually inadvisable:
+The integral is expensive to compute.
+Most MCMC methods are cleverly designed to \emph{avoid} computing this
+integral; so it usually adds expense to the project.
+At the end of the day, it returns only a single, scalar number, with
+no guidance about how to use it (and indeed there are only heuristics
+to reach for).
+If your model is wrong, the richer tests of visualization of residuals
+or sensitivity to assumptions are much more likely to lead to insight
+about what changes to make or what new directions to explore.
 
@@
 \clearpage
+\raggedright
 \begin{thebibliography}{}\markright{References}
+\bibitem[Box \& Draper(1987)]{box}
+  Box,~G.~E.~P. \& Draper,~N.~R.\ 1987,
+  \textit{Empirical Model Building and Response Surfaces},
+  John Wiley \& Sons, New York, NY
+  \isbn{978-0471810339}.
 \bibitem[Brooks \etal(2011)]{mcmchandbook}
@@
   \project{emcee}: The MCMC Hammer,
+  \textit{Pubs.\,Astron.\,Soc.\,Pac.}\ \textbf{125} 306--312
+  \doi{10.1086/670067}
   (also \arxiv{1202.3665}).

### 2015-11-29 mcmc.tex
tiny tweaks

@@
 % -----
-% - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``...''
+% - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``CITE'' and ``...''
 % - Should we cite the Brooks book or just chapters within it?  I think the latter!
@@
 % - mention JAGS and STAN in methods section
-% - when we say ``below'' also give a Section Number.
+% - when we say ``below'' or ``above'' also give a Section Number.
 % - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
@@
 
-\section{When do you need MCMC?}
+\section{When do you need MCMC?}\label{sec:when}
 
@@
 
-As we noted above, in the context of MCMC, we are often sampling or
+As we noted above (\sectionname~\ref{sec:when}), in the context of MCMC, we are often sampling or
 using some \emph{badly normalized} function $f(\pars)$.
@@
 Most MCMC methods are cleverly designed to \emph{avoid} computing this
-integral; so it usually adds expense to the project.
+integral (as we note above in \sectionname~\ref{sec:when}); so it usually adds
+expense to the project.
 At the end of the day, it returns only a single, scalar number, with

### 2015-12-05 mcmc.tex
words about initialization

@@
 % - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
-% - should we use the word ``walker'' anywhere, everywhere, or never?
+% - should we use the word ``walker'' everywhere and do we?
 
@@
 
+DWH: ...Why does M--H MCMC work (briefly)?  Here or later?
+
 This algorithm---and indeed any MCMC algorithm---produces a
@@
 and diagnose problems (\sectionname~\ref{sec:troubleshooting}).
-
-...Why does M--H MCMC work (briefly)? ...Or put this in a note on an earlier paragraph?
 
@@
 
-Just as most (though not all) MCMC methods require a choice of proposal distribution,
-  most (though not all) require a choice about initialization:
-The walker needs (or walkers need) to be started somewhere in the parameter space.
-In many use cases for MCMC,
-  the investigator wants to use it to obtain uncertainty information about
-  or propagate uncertainty into parameter estimates.
-In these cases, it makes sense to initialize the walker or walkers at sensible parameter estimates,
-  found by optimizing a likelihood or a posterior PDF in advance of sampling.
-In extremely high dimensions (large numbers of parameters), this can be a bad idea,
-  since the optimal parameters are not necessarily near typical posterior samples\note{Make the counterintuitive point.},
-  but in low dimensions (few to tens) this is often sensible.
-
-In other use cases,
-  the investigator wants to use MCMC to search all of parameter space
-  for good models, and also sample.
-In this case, it makes sense to be more catholic with initialization...
-
-...but note that MCMC is not the ideal search algorithm for any parameter space.
-
-...Start from different locations
-% BJB
-% What different locations?
-% Drawn from the prior can be useful. Then you can
-% look at the chains and see which ones "made it" and
-% which ones "got stuck" and delete the latter
-
-...Dealing with multi-modal problems
-% Starting from different locations is good
-% for DIAGNOSING a problem as multimodal but it's
-% terrible for quantifying the modes. Explain why,
-% it's a good teaching point.
+Just as most (though not all) MCMC methods require a choice of
+proposal distribution, most (though not all) require a choice about
+initialization:
+The walker needs (or walkers need) to be started somewhere in the
+parameter space.
+In many use cases for MCMC, the investigator wants to use it to obtain
+uncertainty information about or propagate uncertainty into parameter
+estimates.
+In these cases, it makes sense to initialize the walker or walkers at
+sensible parameter estimates, found by optimizing a likelihood or a
+posterior PDF in advance of sampling.
+In extremely high dimensions (large numbers of parameters), this can
+be a bad idea, since the optimal parameters are not necessarily near
+typical posterior samples\note{When the dimensionality of the
+  parameter space gets very large, almost all samples will be
+  \emph{very far} from the optimum of the posterior pdf.  One way to
+  think about this is to think of a random Gaussian draw in $D$
+  dimensions: A typical draw will be $\sqrt{D}$ standard deviations
+  Euclidean distance from the maximum of the Gaussian.  That's a long
+  way when $D=100$ or $1000$!  Another way to think about it is that
+  there are a lot of ways to move \emph{away} from the point you care
+  about, but very few ways to move close to it.  The upshot is that
+  even with an \emph{enormous} sampling, if you are in large
+  dimensions, not a single sample will be very near the optimum.}, but
+in low dimensions (few to tens) this is often sensible.
+
+Ideally, you will initialize the walker not at a completely irrelevant
+point, nor at the optimum of the posterior pdf, but at a typical or
+pretty good place in the posterior pdf.
+That would minimize burn-in!
+But, in general, being at the optimum or anywhere near a good place is
+usually better than a mindless initialization.
+
+If you suspect, or if it is even possible, that your problem is badly
+multi-modal, then you will have to start at multiple points in
+parameter space and compare the resulting chains.
+By ``badly multi-modal'' we mean that there are peaks in the posterior
+pdf that are connected by low-enough valleys that it is very unlikely
+or takes a long time for a walker to traverse from one peak to the
+next.
+If this might be an issue, then the best diagnosis is to start many
+chains in parallel and check that they lead to identical (or very
+similar) posterior inferences (same pdf location and shape and
+statistics).
+
+Of course if you find out the worst---if you find out that different
+initializations lead to different posterior inferences---you are in
+trouble:
+There is no trivial way to combine together the samplings you get of
+different modes!
+You can either wait a \emph{very long time} so that you see a single
+walker in a single MCMC run traverse from mode to mode enough times to
+make a representative sampling (dozens or hundreds of times, ideally),
+or you can use a high-end sampler (like nested sampling) that is
+designed for such problems.
+In principle there are methods that involve splitting the space into
+two spaces, sampling them separately, and then combining the chains
+according to their relative evidence afterwards.
+That's a research project beyond the scope of this \documentname.
+
+DFM: IS THIS PARAGRAPH USEFUL OR CLEAR?
+Of course there are some kinds of multi-modalities that aren't a
+problem.
+For example, you can have two solutions to a problem that differ only
+in the \emph{labeling} of the components---say the Gaussians making up
+a mixture of Gaussians---that make up the model.
+One solution has the $K$ Gaussian components in one order, and the
+other in another order, but they are identical Gaussians in all
+other respects.
+Because these two solutions make the same predictions for any data, and
+only differ in their irrelevant, latent, internal ``naming'' of components,
+they are not really different solutions.
+This suggests that when you ask whether the outputs of two MCMC chains
+are consistent with one another, you should do so in the realm of the
+parameters you \emph{care about}, not irrelevant parameters that do
+not have impact on any present or future data.
+
+But it is worth remembering that just as MCMC is not a good optimizer
+(generically, samples will not lie close to the maximum of the posterior pdf),
+it is also not a good \emph{search algorithm}.
+There is no sense in which (standard) MCMC methods are efficient at
+searching, or engineered to search, all of parameter space.
+If you really need to check all of parameter space, you should put
+MCMC aside, and do an exhaustive search.\note{You might want to open a
+  bottle of wine or two while the code runs.}
 

### 2015-12-05 mcmc.tex
notes
some content, notes to selves

@@
 % - when we say ``below'' or ``above'' also give a Section Number.
-% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
 % - need to mention case where there is an MC integral INSIDE the likelihood function
@@
   where there are always parameters with very different levels of importance to the scientific conclusions.
-This point will return again below, either explicitly or implicitly. % HOGG: does it?
+This point will return again below, either explicitly or implicitly.
+[HOGG: does it?]
 
@@
 
-DWH: ...Why does M--H MCMC work (briefly)?  Here or later?
+HOGG: ...Why does M--H MCMC work (briefly)?  Here or later?
 
@@
 \end{eqnarray}
-%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING?
+%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING? OR A SUM FROM -infty to +infty?
 Heuristically, the autocorrelation time is the number of samples you
@@
 serious problem would do well to consult with the applied-mathematics
-or statistics literature.\note{HOGG: Your best interface to these
-  literatures is a colleague in another department.  Break down those walls!}
+or statistics literature.\note{Your best interface to these
+  literatures is often a colleague in another department, like applied
+  mathematics, statistics, or computer science.  Break down those
+  walls!}
 The different methods to which we refer here are (in general) very
@@
 We will try to name some current implementations in the notes, but any
-list compiled today will be incomplete tomorrow.
+list compiled today will be incomplete tomorrow.\note{One check-point
+  in this literature is the compliation volume by
+  \citet{mcmchandbook}, but much has been developed and changed even
+  since that was written.}
 The spaces of MCMC methods and MCMC method implementations are both
@@
 
-...Note a good general reference here?  MacKay?
+All we are going to do here is give a cursory mention to some methods
+we know about, with some keywords that can be used to search more
+deeply.
 
@@
 
-... HOGG more here!  For example, does this work better in large
-numbers of dimensions?  If so, we should say so.
+DFM: more here!  For example, does this work better in large numbers
+of dimensions?  If so, we should say so.  And references, please.
 
@@
 like this:
+% DFM: Do you have a preferred environment for code other than verbatim?
 \begin{verbatim}
@@
 
-...It is a good idea to have proper priors.
+HOGG: ...It is a good idea to have proper priors.
 It is a good test of your priors that you can sample with a ``null'' likelihood function.
@@
 
-...You must describe your priors in the same parameterization that you ``step'' in.
+HOGG: ...You must describe your priors in the same parameterization that you ``step'' in.
 That is, if you change your parameters, you \emph{must} change your priors correspondingly.
@@
 \paragraph{visualization:}
-...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?
+HOGG:...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?
 
@@
 
-...SHOW EXAMPLES?
+DFM:...SHOW EXAMPLES?
 
@@
 
-SHOW EXAMPLES OF GOOD AND BAD?
+DFM: SHOW EXAMPLES OF GOOD AND BAD?
 

### 2016-01-07 mcmc.tex
uncertainty and tau

@@
 
+% HOGG: \Var NEEDS TO BE DEFINED:
+\newcommand{\Var}{\mathrm{Var}}
 \newcommand{\data}{D}
@@
 E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
+\label{eq:expectationvalue}
 \quad .
@@
   given the simplicity of the algorithm,
-  and given the intuition you gain by doing it (at least once) yourself.\note{HOGG Refer again to the DAR on fitting a line for an example project.}
+  and given the intuition you gain by doing it (at least once) yourself.\note{%
+    HOGG Refer again to the DAR on fitting a line for an example project.}
 
@@
 Heuristically, the autocorrelation time is the number of samples you
-need to take before you get an independent sample.
+need to take before you get a new independent sample.
 More precisely, the autocorrelation time is the factor by which you
 have to reduce the number of samples when you compute the variance on
-a sampling-based expectation value estimate.\note{HOGG Explain more fully.  Maybe also cite Fengji.}
+a sampling-based expectation value estimate.
+
+That is, if the expectation value $E[q]$ is computed by the sampling
+approximation given in \equationname~(\ref{eq:expectationvalue}),
+there is an uncertainty on that expectation given by
+\begin{eqnarray}
+  \mbox{$E[q]$ uncertainty} &=& \left[\frac{\tau}{K}\,\Var[q]\right]^{1/2}
+  \\
+  &\approx& \left[\frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2\right]^{1/2}
+  \quad ,
+\end{eqnarray}
+where we are being a bit loose with terminology because this is the
+uncertainty not on $E[q]$ but on our sampling-based \emph{estimator} for
+$E[q]$, and the $E[q]$ inside the sum is that estimator too.
 It is slightly off-topic here, but related to this, if you have a long

### 2016-01-07 mcmc.tex
gibbs implementation note

@@
     Any reasonable optimizer is far, far better than this (in terms of compute time).}
-For three, if you just want an optimum, use an optimizer!%
-  \note{HOGG, something?}
+For three, if you just want an optimum, use an optimizer!
 That is, we \emph{strongly advise against} reporting,
@@
 Either \textsl{(1)}~it is taking too long---it is taking too many steps or
-executing too many calls of the function $f(\pars)$ to get independent samples---or \textsl{(2)}~it is not
+executing too many calls of the function $f(\pars)$ to get independent
+samples---or \textsl{(2)}~it is not
 exploring the full parameter space---there are local optima (or bad
@@
 implementation to capitalize on simpler calculations for the simpler
-parameter directions.
-
-%% HOGG:  Note here that this requires writing the f(\pars) code to capitalize on the axis-aligned moves!!
+parameter directions.\note{%
+  The point is that sometimes there are some parameters that only
+  affect a very small number of data points, or parameters that change
+  the model or the likelihood function and prior values in a very
+  simple-to-compute way.  If such simple parameters exist, it makes
+  sense to write your likelihood and prior code to capitalize on these
+  simplicities.  This is not always trivial, because it may involve
+  caching parts of the calculation and so on.  Details would be beyond
+  our scope.}
 

### 2016-01-25 mcmc.tex
testing
years and author order

@@
 %   which is (in turn) part of the Data Analysis Recipes project.
-% Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).
+% Copyright 2013, 2014, 2015, 2016 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).
 
@@
 % -----
+% - [delete this]
 % - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``CITE'' and ``...''
@@
 \noindent
-Dan~Foreman-Mackey\\
-\affil{Department~of~Astronomy, University~of~Washington}\\
-\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
-       New~York~University}\\[1ex]
 David~W.~Hogg\\
@@
        New~York~University}\\
-\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}
+\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}\\[1ex]
+Dan~Foreman-Mackey\\
+\affil{Department~of~Astronomy, University~of~Washington}
 

### 2016-01-25 mcmc.tex
author, abstract, and intro tweaks

@@
 % - mention JAGS and STAN in methods section
+% - follow up the idea that burn-in doesn't exist at all!
 % - when we say ``below'' or ``above'' also give a Section Number.
@@
 Markov Chain Monte Carlo (MCMC) methods for sampling probability
-distribution functions, plus abundant computational resources,
+distribution functions (combined with abundant computational resources)
 have transformed the sciences.
-Here we give a fast overview of basic MCMC methods and then turn to
+Here we give a brief overview of basic MCMC methods and then turn to
 practical advice for their use in real inference problems.
 We give advice on method choice, tuning for performance,
-initialization and burn-in, judging convergence, and use of the chain
+methods for initialization, tests of convergence, and use of the chain
 output to produce parameter estimates with associated uncertainties.
-[Insert some generally useful point here!]
 \end{abstract}
@@
 In these contexts, the constant $Z$, sometimes written as
-$p(\data)$, is known by the names ``evidence'', ``marginal likelihood''
-the ``Bayes integral'' and ``prior predictive probability'', and is usually
-\emph{extremely hard to
-  calculate}.\note{Some would call
+$p(\data)$, is known by the names ``evidence'', ``marginal likelihood'',
+``Bayes integral'', and ``prior predictive probability'', and is usually
+\emph{extremely hard to calculate}.\note{Some would call
   the factor $Z$ in \equationname~(\ref{eq:bayes}) something like
@@
 
-Say you are in this situation.
+Say you are in this situation:
 You have a huge blob of data $\data$ (think of this as a vector or
@@
   to vary.}) $p(\data\given\pars)$.
-Furthermore, say you also can write down some kind of informative or
-vague prior pdf $p(\pars)$ for the parameter blob.
+Furthermore, say also that you can write down some kind of informative or
+vague prior pdf $p(\pars)$ for the parameter blob $\pars$.
 If all these things are true, then---even if you can't compute
@@
   most uses of the methods in question are not appropriate!}, we are
-compelled to note that MCMC is in fact \emph{over-used}.
+compelled to note at the outset that MCMC is in fact \emph{over-used}.
 Because MCMC provably (under assumptions\note{Hint at assumptions
@@
 \emph{search the parameter space for good models}.
+This is not a good reason to use MCMC!
+Another bad use case is the following:
 Because MCMC samples the parameter representatively, it spends most of
@@
   back something far more useful than the output of any optimizer!}
+This is another bad reason!
 

### 2016-01-25 mcmc.tex
name

@@
 
-% HOGG: \Var NEEDS TO BE DEFINED:
 \newcommand{\Var}{\mathrm{Var}}
@@
 \section*{Data analysis recipes:\\ \chaptertitle\footnotemark}
-
+%
 \footnotetext{%
@@
     license\note{\label{note:first}
-        Copyright 2013, 2014, 2015 by the authors. This work is licensed under a
+        Copyright 2013, 2014, 2015, 2016 the authors. This work is licensed under a
         \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
@@
           Dustin Lang (CMU)
-        for valuable advice and comments.}.
+        for valuable advice and comments over the years.}.
 }
@@
 \affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}\\[1ex]
-Dan~Foreman-Mackey\\
+Daniel~Foreman-Mackey\\
 \affil{Department~of~Astronomy, University~of~Washington}

### 2016-01-25 mcmc.tex
remove one HOGG

@@
   Perhaps more importantly, we also have views about why you probably never really
-  \emph{want} to compute it (HOGG CITE).
-  The computation of $Z$ is outside the scope of this \documentname,
+  \emph{want} to compute it.
+  These views relate to the fact that the value of this integral
+  very rarely delivers the answer to a question you care about (see future \documentname s in this series).
+  These philosophical matters and the computational issues are both
+  outside the scope of this \documentname,
   but suffice it to say:
-  If the computation of $Z$ is in your travel plans, you will need to
+  If the computation of $Z$ is in your travel plans, you will need yto
   think about implementing some non-trivial integrating or sampling

### 2016-01-25 mcmc.tex
placeholders for problems
first problem / exercise

@@
 
+\begin{problem}
+Look up (or choose) definitions for the mean, variance, skewness,
+and kurtosis of a distribution.
+Also look up or compute the analytic values of these for a top-hat
+(uniform) distribution.
+Write a computer program that uses some standard package (such as
+\project{numpy}\note{HOGG CITE NUMPY}) to generate $K$ random numbers
+$x$ from a uniform distribution in the interval $0<x<1$.
+Now use those $K$ numbers to compute a the sampling estimate of the
+mean, variance, skewness, and kurtosis (four estimates)
+Make four plot of these four estimates as a function of $\ln K$, for
+$K=2^n$ for $n=8$ up to $n=20$ (that is, $K=256$, $K=512$, and so on
+up to $K=1048576$).
+Over-plot the analytic answers.
+What can you conclude?
+\end{problem}
+
 \section{Metropolis--Hastings MCMC}
@@
     HOGG Refer again to the DAR on fitting a line for an example project.}
+
+\begin{problem}
+HOGG
+\end{problem}
 
@@
 
+\begin{problem}
+HOGG
+\end{problem}
+
 \section{Tuning}\label{sec:tuning}
@@
 Between them, they ought to span the literature.
+
+\begin{problem}
+HOGG
+\end{problem}
 
@@
   bottle of wine or two while the code runs.}
+
+\begin{problem}
+HOGG
+\end{problem}
 
@@
 
+\begin{problem}
+HOGG
+\end{problem}
+
 \section{Gibbs, ensemble, Hamiltonian, and so on}\label{sec:methods}
@@
 
+\begin{problem}
+HOGG
+\end{problem}
+
 \section{Likelihoods and priors}
@@
 
+\begin{problem}
+HOGG
+\end{problem}
+
 \section{Troubleshooting and advice}\label{sec:troubleshooting}

### 2016-01-25 mcmc.tex
problem tweaks

@@
 and kurtosis of a distribution.
-Also look up or compute the analytic values of these for a top-hat
-(uniform) distribution.
+Also look up or compute the analytic values of these four statistics
+for a top-hat (uniform) distribution.
 Write a computer program that uses some standard package (such as
@@
 mean, variance, skewness, and kurtosis (four estimates)
-Make four plot of these four estimates as a function of $\ln K$, for
-$K=2^n$ for $n=8$ up to $n=20$ (that is, $K=256$, $K=512$, and so on
-up to $K=1048576$).
+Make four plot of these four estimates as a function of $1/K$ or
+perhaps $\log_2 K$, for $K=4^n$ for $n=1$ up to $n=10$ (that is,
+$K=4$, $K=16$, and so on up to $K=1048576$).
 Over-plot the analytic answers.

### 2016-01-26 mcmc.tex
problems, reference, notes to self

@@
 % -----
-% - [delete this]
 % - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``CITE'' and ``...''
@@
 % - mention JAGS and STAN in methods section
-% - follow up the idea that burn-in doesn't exist at all!
+% - follow up the idea that burn-in doesn't exist at all!  It does for emcee.
 % - when we say ``below'' or ``above'' also give a Section Number.
@@
 % -----------
-% - These documents are ``Chapters'', no?
+% - These documents are ``Chapters'', no?  Use \documentname.
 % - [LaTeX] newline after every full stop for proper git diffing
@@
     license\note{\label{note:first}
-        Copyright 2013, 2014, 2015, 2016 the authors. This work is licensed under a
+        Copyright 2013, 2014, 2015, 2016 the authors (OMG this took us a long time).
+        This work is licensed under a
         \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
@@
           Jonathan Goodman (NYU),
-          Fengji Hou (NYU), and
-          Dustin Lang (CMU)
+          Fengji Hou (NYU),
+          Dustin Lang (CMU), and
+          Iain Murray (Edinburgh)
         for valuable advice and comments over the years.}.
@@
   where there are always parameters with very different levels of importance to the scientific conclusions.
-This point will return again below, either explicitly or implicitly.
-[HOGG: does it?]
+This point will return again below in \sectionname~\ref{sec:results}.
 
@@
 
+\begin{problem}\label{prob:firstMH}
+In your scientific programming language of choice, write a very simple
+M-H MCMC sampler.
+Sample in a single parameter $x$ and give the sampler as its density
+function $p(x)$ a Gaussian density with mean 2 and variance 2.
+Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
+pdf with mean zero and variance 1.
+Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
+Plot the results as a histogram, with the true density over-plotted sensibly.
+\end{problem}
+
 \begin{problem}
-HOGG
+Re-do \problemname~\ref{prob:firstMH} but now with an input density
+that is uniform on $3<x<7$ and zero everywhere else.
+What change did you have to make to the initialization, and why?
+\end{problem}
+
+\begin{problem}
+Re-do \problemname~\ref{prob:firstMH} but with different values for
+the variance of the proposal distribution $q(x'\given x)$.
+What happens when you go to very extreme values (like for instance
+$0.00001$ or $1000.0$)?
+\end{problem}
+
+\begin{problem}
+Why, in \problemname~\ref{prob:firstMH}, did we give the proposal
+distribution $q(x'\given x)$ zero mean?
+What would be bad if we hadn't done that?
+Try setting the mean of $q(x'\given x)$ to something non-zero and see
+what happens.
 \end{problem}
@@
 
-\section{Results, error bars, and figures}
+\section{Results, error bars, and figures}\label{sec:results}
 
@@
   (also \arxiv{1202.3665}).
+\bibitem[Hogg \etal(2010)]{fittingaline}
+  Hogg,~D.~W., Bovy,~J., \& Lang,~D., 2010,
+  Data analysis recipes: Fitting a model to data,
+  \arxiv{1008.4686v1}.
 \bibitem[Hou \etal(2014)]{fengji}

### 2016-01-27 mcmc.tex
problemtweaks
two problems

@@
 % -----
+% - Kill all issues in this list, below:
 % - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``CITE'' and ``...''
@@
 % - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
+% - mention ESJD
 % - mention Gelman--Rubin statistic
@@
 %   - Gelman
+%   - Goodman
 %   - Lang
@@
   violate detailed balance in this way and has a compensated
-  acceptance ratio.  Read the paper [HOGG CITE] for more details.}
+  acceptance fraction.  Read the paper [HOGG CITE] for more details.}
 If the detailed balance condition frightens you (as it frightens us),
@@
 What happens when you go to very extreme values (like for instance
-$0.00001$ or $1000.0$)?
+$10^{-8}$ or $10^6$)?
 \end{problem}
@@
 \begin{problem}
-HOGG
+Re-do \problemname~\ref{prob:firstMH} but now look at convergence:
+Plot the $x$ chain as a function of timestep.  Also split the chain
+into four contiguous segments (the first, second, third, and fourth
+quarters of the chain).  In each of these four, compute the empirical
+mean and empirical variance of $x$.  What do you conclude about
+convergence from these heuristics?
+\end{problem}
+
+\begin{problem}
+Write a piece of code that computes the empirical autocorrelation
+function and the empirical autocorrelation time from a chain.  Run
+this on the chain you obtained from \problemname~\ref{prob:firstMH}.
+Plot the autocorrelation function you get and over-plot the
+autocorrelation time as an indicator on the plot.  What do you
+conclude?
 \end{problem}

### 2016-01-27 mcmc.tex
cite numpy

@@
 Write a computer program that uses some standard package (such as
-\project{numpy}\note{HOGG CITE NUMPY}) to generate $K$ random numbers
+\project{numpy}\note{There isn't a full citation for this package but there is \cite{numpy}.})
+to generate $K$ random numbers
 $x$ from a uniform distribution in the interval $0<x<1$.
@@
   \arxiv{1401.6128}.
+\bibitem[van~der~Walt \etal(2011)]{numpy}
+  van~der~Walt,~S., Colbert,~S.~C., and Varoquaux,~G., 2011,
+  The NumPy Array: A structure for efficient numerical computation,
+  \textit{Computing in Sci.\ \& Eng.}\ \textbf{13}, 22--30
+  \doi{10.1109/MCSE.2011.37}.
 \end{thebibliography}

### 2016-01-28 mcmc.tex
problems

@@
 
+\begin{problem}\label{prob:twod}
+Re-do \problemname~\ref{prob:firstMH} but now with an input density
+that is a function of two variables $(x, y)$.
+For the function use a covariant Gaussian density with variance tensor HOGG,
+and then a top-hat function that is uniform on HOGG and zero elsewhere.
+For the proposal distribution $q(x'\given x)$ a two-dimensional Gaussian
+density with zero mean and variance tensor set to the two-dimensional
+identity matrix.
+Plot the two one-dimensional histograms and also a two-dimensional scatter
+plot for each sampling.
+\end{problem}
+
 \begin{problem}
@@
 what happens.
+Bonus points: Modify the acceptance--rejection criterion to deal with
+the messed-up $q(x'\given x)$ and show that everything works once
+again.
 \end{problem}
@@
 \begin{problem}
-HOGG
+Look at the source code of HOGG: REFERENCE EMCEE, and see if you can
+re-code or extract the autocorrelation-time estimator called HOGG.
+This code should take a MCMC chain of parameter vectors as input and
+return a vector of autocorrelation times as output.
+\end{problem}
+
+\begin{problem}
+Run the MCMC sampling of \probemname~\ref{prob:twod} to...HOGG
 \end{problem}

### 2016-01-28 mcmc.tex
typo
problems

@@
 
+\begin{problem}\label{prob:tuning}
+Run the MCMC sampling of \problemname~\ref{prob:twod} with the
+covariant Gaussian density.
+Give the proposal density $q(x'\given x)$ a diagonal variance
+tensor that is $Q$ times the two-dimensional identity matrix.
+Assess the acceptance fraction as a function of $Q$.
+Find (very roughly) the value of $Q$ that gives an acceptance
+fraction of about 0.25.
+\end{problem}
+
 \begin{problem}
-Run the MCMC sampling of \probemname~\ref{prob:twod} to...HOGG
+Re-do \problemname~\ref{prob:tuning} but instead of trying to reach a
+certain acceptance fraction, try to minimize the autocorrelation time.
+What do you get as the best value of $Q$?
 \end{problem}

### 2016-01-29 mcmc.tex
M-H name

@@
 The simplest algorithm for MCMC is the Metropolis--Hastings algorithm (M--H MCMC).\note{%
-  HOGG put here some short discussion of the name of the algorithm and the original reference.}
+  There are many claims that this algorithm is incorrectly named,
+  with claims that it should be credited to Enrico Fermi or Stan Ulam.
+  We don't have any opinions of this matter, but encourage the reader to follow this up.
+  The original paper is \citet{metropolis}, and there are a few sketchy historical notes
+  in \citet{geyer}.}
 It is so simple that we recommend that any reader of this document
@@
 certain acceptance fraction, try to minimize the autocorrelation time.
-What do you get as the best value of $Q$?
+(This, by the way, is the Right Thing To Do, but often expensive.)
+What do you get as the best value of $Q$ in this case?
 \end{problem}
@@
 \begin{problem}
+
 HOGG
@@
   (also \arxiv{1202.3665}).
+\bibitem[Geyer(2011)]{geyer}
+  Geyer,~C.~J., 2011,
+  Introduction to Markov Chain Monte Carlo,
+  in \citet{mcmchandbook}.
 \bibitem[Hogg \etal(2010)]{fittingaline}
@@
   \arxiv{1401.6128}.
+\bibitem[Metropolis \etal(1953)]{metropolis}
+  Metropolis,~N., Rosenbluth,~A.~W., Rosenbluth,~M.~N., Teller,~A.~H., \& Teller,~E., 1953,
+  Equation of state calculations by fast computing machines,
+  \textit{J.\ Chem.\ Phys.}, \textbf{21} 1087--1092.
 \bibitem[van~der~Walt \etal(2011)]{numpy}
-  van~der~Walt,~S., Colbert,~S.~C., and Varoquaux,~G., 2011,
+  van~der~Walt,~S., Colbert,~S.~C., \& Varoquaux,~G., 2011,
   The NumPy Array: A structure for efficient numerical computation,

### 2016-01-29 mcmc.tex
grants

@@
           Iain Murray (Edinburgh)
-        for valuable advice and comments over the years.}.
+        for valuable advice and comments over the years.
+        HOGG: GRANTS AND GRANT NUMBERS}.
 }

### 2016-01-31 mcmc.tex
grantNumbers
problems

@@
         for valuable advice and comments over the years.
-        HOGG: GRANTS AND GRANT NUMBERS}.
+        This project benefitted from various research grants, especially
+          NSF grants AST-0908357, IIS-1124794, and AST-1517237,
+          NASA grant NNX12AI50G,
+          and the The Moore-Sloan Data Science Environment at NYU.}
 }
@@
   in the context of some simple problem.\note{%
-    HOGG put reference to fitting a line here and point inside to the simple problem suggestion.}
+    HOGG put reference to fitting a line here and point inside to the simple problem suggestion.
+    HOGG connect to the actual \problemname s in this document!!}
 
@@
 
+\begin{problem}\label{prob:initialization}
+Re-do \problemname~\ref{prob:firstMH} but with different starting
+positions.
+What happens as you make the starting position extremely far from the
+origin?
+Did you write the algorithm in terms of $\ln p(x)$ or $p(x)$?
+This makes a difference for extreme initializations!
+What is the scaling:  As you move the initialization further away,
+how much longer does it take for the sampler to reach reasonable
+density?
+\end{problem}
+
 \begin{problem}
-
-HOGG
+Check the scaling you found in \problemname~\ref{prob:initialization}
+with a higher-dimensional Gaussian (try, for example, a 10-d Gaussian).
+The same or worse?
+\end{problem}
+
+\begin{problem}
+Import (or write) an optimizer, and upgrade the code you wrote for
+\problemname~\ref{prob:initialization} to begin by optimizing $\ln
+p(x)$ and only then start the MCMC sampler from that optimum.
+Use a sensible optimizer.
+Compare the scaling you found in
+\problemname~\ref{prob:initialization} to the same scaling for the
+optimizer.
+To make this test fair, don't use the awesome math you know about
+Gaussians to help you here; pretend that $p(x)$ is an unknown function
+with unknown derivatives.
 \end{problem}

### 2016-01-31 mcmc.tex
typo
problems

@@
 fraction of about 0.25.
+Don't try to optimize precisely; just evaluate the acceptance fraction
+on a logarithmic grid of $Q$ with values of $Q$ separated by factors
+of 2.
 \end{problem}
@@
 What do you get as the best value of $Q$ in this case?
+Again, just evaluate on a coarse logarithmic grid.
+\end{problem}
+
+\begin{problem}
+In \problemname~\ref{prob:tuning} you varied only the parameter $Q$, but
+really there are three free parameters (two variances and a covariance).
+If the problem was $D$-dimensional, how many tuning parameters would there
+be, in principle?
 \end{problem}
@@
 \emph{publish}, we are not permitted (not now, at least) to publish a
-\emph{probability distribution over pubications}; we have to publish
+\emph{probability distribution over publications}; we have to publish
 \emph{one single, deterministic text}; we have to make a decision about what

### 2016-02-01 mcmc.tex
problems

@@
 
-\begin{problem}\label{prob:firstMH}
+\begin{problem}\label{prob:MH}
 In your scientific programming language of choice, write a very simple
@@
 \begin{problem}
-Re-do \problemname~\ref{prob:firstMH} but now with an input density
+Re-do \problemname~\ref{prob:MH} but now with an input density
 that is uniform on $3<x<7$ and zero everywhere else.
@@
 \begin{problem}\label{prob:twod}
-Re-do \problemname~\ref{prob:firstMH} but now with an input density
+Re-do \problemname~\ref{prob:MH} but now with an input density
 that is a function of two variables $(x, y)$.
@@
 \begin{problem}
-Re-do \problemname~\ref{prob:firstMH} but with different values for
+Re-do \problemname~\ref{prob:MH} but with different values for
 the variance of the proposal distribution $q(x'\given x)$.
@@
 \begin{problem}
-Why, in \problemname~\ref{prob:firstMH}, did we give the proposal
+Why, in \problemname~\ref{prob:MH}, did we give the proposal
 distribution $q(x'\given x)$ zero mean?
@@
 \begin{problem}
-Re-do \problemname~\ref{prob:firstMH} but now look at convergence:
+Re-do \problemname~\ref{prob:MH} but now look at convergence:
 Plot the $x$ chain as a function of timestep.  Also split the chain
@@
 function and the empirical autocorrelation time from a chain.  Run
-this on the chain you obtained from \problemname~\ref{prob:firstMH}.
+this on the chain you obtained from \problemname~\ref{prob:MH}.
 Plot the autocorrelation function you get and over-plot the
@@
 \begin{problem}\label{prob:initialization}
-Re-do \problemname~\ref{prob:firstMH} but with different starting
+Re-do \problemname~\ref{prob:MH} but with different starting
 positions.
@@
 DFM: more here!  For example, does this work better in large numbers
-of dimensions?  If so, we should say so.  And references, please.
+of dimensions?  If so, we should say so.  And cite \citep{hamiltonian}.
 
@@
 \begin{problem}
-HOGG
+Run your M-H MCMC sampler from \problemname~\ref{prob:MH}, but now
+with a density function that is precisely unity \emph{everywhere}
+(that is, at any input value of $x$ it returns unity).  That is, an
+improper function.  Run it for longer and longer and plot the chain
+value $x$ as a function of timestep.  What happens?
+\end{problem}
+
+\begin{problem}
+Modify the sampler you wrote in \problemname~\ref{prob:MH} to take
+steps not in $x$ but in $\ln x$.  That is, replace the Gaussian
+proposal distribution $q(x'\given x)$ with a Gaussian distribution in
+$\ln x$ $q(\ln x'\given \ln x)$.  What about your answers change?  How
+wrong is your answer?  Bonus points: Compute the analytic function
+that you have sampled from (that is, not $p(x)$) and over-plot it on
+your histogram.
 \end{problem}
@@
   Brooks,~S., Gelman~A., Jones~G., \& Meng,~X.-L., eds.\ 2011,
-  \textit{Handbook of Markov Chain Monte Carlo},
+  \href{http://www.mcmchandbook.net/HandbookTableofContents.html}{\textit{Handbook of Markov Chain Monte Carlo}},
   Chapman and Hall~/~CRC Press
@@
   Geyer,~C.~J., 2011,
-  Introduction to Markov Chain Monte Carlo,
+  \href{http://www.mcmchandbook.net/HandbookChapter1.pdf}{Introduction to Markov Chain Monte Carlo},
   in \citet{mcmchandbook}.
@@
   \textit{J.\ Chem.\ Phys.}, \textbf{21} 1087--1092.
+\bibitem[Neal \etal(2011)]{hamiltonian}
+  Neal,~R.~M., 2011,
+  \href{http://www.mcmchandbook.net/HandbookChapter5.pdf}{MCMC Using Hamiltonian Dynamics},
+  in \citet{mcmchandbook}.
 \bibitem[van~der~Walt \etal(2011)]{numpy}

### 2016-02-02 mcmc.tex
another problem, another citation

@@
   in the context of some simple problem.\note{%
-    HOGG put reference to fitting a line here and point inside to the simple problem suggestion.
-    HOGG connect to the actual \problemname s in this document!!}
+    We request this in a trivial case below in
+    \problemname~\ref{prob:MH} and the following problems in this
+    \sectionname.
+    We make the same request in a less trivial context in our
+    model-fitting \documentname\ (\citealt{fittingaline},
+    \problemname~6 and 7), where we ask the Reader to implement MCMC for
+    a useful mixture model.}
 
@@
 \begin{problem}
-HOGG
+Look up the Rosenbrock density used as the demonstration case for the
+\project{emcee} sampler (see \citealt{emcee}).
+Tune up a M--H MCMC sampler to sample this density (this isn't trivial),
+or just install and use \project{emcee} to sample it.
+Now plot the $x$ and $y$ histograms of a 10,000-point sampling of this
+distribution (you might have to sample more than 10,000 and thin the
+chain), and also plot the two-dimensional scatter plot of $x,y$
+samples.
+Overplot on all three plots an indicator of the means and medians of
+the samples along the $x$ and $y$ directions.
+Overplot on all three plots the (above) recommended quantiles of the
+samples.
+Comment on the results.
 \end{problem}

### 2016-02-02 mcmc.tex
abstract tweaks

@@
 have transformed the sciences.
-Here we give a brief overview of basic MCMC methods and then turn to
-practical advice for their use in real inference problems.
+Here we give a brief overview of the most basic MCMC method and then turn to
+practical advice for the use of MCMC in real inference problems.
 We give advice on method choice, tuning for performance,
-methods for initialization, tests of convergence, and use of the chain
-output to produce parameter estimates with associated uncertainties.
+methods for initialization, tests of convergence, troubleshooting, and use of the chain
+output to produce or report parameter estimates with associated uncertainties.
 \end{abstract}

### 2016-02-04 mcmc.tex
phil

@@
           Fengji Hou (NYU),
-          Dustin Lang (CMU), and
+          Dustin Lang (CMU),
+          Phil Marshall (KIPAC), and
           Iain Murray (Edinburgh)

### 2016-02-04 mcmc.tex
autodiff

@@
   violate detailed balance in this way and has a compensated
-  acceptance fraction.  Read the paper [HOGG CITE] for more details.}
+  acceptance fraction.  Read the paper \cite{emcee} for more details.}
 If the detailed balance condition frightens you (as it frightens us),
@@
 auto-differentiation systems that deliver automatically code that
-produces the gradient of any function you can write.\note{HOGG cite
-  autodiff systems}
+produces the gradient of any function you can write.\note{%
+  The idea is that for (almost) any function you can write,
+  a robot can write the function that is the derivative of that
+  function, using simple differentiation rules andthe magic of the
+  chain rule. Differentiation is a typographic operation on computer
+  functions!  The idea is \emph{not} to take numerical (small
+  difference) derivatives of the code; there is no way that
+  finite-difference differentiation could help your sampling in the
+  long run.}
 There is a class of MCMC methods that make use of gradient information
@@
 trained professional.
-
-\begin{problem}
-HOGG
-\end{problem}
 

### 2016-02-04 mcmc.tex
problems

@@
 
+\begin{problem}\label{prob:rosenbrock}
+Look up the Rosenbrock density used as the demonstration case for the
+\project{emcee} sampler (see \citealt{emcee}).
+Tune the Gaussian proposal distribution in your M--H MCMC sampler to sample this density efficiently.
+What autocorrelation time do you get?
+Compare to what \project{emcee} gets.
+\end{problem}
+
 \section{Initialization and burn-in}
@@
 \begin{problem}
-Look up the Rosenbrock density used as the demonstration case for the
-\project{emcee} sampler (see \citealt{emcee}).
-Tune up a M--H MCMC sampler to sample this density (this isn't trivial),
-or just install and use \project{emcee} to sample it.
+Execute \problemname~\ref{prob:rosenbrock}, or, if you are lazy,
+or just install and use \project{emcee} to do the hard work.
 Now plot the $x$ and $y$ histograms of a 10,000-point sampling of this

### 2016-02-04 mcmc.tex
typos
citations

@@
   (though not all\note{We are promoters of an ensemble method that has a
-  nearly-fixed proposal distribution and requires no tuning (CITE EMCEE).})
+  nearly-fixed proposal distribution and requires no tuning \citep{emcee}.
+  There are many problems for which this is \emph{not} the best choice
+  of sampler, but since it obviates tuning, it can be a good first choice.})
   make use of something like a ``proposal distribution''
@@
 If you are accepting almost none, your step sizes are too large.
-The Goldilocks value is around a quarter,
-  with an argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}
+The Goldilocks value is between a half and about a quarter,
+  with an argument floating around that it should be 0.234 for best
+  performance in high dimensional problems\note{%
+    The argument and assumptions underlying the 0.234 fraction
+    (and higher acceptance fractions at lower numbers of dimensions)
+    are laid out in \citet{gelman234}.}
   (though you could never tune it precisely enough to warrant that third digit of accuracy).
@@
 \begin{problem}
-Execute \problemname~\ref{prob:rosenbrock}, or, if you are lazy,
-or just install and use \project{emcee} to do the hard work.
+Execute \problemname~\ref{prob:rosenbrock}, or---if you are lazy---%
+just install and use \project{emcee} to do the hard work.
 Now plot the $x$ and $y$ histograms of a 10,000-point sampling of this
@@
 Not a few times; many times.
-If you are using an ensemble sampler\note{CITE EMCEE},
+If you are using an ensemble sampler\note{Such as \project{emcee} \citep{emcee};
+  one nice thing about the ensemble sampler is that it gives you great diagnostic information automatically.}
   make plots showing all $M$ walkers, each a different color.
@@
   (also \arxiv{1202.3665}).
+\bibitem[Gelman \etal(1996)]{gelman234}
+  Gelman,~A., Roberts,~G.~O., \& Gilks,~W.~R., 1996,
+  Efficient Metropolis jumping rules,
+  \textit{Bayesian Stat.}\ \textbf{5} 599--607.
 \bibitem[Geyer(2011)]{geyer}

### 2016-02-04 mcmc.tex
ltfdfcf

@@
 discuss the economic explanation of how we can make hard decisions in
-the context of probabilistic reasoning\note{CITE DECISION DOCUMENT}.
-But without going into that ``economic'' model, it is fair to say that a
+the context of probabilistic reasoning\note{We intend to write a subsequent
+  \documentname\ in this series about these issues.  The idea is that you
+  can make principled decisions by optimizing the expected utility under
+  the posterior, given the data.  This is a great idea!  Of course we also
+  have very deep, fundamental reasons that you \emph{can't know your utility
+    precisely}.  The big issue is that your only sensible utility involves
+  an integral out to the ``long term'', and the long term (by definition)
+  includes outcomes that are outside your present-day quantitative model.}
+But without going into that economic model, it is fair to say that a
 substantial problem with being a committed probabilist is that when we

### 2016-02-04 mcmc.tex
refs

@@
   Gelman,~A., Roberts,~G.~O., \& Gilks,~W.~R., 1996,
-  Efficient Metropolis jumping rules,
-  \textit{Bayesian Stat.}\ \textbf{5} 599--607.
+  \href{http://www.stat.columbia.edu/~gelman/research/published/baystat5.pdf}{Efficient Metropolis jumping rules},
+  in Bernardo,~J.~M, Berger,~J.~O., Dawid,~A.~P., \& Smith,~A.~F.~M, eds.,
+  \textit{Bayesian Statistics} \textbf{5}, Oxford University Press
+  599--607.
 \bibitem[Geyer(2011)]{geyer}
@@
   \href{http://www.mcmchandbook.net/HandbookChapter1.pdf}{Introduction to Markov Chain Monte Carlo},
-  in \citet{mcmchandbook}.
+  in \citet{mcmchandbook}, 3--48.
 \bibitem[Hogg \etal(2010)]{fittingaline}
@@
   \href{http://www.mcmchandbook.net/HandbookChapter5.pdf}{MCMC Using Hamiltonian Dynamics},
-  in \citet{mcmchandbook}.
+  in \citet{mcmchandbook}, 113--162.
 \bibitem[van~der~Walt \etal(2011)]{numpy}

### 2016-02-05 mcmc.tex
citations

@@
 Another proxy for autocorrelation time useful for tuning is the
-  Expected Squared Jump Distance.\note{(CITE GELMAN)}
+  Expected Squared Jump Distance.\note{See \citet{esjd}.}
 This is the mean squared distance the walker moves, per step.
@@
 Again, such transformations require also prior transformations, which must be made with care.
-There are some affine invariant samplers\note{CITE EMCEE AGAIN HERE} that are invariant to such transformations,
+There are some affine invariant samplers\note{Note how much we like to cite \citet{emcee}!} that are invariant to such transformations,
   but if you aren't using such a sampler,
@@
   effects and noises for which there is (and can be, for deep reasons,
-  discussed elsewhere by us; CITE) no extremely precise model.  So in
+  to be discussed in other \documentname s in this series) no extremely precise model.  So in
   detail, any model of any specific observation must be wrong, even in
@@
   in \citet{mcmchandbook}, 113--162.
+\bibitem[Pasarica \& Gelman(2010)]{esjd}
+  Pasarica,~C. \& Gelman,~A., 2010
+  \href{http://www.jstor.org/stable/24308995}{Adaptively scaling the Metropolis algorithm using expected squared jumped distance},
+  \textit{Statistica Sinica}, \textbf{20}(1) 343--364.
 \bibitem[van~der~Walt \etal(2011)]{numpy}

### 2016-02-05 mcmc.tex
all CITE requests done

@@
 % - Kill all issues in this list, below:
-% - Deal with all mentions of ``DFM'' or ``HOGG'' and all ``CITE'' and ``...''
+% - Deal with all mentions of ``DFM'' or ``HOGG'' and ``...''
 % - Should we cite the Brooks book or just chapters within it?  I think the latter!
@@
 a rich way, or confirm that certain assumptions are bad.
-Cross-validation\note{CITE SOMETHING} is a framework for constructing
-valuable tests of this form, as is posterior predictive checks\note{CITE SOMETHING}.
+Cross-validation\note{Cross-validation is an incredibly valuable set
+  of techniques that all astronomers should know.
+  it is not clear what to cite here, but one possibility is \citet{crossval}.}
+is a framework for constructing
+valuable tests of this form, as is posterior predictive checks\note{%
+  The posterior predictive check is one of the Bayesian equivalents of
+  a goodness-of-fit test \citep{ppcheck}.}
 Importantly, you want to find ways to compare your model to your data
@@
   (also \arxiv{1202.3665}).
-\bibitem[Gelman \etal(1996)]{gelman234}
-  Gelman,~A., Roberts,~G.~O., \& Gilks,~W.~R., 1996,
+\bibitem[Geisser(1993)]{crossval}
+  Geisser,~S., 1993,
+  \textit{Predictive Inference: An Introduction},
+  Chapman \& Hall
+  \isbn{978-0412034718}.
+\bibitem[Gelman \etal(1996a)]{ppcheck}
+  Gelman,~A., Meng,~X.-L., \& Stern,~H., 1996a,
+  \href{http://www.jstor.org/stable/24306036}{Posterior predictive assessment of model fitness via realized discrepancies},
+  \textit{Statistica Sinica}, \textbf{6}(4) 733--760.
+\bibitem[Gelman \etal(1996b)]{gelman234}
+  Gelman,~A., Roberts,~G.~O., \& Gilks,~W.~R., 1996b,
   \href{http://www.stat.columbia.edu/~gelman/research/published/baystat5.pdf}{Efficient Metropolis jumping rules},

### 2016-02-07 mcmc.tex
eradicated four HOGGs

@@
 % - Kill all issues in this list, below:
-% - Deal with all mentions of ``DFM'' or ``HOGG'' and ``...''
-% - Should we cite the Brooks book or just chapters within it?  I think the latter!
+% - Deal with all mentions of ``DFM'' or ``HOGG''
+% - Go through github issues, if any.
 % - Should we make an \acronym{} environment and use it religiously?  I think yes.
@@
 % -----------
+% - \MCMC not MCMC
 % - These documents are ``Chapters'', no?  Use \documentname.
@@
 
+\newcommand{\MCMC}{\acronym{MCMC}}
+\newcommand{\MH}{\acronym{M--H}}
 \newcommand{\Var}{\mathrm{Var}}
@@
 \begin{abstract}
-Markov Chain Monte Carlo (MCMC) methods for sampling probability
+Markov Chain Monte Carlo (\MCMC) methods for sampling probability
 distribution functions (combined with abundant computational resources)
 have transformed the sciences.
-Here we give a brief overview of the most basic MCMC method and then turn to
-practical advice for the use of MCMC in real inference problems.
+Here we give a brief overview of the most basic \MCMC\ method and then turn to
+practical advice for the use of \MCMC\ in real inference problems.
 We give advice on method choice, tuning for performance,
@@
 
-Markov Chain Monte Carlo (MCMC) methods are methods for sampling
+Markov Chain Monte Carlo (\MCMC) methods are methods for sampling
 probability distribution functions or probability density functions (pdfs).
@@
 concentrate on the latter in this \documentname.
-MCMC methods don't require that you have a full analytic description of the
+\MCMC\ methods don't require that you have a full analytic description of the
 properly normalized pdf for sampling to proceed; they only require
 that you be able to compute ratios of the pdf at pairs of locations.
-This makes MCMC methods ideal for sampling \emph{posterior
+This makes \MCMC\ methods ideal for sampling \emph{posterior
   pdfs} in probabilistic inferences:
@@
 
-In addition to this normalization-insensitive property of MCMC, in its
+In addition to this normalization-insensitive property of \MCMC, in its
 simplest forms it can be run without computing any derivatives or
@@
 forms it is \emph{extremely easy to implement}.
-For all these reasons, MCMC is ideal for sampling posterior pdfs in
+For all these reasons, \MCMC\ is ideal for sampling posterior pdfs in
 the real situations in which scientists find themselves.
@@
 If all these things are true, then---even if you can't compute
-anything else---in principle a trivial-to-implement MCMC can give you
+anything else---in principle a trivial-to-implement \MCMC\ can give you
 a fair sampling of the posterior pdf.
-That is, you can run MCMC
+That is, you can run \MCMC\ 
 (for a very long time---see \sectionname~\ref{sec:convergence} for how long)
@@
   most uses of the methods in question are not appropriate!}, we are
-compelled to note at the outset that MCMC is in fact \emph{over-used}.
-Because MCMC provably (under assumptions\note{Hint at assumptions
+compelled to note at the outset that \MCMC\ is in fact \emph{over-used}.
+Because \MCMC\ provably (under assumptions\note{Hint at assumptions
   here.}, some of which will be discussed below) samples the full
-posterior pdf in all of parameter space, many investigators use MCMC
+posterior pdf in all of parameter space, many investigators use \MCMC\ 
 because it will sample \emph{all} of the parameter space
 ($\pars$-space).
-That is, they are using MCMC because they want to
+That is, they are using \MCMC\ because they want to
 \emph{search the parameter space for good models}.
-This is not a good reason to use MCMC!
+This is not a good reason to use \MCMC!
 Another bad use case is the following:
-Because MCMC samples the parameter representatively, it spends most of
+Because \MCMC\ samples the parameter representatively, it spends most of
 its time near very good models; models that (within the confines of
 the prior pdf) do a good job of explaining the data.
-For this reason, many investigators are using MCMC because it
+For this reason, many investigators are using \MCMC\ because it
 effectively \emph{optimizes the posterior pdf}, or, for certain
@@
 
-Both of these reasons for using MCMC---that it is a parameter-space
+Both of these reasons for using \MCMC---that it is a parameter-space
 search algorithm, and that it is a simple-to-code effective
-optimizer---are \emph{not good reasons}.  MCMC is first and foremost a
+optimizer---are \emph{not good reasons}.  \MCMC\ is first and foremost a
 \emph{sampler}.  If you are trying to find the optimum of the
@@
 not a sampler.  If you want to make sure you search all of parameter
-space, you should use a \emph{search algorithm}, not a sampler.  MCMC
+space, you should use a \emph{search algorithm}, not a sampler.  \MCMC\ 
 is good at one thing, and one thing only: Sampling ill-normalized (or
@@
 In what follows, we are going to provide a kind of ``user manual'' or
-advice document or folklore capture regarding the use of MCMC for
+advice document or folklore capture regarding the use of \MCMC\ for
 data analysis.
-This will not be a detailed description of multiple MCMC methods
+This will not be a detailed description of multiple \MCMC\ methods
 (indeed, we will only explain one method in detail), and it will not
@@
 methods.
-It will be about how to use MCMC, including diagnosis and
+It will be about how to use \MCMC, including diagnosis and
 trouble-shooting.
@@
 The first couple of \sectionname s will describe what a sampling is, and how
-the simplest MCMC method, the Metropolis-Hastings algorithm, can
+the simplest \MCMC\ method, the Metropolis-Hastings algorithm, can
 provide one.
 The next few \sectionname s will provide ideas about how to
-initialize, tune and operate MCMC methods for good performance.
+initialize, tune and operate \MCMC\ methods for good performance.
 The last few \sectionname s will provide advice for making decisions
-among the myriad MCMC methods implementations, how to implement a good
+among the myriad \MCMC\ methods implementations, how to implement a good
 likelihood function and prior pdf function for inference, and how to
-trouble-shoot standard kinds of problems that arise in operating MCMC
+trouble-shoot standard kinds of problems that arise in operating \MCMC\ 
 methods on real problems.
@@
 
-As we noted above (\sectionname~\ref{sec:when}), in the context of MCMC, we are often sampling or
+As we noted above (\sectionname~\ref{sec:when}), in the context of \MCMC, we are often sampling or
 using some \emph{badly normalized} function $f(\pars)$.
@@
 \emph{never learn that normalization}.
-When we run MCMC sampling on $f(\pars)$, a function that differs from a
+When we run \MCMC\ sampling on $f(\pars)$, a function that differs from a
 pdf $p(\pars)$ by some unknown normalization constant, then the sampling
@@
   computing integrals.
-There are many uses of the MCMC sampling, some of them good and some of them bad.
+There are many uses of the \MCMC\ sampling, some of them good and some of them bad.
 Most of the good or sensible uses will somehow involve integration.
@@
 For this reason (and others), in what follows, when we talk about what to report
-  about the outcome of your MCMC sampling,
+  about the outcome of your \MCMC\ sampling,
   we will advise in favor of mean, median, and quantiles,
@@
 Although the discussion in this \documentname\ is general, the most
-common use of MCMC sampling (for us, anyway) is in probabilistic
+common use of \MCMC\ sampling (for us, anyway) is in probabilistic
 inference.
@@
 
-The simplest algorithm for MCMC is the Metropolis--Hastings algorithm (M--H MCMC).\note{%
+The simplest algorithm for \MCMC\ is the Metropolis--Hastings algorithm (\MH\ \MCMC).\note{%
   There are many claims that this algorithm is incorrectly named,
@@
     model-fitting \documentname\ (\citealt{fittingaline},
-    \problemname~6 and 7), where we ask the Reader to implement MCMC for
+    \problemname~6 and 7), where we ask the Reader to implement \MCMC\ for
     a useful mixture model.}
 
-The M--H MCMC algorithm requires two inputs.
+The \MH\ \MCMC\ algorithm requires two inputs.
 The first is a handle to the function $f(\pars)$ that is the function to be sampled,
@@
 
-HOGG: ...Why does M--H MCMC work (briefly)?  Here or later?
-
-This algorithm---and indeed any MCMC algorithm---produces a
+HOGG: ...Why does \MH\ \MCMC\ work (briefly)?  Here or later?
+
+This algorithm---and indeed any \MCMC\ algorithm---produces a
   \emph{biased random walk} through parameter space.
@@
 
-The principal user-settable knob in M--H MCMC is the proposal
+The principal user-settable knob in \MH\ \MCMC\ is the proposal
   pdf $q(\pars'\given\pars)$.
@@
 As we discuss briefly below in \sectionname~\ref{sec:methods},
-  there are many MCMC methods more advanced than M--H.
+  there are many \MCMC\ methods more advanced than M--H.
 However, they all share characteristics with M--H
   (initialization, tuning, judging convergence),
-  such that it is very valuable to understand M--H well before using anything more advanced.
-Furthermore, a scientist new to MCMC benefits enormously from
-  building, tuning, and using her or his own MCMC software.
+  such that it is very valuable to understand \MH\ well before using anything more advanced.
+Furthermore, a scientist new to \MCMC\ benefits enormously from
+  building, tuning, and using her or his own \MCMC\ software.
 One piece of advice we give then,
-  to the new user of MCMC,
-  is to code up, tune, and use a M--H MCMC sampler for a scientific project.
+  to the new user of \MCMC,
+  is to code up, tune, and use a \MH\ \MCMC\ sampler for a scientific project.
 A huge amount is learned in doing this,
-  and it is very often the case that the home-built M--H MCMC does everything needed;
+  and it is very often the case that the home-built \MH\ \MCMC\ does everything needed;
   you often don't need any more advanced tool.
-Only after you have concluded that your home-built M--H MCMC sampler
+Only after you have concluded that your home-built \MH\ \MCMC\ sampler
   is not suited to your project
@@
   and given the intuition you gain by doing it (at least once) yourself.\note{%
-    HOGG Refer again to the DAR on fitting a line for an example project.}
+    We also make this point in a previous \documentname\ (\citealt{fittingaline};
+    the ambitious reader will find in that \documentname\ a solid example project
+    for using \MCMC\ in a real context.}
 
@@
 In your scientific programming language of choice, write a very simple
-M-H MCMC sampler.
+M-H \MCMC\ sampler.
 Sample in a single parameter $x$ and give the sampler as its density
 function $p(x)$ a Gaussian density with mean 2 and variance 2.
+(Note that variance is the \emph{square} of the standard deviation.)
 Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
@@
 that is a function of two variables $(x, y)$.
-For the function use a covariant Gaussian density with variance tensor HOGG,
-and then a top-hat function that is uniform on HOGG and zero elsewhere.
-For the proposal distribution $q(x'\given x)$ a two-dimensional Gaussian
+For the density function use two different functions.
+The first density function is a covariant two-dimensional Gaussian density with variance tensor
+\begin{eqnarray}
+V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
+\quad .
+\end{eqnarray}
+The second density function is a rectangular top-hat function that is uniform on
+the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
+For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
 density with zero mean and variance tensor set to the two-dimensional
@@
 
-A key question for an MCMC operator---\emph{the} key question in some
+A key question for an \MCMC\ operator---\emph{the} key question in some
 sense---is \emph{how long to run} to be sure of having reliable
@@
 In this context,
-  it is \emph{simply not fair} to require an MCMC run to have fully and completely
+  it is \emph{simply not fair} to require an \MCMC\ run to have fully and completely
   sampled the posterior pdf,
@@
   despite many statements in the literature to the contrary.%
-  \note{Give example statements... Mention cases in which you \emph{might} be able to claim this...}
+  \note{Give example statements... HOGG: Mention cases in which you \emph{might} be able to claim this...}
 The upshot of this is that we don't and can't require any kind of absolute convergence;
@@
 In the above equation, the assumption is that the samples $\pars_k$ are
-given in the order that they were generated by the MCMC algorithm.
+given in the order that they were generated by the \MCMC\ algorithm.
 By construction
@@
 The chain is likely to be converged only if the random-walk process
-crossed the domain of $\pars$ fully many times during the MCMC run.
+crossed the domain of $\pars$ fully many times during the \MCMC\ run.
 Often some parameter directions are much worse than others; it is
@@
 
-Most MCMC methods
+Most \MCMC\ methods
   (though not all\note{We are promoters of an ensemble method that has a
@@
 This is easy to state, but hard to use in practice;
-  being a second-order statistic of the MCMC chain,
+  being a second-order statistic of the \MCMC\ chain,
   the autocorrelation time is a hard thing to measure without a lot of data,
@@
   (though you could never tune it precisely enough to warrant that third digit of accuracy).
-In the burn-in phase (discussed below) of an MCMC run,
+In the burn-in phase (discussed below) of an \MCMC\ run,
   it makes sense to track the acceptance ratio,
@@
 This process can be automated easily;
-  such automation is part of many projects that use MCMC.
+  such automation is part of many projects that use \MCMC.
 
@@
   is that tuning can \emph{only} take place during the burn-in phase;
-  you cannot tune \emph{while} you run your final MCMC run.
+  you cannot tune \emph{while} you run your final \MCMC\ run.
 Why not?
@@
   when you violate it,
-  you lose all the provable properties of MCMC
+  you lose all the provable properties of \MCMC\ 
   on which all our righteous power is based.
@@
   the algorithm looks like:
-Take an MCMC step in the hard parameters,
+Take an \MCMC\ step in the hard parameters,
   then take an exact step in the easy parameters,
@@
 re-code or extract the autocorrelation-time estimator called HOGG.
-This code should take a MCMC chain of parameter vectors as input and
+This code should take a \MCMC\ chain of parameter vectors as input and
 return a vector of autocorrelation times as output.
@@
 \begin{problem}\label{prob:tuning}
-Run the MCMC sampling of \problemname~\ref{prob:twod} with the
+Run the \MCMC\ sampling of \problemname~\ref{prob:twod} with the
 covariant Gaussian density.
@@
 \project{emcee} sampler (see \citealt{emcee}).
-Tune the Gaussian proposal distribution in your M--H MCMC sampler to sample this density efficiently.
+Tune the Gaussian proposal distribution in your \MH\ \MCMC\ sampler to sample this density efficiently.
 What autocorrelation time do you get?
@@
 
-Just as most (though not all) MCMC methods require a choice of
+Just as most (though not all) \MCMC\ methods require a choice of
 proposal distribution, most (though not all) require a choice about
@@
 parameter space.
-In many use cases for MCMC, the investigator wants to use it to obtain
+In many use cases for \MCMC, the investigator wants to use it to obtain
 uncertainty information about or propagate uncertainty into parameter
@@
 You can either wait a \emph{very long time} so that you see a single
-walker in a single MCMC run traverse from mode to mode enough times to
+walker in a single \MCMC\ run traverse from mode to mode enough times to
 make a representative sampling (dozens or hundreds of times, ideally),
@@
 they are not really different solutions.
-This suggests that when you ask whether the outputs of two MCMC chains
+This suggests that when you ask whether the outputs of two \MCMC\ chains
 are consistent with one another, you should do so in the realm of the
@@
 
-But it is worth remembering that just as MCMC is not a good optimizer
+But it is worth remembering that just as \MCMC\ is not a good optimizer
 (generically, samples will not lie close to the maximum of the posterior pdf),
 it is also not a good \emph{search algorithm}.
-There is no sense in which (standard) MCMC methods are efficient at
+There is no sense in which (standard) \MCMC\ methods are efficient at
 searching, or engineered to search, all of parameter space.
 If you really need to check all of parameter space, you should put
-MCMC aside, and do an exhaustive search.\note{You might want to open a
+\MCMC\ aside, and do an exhaustive search.\note{You might want to open a
   bottle of wine or two while the code runs.}
@@
 \problemname~\ref{prob:initialization} to begin by optimizing $\ln
-p(x)$ and only then start the MCMC sampler from that optimum.
+p(x)$ and only then start the \MCMC\ sampler from that optimum.
 Use a sensible optimizer.
@@
 to write in the title, the abstract, the tables, figures, and results
-section.  Given a MCMC sampling of the posterior pdf, what do we
+section.  Given a \MCMC\ sampling of the posterior pdf, what do we
 report as our \emph{results}?
@@
   which is that samplings are good for doing \emph{integrals}.
-We should endeavor to use as our ``results'' outputs from the MCMC
+We should endeavor to use as our ``results'' outputs from the \MCMC\ 
   that are based on integrals computed with the sampling.
@@
 This does \emph{not} include the ``best'' sample or a mode or optimum.
-The latter things are not necessarily illegitimate outputs of the MCMC,
+The latter things are not necessarily illegitimate outputs of the \MCMC,
   but they do not make best use of the fact that the sampling is a tool for integration,
@@
 For example,
-  imagine that by MCMC you have generated
+  imagine that by \MCMC\ you have generated
   $K$ samples $\pars_k$ of a parameter vector $\pars$.
@@
 This is not usually a good idea.
-For one, MCMC is a sampling algorithm, not an optimization algorithm;
+For one, \MCMC\ is a sampling algorithm, not an optimization algorithm;
   there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.%
@@
   it is like a maximum-likelihood value but regularized by the prior.
-It is an estimator with some good (and some bad) properties, but it is not the point of MCMC,
+It is an estimator with some good (and some bad) properties, but it is not the point of \MCMC,
   and therefore outside the scope of this \documentname.
@@
 ``banana-shaped'' posterior pdf.
-This all relates to the fact that MCMC is not an optimizer, it is a
+This all relates to the fact that \MCMC\ is not an optimizer, it is a
 sampler.
-It is important, when reporting the output of an MCMC run by giving
+It is important, when reporting the output of an \MCMC\ run by giving
 means or medians of the posterior PDF to remind the reader or user of
@@
 
-As we have discussed above, MCMC methods are proven to be correct ``in the limit''; that is, when
+As we have discussed above, \MCMC\ methods are proven to be correct ``in the limit''; that is, when
 an infinite number of samples have been taken.
@@
 can approach this limit very slowly (or not at all\note{HOGG: discuss
-  situations in which MCMC will not work---related to support.}).
+  situations in which \MCMC\ will not work---related to support.}).
 For this reason, in the finite duration of a scientific project, most
-users of MCMC have one of two (related) problems:
+users of \MCMC\ have one of two (related) problems:
 Either \textsl{(1)}~it is taking too long---it is taking too many steps or
@@
 that might help in different circumstances.
-We will not give a full description of all the MCMC algorithms out
+We will not give a full description of all the \MCMC\ algorithms out
 there, but just provide some field notes and pointers to references.
-New MCMC methods are being developed all the time, so a reader with a
+New \MCMC\ methods are being developed all the time, so a reader with a
 serious problem would do well to consult with the applied-mathematics
@@
   since that was written.}
-The spaces of MCMC methods and MCMC method implementations are both
-growing rapidly, in part because MCMC has become such a core
+The spaces of \MCMC\ methods and \MCMC\ method implementations are both
+growing rapidly, in part because \MCMC\ has become such a core
 technology in the empirical sciences.
@@
 As we discuss more below, one of the critical challenges in the use of
-MCMC is tuning the proposal pdf $q(\pars'\given\pars)$.
+\MCMC\ is tuning the proposal pdf $q(\pars'\given\pars)$.
 In problems in which all of the parameters (all of the components or entries of $\pars$) are somehow ``equivalent'',
@@
   ref emcee}, which is popular in the astrophysical sciences.
-Because ensemble methods are updating independent MCMC chains, they
+Because ensemble methods are updating independent \MCMC\ chains, they
 are often easy to parallelize to make use of multiple cores.
@@
 A trivial kind of Gibbs sampling was mentioned above in the
-discussion of M--H MCMC:
-Even when running vanilla M--H MCMC, it is often useful to update only
+discussion of \MH\ \MCMC:
+Even when running vanilla \MH\ \MCMC, it is often useful to update only
 one parameter at a time; that is, do only axis-aligned moves in the
@@
 
-\paragraph{Hamiltonian:}
-[HOGG, DFM: THIS HAS SOME OTHER NAME; WHAT?]
+\paragraph{Hamiltonian or Hybrid:}
 In many cases of interest, the function $f(\pars)$ can be analytically
@@
   long run.}
-There is a class of MCMC methods that make use of gradient information
+There is a class of \MCMC\ methods that make use of gradient information
 to effectively inform the proposal distribution and thus speed sampling.
@@
 that is close to the target function $f(\pars)$ you care about but
-hard to sample efficiently with MCMC.
+hard to sample efficiently with \MCMC.
 An example is a problem that has a simple (Gaussian, say) likelihood
@@
 It is easy to sample from a Gaussian likelihood times a Gaussian
-prior, even without MCMC, but it might be hard to do the same sampling
+prior, even without \MCMC, but it might be hard to do the same sampling
 under a more complex prior.
-Importance sampling is a direct sampling method (not really an MCMC
+Importance sampling is a direct sampling method (not really an \MCMC\ 
 method at all):
@@
 approximate function as a weight or probability.
-We use this quite a bit in our hierarchical inferences.\note{HOGG:
-  Citations}
+We use this quite a bit in our hierarchical inferences.\note{%
+  For example, we use importance sampling to convert individual-planet
+  exoplanet inferences (all performed with dumb priors) into a full-population
+  consistent hierarchical model in \citet{eccentricity}.}
 
@@
 \paragraph{Slice:}
-... When is slice sampling a good idea?  DFM, can you help here?
+DFM ... When is slice sampling a good idea?  DFM, can you help here?
 
@@
 \paragraph{Tempering:}
-A huge problem for MCMC samplers---which make local moves based on
+A huge problem for \MCMC\ samplers---which make local moves based on
 local positioning in the parameter space---is that there can be
@@
 zero or near-zero density.
-A vanilla MCMC method cannot easily cross such valleys with local
+A vanilla \MCMC\ method cannot easily cross such valleys with local
 moves.
@@
 out'' the function $f(\pars)$ or reduce its dynamic range, to make the
-peaks less tall and the valleys less deep, to permit the MCMC random
+peaks less tall and the valleys less deep, to permit the \MCMC\ random
 walk to cross the valleys.
@@
 In these cases you can only use certain kinds of samplers (though M--H
-MCMC is one you can choose), and, in addition, you need to make
+\MCMC\ is one you can choose), and, in addition, you need to make
 special kinds of proposals that permit the system to jump from one
@@
 
-MCMC is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
+\MCMC\ is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
   or a badly normalized all-positive function $f(\pars)$
@@
 In the context of data analysis,
-  MCMC is usually being used to obtain samples $\pars_k$
+  \MCMC\ is usually being used to obtain samples $\pars_k$
   that are \emph{parameter values} $\pars$ for a probabilistic model of some data.
-That is, MCMC is sampling a pdf \emph{for the parameters}.
+That is, \MCMC\ is sampling a pdf \emph{for the parameters}.
 If it is returning parameter values, then it is not sampling a pdf for the data.
-That is, MCMC can only sample a posterior (or prior) pdf for the parameters.
+That is, \MCMC\ can only sample a posterior (or prior) pdf for the parameters.
 \emph{It cannot sample a likelihood.}%
-\note{Well, technically, MCMC \emph{can} be used to sample a likelihood function,
+\note{Well, technically, \MCMC\ \emph{can} be used to sample a likelihood function,
   but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
@@
   (or log-likelihood function)
-  in as the input to the MCMC code, where the probability function
+  in as the input to the \MCMC\ code, where the probability function
   (or log-probability function) should go.
@@
 
-Ideally, if you are using MCMC for inference, your code should input
-to the MCMC function or routine a probability function which is a
+Ideally, if you are using \MCMC\ for inference, your code should input
+to the \MCMC\ function or routine a probability function which is a
 product of a prior times a likelihood.
 In terms of implementation, it is usually advisable to work in the
-logarithm of the density function, so the function input to the MCMC code
+logarithm of the density function, so the function input to the \MCMC\ code
 would be called something like \code{ln_f()}.
@@
 \code{ln_likelihood()}.
-(When working in logarithmic probability, in M--H-like MCMC methods,
+(When working in logarithmic probability, in M--H-like \MCMC\ methods,
 the accept--reject step would change from a comparison of a random
@@
 Of course the above pseudo-code presumes that when you perform the
-accept--reject step of the MCMC method, the programming language handles
+accept--reject step of the \MCMC\ method, the programming language handles
 properly the \code{-Inf} values.\footnote{If you are working in a language
@@
 \begin{problem}
-Run your M-H MCMC sampler from \problemname~\ref{prob:MH}, but now
+Run your M-H \MCMC\ sampler from \problemname~\ref{prob:MH}, but now
 with a density function that is precisely unity \emph{everywhere}
@@
 
+\begin{problem}
+For a real-world inference problem, read enough of
+\citet{fittingaline} to understand and execute Exercise~6 in that
+document.
+\end{problem}
+
 \section{Troubleshooting and advice}\label{sec:troubleshooting}
 
-When issues arise with MCMC sampling,
+When issues arise with \MCMC\ sampling,
   they are sometimes difficult to diagnose:
-Is it the MCMC code, the priors, the likelihood function,
+Is it the \MCMC\ code, the priors, the likelihood function,
   the initialization, the tuning, or something else?
@@
   (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
-  we recommend building some end-to-end functional tests that permit you to perform MCMC
+  we recommend building some end-to-end functional tests that permit you to perform \MCMC\ 
   runs with output that must meet certain expectations.
@@
   and check that the empirical covariance comes out as expected.
-This is an easy test to fail, so a success builds confidence in any MCMC code.
+This is an easy test to fail, so a success builds confidence in any \MCMC\ code.
 
@@
 Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
-Set up the MCMC sampler with the Gaussian function with this covariance matrix
+Set up the \MCMC\ sampler with the Gaussian function with this covariance matrix
   as the probability function (or the log of it as the log-probability function).
@@
   if you structure your code such that the log probability function \code{lnprob}
-  input to the MCMC scheme
+  input to the \MCMC\ scheme
   is set up as a sum of a log likelihood function \code{lnlikelihood}
@@
 
-MCMC should run well in this flat-likelihood case (priors are usually easy to sample)
+\MCMC\ should run well in this flat-likelihood case (priors are usually easy to sample)
   and the output sample histograms should look like what you expect
@@
 If you find that you are getting low acceptance fraction
-  (in standard M-H MCMC or other varieties for which there is such a concept)
+  (in standard M-H \MCMC\ or other varieties for which there is such a concept)
   you can simply reduce the sizes of the steps you consider to increase the fraction that are accepted.
@@
 \paragraph{initialization-dependence:}
-If you initialize your MCMC runs in different places in parameter space,
+If you initialize your \MCMC\ runs in different places in parameter space,
   do you get the same (or very similar) final parameter samplings,
@@
   then your posterior PDF is badly multi-modal;
-  trivial MCMC methods probably won't fix your problem.
+  trivial \MCMC\ methods probably won't fix your problem.
 Not to put too fine a point on it:
 If your results are initialization-dependent,
-  then your MCMC runs \emph{are not converged}.
+  then your \MCMC\ runs \emph{are not converged}.
 
@@
   and to do your best to express in your results the multiple optima.
-One of the miracles of MCMC is that if you \emph{can} fairly sample
+One of the miracles of \MCMC\ is that if you \emph{can} fairly sample
   as complex posterior PDF,
@@
 There are model checking methods (like varying the prior and
-re-running MCMC) that test the sensitivity of results to assumptions.
+re-running \MCMC) that test the sensitivity of results to assumptions.
 Not only are these very valuable for testing the model, they are
@@
 The integral is expensive to compute.
-Most MCMC methods are cleverly designed to \emph{avoid} computing this
+Most \MCMC\ methods are cleverly designed to \emph{avoid} computing this
 integral (as we note above in \sectionname~\ref{sec:when}); so it usually adds
@@
   Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2012,
-  \project{emcee}: The MCMC Hammer,
+  \project{emcee}: The \MCMC\ Hammer,
   \textit{Pubs.\,Astron.\,Soc.\,Pac.}\ \textbf{125} 306--312
@@
   in \citet{mcmchandbook}, 3--48.
-\bibitem[Hogg \etal(2010)]{fittingaline}
-  Hogg,~D.~W., Bovy,~J., \& Lang,~D., 2010,
+\bibitem[Hogg \etal(2010a)]{fittingaline}
+  Hogg,~D.~W., Bovy,~J., \& Lang,~D., 2010a,
   Data analysis recipes: Fitting a model to data,
   \arxiv{1008.4686v1}.
+\bibitem[Hogg \etal(2010b)]{eccentricity}
+  Hogg,~D.~W., Myers,~A.~D., \& Bovy,~J., 2010b,
+  \href{http://adsabs.harvard.edu/abs/2010ApJ...725.2166H}{Inferring the eccentricity distribution},
+  \textit{Astrophys.\ J.}\ \textbf{725}, 2166--2175.
 \bibitem[Hou \etal(2014)]{fengji}
@@
   Neal,~R.~M., 2011,
-  \href{http://www.mcmchandbook.net/HandbookChapter5.pdf}{MCMC Using Hamiltonian Dynamics},
+  \href{http://www.mcmchandbook.net/HandbookChapter5.pdf}{\MCMC\ Using Hamiltonian Dynamics},
   in \citet{mcmchandbook}, 113--162.

### 2016-02-08 mcmc.tex
cross-refs within text

@@
 % - Go through github issues, if any.
-% - Should we make an \acronym{} environment and use it religiously?  I think yes.
-% - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
+% - Systematically cite chapters of Brooks \etal in each section.
 % - mention ESJD
@@
 % - follow up the idea that burn-in doesn't exist at all!  It does for emcee.
-% - when we say ``below'' or ``above'' also give a Section Number.
 % - need to mention case where there is an MC integral INSIDE the likelihood function
-% - need to mention testing the likelihood function
+% - need to mention testing the likelihood function, by, say, 1-d scans back and forth.
 % - would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
@@
 simplest forms it can be run without computing any derivatives or
-integrals of the function, and (as we will show below) in its simplest
+integrals of the function, and (as we will show below in \sectionname~\ref{sec:MH}) in its simplest
 forms it is \emph{extremely easy to implement}.
@@
 compelled to note at the outset that \MCMC\ is in fact \emph{over-used}.
-Because \MCMC\ provably (under assumptions\note{Hint at assumptions
-  here.}, some of which will be discussed below) samples the full
+Because \MCMC\ provably (under assumptions\note{HOGG: Hint at assumptions
+  here.}, some of which will be discussed) samples the full
 posterior pdf in all of parameter space, many investigators use \MCMC\ 
@@
 
-\section{What is a sampling?}
+\section{What is a sampling?}\label{sec:sampling}
 
@@
 \end{eqnarray}
-That is, the sampling can be constructed (as we will show below) from
-$f(\pars)$ directly, and it permits you to compute expectation values
+That is, the sampling can be constructed (as we will show below in \sectionname~\ref{sec:MH}) from
+evaluations of $f(\pars)$ directly, and it permits you to compute expectation values
 without ever requiring you to integrate either the numerator integral
@@
 
-The ``correctness'' of a sampling is defined (above) in terms of its use in
+The ``correctness'' of a sampling is defined (above in \sectionname~\ref{sec:sampling}) in terms of its use in
   performing integrals or approximate computation of integrals.
@@
 Most of the good or sensible uses will somehow involve integration.
-
-% in the next paragraph I adopt $D$ to be the dimensionality of the space
 
@@
   where there are always parameters with very different levels of importance to the scientific conclusions.
-This point will return again below in \sectionname~\ref{sec:results}.
+This point will return again below (\sectionname~\ref{sec:results}).
 
@@
 the context of advanced sampling techniques like tempering or
-nested sampling, both mentioned briefly below.
+nested sampling, both mentioned briefly below (\sectionname~\ref{sec:methods}).
 
@@
 
-\section{Metropolis--Hastings MCMC}
+\section{Metropolis--Hastings MCMC}\label{sec:MH}
 
@@
   arbitrary run time, and two samples are only independent when they are
-  sufficiently separated in the chain (more on this below).
+  sufficiently separated in the chain (more on this below in \sectionname~\ref{sec:convergence}).
 
@@
 We will discuss the choice and tuning of this proposal distribution
-  below.
+  below (\sectionname~\ref{sec:tuning} and \sectionname~\ref{sec:trouble}).
 
@@
 It will help us tune the proposal pdf (\sectionname~\ref{sec:tuning})
-and diagnose problems (\sectionname~\ref{sec:troubleshooting}).
+and diagnose problems (\sectionname~\ref{sec:trouble}).
 
@@
   all in precisely the same parameterization.
-We will return to this point below in our comments on functional testing.
-
-As we discuss briefly below in \sectionname~\ref{sec:methods},
+We will return to this point below in our comments on testing (\sectionname~\ref{sec:trouble}).
+
+As we discuss briefly below (\sectionname~\ref{sec:methods}),
   there are many \MCMC\ methods more advanced than M--H.
@@
 In general, the best sampler will be different for different problems,
-and we will (below) tune the samplers we have to do the best on the
+and we will (below; \sectionname~\ref{sec:methods}) tune the samplers we have to do the best on the
 problems we have; this tuning will also be problem-specific.
@@
   when choosing proposal distribution (or, loosely speaking, step size),
-  is the autocorrelation time, described above.
+  is the autocorrelation time, described above (\sectionname~\ref{sec:convergence}).
 The optimal step size is the step size that makes for the shortest autocorrelation time.
@@
   (though you could never tune it precisely enough to warrant that third digit of accuracy).
-In the burn-in phase (discussed below) of an \MCMC\ run,
+In the burn-in phase (discussed below in \sectionname~\ref{sec:initialization}) of an \MCMC\ run,
   it makes sense to track the acceptance ratio,
@@
 
-\section{Initialization and burn-in}
+\section{Initialization and burn-in}\label{sec:initialization}
 
@@
 
-One thing we can keep in mind to guide us in this is what we said above,
+One thing we can keep in mind to guide us in this is what we said above (\sectionname{sec:sampling}),
   which is that samplings are good for doing \emph{integrals}.
@@
   the parameter value or values for the best sample.
-Giving some posterior samples is a good idea (see below).
+Giving some posterior samples is a good idea (we return to this below).
 If you feel drawn to give the best sample,
@@
 good-fitting models is a few---randomly chosen---example samples.
-We strongly recommend this\note{HOGG:  And we do this WHERE? Citation!};
+We strongly recommend this\note{HOGG:  And we do this WHERE? Cite Lang, cite eccentricity; etc??};
 in particular this is a better thing to do
@@
 
-As we have discussed above, \MCMC\ methods are proven to be correct ``in the limit''; that is, when
+As we have discussed above (\sectionname~\ref{sec:MH}),
+\MCMC\ methods are proven to be correct ``in the limit''; that is, when
 an infinite number of samples have been taken.
@@
 \paragraph{Ensembles:}
-As we discuss more below, one of the critical challenges in the use of
+As we discussed (\sectionname~\ref{sec:tuning}), one of the critical challenges in the use of
 \MCMC\ is tuning the proposal pdf $q(\pars'\given\pars)$.
@@
 
-A trivial kind of Gibbs sampling was mentioned above in the
+A trivial kind of Gibbs sampling was mentioned above (\sectionname~\ref{sec:MH}) in the
 discussion of \MH\ \MCMC:
@@
 That is, it is a posterior probability for some implicit (and improper) ``flat'' priors.
-
-% HOGG say:  I have no idea what the comment here (below) means, even though I wrote it!
-% ---
-% Give a sense of the difficulties you get into if you fail here; like
-% discussion with Sven Kreiss on 2013-02-14...
-% ---
 
@@
 
-\section{Troubleshooting and advice}\label{sec:troubleshooting}
+\section{Troubleshooting and advice}\label{sec:trouble}
 

### 2016-02-08 mcmc.tex
assumptions

@@
 compelled to note at the outset that \MCMC\ is in fact \emph{over-used}.
-Because \MCMC\ provably (under assumptions\note{HOGG: Hint at assumptions
-  here.}, some of which will be discussed) samples the full
+Because \MCMC\ provably (under assumptions\note{The assumptions include
+  things like:  The algorithm is run ``long enough'', where this phrase
+  is undefined (since the convergence requirements are set by precision
+  requirements on particular integrals), and that the density that is
+  being sampled has some connectedness properties:  There aren't distant islands of finite density
+  separated by regions of zero (or exceedingly low) density.}, some of which will be discussed) samples the full
 posterior pdf in all of parameter space, many investigators use \MCMC\ 

### 2016-02-08 mcmc.tex
autocorrelation time

@@
 
-The algorithm is the following:
-We have generated some set of samples, the most recent of which is $\pars_k$
-To generate the next sample $\pars_{k+1}$ do the following:
+The algorithm\note{Technically this is the \emph{Metropolis} algorithm
+  rather than the Metropolis--Hastings algorithm.  The difference is
+  that in the Metropolis--Hastings algorithm, we permit the proposal
+  distribution to disobey detailed balance and then we correct the
+  acceptance/rejection step to recover detailed balance.} is the
+following: We have generated some set of samples, the most recent of
+which is $\pars_k$ To generate the next sample $\pars_{k+1}$ do the
+following:
 \begin{itemize}
@@
 
-HOGG: ...Why does \MH\ \MCMC\ work (briefly)?  Here or later?
+Why does this algorithm work?  The answer is not absolutely
+trivial\note{There is a very mathematical discussion in \citet{geyer}
+  that we do not fully understand.  There is a more heuristic answer
+  on \project{Wikipedia} that we are happier with!},
+but there are two components to the argument:
+The first is that the Markov process delivers a unique stationary
+distribution.
+The second is that that stationary distribution is proportional to the
+density function $f(\pars)$.
 
@@
 You can break this property, if you like, and then adjust the
-  acceptance condition accordingly, but we \emph{do not recommend this}
+  acceptance condition accordingly (which is truly the Metropolis--Hastings algorithm), but we \emph{do not recommend this}
   except under the supervision of a trained professional.\note{Actually,
@@
   acceptance fraction.  Read the paper \cite{emcee} for more details.}
+The reason is:  It is one thing to draw samples from $q(x'\given x)$.
+It is another thing to correctly write down $q(x'\given x)$.
+  If you violate detailed balance in $q(x'\given x)$ then you have to be
+  able to \emph{both} draw from \emph{and} write down $q(x'\given x)$
+  and you are subject to a set of new possible bugs for your code.
 If the detailed balance condition frightens you (as it frightens us),
@@
 That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
-  despite many statements in the literature to the contrary.%
-  \note{Give example statements... HOGG: Mention cases in which you \emph{might} be able to claim this...}
+  despite many statements in the literature to the contrary.\note{%
+For example, you can (almost) never say that your chains are
+definitely converged, or that you have the posterior pdf correct to
+some given level of accuracy.  The reason is related to the ``no free
+lunch'' theorem of discrete optimization: In most problems there is no
+way you could have searched your parameter space finely enough to know
+this.  There are some exceptions of course.  In one kind of exception,
+your problem is convex or Gaussian, and you can analytically show that
+there is exactly one mode in the density and where it is.  Of course
+in these cases you rarely need \MCMC\ to solve your problems!  In
+another kind of exception, you can say something about the finite
+width or shape of the modes of the likelihood function (and prior
+pdf).  For example, sometimes the Cram\'er--Rao bound tells you that
+modes of the density must be smoother than some finite amount.  Then
+an exhaustive search of parameter space followed by \MCMC\ can in
+principle have provable properties.  But, as we say, these situations
+are rare.}
 The upshot of this is that we don't and can't require any kind of absolute convergence;
@@
 that separates good samplers from bad ones and is the ultimate basis
-on which we compare performance.\note{HOGG: Bash Dunkley and anyone
-  else who has used something else to compare.}
+on which we compare performance.\note{%
+  We sometimes see \MCMC\ methods compared according to burn-in times,
+  which, for one, depends strongly on initialization, and, for two,
+  depends strongly on tuning and dynamics.  Similarly we often see
+  comparisons in terms of acceptance ratio.  In principle the only
+  question is how precise are one's inferences given a certain amount
+  of computation.  This is set by the autocorrelation time and (pretty
+  much) nothing else, for reasonably converged chains.  In the real
+  world, of course, one must add to the CPU time the investigator time
+  spent tuning the method (and thinking about initialization) but
+  there is never (to our knowledge) a condition in which burn-in time
+  is the dominant consideration in choosing an MCMC method.}
 

### 2016-02-08 mcmc.tex
bits

@@
 % - Systematically cite chapters of Brooks \etal in each section.
-% - mention ESJD
 % - mention Gelman--Rubin statistic
@@
     (and higher acceptance fractions at lower numbers of dimensions)
-    are laid out in \citet{gelman234}.}
+    are laid out in \citet{gelman234}.\label{note:234}}
   (though you could never tune it precisely enough to warrant that third digit of accuracy).
@@
 
-% HOGG:  Is the following a repeat from previously and/or does it mesh with what's earlier?
-
 A very common---%
@@
   kind of proposal distribution is one that cycles through parameters,
-  taking a random step in just one parameter at a time.
+  taking a random step in just one parameter at a time.\note{This has a lot to do with Gibbs sampling, which is discussed in \sectionname~\ref{sec:methods}.}
 Aside from involving some persistence of state,

### 2016-02-09 mcmc.tex
problems, and conversions of HOGG to DFM

@@
 \begin{eqnarray}
-\tau = \sum_{\Delta = 0}^{\infty} C(\Delta)
+\tau = 2\,\sum_{\Delta = 0}^{\infty} C(\Delta)
 \label{eq:autocorrelationtime}\quad .
@@
 a sampling-based expectation value estimate.
+
+DFM: Here or in a note, can you give an empirical estimator of the
+autocorrelation time?  The naive estimator is good to give, and we
+could also give a good estimator too.
 
@@
   there is never (to our knowledge) a condition in which burn-in time
-  is the dominant consideration in choosing an MCMC method.}
+  is the dominant consideration in choosing an \MCMC\ method.}
 
@@
 
-\begin{problem}
+\begin{problem}\label{prob:estimatetau}
 Write a piece of code that computes the empirical autocorrelation
-function and the empirical autocorrelation time from a chain.  Run
+function and the empirical autocorrelation time from a chain using the na\"ive method DFM:WHERE?  Run
 this on the chain you obtained from \problemname~\ref{prob:MH}.
@@
 conclude?
+\end{problem}
+
+\begin{problem}
+Look at the source code of \project{emcee} \citep{emcee}, and see if you can
+re-code or extract the autocorrelation-time estimator called DFM:WHAT?
+This code should take a \MCMC\ chain of parameter vectors as input and
+return a vector of autocorrelation times as output.
 \end{problem}
@@
 
-\begin{problem}
-Look at the source code of HOGG: REFERENCE EMCEE, and see if you can
-re-code or extract the autocorrelation-time estimator called HOGG.
-This code should take a \MCMC\ chain of parameter vectors as input and
-return a vector of autocorrelation times as output.
-\end{problem}
-
 \begin{problem}\label{prob:tuning}
@@
 certain acceptance fraction, try to minimize the autocorrelation time.
+You will need one of the autocorrelation-time estimators you might have
+built in a previous \problemname.
 (This, by the way, is the Right Thing To Do, but often expensive.)
@@
 sensible parameter estimates, found by optimizing a likelihood or a
-posterior PDF in advance of sampling.
+posterior pdf in advance of sampling.
 In extremely high dimensions (large numbers of parameters), this can
@@
 If you feel drawn to give the best sample,
-  instead give the parameter values found by optimizing the posterior PDF,
+  instead give the parameter values found by optimizing the posterior pdf,
   starting at the best sample as an initialization
@@
 It is important, when reporting the output of an \MCMC\ run by giving
-means or medians of the posterior PDF to remind the reader or user of
+means or medians of the posterior pdf to remind the reader or user of
 that output that it does not necessarily represent (collectively) a
@@
 If you don't,
-  then your posterior PDF is badly multi-modal;
+  then your posterior pdf is badly multi-modal;
   trivial \MCMC\ methods probably won't fix your problem.
@@
   as we mentioned above, it is provably impossible
-  to explore all of parameter space if the posterior PDF is
+  to explore all of parameter space if the posterior pdf is
   complex in morphology.
 We don't have very useful advice here, except to
-  spend time learning about the multiple modes in the posterior PDF
+  spend time learning about the multiple modes in the posterior pdf
   and what each ``means'' or corresponds to,
@@
 One of the miracles of \MCMC\ is that if you \emph{can} fairly sample
-  as complex posterior PDF,
+  as complex posterior pdf,
   the fraction of samples in each mode
@@
   to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
-Of course this re-parameterization changes significantly the \emph{form} of the prior PDF
+Of course this re-parameterization changes significantly the \emph{form} of the prior pdf
   (the transformation brings in a Jacobian);

### 2016-02-09 mcmc.tex
2 hoggs

@@
 We usually use quantiles.
-That is, the ``one-sigma'' error bar can be taken to be the half-size of the central (or smallest%
-  \note{HOGG shizzle about choosing quantiles}) interval of the parameter
-  that contains 68 percent of the posterior samples.
+That is, the ``one-sigma'' error bar can be taken to be the half-size of the central (or smallest\note{%
+  We usually create a 68-percent (or 95-percent) interval or region by
+  excluding the top and bottom 16 percent (or 2.5 percent) of the
+  posterior samples.  Some more aggressive investigators find the
+  smallest interval that contains 68 (or 95) percent of the samples.
+  Both are legitimate, and they are near-identical for distributions
+  that are near-symmetric around the mean.})
+  interval of the parameter that contains 68 percent of the posterior
+  samples.
 The ``two-sigma'' error bar would be the same but for the 95-percent interval.
@@
 Again, to form these limits, $q(\pars)$ must be a scalar function (so sorting is well defined).
+Because they are so afraid of being confused with frequentists,
+Bayesians often call these regions ``credible'' rather than
+``confidence''\note{We don't like this ``credible'' terminology,
+  though we sometimes adopt it to avoid confusion.}
 
@@
 good-fitting models is a few---randomly chosen---example samples.
-We strongly recommend this\note{HOGG:  And we do this WHERE? Cite Lang, cite eccentricity; etc??};
-in particular this is a better thing to do
-than to give the ``best sample'', which isn't even guaranteed to be
-near the bulk of the posterior pdf or the samples therefrom.
+We strongly recommend this\note{And we do it ourselves.  For examples,
+  you can look at \citet{cometholmes} and \citet{exopop}.}; in
+particular this is a better thing to do than to give the ``best
+sample'', which isn't even guaranteed to be near the bulk of the
+posterior pdf or the samples therefrom.
 
@@
   \project{emcee}: The \MCMC\ Hammer,
-  \textit{Pubs.\,Astron.\,Soc.\,Pac.}\ \textbf{125} 306--312
+  \textit{Pubs.\ Astron.\ Soc.\ Pac.}\ \textbf{125} 306--312
   \doi{10.1086/670067}
   (also \arxiv{1202.3665}).
+\bibitem[Foreman-Mackey \etal(2014)]{exopop}
+  Foreman-Mackey,~D., Hogg,~D.~W., \& Morton,~T.~D., 2014,
+  \href{http://adsabs.harvard.edu/abs/2014ApJ...795...64F}{Exoplanet population
+    inference and the abundance of earth analogs from noisy, incomplete catalogs},
+  \textit{Astrophys.\ J.}\ \textbf{795} 64
+  \doi{10.1088/0004-637X/795/1/64}.
 \bibitem[Geisser(1993)]{crossval}
@@
   \href{http://adsabs.harvard.edu/abs/2010ApJ...725.2166H}{Inferring the eccentricity distribution},
-  \textit{Astrophys.\ J.}\ \textbf{725}, 2166--2175.
+  \textit{Astrophys.\ J.}\ \textbf{725} 2166--2175.
 \bibitem[Hou \etal(2014)]{fengji}
@@
   \arxiv{1401.6128}.
+\bibitem[Lang \& Hogg(2012)]{cometholmes}
+  Lang,~D. \& Hogg,~D.~W., 2012,
+  \href{http://adsabs.harvard.edu/abs/2012AJ....144...46L}{Searching for comets on the World Wide Web:
+    The orbit of 17P/Holmes from the behavior of photographers},
+  \textit{Astron.\ J.}\ \textbf{144} 46
+  \doi{10.1088/0004-6256/144/2/46}.
 \bibitem[Metropolis \etal(1953)]{metropolis}

### 2016-02-09 mcmc.tex
priors wrong

@@
 % -----
-% - Kill all issues in this list, below:
+% - Kill all issues in this list, below
 % - Deal with all mentions of ``DFM'' or ``HOGG''
@@
 % - Systematically cite chapters of Brooks \etal in each section.
+% - Skim MacKay and make sure we cite him in any relevant place.
 % - mention Gelman--Rubin statistic
@@
 need many samples; a dozen suffices to give you a reasonable posterior
-pdf mean and variance.\note{HOGG cite Mackay.}
+pdf mean and variance.\note{This point is made well by \citet{mackay}.}
 If you care about $K$ parameters or dimensions of the parameter space,
@@
   tests that your priors really are what you think they are
-  (and there are so many ways for them \emph{not} to be\note{HOGG}),
-  and tests that your prior pdfs are proper
-  (functions that can be normalized).
+  (and there are so many ways for them \emph{not} to be\note{It is
+    easy to think you have flat priors when in practice your sampler
+    has flat priors in the logarithms of the parameters or inverses of
+    the parameters.  Also, it is easy to have improper priors when you
+    think they are proper, because limits are not working, or you were
+    wrong about the convergence of some integral.  Sometimes the prior pdf in practice has
+    different edges than you think because there is some censoring
+    happening that you haven't considered.  And then, of course, it is
+    possible for your sampler itself to be buggy.}), and tests that
+  your prior pdfs are proper (functions that can be normalized).
 This test is most effective as a true end-to-end test of your model
@@
   \doi{10.1088/0004-6256/144/2/46}.
+\bibitem[MacKay(2003)]{mackay}
+  MacKay,~D.~J.~C., 2003,
+  \href{http://www.inference.phy.cam.ac.uk/mackay/itila/book.html}{\textit{Inference, Information Theory, and Learning Algorithms}},
+  Cambridge University Press
+  \isbn{9780521642989}.
 \bibitem[Metropolis \etal(1953)]{metropolis}

### 2016-02-10 mcmc.tex
notes to self
nope

@@
 % - mention Gelman--Rubin statistic
-% - mention JAGS and STAN in methods section
+% - mention & cite JAGS and STAN in methods section
 % - follow up the idea that burn-in doesn't exist at all!  It does for emcee.
+% - do we explain how to initialize emcee; do we need to?
 % - need to mention case where there is an MC integral INSIDE the likelihood function
@@
 in general the need for samples grows exponentially (or perhaps even
-factorially!) with the number of dimensions.\note{HOGG cite something.}
+factorially!)\ with the number of dimensions.
 So if you have a ten-dimensional space, expect to be publishing your

### 2016-02-10 mcmc.tex
two HOGGs

@@
 \\
-Z &=& \int f(\pars)\,\dd \pars
+Z &=& \int f(\pars)\,\dd \pars \label{eq:proper}
 \quad .
@@
 samples in an electronic table on the web!
+
 If you \emph{do} publish sufficient numbers of posterior samples, you
-might find users who want to make serious use of them.\note{HOGG cite
-  our importance-sampling papers with some explanation here.}
+might find users who want to make serious use of them.\note{We have
+  been pushing this approach in our own work; our hierarchical
+  inferences \citep{eccentricity, exopop} make use of importance
+  samplings, starting at posterior samplings from interim prior pdfs.
+  We have used this technology for computational convenience, but it
+  is a framework for building subsequent analyses on the posterior
+  samplings provided by other investigators.  One key---if you want
+  your posterior samplings to be useful---is that you must release
+  (along with your posterior sampling), the value of the prior pdf at
+  the locations of the samples.  That is, you must ``decorate'' the
+  samples (augment them) with prior pdf values or calls, or else
+  release an executable version of your prior pdf, into which all your
+  samples can be put as inputs.  In \emph{principle} it is enough to
+  release a \emph{description} of your prior pdf, but in practice it
+  is rarely correct or specific enough for another investigator to
+  duplicate exactly.  So just augment your samples!}
 
@@
 
-HOGG: ...It is a good idea to have proper priors.
-It is a good test of your priors that you can sample with a ``null'' likelihood function.
-You can often have your priors wrong despite having beautiful code.
+It should be outside of the scope of this document to note here that
+it is a good idea to have proper priors.\note{This point---that you
+  should be using proper priors---is just basic Bayesian good
+  practice, often violated but never for good reasons.}
+Proper priors obey the integral constraint (\ref{eq:proper}),
+with $Z$ finite.
+It isn't a requirement that priors be proper for the posterior to 
+be proper, so many investigators and projects violate this rule.
+This is outside the current scope, except for the following point:
+It is a great functional test of your sampler and your data analysis
+setup to take the likelihood function to a very small power (much less
+than one) or multiply the log-likelihood by a very small number (much
+less than one) and check that the sampler samples, properly and
+correctly, the prior pdf.
+This test is only possible if the prior pdf is proper.
 
@@
 
-%% ...Initialization and burn-in advice we haven't addressed earlier?
-
 %% ...Finally, to repeat: You can't sample the likelihood!  You can only
 %% sample a posterior pdf!
-
-%% % \begin{figure}[htbp]
-%% % \exampleplot{ex17}
-%% % \caption{Partial solution to \problemname~\ref{prob:bayesintrinsic}:
-%% % The marginalized posterior probability distribution for the intrinsic
-%% % variance.}\label{fig:bayesintrinsic}
-%% % \end{figure}
 

### 2016-02-10 mcmc.tex
found more places to fix

@@
 
-For example, one magical property of a sampling (in a $D$-dimensional space)
-  is that a \emph{histogram} (a $D$-dimensional histogram) of the samples
-  (divided, if you like, by the number of samples in each bin and the bin width\note{something about units and normalization})
+For example, one magical property of a sampling (in a $D$-dimensional
+space) is that a \emph{histogram} (a $D$-dimensional histogram) of the
+samples (divided, if you like, by the number of samples in each bin
+and the bin width\note{HOGG: something about units and normalization})
 looks very much like the pdf from which the samples were drawn.
@@
 be able to solve the famously difficult discrete optimization
-problem.\note{explain this remark and cite something here}
+problem.\note{HOGG: explain this remark and cite something here}
 Another way to put it is that you might have two modes---two separated
@@
   starting at the best sample as an initialization
-  (and sacrifice your probabilism\note{something}).
+  (and sacrifice your probabilism\note{HOGG: something}).
 This optimal-posterior parameter value is called the

### 2016-02-10 mcmc.tex
two more HOGGs, burn in

@@
 % - mention & cite JAGS and STAN in methods section
-% - follow up the idea that burn-in doesn't exist at all!  It does for emcee.
 % - do we explain how to initialize emcee; do we need to?
@@
 samples (divided, if you like, by the number of samples in each bin
-and the bin width\note{HOGG: something about units and normalization})
-looks very much like the pdf from which the samples were drawn.
+and the bin width\note{We are referring here to the point that if you
+  want a histogram of samples to look just like the posterior pdf from
+  which the samples are drawn, the histogram (thought of as a step
+  function) must integrate to unity.}  looks very much like the pdf
+from which the samples were drawn.
 This is a way to ``reconstruct'' the pdf from the sampling:
@@
 be able to solve the famously difficult discrete optimization
-problem.\note{HOGG: explain this remark and cite something here}
+problem.\note{Famously, there is ``no free lunch''
+  \citep{nofreelunch}: You can't find the global optimum of a general
+  optimization problem without exhastive search of all possibilities.
+  This is very closely related---somehow---to the difficulty of sampling.}
 Another way to put it is that you might have two modes---two separated
@@
   because it is relevant to tuning---%
-  is that tuning can \emph{only} take place during the burn-in phase;
+  is that tuning can \emph{only} take place during the burn-in phase
+  (that is, some part of your chain you will discard later);
   you cannot tune \emph{while} you run your final \MCMC\ run.
@@
 usually better than a mindless initialization.
+
+As to burn-in:
+If you have started your sampler in a non-typical place---or if you
+are concerned that you \emph{might} have started your sampler in a
+non-typical place---then you should discard the beginning of your
+\MCMC\ run before you do your inferences.
+This discarded part is called the ``burn-in''.
+Some practitioners insist that burn-in cannot exist\note{\citet{geyer}
+  takes this view.}: So long as the initial point is a conceivable
+sample, you are fine!
+Ensemble methods (like \project{emcee}\note{\citet{emcee}.}; and
+mentioned below in \sectionname~\ref{sec:methods}) require a burn-in
+phase and discard, because the whole point is that the ensemble must
+grow (or shrink) to fill the posterior pdf volume, and you can't
+initialize the ensemble sensibly if you don't know that volume
+\foreign{a priori}.
 
@@
   \doi{10.1109/MCSE.2011.37}.
+\bibitem[Wolpert \& Macready(1997)]{nofreelunch}
+  Wolpert,~D.~H. \& Macready,~W.~G., 1997,
+  No Free Lunch Theorems for Optimization,
+  \textit{IEEE Trans.\ Evolutionary Computation}\ \textbf{1} 67.
 \end{thebibliography}

### 2016-02-10 mcmc.tex
not best sample

@@
 Giving some posterior samples is a good idea (we return to this below).
-If you feel drawn to give the best sample,
-  instead give the parameter values found by optimizing the posterior pdf,
-  starting at the best sample as an initialization
-  (and sacrifice your probabilism\note{HOGG: something}).
+
+If you feel drawn to give the best sample, instead give the parameter
+values found by optimizing the posterior pdf, starting at the best
+sample as an initialization (and sacrifice your probabilism\note{We
+  choose not to count the number of papers out there that claim to be
+  ``Bayesian'' but then deliver the optimum of a posterior PDF.  That
+  sacrifices probabilism, but also all the useful things you get from
+  delivering a posterior, like protection from over-fitting, and
+  non-approximate uncertainty propagation.  You certainly don't need
+  to be a Bayesian---you can do most of the same science as a
+  frequentist---but if you are using \MCMC\ you probably are doing so
+  because you want the benefits of Bayesianism.}).
 This optimal-posterior parameter value is called the

### 2016-02-12 mcmc.tex
kill two HOGGs, add one back

@@
 % - Skim MacKay and make sure we cite him in any relevant place.
-% - mention Gelman--Rubin statistic
-% - mention & cite JAGS and STAN in methods section
-% - do we explain how to initialize emcee; do we need to?
-% - need to mention case where there is an MC integral INSIDE the likelihood function
-% - need to mention testing the likelihood function, by, say, 1-d scans back and forth.
-% - would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
-% - send to friendlies for comments
+% - Mention & cite JAGS and STAN in methods section
+% - Do we explain how to initialize emcee; do we need to?
+% - Need to mention case where there is an MC integral INSIDE the likelihood function
+% - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
+% - Would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
+% - Send to friendlies for comments
 %   - Gelman
@@
 %   - Brewer
-% - post on arXiv
+% - Post on arXiv
+
+% low priority
+% ------------
+% - Make endnote numbers clickable links to endnotes and back.  Relevant ideas here:
+%   http://tex.stackexchange.com/questions/8452/making-endnotes-clickable-links-with-hyperref
 
@@
 % - pdf not PDF
-% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
+% - Is it ``a MCMC sampling'' or ``an MCMC sampling''?
 % - \note{} right after word if you are endnoting a phrase,
 %   but after the full stop if endnoting the sentence or paragraph.
-% - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
-% - should we use the word ``walker'' everywhere and do we?
+% - Always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
+% - Should we use the word ``walker'' everywhere and do we?
 
@@
 
+HOGG: Introduce and explain the Gelman--Rubin
+diagnostic\note{\citet{gelmanrubin}.}, which is also a good idea, and
+which tests something slightly different than the autocorrelation
+time.
+
 Finally one last point about convergence:
@@
 If you are clever, all these plots can be arranged into a lovely and
-informative triangle.\note{HOGG cite corner.py and put an example in
-  the note!}
+informative triangle.\note{Our favorite package these days is our own \project{corner.py}.  Find it on the internets.}
 These two-dimensional plots do not in any sense contain all the
@@
 When the function $f(\pars)$ has challenging properties, the results
-can approach this limit very slowly (or not at all\note{HOGG: discuss
-  situations in which \MCMC\ will not work---related to support.}).
+can approach this limit very slowly (or not at all\note{You can get
+  essentially infinite autocorrelation times if the function has
+  isolated regions of finite density separated by large seas of zero
+  or near-zero density.  One disturbing thing about these situations
+  is that any empirical measure of autocorrelation time will give a
+  finite answer, but the density won't be properly sampled, ever.  In
+  principle you can find these with multiple re-starts of the chain
+  and (things like) the Gelman--Rubin diagnostic \citep{gelmanrubin}.}).
 For this reason, in the finite duration of a scientific project, most
@@
   \isbn{978-0412034718}.
+\bibitem[Gelman \& Rubin(1992)]{gelmanrubin}
+  Gelman,~A., \& Rubin,~D.~B., 1992,
+  Inference from iterative simulation using multiple sequences,
+  \textit{Statist.\ Sci.}\ \textbf{4} 457--472.
 \bibitem[Gelman \etal(1996a)]{ppcheck}

### 2016-02-12 mcmc.tex
HOGGs

@@
 % -----
-% - Kill all issues in this list, below
 % - Deal with all mentions of ``DFM'' or ``HOGG''
@@
 % - Skim MacKay and make sure we cite him in any relevant place.
-% - Mention & cite JAGS and STAN in methods section
-% - Do we explain how to initialize emcee; do we need to?
+% - Mention & cite STAN in methods section
 % - Need to mention case where there is an MC integral INSIDE the likelihood function
@@
 %   - Brewer
+% - Kill all issues in this list, above, and then:
 %   - Post on arXiv
+%   - Blog about it
 
@@
 reducing the tuning requirements.
-We are co-authors on one of these methods, \project{emcee}\note{HOGG:
-  ref emcee}, which is popular in the astrophysical sciences.
+We are co-authors on one of these methods,
+\project{emcee}\note{\citet{emcee}}, which is popular in the
+astrophysical sciences.
+It is worty of note that ensemble methods have particular
+peculiarities about initialization and convergence diagnostics that
+are worth understanding before use.\note{For example, we find
+  empirically that it is better to initialize the ensemble in a ball
+  that is smaller than the full posterior pdf width, not larger.  For
+  another example, the ensemble produces a set of chains that can be
+  used to compute the Gelman--Rubin diagnostic \citep{gelmanrubin},
+  but this is not a conservative test if the initialization was made
+  in a small ball.}
 Because ensemble methods are updating independent \MCMC\ chains, they
@@
 For problems with these structures, Gibbs samplers are ideal, and
-there are good implementations.\note{HOGG cite JAGS}
+there are good implementations.\note{For example, \project{JAGS}; find it on the internets.}
 
@@
 a whole.
-Parallelization is built in to some Gibbs implementations.\note{HOGG}
+Parallelization is built in to some Gibbs implementations.\note{For
+  example, \project{JAGS} automatically
+  parallelizes Gibbs problems.}
 
@@
 \paragraph{Data augmentation:}
-...HOGG?
+...HOGG Look at the relevant chapter in Brooks.
 
@@
 not of fixed dimension.
+
 For example, if you are modeling an astronomical image with a
@@
 case the number of stars itself is a parameter, so the number of
-parameters is itself a function of the parameters.\note{HOGG Cite
-  Brewer et al}
+parameters is itself a function of the parameters.\note{Our attempt to sample this very problem is \citet{brewer}.}
 In these cases you can only use certain kinds of samplers (though M--H
@@
   \isbn{978-0471810339}.
+\bibitem[Brewer \etal(2013)]{brewer}
+  Brewer,~B.~J., Foreman-Mackey,~D., \& Hogg,~D.~W., 2013,
+  {Probabilistic catalogs for crowded stellar fields},
+  \textit{Astron.\,J.} \textbf{146} 7
+  \doi{10.1088/0004-6256/146/1/7}.
 \bibitem[Brooks \etal(2011)]{mcmchandbook}

### 2016-02-12 mcmc.tex
clarification

@@
 proposal distribution $q(x'\given x)$ with a Gaussian distribution in
-$\ln x$ $q(\ln x'\given \ln x)$.  What about your answers change?  How
-wrong is your answer?  Bonus points: Compute the analytic function
-that you have sampled from (that is, not $p(x)$) and over-plot it on
-your histogram.
+$\ln x$ $q(\ln x'\given \ln x)$, but---incorrectly---make no other
+changes.  What about your answers change?  How wrong is your answer?
+Bonus points: Compute the analytic function that you have sampled from
+(that is, not $p(x)$) and over-plot it on your histogram.
 \end{problem}

### 2016-02-12 mcmc.tex
Fix minor typos.

@@
   but suffice it to say:
-  If the computation of $Z$ is in your travel plans, you will need yto
+  If the computation of $Z$ is in your travel plans, you will need to
   think about implementing some non-trivial integrating or sampling
@@
 \emph{sampler}.  If you are trying to find the optimum of the
-likelihood or the posterior pdf, you should use an \emph{optimzer},
+likelihood or the posterior pdf, you should use an \emph{optimizer},
 not a sampler.  If you want to make sure you search all of parameter
@@
   \citep{nofreelunch}: You can't find the global optimum of a general
-  optimization problem without exhastive search of all possibilities.
+  optimization problem without exhaustive search of all possibilities.
   This is very closely related---somehow---to the difficulty of sampling.}

### 2016-02-12 mcmc.tex
Fix a typo (?).

--- 2016-02-12 mcmc.tex
@@
   and given the intuition you gain by doing it (at least once) yourself.\note{%
-    We also make this point in a previous \documentname\ (\citealt{fittingaline};
+    We also make this point in a previous \documentname\ (\citealt{fittingaline});
     the ambitious reader will find in that \documentname\ a solid example project

### 2016-02-13 mcmc.tex
Fix typos.

@@
   but it will take a long time to move anywhere because of timidity.
-If the propsoal distribution is too wide---%
+If the proposal distribution is too wide---%
   it proposes steps too large---%
@@
   to $D$ one-dimensional problems:
-In this form of proposal distrubution,
+In this form of proposal distribution,
   it is possible to track a separate acceptance fraction for every parameter.
@@
   of the posterior samples,
-  with the relvant quantiles (2.5, 16, 50, 84, and 97.5~percent) indicated.
+  with the relevant quantiles (2.5, 16, 50, 84, and 97.5~percent) indicated.
 

### 2016-02-13 mcmc.tex
Fix typo (?)

--- 2016-02-13 mcmc.tex
@@
 list compiled today will be incomplete tomorrow.\note{One check-point
-  in this literature is the compliation volume by
+  in this literature is the compilation volume by
   \citet{mcmchandbook}, but much has been developed and changed even

### 2016-02-13 mcmc.tex
Fix reference to section 'sec:sampling'.

--- 2016-02-13 mcmc.tex
@@
 
-One thing we can keep in mind to guide us in this is what we said above (\sectionname{sec:sampling}),
+One thing we can keep in mind to guide us in this is what we said above (\sectionname~\ref{sec:sampling}),
   which is that samplings are good for doing \emph{integrals}.

### 2016-02-13 mcmc.tex
Switch from 'endnotes' package to 'enotez'.

@@
 \documentclass[12pt,twoside,pdftex]{article}
-\usepackage{styles/dar_endnotes}
+\usepackage[backref=true]{enotez}
 
@@
 \newcommand{\better}{{\mathrm{(better)}}}
+
+\let\note=\endnote
+
+\DeclareInstance{enotez-list}{custom}{paragraph}
+{
+  number = \makebox[0pt][r]{\textsuperscript{#1}},
+  format = \indent\normalfont,
+  number-format = \bfseries,
+}
 
@@
 
-One thing we can keep in mind to guide us in this is what we said above (\sectionname~\ref{sec:sampling}),
+One thing we can keep in mind to guide us in this is what we said above (\sectionname{sec:sampling}),
   which is that samplings are good for doing \emph{integrals}.
@@
 \clearpage
-\markright{Notes}\theendnotes
+\markright{Notes}\printendnotes[custom]
 

### 2016-02-14 mcmc.tex
Fix typo.

@@
 \documentclass[12pt,twoside,pdftex]{article}
-\usepackage[backref=true]{enotez}
+\usepackage{styles/dar_endnotes}
 
@@
 \newcommand{\better}{{\mathrm{(better)}}}
-
-\let\note=\endnote
-
-\DeclareInstance{enotez-list}{custom}{paragraph}
-{
-  number = \makebox[0pt][r]{\textsuperscript{#1}},
-  format = \indent\normalfont,
-  number-format = \bfseries,
-}
 
@@
   a robot can write the function that is the derivative of that
-  function, using simple differentiation rules andthe magic of the
+  function, using simple differentiation rules and the magic of the
   chain rule. Differentiation is a typographic operation on computer
@@
 \clearpage
-\markright{Notes}\printendnotes[custom]
+\markright{Notes}\theendnotes
 

### 2016-02-15 mcmc.tex
thank AG

@@
           Iain Murray (Edinburgh)
-        for valuable advice and comments over the years.
+        for valuable advice and comments over the years, and
+          Alessandro Gentilini
+        for a very close read.
         This project benefitted from various research grants, especially

### 2016-02-15 mcmc.tex
lam citation

@@
 It is disappointing and annoying to many that there is no extremely
-simple and reliable answer to this question.
+simple and reliable answer to this question.\note{There is very good
+  coverage of most of the points in this \sectionname, and more, in a
+  set of lecture notes by Patrick Lam up at
+  \url{http://www.people.fas.harvard.edu/~plam/teaching/methods/convergence/convergence_print.pdf}.}
 

### 2016-02-16 mcmc.tex
Gelman--Rubin

@@
 
-HOGG: Introduce and explain the Gelman--Rubin
-diagnostic\note{\citet{gelmanrubin}.}, which is also a good idea, and
-which tests something slightly different than the autocorrelation
-time.
+One simple and sensible test of convergence is the Gelman--Rubin
+diagnostic \note{\citet{gelmanrubin}.}, which compares the variance
+(in one parameter, or your most important parameter, or all
+parameters) \emph{within} a chain to the variance \emph{across}
+chains.
+This requires running multiple chains, and looking at the empirical
+variance of the parameter away from its mean within each individual
+chain, and comparing it to the variance in the mean of that parameter
+across chains, inflated to be a mean per-sample variance.
+What Gelman and Rubin do with these variances specifically is
+sensible, but the important question of convergence is whether, as the
+chains get longer, these two variances asymptotically reach stable
+values, and that those two values agree.
+The Gelman--Rubin diagnostic is related to the autocorrelation time,
+in that it will only deliver success when the chains are much longer
+than an autocorrelation time.
+It can also be used to assess the accuracy of the chain for performing
+integrals---by replacing the mean of the parameter with whatever
+integral is of most interest.
 

### 2016-02-24 mcmc.tex
notes to self
problems, a comment, and I like my expectations specific

@@
 
-Given this pdf $p(\pars)$, we can define \emph{expectation values} $E[\pars]$
+Given this pdf $p(\pars)$, we can define \emph{expectation values}
+$E_{p(\pars)}[\pars]$
 for $\pars$ or for any quantity that can be expressed as a function $q(\pars)$
@@
 \begin{eqnarray}
-E[\pars] &\equiv& \int \pars\,p(\pars)\,\dd \pars
+E_{p(\pars)}[\pars] &\equiv& \int \pars\,p(\pars)\,\dd \pars
 \\
-E[q] &\equiv& \int q(\pars)\,p(\pars)\,\dd \pars
+E_{p(\pars)}[q(\pars)] &\equiv& \int q(\pars)\,p(\pars)\,\dd \pars
 \quad ,
@@
 \begin{eqnarray}
-E[\pars] &\approx& \frac{1}{K}\,\sum_{k=1}^K \pars_k
+E_{p(\pars)}[\pars] &\approx& \frac{1}{K}\,\sum_{k=1}^K \pars_k
 \\
-E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
+E_{p(\pars)}[q(\pars)] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
 \quad .
@@
 
+DFM says: I don't agree with this section. Using MCMC you \emph{are} sampling
+from $p(\pars)$\ldots you \emph{can't} sample from $f(\pars)$.
+
 When the sampling $\setof{\pars_k}_{k=1}^K$ is of one of these badly
@@
 \begin{eqnarray}
-E[q] &\equiv& \frac{\int q(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
+E_{p(\pars)}[q(\pars)] &\equiv& \frac{\int q(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
 \\
-E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
+E_{p(\pars)}[q(\pars)] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
 \label{eq:expectationvalue}
@@
 \paragraph{visualization:}
+HOGG: What plots do you make to diagnose problems?
+
+HOGG: Plot parameters vs ``time'' in the chain.
+
+HOGG: Plot a cut through the density function and the samples near that cut.
+
 HOGG:...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?

### 2016-02-24 problems/p1.py

@@ -1,0 +1,53 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+from plot_setup import SQUARE_FIGSIZE, COLORS
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+np.random.seed(42)
+
+
+def compute_stats(K, a=0.0, b=1.0):
+    a, b = sorted((a, b))
+    x = np.random.uniform(a, b, size=K)
+
+    # Analytic results:
+    p = 1. / (b - a)
+    a_mu = 0.5 * p * (b**2 - a**2)
+    a_var = p * ((b - a_mu)**3 - (a - a_mu)**3) / 3.
+    a_skew = 0.25 * p * ((b-a_mu)**4-(a-a_mu)**4) / a_var**(3./2)
+    a_kurt = p * ((b-a_mu)**5-(a-a_mu)**5) / a_var**2 / 5.0
+
+    # Sample estimates of stats:
+    s_mu = np.mean(x)
+    s_var = np.var(x)
+    s_skew = np.mean((x - s_mu)**3) / s_var**(3./2)
+    s_kurt = np.mean((x - s_mu)**4) / s_var**2
+
+    return (
+        (a_mu, a_var, a_skew, a_kurt),
+        (s_mu, s_var, s_skew, s_kurt),
+    )
+
+
+K = 4**np.arange(1, 11)
+a_stats, s_stats = map(np.array, zip(*(compute_stats(k) for k in K)))
+
+fig, axes = pl.subplots(2, 2, sharex=True, figsize=2*SQUARE_FIGSIZE)
+
+for i, (ax, title) in enumerate(zip(
+        axes.flat, ("mean", "variance", "skewness", "kurtosis"))):
+    mu = a_stats[0, i]
+    ax.axhline(mu, color=COLORS["DATA"])
+    ax.plot(np.log2(K), s_stats[:, i], "-o", color=COLORS["MODEL_1"], ms=3)
+    d = np.max(np.abs(np.array(ax.get_ylim()) - mu))
+    ax.set_ylim(mu + d * np.array([-1, 1]))
+    ax.set_ylabel(title)
+
+[ax.set_xlabel(r"$\log_2 K$") for ax in axes[1]]
+
+pl.savefig("p1.pdf")

### 2016-02-24 problems/plot_setup.py

@@ -1,0 +1,27 @@
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+from cycler import cycler
+from matplotlib import rcParams
+
+__all__ = ["COLORS"]
+
+rcParams["font.size"] = 20
+rcParams["font.family"] = "serif"
+rcParams["font.serif"] = "Computer Sans"
+rcParams["text.usetex"] = True
+rcParams["figure.autolayout"] = True
+rcParams["axes.prop_cycle"] = cycler("color", (
+    "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
+    "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
+))  # d3.js color cycle
+
+COLORS = dict(
+    DATA="k",
+    MODEL_1="#1f77b4",
+    MODEL_2="#ff7f0e",
+)
+
+SQUARE_FIGSIZE = np.array((4, 4))

### 2016-02-24 mcmc.tex
problems, a comment, and I like my expectations specific

@@
   of this \sectionname\ and implement it forthwith, in a short piece of computer code,
-  in the context of some simple problem.\note{%
+  in the context of some simple problems.\note{%
     We request this in a trivial case below in
@@
 \item Draw a random number $0<r<1$ from the uniform distribution.
-\item If $f(\pars') / f(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
+\item If $f(\pars') \, q(\pars_k\given\pars') / f(\pars_k) \, q(\pars'\given\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
       otherwise $\pars_{k+1} \leftarrow \pars_k$.

### 2016-02-24 mcmc.tex
acceptance probability not fraction

@@
 \item Draw a random number $0<r<1$ from the uniform distribution.
-\item If $f(\pars') \, q(\pars_k\given\pars') / f(\pars_k) \, q(\pars'\given\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
+\item If $f(\pars') / f(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
       otherwise $\pars_{k+1} \leftarrow \pars_k$.
@@
   violate detailed balance in this way and has a compensated
-  acceptance fraction.  Read the paper \cite{emcee} for more details.}
+  acceptance probability.  Read the paper \cite{emcee} for more details.}
 The reason is:  It is one thing to draw samples from $q(x'\given x)$.
@@
   but you find it easier to run your sampler by proposing steps in $\ln \pars$,
-  if you don't modify your acceptance fraction by the Jacobian,
+  if you don't modify your acceptance probability by the Jacobian,
   your \emph{real} prior won't be flat in $\pars$.
@@
     for using \MCMC\ in a real context.}
+
+DFM says: should we say something about using the log density here?
 

### 2016-02-24 problems/p2.py

@@ -1,0 +1,29 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+from plot_setup import SQUARE_FIGSIZE, COLORS
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+np.random.seed(42)
+
+
+def log_p(x):
+    return -0.5 * (x - 2)**2 / 2.0  # Mean = 2, Variance = 2
+
+
+x = 0.0
+lp = log_p(x)
+chain = np.empty(2e4)
+for step in range(len(chain)):
+    x_prime = x + np.random.randn()
+    lp_prime = log_p(x_prime)
+    if np.random.rand() < np.exp(lp_prime - lp):
+        x = x_prime
+        lp = lp_prime
+    chain[step] = x
+
+

### 2016-02-24 problems/p1.py
adding problem 3

@@ -50,4 +50,4 @@ for i, (ax, title) in enumerate(zip(
 
 [ax.set_xlabel(r"$\log_2 K$") for ax in axes[1]]
 
-pl.savefig("p1.pdf")
+fig.savefig("p1.pdf")

### 2016-02-24 problems/p3.py

@@ -11,11 +11,13 @@ import matplotlib.pyplot as pl
 np.random.seed(42)
 
 
-def log_p(x, mean=2.0, variance=2.0):
-    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
+def log_p(x, a=3.0, b=7.0):
+    if a <= x <= b:
+        return -np.log(b-a)
+    return -np.inf
 
 
-x = 0.0
+x = 5.0
 lp = log_p(x)
 chain = np.empty(2e4)
 for step in range(len(chain)):
@@ -28,11 +30,14 @@ for step in range(len(chain)):
 
 
 fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
-ax.hist(chain, 100, histtype="step", color=COLORS["DATA"], normed=True)
-x = 2 + np.linspace(-4.5, 4.5, 5000)
-ax.plot(x, np.exp(log_p(x)), color=COLORS["MODEL_1"])
+ax.axhline(0.0, color="k")
+ax.hist(chain, 50, histtype="step", color=COLORS["DATA"], normed=True)
+x = np.linspace(2.5, 7.5, 5000)
+y = np.exp([log_p(_) for _ in x])
+ax.plot(x, y, color=COLORS["MODEL_1"])
 ax.set_xlim(x.min(), x.max())
+ax.set_ylim(-0.01, y.max() + 0.05)
 ax.set_xlabel("$x$")
 ax.set_ylabel("$p(x)$")
 
-fig.savefig("p2.pdf")
+fig.savefig("p3.pdf")

### 2016-02-24 problems/p1.py
Adding problems 4 and 5

@@ -3,10 +3,12 @@
 
 from __future__ import division, print_function
 
-from plot_setup import SQUARE_FIGSIZE, COLORS
 
 import numpy as np
 import matplotlib.pyplot as pl
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+
+setup()  # initialize the plotting styles
 
 np.random.seed(42)
 
@@ -47,6 +49,8 @@ for i, (ax, title) in enumerate(zip(
     d = np.max(np.abs(np.array(ax.get_ylim()) - mu))
     ax.set_ylim(mu + d * np.array([-1, 1]))
     ax.set_ylabel(title)
+    ax.yaxis.set_label_coords(-0.3, 0.5)
+    ax.xaxis.set_major_locator(pl.MaxNLocator(6))
 
 [ax.set_xlabel(r"$\log_2 K$") for ax in axes[1]]
 

### 2016-02-24 problems/p3.py

@@ -3,11 +3,12 @@
 
 from __future__ import division, print_function
 
-from plot_setup import SQUARE_FIGSIZE, COLORS
-
 import numpy as np
 import matplotlib.pyplot as pl
 
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+
+setup()  # initialize the plotting styles
 np.random.seed(42)
 
 

### 2016-02-24 problems/p4.py

@@ -1,0 +1,51 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import corner
+import numpy as np
+
+from plot_setup import setup
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p_gauss(x):
+    V = np.array([[2.0, 1.2], [1.2, 2.0]])
+    alpha = np.linalg.solve(V, x)
+    return -0.5 * np.dot(x, alpha)
+
+
+def log_p_uniform(x):
+    if 3 <= x[0] <= 7 and 1 <= x[1] <= 9:
+        return 0.0
+    return -np.inf
+
+
+def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
+    lp = log_p_gauss(x)
+    chain = np.empty((nsteps, len(x)))
+    for step in range(len(chain)):
+        x_prime = np.array(x)
+        x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
+        lp_prime = log_p(x_prime)
+        if np.random.rand() <= np.exp(lp_prime - lp):
+            x[:] = x_prime
+            lp = lp_prime
+        chain[step] = x
+    return chain
+
+
+chain = run_mcmc(log_p_gauss, np.array([0.0, 0.0]))
+fig = corner.corner(chain, labels=["$x$", "$y$"],
+                    range=[(-4.5, 4.5), (-4.5, 4.5)],
+                    plot_density=False, plot_contours=False)
+fig.savefig("p4a.pdf", dpi=300)
+
+chain = run_mcmc(log_p_uniform, np.array([5.0, 5.0]), nsteps=1e5)
+fig = corner.corner(chain, labels=["$x$", "$y$"],
+                    range=[(2.5, 7.5), (0.5, 9.5)],
+                    plot_density=False, plot_contours=False)
+fig.savefig("p4b.pdf", dpi=300)

### 2016-02-24 problems/p5.py

@@ -1,0 +1,37 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from p4 import run_mcmc, log_p_gauss
+from plot_setup import setup, COLORS, SQUARE_FIGSIZE
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+x0 = [-4.0, 5.0]
+
+s = SQUARE_FIGSIZE
+s[0] *= 3
+fig, axes = pl.subplots(1, 3, sharex=True, sharey=True, figsize=s)
+
+ax = axes[0]
+for n, ax in zip([0.0, -1.0, 2.0], axes):
+    chain = run_mcmc(log_p_gauss, np.array(x0), nsteps=2e3, prop_sigma=10**n)
+    ax.plot(chain[:, 0], chain[:, 1], "o-", color=COLORS["DATA"], ms=2)
+    ax.plot(x0[0], x0[1], "o", color=COLORS["MODEL_1"])
+    ax.set_xlim(-6.3, 6.3)
+    ax.set_ylim(-6.3, 6.3)
+    ax.set_xlabel("$x$")
+    ax.annotate(r"$\sigma = 10^{{{0:.0f}}}$".format(n),
+                (1, 0), xycoords="axes fraction",
+                xytext=(-5, 5), textcoords="offset points",
+                ha="right", va="bottom")
+
+axes[0].set_ylabel("$y$")
+
+fig.savefig("p5.pdf")

### 2016-02-24 problems/plot_setup.py

@@ -6,17 +6,7 @@ import numpy as np
 from cycler import cycler
 from matplotlib import rcParams
 
-__all__ = ["COLORS"]
-
-rcParams["font.size"] = 20
-rcParams["font.family"] = "serif"
-rcParams["font.serif"] = "Computer Sans"
-rcParams["text.usetex"] = True
-rcParams["figure.autolayout"] = True
-rcParams["axes.prop_cycle"] = cycler("color", (
-    "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
-    "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
-))  # d3.js color cycle
+__all__ = ["COLORS", "SQUARE_FIGSIZE", "setup"]
 
 COLORS = dict(
     DATA="k",
@@ -25,3 +15,15 @@ COLORS = dict(
 )
 
 SQUARE_FIGSIZE = np.array((4, 4))
+
+
+def setup():
+    rcParams["font.size"] = 20
+    rcParams["font.family"] = "serif"
+    rcParams["font.serif"] = "Computer Sans"
+    rcParams["text.usetex"] = True
+    rcParams["figure.autolayout"] = True
+    rcParams["axes.prop_cycle"] = cycler("color", (
+        "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
+        "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
+    ))  # d3.js color cycle

### 2016-02-24 problems/p4.py
adding problem 6

@@ -25,7 +25,7 @@ def log_p_uniform(x):
 
 
 def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
-    lp = log_p_gauss(x)
+    lp = log_p(x)
     chain = np.empty((nsteps, len(x)))
     for step in range(len(chain)):
         x_prime = np.array(x)

### 2016-02-24 problems/p5.py

@@ -18,8 +18,6 @@ x0 = [-4.0, 5.0]
 s = SQUARE_FIGSIZE
 s[0] *= 3
 fig, axes = pl.subplots(1, 3, sharex=True, sharey=True, figsize=s)
-
-ax = axes[0]
 for n, ax in zip([0.0, -1.0, 2.0], axes):
     chain = run_mcmc(log_p_gauss, np.array(x0), nsteps=2e3, prop_sigma=10**n)
     ax.plot(chain[:, 0], chain[:, 1], "o-", color=COLORS["DATA"], ms=2)
@@ -27,7 +25,7 @@ for n, ax in zip([0.0, -1.0, 2.0], axes):
     ax.set_xlim(-6.3, 6.3)
     ax.set_ylim(-6.3, 6.3)
     ax.set_xlabel("$x$")
-    ax.annotate(r"$\sigma = 10^{{{0:.0f}}}$".format(n),
+    ax.annotate(r"$\sigma_q = 10^{{{0:.0f}}}$".format(n),
                 (1, 0), xycoords="axes fraction",
                 xytext=(-5, 5), textcoords="offset points",
                 ha="right", va="bottom")

### 2016-02-24 problems/p6.py

@@ -1,0 +1,60 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from p4 import run_mcmc, log_p_gauss
+from plot_setup import setup, COLORS, SQUARE_FIGSIZE
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def run_broken_mcmc(log_p, x, prop_sigma=1.0, prop_mean=0.5, nsteps=2e3,
+                    broken=True):
+    lp = log_p(x)
+    chain = np.empty((nsteps, len(x)))
+    for step in range(len(chain)):
+        x_prime = np.array(x)
+        ind = np.random.randint(len(x))
+        x_prime[ind] += prop_mean + prop_sigma * np.random.randn()
+        lp_prime = log_p(x_prime)
+        if broken:
+            if np.random.rand() <= np.exp(lp_prime - lp):
+                x[:] = x_prime
+                lp = lp_prime
+        else:
+            r = lp_prime - lp
+            delta = x[ind] - x_prime[ind]
+            r += -0.5 * ((delta - 0.5)**2 - (delta + 0.5)**2)
+            if np.random.rand() <= np.exp(r):
+                x[:] = x_prime
+                lp = lp_prime
+        chain[step] = x
+    return chain
+
+
+x0 = [-4.0, 5.0]
+
+s = SQUARE_FIGSIZE
+s[0] *= 2
+fig, axes = pl.subplots(1, 2, sharex=True, sharey=True, figsize=s)
+for broken, ax in zip([True, False], axes):
+    chain = run_broken_mcmc(log_p_gauss, np.array(x0), broken=broken)
+    chain = np.concatenate(([x0], chain))
+    ax.plot(chain[:, 0], chain[:, 1], "o-", color=COLORS["DATA"], ms=2)
+    ax.plot(x0[0], x0[1], "o", color=COLORS["MODEL_1"])
+    ax.set_xlim(-6.3, 6.3)
+    ax.set_ylim(-6.3, 6.3)
+    ax.set_xlabel("$x$")
+    ax.annotate("incorrect acceptance" if broken else "corrected acceptance",
+                (1, 0), xycoords="axes fraction",
+                xytext=(-5, 5), textcoords="offset points",
+                ha="right", va="bottom")
+
+axes[0].set_ylabel("$y$")
+
+fig.savefig("p6.pdf")

### 2016-02-24 problems/p4a.py
problems Makefile

@@ -18,12 +18,6 @@ def log_p_gauss(x):
     return -0.5 * np.dot(x, alpha)
 
 
-def log_p_uniform(x):
-    if 3 <= x[0] <= 7 and 1 <= x[1] <= 9:
-        return 0.0
-    return -np.inf
-
-
 def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
     lp = log_p(x)
     chain = np.empty((nsteps, len(x)))
@@ -43,9 +37,3 @@ fig = corner.corner(chain, labels=["$x$", "$y$"],
                     range=[(-4.5, 4.5), (-4.5, 4.5)],
                     plot_density=False, plot_contours=False)
 fig.savefig("p4a.pdf", dpi=300)
-
-chain = run_mcmc(log_p_uniform, np.array([5.0, 5.0]), nsteps=1e5)
-fig = corner.corner(chain, labels=["$x$", "$y$"],
-                    range=[(2.5, 7.5), (0.5, 9.5)],
-                    plot_density=False, plot_contours=False)
-fig.savefig("p4b.pdf", dpi=300)

### 2016-02-24 problems/p4b.py

@@ -1,0 +1,26 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import corner
+import numpy as np
+
+from p4a import run_mcmc
+from plot_setup import setup
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p_uniform(x):
+    if 3 <= x[0] <= 7 and 1 <= x[1] <= 9:
+        return 0.0
+    return -np.inf
+
+
+chain = run_mcmc(log_p_uniform, np.array([5.0, 5.0]), nsteps=1e5)
+fig = corner.corner(chain, labels=["$x$", "$y$"],
+                    range=[(2.5, 7.5), (0.5, 9.5)],
+                    plot_density=False, plot_contours=False)
+fig.savefig("p4b.pdf", dpi=300)

### 2016-02-24 problems/p5.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from p4 import run_mcmc, log_p_gauss
+from p4a import run_mcmc, log_p_gauss
 from plot_setup import setup, COLORS, SQUARE_FIGSIZE
 
 setup()  # initialize the plotting styles

### 2016-02-24 problems/p6.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from p4 import run_mcmc, log_p_gauss
+from p4a import log_p_gauss
 from plot_setup import setup, COLORS, SQUARE_FIGSIZE
 
 setup()  # initialize the plotting styles

### 2016-03-17 problems/p7.py
adding problem 7

@@ -1,0 +1,53 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p(x, mean=2.0, variance=2.0):
+    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
+
+
+x = 0.0
+lp = log_p(x)
+chain = np.empty(2e4)
+for step in range(len(chain)):
+    x_prime = x + np.random.randn()
+    lp_prime = log_p(x_prime)
+    if np.random.rand() <= np.exp(lp_prime - lp):
+        x = x_prime
+        lp = lp_prime
+    chain[step] = x
+
+
+if __name__ == "__main__":
+    s = SQUARE_FIGSIZE
+    s[0] *= 2
+    fig, (ax1, ax2) = pl.subplots(1, 2, figsize=s)
+
+    ax1.plot(np.arange(len(chain))/1e3, chain, color=COLORS["DATA"],
+             rasterized=True)
+    ax1.set_xlim(0, len(chain) / 1e3)
+    ax1.set_ylim(2-5.5, 2+5.5)
+    ax1.set_ylabel("$x$")
+    ax1.set_xlabel("thousands of steps")
+
+    for i in range(4):
+        a = i*len(chain)/4
+        b = (i+1)*len(chain)/4
+        print(i+1, np.mean(chain[a:b]), np.var(chain[a:b]))
+        ax2.hist(chain[a:b], 50, histtype="step")
+    ax2.set_xlim(2-5.5, 2+5.5)
+    ax2.set_yticklabels([])
+    ax2.set_ylabel("$p(x)$")
+    ax2.set_xlabel("$x$")
+
+    fig.savefig("p7.pdf", dpi=300)

### 2016-03-17 problems/p8.py
adding problem 8

@@ -1,0 +1,41 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import SQUARE_FIGSIZE, COLORS
+
+from p7 import chain
+
+
+def acor_function_naive(x):
+    xmu = x - np.mean(x)
+    var = np.mean(xmu**2)
+    f = np.array([var] + [np.mean(xmu[:-delta] * xmu[delta:])
+                          for delta in range(1, len(x))]) / var
+    return f
+
+
+def acor_function_fft(x):
+    n = len(x)
+    f = np.fft.fft(x-np.mean(x), n=2*n)
+    acf = np.fft.ifft(f * np.conjugate(f))[:n].real
+    return acf / acf[0]
+
+
+if __name__ == "__main__":
+    fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+
+    acf1 = acor_function_naive(chain)
+    acf2 = acor_function_fft(chain)
+
+    ax.plot(acf1[:60])
+    ax.plot(acf2[:60])
+    ax.set_xlim(0, 60)
+    ax.set_ylabel("$C_x(\Delta)$")
+    ax.set_xlabel("$\Delta$")
+
+    fig.savefig("p8.pdf")

### 2016-03-18 problems/p7.py
Adding code for new problem 9

@@ -18,7 +18,7 @@ def log_p(x, mean=2.0, variance=2.0):
 
 x = 0.0
 lp = log_p(x)
-chain = np.empty(2e4)
+chain = np.empty(1e6)
 for step in range(len(chain)):
     x_prime = x + np.random.randn()
     lp_prime = log_p(x_prime)

### 2016-03-18 problems/p8.py

@@ -6,22 +6,14 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from plot_setup import SQUARE_FIGSIZE, COLORS
+from plot_setup import SQUARE_FIGSIZE
 
 from p7 import chain
 
 
-def acor_function_naive(x):
-    xmu = x - np.mean(x)
-    var = np.mean(xmu**2)
-    f = np.array([var] + [np.mean(xmu[:-delta] * xmu[delta:])
-                          for delta in range(1, len(x))]) / var
-    return f
-
-
-def acor_function_fft(x):
+def autocorr_function(x):
     n = len(x)
-    f = np.fft.fft(x-np.mean(x), n=2*n)
+    f = np.fft.fft(x - np.mean(x), n=2*n)
     acf = np.fft.ifft(f * np.conjugate(f))[:n].real
     return acf / acf[0]
 
@@ -29,11 +21,9 @@ def acor_function_fft(x):
 if __name__ == "__main__":
     fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
 
-    acf1 = acor_function_naive(chain)
-    acf2 = acor_function_fft(chain)
+    acf = autocorr_function(chain)
 
-    ax.plot(acf1[:60])
-    ax.plot(acf2[:60])
+    ax.plot(acf[:61])
     ax.set_xlim(0, 60)
     ax.set_ylabel("$C_x(\Delta)$")
     ax.set_xlabel("$\Delta$")

### 2016-03-18 problems/p9.py

@@ -1,0 +1,51 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import SQUARE_FIGSIZE, COLORS
+
+from p8 import chain, autocorr_function
+
+
+def autocorr_time_simple(acf, window):
+    return 1 + 2 * np.sum(acf[1:window])
+
+
+def autocorr_time_iterative(acf, c=10, low=10):
+    high = int(len(acf) / c)
+    for M in np.arange(low, high).astype(int):
+        tau = autocorr_time_simple(acf, M)
+        if M > c * tau:
+            return tau
+    raise RuntimeError("chain too short to estimate tau reliably")
+
+
+if __name__ == "__main__":
+    m = len(chain)
+    N = 2 ** np.arange(2, 19)
+
+    fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+
+    for s in [slice(0, m//2), slice(m//4, 3*m//4), slice(m//2, m),
+              slice(None)]:
+        acf = autocorr_function(chain[s])
+        taus = np.array([autocorr_time_simple(acf, n) for n in N])
+        tau = autocorr_time_iterative(acf)
+        ax.plot(N, taus, "-", color=COLORS["DATA"])
+        print("tau: {0}".format(tau))
+
+    ax.plot(N, taus, ".-", color=COLORS["MODEL_2"])
+    ax.axhline(tau, color=COLORS["MODEL_2"], ls="dashed")
+
+    ax.set_title(r"$\hat{{\tau}}_\mathrm{{int}} = {0:.2f}$".format(tau))
+    ax.set_ylim(0, 25)
+    ax.set_xlim(2, 5e5)
+    ax.set_xscale("log")
+    ax.set_ylabel(r"$\hat{\tau}_\mathrm{int}$")
+    ax.set_xlabel("window size")
+
+    fig.savefig("p9.pdf")

### 2016-03-18 mcmc.tex
note to self
words on convergence and a new problem 9

@@
 % - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
+% - Make sure we say ``don't use MCMC for 1-d problems or if you have alternatives to directly integrate!''
 % - Would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
@@
 \begin{eqnarray}
-\tau = 2\,\sum_{\Delta = 0}^{\infty} C(\Delta)
+\tau = 1 + 2\,\sum_{\Delta = 1}^{\infty} C(\Delta)
 \label{eq:autocorrelationtime}\quad .
 \end{eqnarray}
-%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING? OR A SUM FROM -infty to +infty?
 Heuristically, the autocorrelation time is the number of samples you
@@
 
-DFM: Here or in a note, can you give an empirical estimator of the
-autocorrelation time?  The naive estimator is good to give, and we
-could also give a good estimator too.
-
 That is, if the expectation value $E[q]$ is computed by the sampling
@@
 \begin{eqnarray}
-  \mbox{$E[q]$ uncertainty} &=& \left[\frac{\tau}{K}\,\Var[q]\right]^{1/2}
+  \Var[E[q]] &=& \frac{\tau}{K}\,\Var[q]
   \\
-  &\approx& \left[\frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2\right]^{1/2}
+  &\approx& \frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2
   \quad ,
@@
 \emph{estimator} for the time.
-Usually, we just take the naive estimator---the ratio of empirical
-expectations implied by \equationname~(\ref{eq:autocorrelationfn}),
-and then perform a naive sum to get the autocorrelation time.
-However, almost all empirical estimators of the autocorrelation time
-will end up being under-estimates (biased low), because the finite
-length of the chain makes any subset seem more representative than it
-truly is; indeed this is a sub-problem of the problem that convergence
-can never be known empirically with good confidence.
+We might first try the na\"ive estimator
+\begin{eqnarray}\label{eq:tauestnaive}
+\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
+\end{eqnarray}
+where the sum is over the autocorrelation function estimated based the chain
+using \equationname~(\ref{eq:autocorrelationfn}).
+In practice, this is a high variance estimator of the integrated
+autocorrelation time and this variance doesn't go to zero for an infinitely
+long chain.\note{%
+  A.~Sokal has a clear discussion of this topic and several proposals for
+  robust autocorrelation time estimation in a set of lecture notes
+  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
+Instead, we can get a much more robust estimate of the autocorrelation time by
+restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
+where
+$\tau \le M \ll N$
+\begin{eqnarray}\label{eq:tauest}
+\tau &=& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
+\end{eqnarray}
+In practice, $M$ can be chosen iteratively to satisfy these constraints but it
+will only be as reliable as the chain itself.
+An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
+trusted if the chain is many times ($\sim 10$) longer than $\tau$.
 
@@
 Write a piece of code that computes the empirical autocorrelation
-function and the empirical autocorrelation time from a chain using the na\"ive method DFM:WHERE?  Run
-this on the chain you obtained from \problemname~\ref{prob:MH}.
-Plot the autocorrelation function you get and over-plot the
-autocorrelation time as an indicator on the plot.  What do you
-conclude?
+function.
+You can speed this computation up considerably by using a fast Fourier
+transform.
+Run this on the chain you obtained from \problemname~\ref{prob:MH}.
+Plot the autocorrelation function you find at short lags ($\Delta < 100$).
 \end{problem}
@@
 \begin{problem}
-Look at the source code of \project{emcee} \citep{emcee}, and see if you can
-re-code or extract the autocorrelation-time estimator called DFM:WHAT?
-This code should take a \MCMC\ chain of parameter vectors as input and
-return a vector of autocorrelation times as output.
+Write a peice of code that estiamtes the integrated autocorrelation time for a
+chain of samples using an estimate of the autocorrelation function and a given
+``window'' size $M$ (see \equationname~\ref{eq:tauest}).
+Plot the estimated $\tau$ as a function of $M$.
+What can you conclude from this plot?
+Implement an iterative procedure for automatically choosing $M$.\note{%
+The recipe given on page 16 of A.~Sokal's notes
+(\url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}) might be helpful.}
+Overplot this estimate on the plot of $\tau(M)$.
 \end{problem}

### 2016-03-22 mcmc.tex
more work for the problem
a note

@@
 ``window'' size $M$ (see \equationname~\ref{eq:tauest}).
-Plot the estimated $\tau$ as a function of $M$.
+Plot the estimated $\tau$ as a function of $M$ for several contiguous segments
+of the chain and overplot the sample function based on the full chain.
 What can you conclude from this plot?
@@
 The recipe given on page 16 of A.~Sokal's notes
-(\url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}) might be helpful.}
+(\url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}) might be helpful.
+Note that the definition of $\tau$ in that document is half of the definition
+that we use.}
 Overplot this estimate on the plot of $\tau(M)$.

### 2016-03-23 mcmc.tex
words about convergence and tuning

@@
 V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
-\quad .
+\quad.
 \end{eqnarray}
@@
 \begin{eqnarray}
-C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
+C_f(\Delta) &\equiv& \frac{E[(f(\pars_k)-\mu_f)\,(f(\pars_{k+\Delta})-\mu_f)]}
+                          {E[(f(\pars)-\mu_f)^2]}
 \label{eq:autocorrelationfn}\\
-\mu &=& E[\pars]
+\mu_f &=& E[f(\pars)]
 \quad ;
 \end{eqnarray}
-that is, it is the covariance of points separated by $\Delta$, taken
-relative to the variance.
+where $f(\pars)$ is a scalar valued function of the parameters.
+In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
+of points in the chain separated by $\Delta$ steps, taken relative to the
+variance.
 In the above equation, the assumption is that the samples $\pars_k$ are
@@
 \begin{eqnarray}
-C(0) &=& 1
+C_f(0) &=& 1
 \\
-\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
+\lim_{\Delta\rightarrow\infty} C_f(\Delta) &=& 0
 \quad ,
@@
 \begin{eqnarray}
-\tau = 1 + 2\,\sum_{\Delta = 1}^{\infty} C(\Delta)
-\label{eq:autocorrelationtime}\quad .
+\tau_f = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_f(\Delta)
+\label{eq:autocorrelationtime}\quad.
 \end{eqnarray}
@@
 \begin{eqnarray}
-  \Var[E[q]] &=& \frac{\tau}{K}\,\Var[q]
+  \Var[E_{p(\pars)}[q(\pars)]] &=& \frac{\tau_q}{K}\,\Var_{p(\pars)}[q(\pars)]
   \\
-  &\approx& \frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2
-  \quad ,
+  &\approx& \frac{\tau_q}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) -
+    E_{p(\pars)}[q(\pars)]\right)^2
+  \quad,
+\label{eq:tauvar}
 \end{eqnarray}
 where we are being a bit loose with terminology because this is the
-uncertainty not on $E[q]$ but on our sampling-based \emph{estimator} for
-$E[q]$, and the $E[q]$ inside the sum is that estimator too.
+variance not on $E_{p(\pars)}[q(\pars)]$ but on our sampling-based
+\emph{estimator} for
+$E_{p(\pars)}[q(\pars)]$, and the $E_{p(\pars)}[q(\pars)]$ inside the sum is
+that estimator too.
 It is slightly off-topic here, but related to this, if you have a long
@@
 had used all $K$ samples.
+
+It is also worth noting that there is not just one autocorrelation function or
+autocorrelation time for an MCMC chain.
+Instead, there is an autocorrelation time $\tau_f$ for every function
+$f(\pars)$ and these will, in general, be different.
+Therefore, the autocorrelation time used in \equationname~(\ref{eq:tauvar}) is
+the time computed on the chain of $q(\pars)$ values instead of the
+autocorrelation time for $\pars$.
 
@@
 known mean and variance).
-In general, the best sampler will be different for different problems,
-and we will (below; \sectionname~\ref{sec:methods}) tune the samplers we have to do the best on the
-problems we have; this tuning will also be problem-specific.
+In general, the best sampler will be different for different problems, and we
+will (below; \sectionname~\ref{sec:methods}) tune the samplers we have to do
+the best on the problems we have; this tuning will also be problem-specific.
 There are many heuristic bases on which different samplers might be
@@
 
-Most \MCMC\ methods
-  (though not all\note{We are promoters of an ensemble method that has a
-  nearly-fixed proposal distribution and requires no tuning \citep{emcee}.
-  There are many problems for which this is \emph{not} the best choice
-  of sampler, but since it obviates tuning, it can be a good first choice.})
-  make use of something like a ``proposal distribution''
+Most \MCMC\ methods make use of something like a ``proposal distribution''
   which determines what kinds of steps the walker can take
@@
   there are various tricks and tips to work through.
-One is to see if any of the parameters has a conditional posterior
-  probability that is guaranteed to be Gaussian
-  (this is true for some linear parameters in least-square problems
-  when the priors are Gaussian, for example).
-These dimensions can often be sampled \emph{exactly},
-  outside of the MCCM framework;
-  the algorithm looks like:
-Take an \MCMC\ step in the hard parameters,
-  then take an exact step in the easy parameters,
-  repeat.
-Another related trick is to analytically marginalize,
-  rather than sample,
-  easy nuisance parameters.
+% One is to see if any of the parameters has a conditional posterior
+%   probability that is guaranteed to be Gaussian
+%   (this is true for some linear parameters in least-square problems
+%   when the priors are Gaussian, for example).
+% These dimensions can often be sampled \emph{exactly},
+%   outside of the \MCMC\ framework;
+%   the algorithm looks like:
+% Take an \MCMC\ step in the hard parameters,
+%   then take an exact step in the easy parameters,
+%   repeat.
+% Another related trick is to analytically marginalize,
+%   rather than sample,
+%   easy nuisance parameters.
 The huge bag of possible tricks is so large,
   it goes way beyond the scope of this introductory \documentname.
-In our experience, it is valuable to make friends with at least one statistician,
-  one applied mathematician, and one computer scientist.
-Between them, they ought to span the literature.
+In our experience, it is valuable to make friends with at least one
+  statistician, one applied mathematician, and one computer scientist.
+Between them, they ought to span the relevant literature.
 

### 2016-03-23 problems/p10.py
adding problem 10

@@ -1,0 +1,49 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p_gauss(x):
+    V = np.array([[2.0, 1.2], [1.2, 2.0]])
+    alpha = np.linalg.solve(V, x)
+    return -0.5 * np.dot(x, alpha)
+
+
+def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
+    lp = log_p(x)
+    chain = np.empty((nsteps, len(x)))
+    acc = 0
+    for step in range(len(chain)):
+        x_prime = np.array(x)
+        x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
+        lp_prime = log_p(x_prime)
+        if np.random.rand() <= np.exp(lp_prime - lp):
+            acc += 1
+            x[:] = x_prime
+            lp = lp_prime
+        chain[step] = x
+    return chain, acc / len(chain)
+
+
+sigs = 2.0 ** np.arange(-2, 5)
+acc_fracs = np.empty_like(sigs)
+for i, sig in enumerate(sigs):
+    _, acc_fracs[i] = run_mcmc(log_p_gauss, np.array([0.0, 0.0]),
+                               prop_sigma=sig)
+
+fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+ax.axhline(0.25, color=COLORS["DATA"])
+ax.plot(sigs, acc_fracs, ".-", color=COLORS["MODEL_2"])
+ax.set_xscale("log")
+ax.set_xlabel(r"$\sigma_q$")
+ax.set_ylabel("acceptance fraction")
+ax.set_xlim(0.2, 20)
+fig.savefig("p10.pdf")

### 2016-03-23 problems/p9.py

@@ -41,11 +41,11 @@ if __name__ == "__main__":
     ax.plot(N, taus, ".-", color=COLORS["MODEL_2"])
     ax.axhline(tau, color=COLORS["MODEL_2"], ls="dashed")
 
-    ax.set_title(r"$\hat{{\tau}}_\mathrm{{int}} = {0:.2f}$".format(tau))
+    ax.set_title(r"$\tau_x = {0:.2f}$".format(tau))
     ax.set_ylim(0, 25)
     ax.set_xlim(2, 5e5)
     ax.set_xscale("log")
-    ax.set_ylabel(r"$\hat{\tau}_\mathrm{int}$")
+    ax.set_ylabel(r"$\tau_x$")
     ax.set_xlabel("window size")
 
     fig.savefig("p9.pdf")

### 2016-05-18 problems/p1.py
switching plots to sans #bringhurst

@@ -50,6 +50,7 @@ for i, (ax, title) in enumerate(zip(
     ax.set_ylim(mu + d * np.array([-1, 1]))
     ax.set_ylabel(title)
     ax.yaxis.set_label_coords(-0.3, 0.5)
+    ax.yaxis.set_major_locator(pl.MaxNLocator(5))
     ax.xaxis.set_major_locator(pl.MaxNLocator(6))
 
 [ax.set_xlabel(r"$\log_2 K$") for ax in axes[1]]

### 2016-05-18 problems/p11.py

@@ -1,0 +1,53 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+
+from p9 import autocorr_time_iterative
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p_gauss(x):
+    V = np.array([[2.0, 1.2], [1.2, 2.0]])
+    alpha = np.linalg.solve(V, x)
+    return -0.5 * np.dot(x, alpha)
+
+
+def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=5e6):
+    lp = log_p(x)
+    chain = np.empty((nsteps, len(x)))
+    acc = 0
+    for step in range(len(chain)):
+        x_prime = np.array(x)
+        x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
+        lp_prime = log_p(x_prime)
+        if np.random.rand() <= np.exp(lp_prime - lp):
+            acc += 1
+            x[:] = x_prime
+            lp = lp_prime
+        chain[step] = x
+    return chain, acc / len(chain)
+
+
+sigs = 2.0 ** np.arange(-2, 5)
+taus = np.empty((len(sigs), 2))
+for i, sig in enumerate(sigs):
+    chain, _ = run_mcmc(log_p_gauss, np.array([0.0, 0.0]), prop_sigma=sig)
+    taus[i, 0] = autocorr_time_iterative(chain[:, 0])
+    taus[i, 1] = autocorr_time_iterative(chain[:, 1])
+    print(taus[i])
+
+fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+ax.plot(sigs, taus[:, 0], ".-", color=COLORS["MODEL_1"])
+ax.plot(sigs, taus[:, 1], ".-", color=COLORS["MODEL_2"])
+ax.set_xscale("log")
+ax.set_xlabel(r"$\sigma_q$")
+ax.set_ylabel(r"$\tau_x$")
+ax.set_xlim(0.2, 20)
+fig.savefig("p11.pdf")

### 2016-05-18 problems/p2.py

@@ -33,6 +33,7 @@ ax.hist(chain, 100, histtype="step", color=COLORS["DATA"], normed=True)
 x = 2 + np.linspace(-4.5, 4.5, 5000)
 ax.plot(x, np.exp(log_p(x)), color=COLORS["MODEL_1"])
 ax.set_xlim(x.min(), x.max())
+ax.yaxis.set_major_locator(pl.MaxNLocator(4))
 ax.set_xlabel("$x$")
 ax.set_ylabel("$p(x)$")
 

### 2016-05-18 problems/p9.py

@@ -8,7 +8,7 @@ import matplotlib.pyplot as pl
 
 from plot_setup import SQUARE_FIGSIZE, COLORS
 
-from p8 import chain, autocorr_function
+from p8 import autocorr_function
 
 
 def autocorr_time_simple(acf, window):
@@ -19,12 +19,14 @@ def autocorr_time_iterative(acf, c=10, low=10):
     high = int(len(acf) / c)
     for M in np.arange(low, high).astype(int):
         tau = autocorr_time_simple(acf, M)
-        if M > c * tau:
+        if tau > 1.0 and M > c * tau:
             return tau
     raise RuntimeError("chain too short to estimate tau reliably")
 
 
 if __name__ == "__main__":
+    from p8 import chain
+
     m = len(chain)
     N = 2 ** np.arange(2, 19)
 

### 2016-05-18 problems/plot_setup.py

@@ -18,10 +18,14 @@ SQUARE_FIGSIZE = np.array((4, 4))
 
 
 def setup():
-    rcParams["font.size"] = 20
-    rcParams["font.family"] = "serif"
-    rcParams["font.serif"] = "Computer Sans"
+    rcParams["agg.path.chunksize"] = 10000  # huh?
+    rcParams["font.size"] = 16
+    rcParams["font.family"] = "sans-serif"
+    rcParams["font.sans-serif"] = ["Computer Modern Sans"]
+    # rcParams["font.family"] = "serif"
+    # rcParams["font.serif"] = ["Computer Modern Roman"]
     rcParams["text.usetex"] = True
+    rcParams["text.latex.preamble"] = r"\usepackage{cmbright}"
     rcParams["figure.autolayout"] = True
     rcParams["axes.prop_cycle"] = cycler("color", (
         "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",

### 2016-05-18 mcmc.tex
makefile builds figures; nicer figures

@@
 Plot the results as a histogram, with the true density over-plotted sensibly.
+The resulting  plot should look something like \figurename~\ref{fig:MH}.
 \end{problem}
 
-\begin{problem}
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/p2}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH}:
+MCMC samples from a Gaussian (black) and the true distribution (blue).}
+\label{fig:MH}
+\end{figure}
+
+\begin{problem}\label{prob:MH2}
 Re-do \problemname~\ref{prob:MH} but now with an input density
 that is uniform on $3<x<7$ and zero everywhere else.
+The plot should look like \figurename~\ref{fig:MH2}.
 What change did you have to make to the initialization, and why?
 \end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/p3}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH2}:
+MCMC samples from a uniform distribution (black) and the true distribution
+(blue).}
+\label{fig:MH2}
+\end{figure}
 
@@
 For the density function use two different functions.
-The first density function is a covariant two-dimensional Gaussian density with variance tensor
+\emph{(a)} The first density function is a covariant two-dimensional Gaussian
+density with variance tensor
 \begin{eqnarray}
@@
 \end{eqnarray}
-The second density function is a rectangular top-hat function that is uniform on
-the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
+\emph{(b)} The second density function is a rectangular top-hat function that
+is uniform on the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
 For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
@@
 
-\begin{problem}
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{problems/p4a}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:twod}a:
+MCMC samples from a two-dimensional Gaussian (scatter plot) and the
+one-dimensional marginalized distributions (histograms).}
+\label{fig:twod}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{problems/p4b}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:twod}b:
+MCMC samples from a two-dimensional top hat distribution (scatter plot) and the
+one-dimensional marginalized distributions (histograms).}
+\label{fig:twod}
+\end{figure}
+
+\begin{problem}\label{prob:MH-sigma}
 Re-do \problemname~\ref{prob:MH} but with different values for
@@
 \end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{problems/p5}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH-sigma}:
+MCMC samples from a two-dimensional top hat distribution (scatter plot) and the
+one-dimensional marginalized distributions (histograms).}
+\label{fig:MH-sigma}
+\end{figure}
 

### 2016-05-18 problems/p1.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 
 import numpy as np
 import matplotlib.pyplot as pl
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
 setup()  # initialize the plotting styles
 
@@ -39,7 +39,8 @@ def compute_stats(K, a=0.0, b=1.0):
 K = 4**np.arange(1, 11)
 a_stats, s_stats = map(np.array, zip(*(compute_stats(k) for k in K)))
 
-fig, axes = pl.subplots(2, 2, sharex=True, figsize=2*SQUARE_FIGSIZE)
+shape = np.array([2, 1.5])*SQUARE_FIGSIZE
+fig, axes = pl.subplots(2, 2, sharex=True, figsize=shape)
 
 for i, (ax, title) in enumerate(zip(
         axes.flat, ("mean", "variance", "skewness", "kurtosis"))):
@@ -55,4 +56,4 @@ for i, (ax, title) in enumerate(zip(
 
 [ax.set_xlabel(r"$\log_2 K$") for ax in axes[1]]
 
-fig.savefig("p1.pdf")
+savefig(fig, "p1.pdf")

### 2016-05-18 problems/p11.py

@@ -19,7 +19,7 @@ def log_p_gauss(x):
     return -0.5 * np.dot(x, alpha)
 
 
-def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=5e6):
+def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=1e5):
     lp = log_p(x)
     chain = np.empty((nsteps, len(x)))
     acc = 0

### 2016-05-18 problems/p2.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -37,4 +37,4 @@ ax.yaxis.set_major_locator(pl.MaxNLocator(4))
 ax.set_xlabel("$x$")
 ax.set_ylabel("$p(x)$")
 
-fig.savefig("p2.pdf")
+savefig(fig, "p2.pdf")

### 2016-05-18 problems/plot_setup.py

@@ -31,3 +31,8 @@ def setup():
         "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
         "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
     ))  # d3.js color cycle
+
+
+def savefig(fig, fn, **kwargs):
+    kwargs["bbox_inches"] = "tight"
+    fig.savefig(fn, **kwargs)

### 2016-05-18 mcmc.tex
fixing problem 11

@@
 Daniel~Foreman-Mackey\\
+\affil{Sagan Fellow}\\
 \affil{Department~of~Astronomy, University~of~Washington}
@@
 plot for each sampling.
+\figurename~\ref{fig:twod-a} shows the expected results for the Gaussian.
+Make a similar plot for the top-hat.
 \end{problem}
@@
 one-dimensional marginalized distributions (histograms).}
-\label{fig:twod}
+\label{fig:twod-a}
 \end{figure}
 
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{problems/p4b}
-\end{center}
-\caption{Solution to \problemname~\ref{prob:twod}b:
-MCMC samples from a two-dimensional top hat distribution (scatter plot) and the
-one-dimensional marginalized distributions (histograms).}
-\label{fig:twod}
-\end{figure}
-
 \begin{problem}\label{prob:MH-sigma}
-Re-do \problemname~\ref{prob:MH} but with different values for
+Re-do \problemname~\ref{prob:twod}a but with different values for
 the variance of the proposal distribution $q(x'\given x)$.
 What happens when you go to very extreme values (like for instance
-$10^{-8}$ or $10^6$)?
+$10^{-1}$ or $10^2$)?
 \end{problem}
 
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{problems/p5}
-\end{center}
-\caption{Solution to \problemname~\ref{prob:MH-sigma}:
-MCMC samples from a two-dimensional top hat distribution (scatter plot) and the
-one-dimensional marginalized distributions (histograms).}
-\label{fig:MH-sigma}
-\end{figure}
-
 \begin{problem}
-Why, in \problemname~\ref{prob:MH}, did we give the proposal
-distribution $q(x'\given x)$ zero mean?
+Why, in all the previous problems, did we give the proposal
+distributions $q(x'\given x)$ zero mean?
 What would be bad if we hadn't done that?
-Try setting the mean of $q(x'\given x)$ to something non-zero and see
-what happens.
+Re-do \problemname~\ref{prob:twod}a with a proposal $q(x'\given x)$ with a
+non-zero mean and see what happens.
 Bonus points: Modify the acceptance--rejection criterion to deal with
@@
 
-\begin{problem}
+\begin{problem}\label{prob:convergence}
 Re-do \problemname~\ref{prob:MH} but now look at convergence:
@@
 Plot the autocorrelation function you find at short lags ($\Delta < 100$).
+This plot should resemble \figurename~\ref{fig:estimatetau}.
 \end{problem}
 
-\begin{problem}
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/p8}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:estimatetau}:
+The autocorrelation function of the chain from \problemname~\ref{prob:MH}.}
+\label{fig:estimatetau}
+\end{figure}
+
+\begin{problem}\label{prob:itertau}
 Write a peice of code that estiamtes the integrated autocorrelation time for a
@@
 that we use.}
-Overplot this estimate on the plot of $\tau(M)$.
+Overplot this estimate on the plot of $\tau(M)$ and the result should look
+like \figurename~\ref{fig:itertau}.
 \end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/p9}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:itertau}:
+Estimates of the integrated autocorrelation time of different segments of the
+MCMC chain from \problemname~\ref{prob:MH} (black lines) and for the full
+chain (orange line) as a function of window size $M$.
+The ``optimized'' value computed using an iterative procedure is overplotted
+as a dashed line and its value is listed in the title.}
+\label{fig:itertau}
+\end{figure}
 

### 2016-05-18 problems/p10.py

@@ -5,45 +5,27 @@ from __future__ import division, print_function
 
 import numpy as np
 import matplotlib.pyplot as pl
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
+
+from p4a import log_p_gauss, run_mcmc
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
 
 
-def log_p_gauss(x):
-    V = np.array([[2.0, 1.2], [1.2, 2.0]])
-    alpha = np.linalg.solve(V, x)
-    return -0.5 * np.dot(x, alpha)
+sigs = 2.0 ** np.arange(-2, 6)
 
+if __name__ == "__main__":
+    acc_fracs = np.empty_like(sigs)
+    for i, sig in enumerate(sigs):
+        _, acc_fracs[i] = run_mcmc(log_p_gauss, np.array([0.0, 0.0]),
+                                   prop_sigma=sig)
 
-def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
-    lp = log_p(x)
-    chain = np.empty((nsteps, len(x)))
-    acc = 0
-    for step in range(len(chain)):
-        x_prime = np.array(x)
-        x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
-        lp_prime = log_p(x_prime)
-        if np.random.rand() <= np.exp(lp_prime - lp):
-            acc += 1
-            x[:] = x_prime
-            lp = lp_prime
-        chain[step] = x
-    return chain, acc / len(chain)
-
-
-sigs = 2.0 ** np.arange(-2, 5)
-acc_fracs = np.empty_like(sigs)
-for i, sig in enumerate(sigs):
-    _, acc_fracs[i] = run_mcmc(log_p_gauss, np.array([0.0, 0.0]),
-                               prop_sigma=sig)
-
-fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
-ax.axhline(0.25, color=COLORS["DATA"])
-ax.plot(sigs, acc_fracs, ".-", color=COLORS["MODEL_2"])
-ax.set_xscale("log")
-ax.set_xlabel(r"$\sigma_q$")
-ax.set_ylabel("acceptance fraction")
-ax.set_xlim(0.2, 20)
[…]

### 2016-05-18 problems/p11.py

@@ -5,43 +5,23 @@ from __future__ import division, print_function
 
 import numpy as np
 import matplotlib.pyplot as pl
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
-from p9 import autocorr_time_iterative
+from p9 import autocorr_function, autocorr_time_iterative
+from p10 import log_p_gauss, run_mcmc, sigs
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
 
 
-def log_p_gauss(x):
-    V = np.array([[2.0, 1.2], [1.2, 2.0]])
-    alpha = np.linalg.solve(V, x)
-    return -0.5 * np.dot(x, alpha)
-
-
-def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=1e5):
-    lp = log_p(x)
-    chain = np.empty((nsteps, len(x)))
-    acc = 0
-    for step in range(len(chain)):
-        x_prime = np.array(x)
-        x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
-        lp_prime = log_p(x_prime)
-        if np.random.rand() <= np.exp(lp_prime - lp):
-            acc += 1
-            x[:] = x_prime
-            lp = lp_prime
-        chain[step] = x
-    return chain, acc / len(chain)
-
-
-sigs = 2.0 ** np.arange(-2, 5)
 taus = np.empty((len(sigs), 2))
 for i, sig in enumerate(sigs):
-    chain, _ = run_mcmc(log_p_gauss, np.array([0.0, 0.0]), prop_sigma=sig)
-    taus[i, 0] = autocorr_time_iterative(chain[:, 0])
-    taus[i, 1] = autocorr_time_iterative(chain[:, 1])
-    print(taus[i])
+    chain, acc = run_mcmc(log_p_gauss, np.array([0.0, 0.0]), prop_sigma=sig,
+                          nsteps=1e6)
+    for j in range(2):
+        acf = autocorr_function(chain[:, j])
+        taus[i, j] = autocorr_time_iterative(acf)
+    print(sig, acc, taus[i])
 
 fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
[…]

### 2016-05-18 problems/p3.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -41,4 +41,4 @@ ax.set_ylim(-0.01, y.max() + 0.05)
 ax.set_xlabel("$x$")
 ax.set_ylabel("$p(x)$")
 
-fig.savefig("p3.pdf")
+savefig(fig, "p3.pdf")

### 2016-05-18 problems/p4a.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import corner
 import numpy as np
 
-from plot_setup import setup
+from plot_setup import setup, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -21,19 +21,22 @@ def log_p_gauss(x):
 def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
     lp = log_p(x)
     chain = np.empty((nsteps, len(x)))
+    acc = 0
     for step in range(len(chain)):
         x_prime = np.array(x)
         x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
         lp_prime = log_p(x_prime)
         if np.random.rand() <= np.exp(lp_prime - lp):
+            acc += 1
             x[:] = x_prime
             lp = lp_prime
         chain[step] = x
-    return chain
+    return chain, acc / len(chain)
 
 
-chain = run_mcmc(log_p_gauss, np.array([0.0, 0.0]))
-fig = corner.corner(chain, labels=["$x$", "$y$"],
-                    range=[(-4.5, 4.5), (-4.5, 4.5)],
-                    plot_density=False, plot_contours=False)
-fig.savefig("p4a.pdf", dpi=300)
+if __name__ == "__main__":
+    chain, _ = run_mcmc(log_p_gauss, np.array([0.0, 0.0]))
+    fig = corner.corner(chain, labels=["$x$", "$y$"],
+                        range=[(-4.5, 4.5), (-4.5, 4.5)],
+                        plot_density=False, plot_contours=False)
+    savefig(fig, "p4a.pdf", dpi=300)

### 2016-05-18 problems/p4b.py

@@ -7,7 +7,7 @@ import corner
 import numpy as np
 
 from p4a import run_mcmc
-from plot_setup import setup
+from plot_setup import setup, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -23,4 +23,4 @@ chain = run_mcmc(log_p_uniform, np.array([5.0, 5.0]), nsteps=1e5)
 fig = corner.corner(chain, labels=["$x$", "$y$"],
                     range=[(2.5, 7.5), (0.5, 9.5)],
                     plot_density=False, plot_contours=False)
-fig.savefig("p4b.pdf", dpi=300)
+savefig(fig, "p4b.pdf", dpi=300)

### 2016-05-18 problems/p5.py

@@ -7,7 +7,7 @@ import numpy as np
 import matplotlib.pyplot as pl
 
 from p4a import run_mcmc, log_p_gauss
-from plot_setup import setup, COLORS, SQUARE_FIGSIZE
+from plot_setup import setup, COLORS, SQUARE_FIGSIZE, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -16,7 +16,8 @@ np.random.seed(42)
 x0 = [-4.0, 5.0]
 
 s = SQUARE_FIGSIZE
-s[0] *= 3
+s[0] *= 2
+s[1] *= 0.8
 fig, axes = pl.subplots(1, 3, sharex=True, sharey=True, figsize=s)
 for n, ax in zip([0.0, -1.0, 2.0], axes):
     chain = run_mcmc(log_p_gauss, np.array(x0), nsteps=2e3, prop_sigma=10**n)
@@ -29,7 +30,9 @@ for n, ax in zip([0.0, -1.0, 2.0], axes):
                 (1, 0), xycoords="axes fraction",
                 xytext=(-5, 5), textcoords="offset points",
                 ha="right", va="bottom")
+    ax.yaxis.set_major_locator(pl.MaxNLocator(5))
+    ax.xaxis.set_major_locator(pl.MaxNLocator(5))
 
 axes[0].set_ylabel("$y$")
 
-fig.savefig("p5.pdf")
+savefig(fig, "p5.pdf")

### 2016-05-18 problems/p6.py

@@ -7,7 +7,7 @@ import numpy as np
 import matplotlib.pyplot as pl
 
 from p4a import log_p_gauss
-from plot_setup import setup, COLORS, SQUARE_FIGSIZE
+from plot_setup import setup, COLORS, SQUARE_FIGSIZE, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -57,4 +57,4 @@ for broken, ax in zip([True, False], axes):
 
 axes[0].set_ylabel("$y$")
 
-fig.savefig("p6.pdf")
+savefig(fig, "p6.pdf")

### 2016-05-18 problems/p7.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -50,4 +50,4 @@ if __name__ == "__main__":
     ax2.set_ylabel("$p(x)$")
     ax2.set_xlabel("$x$")
 
-    fig.savefig("p7.pdf", dpi=300)
+    savefig(fig, "p7.pdf", dpi=300)

### 2016-05-18 problems/p8.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from plot_setup import SQUARE_FIGSIZE
+from plot_setup import SQUARE_FIGSIZE, savefig
 
 from p7 import chain
 
@@ -28,4 +28,4 @@ if __name__ == "__main__":
     ax.set_ylabel("$C_x(\Delta)$")
     ax.set_xlabel("$\Delta$")
 
-    fig.savefig("p8.pdf")
+    savefig(fig, "p8.pdf")

### 2016-05-18 problems/p9.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from plot_setup import SQUARE_FIGSIZE, COLORS
+from plot_setup import SQUARE_FIGSIZE, COLORS, savefig
 
 from p8 import autocorr_function
 
@@ -50,4 +50,4 @@ if __name__ == "__main__":
     ax.set_ylabel(r"$\tau_x$")
     ax.set_xlabel("window size")
 
-    fig.savefig("p9.pdf")
+    savefig(fig, "p9.pdf")

### 2016-05-19 mcmc.tex
a few more figures

@@
 
-\begin{problem}
+\begin{problem}\label{prob:tuningtau}
 Re-do \problemname~\ref{prob:tuning} but instead of trying to reach a
@@
 
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.48\textwidth]{problems/p10}
+\includegraphics[width=0.48\textwidth]{problems/p11}
+\end{center}
+\caption{Solutions to \problemname s~\ref{prob:tuning} and~\ref{prob:tuningtau}:
+Two methods of tuning the M--H proposal parameters.
+\emph{left:} The acceptance fraction as a function of proposal scale for the
+distribution from \problemname~\ref{prob:twod}a.
+\emph{right:} The integrated autocorrelation time for each parameter
+(indicated by the different colors) as a function of proposal scale parameter.}
+\label{fig:tuning}
+\end{figure}
+
 \begin{problem}
@@
 Look up the Rosenbrock density used as the demonstration case for the
-\project{emcee} sampler (see \citealt{emcee}).
+\project{emcee} sampler (see DFM: should cite G\&W here).
 Tune the Gaussian proposal distribution in your \MH\ \MCMC\ sampler to sample this density efficiently.
@@
   \isbn{978-1-4200-7941-8}.
-\bibitem[Foreman-Mackey \etal(2012)]{emcee}
-  Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2012,
+\bibitem[Foreman-Mackey \etal(2013)]{emcee}
+  Foreman-Mackey,~D., Hogg,~D.~W., Lang,~D., \& Goodman,~J.\ 2013,
   \project{emcee}: The \MCMC\ Hammer,

### 2016-05-19 problems/p11.py

@@ -17,7 +17,7 @@ np.random.seed(42)
 taus = np.empty((len(sigs), 2))
 for i, sig in enumerate(sigs):
     chain, acc = run_mcmc(log_p_gauss, np.array([0.0, 0.0]), prop_sigma=sig,
-                          nsteps=1e6)
+                          nsteps=2e5)
     for j in range(2):
         acf = autocorr_function(chain[:, j])
         taus[i, j] = autocorr_time_iterative(acf)
@@ -30,4 +30,6 @@ ax.set_xscale("log")
 ax.set_xlabel(r"$\sigma_q$")
 ax.set_ylabel(r"$\tau_x$")
 ax.set_xlim(0.2, 40)
+ax.set_ylim(0, 450)
+ax.yaxis.set_major_locator(pl.MaxNLocator(5))
 savefig(fig, "p11.pdf")

### 2016-05-19 problems/p4a.py

@@ -18,20 +18,22 @@ def log_p_gauss(x):
     return -0.5 * np.dot(x, alpha)
 
 
-def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4):
+def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4, thin=1):
     lp = log_p(x)
-    chain = np.empty((nsteps, len(x)))
+    chain = np.empty((nsteps // thin, len(x)))
     acc = 0
     for step in range(len(chain)):
-        x_prime = np.array(x)
-        x_prime[np.random.randint(len(x))] += prop_sigma * np.random.randn()
-        lp_prime = log_p(x_prime)
-        if np.random.rand() <= np.exp(lp_prime - lp):
-            acc += 1
-            x[:] = x_prime
-            lp = lp_prime
-        chain[step] = x
-    return chain, acc / len(chain)
+        for i in range(thin):
+            x_prime = np.array(x)
+            x_prime[np.random.randint(len(x))] += prop_sigma*np.random.randn()
+            lp_prime = log_p(x_prime)
+            if np.random.rand() <= np.exp(lp_prime - lp):
+                acc += 1
+                x[:] = x_prime
+                lp = lp_prime
+            if i == thin - 1:
+                chain[step] = x
+    return chain, acc / (len(chain) * thin)
 
 
 if __name__ == "__main__":

### 2016-07-19 mcmc.tex
moving furniture

@@
           Brendon Brewer (Auckland),
+          Will Farr (Birmingham),
           Jonathan Goodman (NYU),
@@
 Not being mathematicians, we don't know the precise definition of a
-pdf\note{Seriously.  We don't.  It has something to do with measure theory.}, but for our purposes here a pdf is
+pdf\note{Seriously.  It has something to do with measure theory.}, but for our purposes here a pdf is
 any single-valued (scalar) function that is non-negative everywhere
@@
 
+\section{Likelihoods and priors}
+
+\MCMC\ is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
+  or a badly normalized all-positive function $f(\pars)$
+  that is different from the pdf by an unknown factor $Z$.
+In the context of data analysis,
+  \MCMC\ is usually being used to obtain samples $\pars_k$
+  that are \emph{parameter values} $\pars$ for a probabilistic model of some data.
+That is, \MCMC\ is sampling a pdf \emph{for the parameters}.
+If it is returning parameter values, then it is not sampling a pdf for the data.
+That is, \MCMC\ can only sample a posterior (or prior) pdf for the parameters.
+\emph{It cannot sample a likelihood.}%
+\note{Well, technically, \MCMC\ \emph{can} be used to sample a likelihood function,
+  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
+  The purist might say that even this is not sampling a likelihood function,
+  because you should only call it a ``likelihood function'' in contexts in which you are
+  treating the data as fixed (at the true data values) and the parameters as variable.
+  In this context, the likelihood function is \emph{not} a pdf for the parameters,
+  so it can't be sampled.}
+
+Despite all this, in many cases,
+  data analysts believe they are sampling the likelihood.
+This is because (we presume) they have put a likelihood function
+  (or log-likelihood function)
+  in as the input to the \MCMC\ code, where the probability function
+  (or log-probability function) should go.
+Then is it the likelihood that is being sampled?
+No, not really;
+  it is a posterior probability that is directly proportional to the likelihood function.
+That is, it is a posterior probability for some implicit (and improper) ``flat'' priors.
+
+Ideally, if you are using \MCMC\ for inference, your code should input
+to the \MCMC\ function or routine a probability function which is a
+product of a prior times a likelihood.
+In terms of implementation, it is usually advisable to work in the
+logarithm of the density function, so the function input to the \MCMC\ code
+would be called something like \code{ln_f()}.
+This function \code{ln_f()} internally would compute and return the
+sum of a log prior pdf \code{ln_prior()} and a log likelihood function
+\code{ln_likelihood()}.
+(When working in logarithmic probability, in M--H-like \MCMC\ methods,
+the accept--reject step would change from a comparison of a random
+deviate with a ratio of probabilities to comparison of the log of a
+random deviate with a difference of log probabilities.)
+
+If you are using ``flat'' (improper) priors, the \code{ln_prior()}
+function can just return a zero no matter what the parameters.
+If it is flat with bounds (that is, proper), the \code{ln_prior()}
+should check the bounds and return \code{-Inf} values if the parameter
+vector is out of bounds.
+The pseudo-code for the \code{ln_f()} function should look something
+like this:
+% DFM: Do you have a preferred environment for code other than verbatim?
+\begin{verbatim}
+def ln_f(pars, data):
+    x = ln_prior(pars)
+    if not is_finite(x):
+        return -Inf
+    return x + ln_likelihood(data, pars)
+\end{verbatim}
+This pseudo-code ensures that you only compute the likelihood when the
+parameters are within the prior bounds; it assumes implicitly that the
+prior pdf is easier to compute than the likelihood.
+If you find yourself in the opposite regime, adjust accordingly.
+Of course the above pseudo-code presumes that when you perform the
+accept--reject step of the \MCMC\ method, the programming language handles
+properly the \code{-Inf} values.\footnote{If you are working in a language
+that doesn't have an \code{Inf} or doesn't evaluate comparisons correctly
+when the \code{Inf} appears, you might have to write some case code, and
+have your \code{ln_f} function return a value \emph{and} some kind of flag
+which indicates ``zero probability'' or $f(\pars)=0$.}
+
+It should be outside of the scope of this document to note here that
+it is a good idea to have proper priors.\note{This point---that you
+  should be using proper priors---is just basic Bayesian good
+  practice, often violated but never for good reasons.}
+Proper priors obey the integral constraint (\ref{eq:proper}),
+with $Z$ finite.
+It isn't a requirement that priors be proper for the posterior to
+be proper, so many investigators and projects violate this rule.
+This is outside the current scope, except for the following point:
+It is a great functional test of your sampler and your data analysis
+setup to take the likelihood function to a very small power (much less
+than one) or multiply the log-likelihood by a very small number (much
+less than one) and check that the sampler samples, properly and
+correctly, the prior pdf.
+This test is only possible if the prior pdf is proper.
+
+\begin{problem}
+Run your M-H \MCMC\ sampler from \problemname~\ref{prob:MH}, but now
+with a density function that is precisely unity \emph{everywhere}
+(that is, at any input value of $x$ it returns unity).  That is, an
+improper function.  Run it for longer and longer and plot the chain
+value $x$ as a function of timestep.  What happens?
+\end{problem}
+
+\begin{problem}
+For a real-world inference problem, read enough of
+\citet{fittingaline} to understand and execute Exercise~6 in that
+document.
+\end{problem}
+
 \section{Metropolis--Hastings MCMC}\label{sec:MH}
@@
 
+HOGG: ...You must describe your priors in the same parameterization that you ``step'' in.
+That is, if you change your parameters, you \emph{must} change your priors correspondingly.
+Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
+
 \begin{problem}\label{prob:MH}
@@
 again.
+\end{problem}
+
+\begin{problem}
+Modify the sampler you wrote in \problemname~\ref{prob:MH} to take
+steps not in $x$ but in $\ln x$.  That is, replace the Gaussian
+proposal distribution $q(x'\given x)$ with a Gaussian distribution in
+$\ln x$ $q(\ln x'\given \ln x)$, but---incorrectly---make no other
+changes.  What about your answers change?  How wrong is your answer?
+Bonus points: Compute the analytic function that you have sampled from
+(that is, not $p(x)$) and over-plot it on your histogram.
 \end{problem}
@@
 trained professional.
-
-\section{Likelihoods and priors}
-
-\MCMC\ is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
-  or a badly normalized all-positive function $f(\pars)$
-  that is different from the pdf by an unknown factor $Z$.
-In the context of data analysis,
-  \MCMC\ is usually being used to obtain samples $\pars_k$
-  that are \emph{parameter values} $\pars$ for a probabilistic model of some data.
-That is, \MCMC\ is sampling a pdf \emph{for the parameters}.
-If it is returning parameter values, then it is not sampling a pdf for the data.
-That is, \MCMC\ can only sample a posterior (or prior) pdf for the parameters.
-\emph{It cannot sample a likelihood.}%
-\note{Well, technically, \MCMC\ \emph{can} be used to sample a likelihood function,
-  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
-  The purist might say that even this is not sampling a likelihood function,
-  because you should only call it a ``likelihood function'' in contexts in which you are
-  treating the data as fixed (at the true data values) and the parameters as variable.
-  In this context, the likelihood function is \emph{not} a pdf for the parameters,
-  so it can't be sampled.}
-
-Despite all this, in many cases,
-  data analysts believe they are sampling the likelihood.
-This is because (we presume) they have put a likelihood function
-  (or log-likelihood function)
-  in as the input to the \MCMC\ code, where the probability function
-  (or log-probability function) should go.
-Then is it the likelihood that is being sampled?
-No, not really;
-  it is a posterior probability that is directly proportional to the likelihood function.
-That is, it is a posterior probability for some implicit (and improper) ``flat'' priors.
-
-Ideally, if you are using \MCMC\ for inference, your code should input
-to the \MCMC\ function or routine a probability function which is a
-product of a prior times a likelihood.
-In terms of implementation, it is usually advisable to work in the
-logarithm of the density function, so the function input to the \MCMC\ code
-would be called something like \code{ln_f()}.
-This function \code{ln_f()} internally would compute and return the
-sum of a log prior pdf \code{ln_prior()} and a log likelihood function
-\code{ln_likelihood()}.
-(When working in logarithmic probability, in M--H-like \MCMC\ methods,
-the accept--reject step would change from a comparison of a random
-deviate with a ratio of probabilities to comparison of the log of a
-random deviate with a difference of log probabilities.)
-
-If you are using ``flat'' (improper) priors, the \code{ln_prior()}
-function can just return a zero no matter what the parameters.
-If it is flat with bounds (that is, proper), the \code{ln_prior()}
-should check the bounds and return \code{-Inf} values if the parameter
-vector is out of bounds.
-The pseudo-code for the \code{ln_f()} function should look something
-like this:
-% DFM: Do you have a preferred environment for code other than verbatim?
-\begin{verbatim}
-def ln_f(pars, data):
-    x = ln_prior(pars)
-    if not is_finite(x):
-        return -Inf
-    return x + ln_likelihood(data, pars)
-\end{verbatim}
-This pseudo-code ensures that you only compute the likelihood when the
-parameters are within the prior bounds; it assumes implicitly that the
-prior pdf is easier to compute than the likelihood.
-If you find yourself in the opposite regime, adjust accordingly.
-Of course the above pseudo-code presumes that when you perform the
-accept--reject step of the \MCMC\ method, the programming language handles
-properly the \code{-Inf} values.\footnote{If you are working in a language
-that doesn't have an \code{Inf} or doesn't evaluate comparisons correctly
-when the \code{Inf} appears, you might have to write some case code, and
-have your \code{ln_f} function return a value \emph{and} some kind of flag
-which indicates ``zero probability'' or $f(\pars)=0$.}
-
-It should be outside of the scope of this document to note here that
-it is a good idea to have proper priors.\note{This point---that you
-  should be using proper priors---is just basic Bayesian good
-  practice, often violated but never for good reasons.}
-Proper priors obey the integral constraint (\ref{eq:proper}),
-with $Z$ finite.
-It isn't a requirement that priors be proper for the posterior to
-be proper, so many investigators and projects violate this rule.
-This is outside the current scope, except for the following point:
-It is a great functional test of your sampler and your data analysis
-setup to take the likelihood function to a very small power (much less
-than one) or multiply the log-likelihood by a very small number (much
-less than one) and check that the sampler samples, properly and
-correctly, the prior pdf.
-This test is only possible if the prior pdf is proper.
-
-HOGG: ...You must describe your priors in the same parameterization that you ``step'' in.
-That is, if you change your parameters, you \emph{must} change your priors correspondingly.
-Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
-
-\begin{problem}
-Run your M-H \MCMC\ sampler from \problemname~\ref{prob:MH}, but now
-with a density function that is precisely unity \emph{everywhere}
-(that is, at any input value of $x$ it returns unity).  That is, an
-improper function.  Run it for longer and longer and plot the chain
-value $x$ as a function of timestep.  What happens?
-\end{problem}
-
-\begin{problem}
-Modify the sampler you wrote in \problemname~\ref{prob:MH} to take
-steps not in $x$ but in $\ln x$.  That is, replace the Gaussian
-proposal distribution $q(x'\given x)$ with a Gaussian distribution in
-$\ln x$ $q(\ln x'\given \ln x)$, but---incorrectly---make no other
-changes.  What about your answers change?  How wrong is your answer?
-Bonus points: Compute the analytic function that you have sampled from
-(that is, not $p(x)$) and over-plot it on your histogram.
-\end{problem}
-
-\begin{problem}
-For a real-world inference problem, read enough of
-\citet{fittingaline} to understand and execute Exercise~6 in that
-document.
-\end{problem}
 

### 2016-07-19 problems/p5.py
bug in p5

@@ -20,7 +20,8 @@ s[0] *= 2
 s[1] *= 0.8
 fig, axes = pl.subplots(1, 3, sharex=True, sharey=True, figsize=s)
 for n, ax in zip([0.0, -1.0, 2.0], axes):
-    chain = run_mcmc(log_p_gauss, np.array(x0), nsteps=2e3, prop_sigma=10**n)
+    chain, _ = run_mcmc(log_p_gauss, np.array(x0), nsteps=2e3,
+                        prop_sigma=10**n)
     ax.plot(chain[:, 0], chain[:, 1], "o-", color=COLORS["DATA"], ms=2)
     ax.plot(x0[0], x0[1], "o", color=COLORS["MODEL_1"])
     ax.set_xlim(-6.3, 6.3)

### 2016-07-19 mcmc.tex
more furniture
yet more furniture rearrangement
infinity safe

@@
 
-\section{Likelihoods and priors}
+\section{Metropolis--Hastings MCMC}\label{sec:MH}
+
+The simplest algorithm for \MCMC\ is the Metropolis--Hastings algorithm (\MH\ \MCMC).\note{%
+  There are many claims that this algorithm is incorrectly named,
+  with claims that it should be credited to Enrico Fermi or Stan Ulam.
+  We don't have any opinions of this matter, but encourage the reader to follow this up.
+  The original paper is \citet{metropolis}, and there are a few sketchy historical notes
+  in \citet{geyer}.}
+It is so simple that we recommend that any reader of this document
+  who has not previously implemented the algorithm take a break at the end
+  of this \sectionname\ and implement it forthwith, in a short piece of computer code,
+  in the context of some simple problems.\note{%
+    We request this in a trivial case below in
+    \problemname~\ref{prob:MH} and the following problems in this
+    \sectionname.
+    We make the same request in a less trivial context in our
+    model-fitting \documentname\ (\citealt{fittingaline},
+    \problemname~6 and 7), where we ask the Reader to implement \MCMC\ for
+    a useful mixture model.}
+
+The \MH\ \MCMC\ algorithm requires two inputs.
+The first is a handle to the function $f(\pars)$ that is the function to be sampled,
+  such that the algorithm can evaluate $f(\pars)$ for any value of the parameters $\pars$.
+In data-analysis contexts,
+  this function would be the prior $p(\pars)$ times the likelihood $p(\data\given\pars)$
+  evaluated at the observed data $\data$.
+The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
+  such that the algorithm can draw a new position $\pars'$ in the parameter space
+  given an ``old'' position $\pars$.
+This second function must meet a symmetry requirement (detailed
+  balance) we discuss further below.
+It permits us to random-walk around the parameter space in a fair way.
+
+The algorithm\note{Technically this is the \emph{Metropolis} algorithm
+  rather than the Metropolis--Hastings algorithm.  The difference is
+  that in the Metropolis--Hastings algorithm, we permit the proposal
+  distribution to disobey detailed balance and then we correct the
+  acceptance/rejection step to recover detailed balance.} is the
+following: We have generated some set of samples, the most recent of
+which is $\pars_k$ To generate the next sample $\pars_{k+1}$ do the
+following:
+\begin{itemize}
+\item Draw a proposal $\pars'$ from the proposal pdf $q(\pars'\given\pars_k)$.
+\item Draw a random number $0<r<1$ from the uniform distribution.
+\item If $f(\pars') / f(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
+      otherwise $\pars_{k+1} \leftarrow \pars_k$.
+\end{itemize}
+That is, at each step, either a new proposed position in the parameter
+  space gets accepted into the list of samples or else the previous sample
+  in the parameter space \emph{gets repeated}.
+The algorithm can be iterated a large number $K$ of times to produce $K$ samples.
+
+Why does this algorithm work?  The answer is not absolutely
+trivial\note{There is a very mathematical discussion in \citet{geyer}
+  that we do not fully understand.  There is a more heuristic answer
+  on \project{Wikipedia} that we are happier with!},
+but there are two components to the argument:
+The first is that the Markov process delivers a unique stationary
+distribution.
+The second is that that stationary distribution is proportional to the
+density function $f(\pars)$.
+
+This algorithm---and indeed any \MCMC\ algorithm---produces a
+  \emph{biased random walk} through parameter space.
+It is a random walk for the same reason that it is ``Markov'':
+The step it makes to position $\pars_{k+1}$ depends only on the state
+  of the sampler at position $\pars_k$ (and no previous state).
+It is a biased random walk, biased by the acceptance algorithm
+  involving the ratios of function values; this acceptance rule biases
+  the random walk such that the amount of time spent in the neighborhood
+  of location $\pars$ is proportional to $f(\pars)$.
+Because of this local (Markov) property, the nearby samples are not
+  independent; the algorithm only produces fair samples in the limit of
+  arbitrary run time, and two samples are only independent when they are
+  sufficiently separated in the chain (more on this below in \sectionname~\ref{sec:convergence}).
+
+The principal user-settable knob in \MH\ \MCMC\ is the proposal
+  pdf $q(\pars'\given\pars)$.
+A typical choice is a multi-variate Gaussian distribution for $\pars'$
+  centered on $\pars$ with some simple (diagonal, perhaps) variance
+  tensor.
+We will discuss the choice and tuning of this proposal distribution
+  below (\sectionname~\ref{sec:tuning} and \sectionname~\ref{sec:trouble}).
+
+Importantly---for the algorithm given above to work correctly---the
+  proposal pdf must satisfy a ``detailed balance'' condition; it
+  must have the property that
+\begin{eqnarray}
+q(\pars'\given\pars) &=& q(\pars\given\pars')
+\quad ;
+\end{eqnarray}
+that is, it must be just as easy to go one way in the parameter space
+  as the other.
+You can break this property, if you like, and then adjust the
+  acceptance condition accordingly (which is truly the Metropolis--Hastings algorithm), but we \emph{do not recommend this}
+  except under the supervision of a trained professional.\note{Actually,
+  our own favorite method, \project{emcee}, has a proposal pdf that does
+  violate detailed balance in this way and has a compensated
+  acceptance probability.  Read the paper \cite{emcee} for more details.}
+The reason is:  It is one thing to draw samples from $q(x'\given x)$.
+It is another thing to correctly write down $q(x'\given x)$.
+  If you violate detailed balance in $q(x'\given x)$ then you have to be
+  able to \emph{both} draw from \emph{and} write down $q(x'\given x)$
+  and you are subject to a set of new possible bugs for your code.
+If the detailed balance condition frightens you (as it frightens us),
+  then just stick with pdfs that are symmetric in $\pars'$ and $\pars$,
+  like the Gaussian, or a centered uniform distribution and forget about
+  it.
+
+In what follows, when we discuss tuning of the proposal pdf, we will
+  often cycle through the parameter blob components and propose changes
+  in only one dimension or element or component at a time.
+That is, at step $k+1$ you might only propose a one-dimensional move
+  in the $i$th dimension, not in all $D$ dimensions of the parameter
+  space.
+That won't change anything significant; but if you are implementing
+  this right now you might want to do that.
+It will help us tune the proposal pdf (\sectionname~\ref{sec:tuning})
+and diagnose problems (\sectionname~\ref{sec:trouble}).
+
+One note to make here
+  is that you must \emph{propose}
+  in the same coordinates (parameterization or transformation of parameters)
+  as that in which your priors are \emph{specified},
+  or else multiply in a (possibly nasty) Jacobian.
+That is, if your prior is ``flat in $\pars$''
+  but you find it easier to run your sampler by proposing steps in $\ln \pars$,
+  if you don't modify your acceptance probability by the Jacobian,
+  your \emph{real} prior won't be flat in $\pars$.
+These think-os can get subtle;
+  the best way to avoid them
+  is to write your code and your likelihood function
+  and your prior pdf function and your proposal distribution
+  all in precisely the same parameterization.
+We will return to this point below in our comments on testing (\sectionname~\ref{sec:trouble}).
+
+As we discuss briefly below (\sectionname~\ref{sec:methods}),
+  there are many \MCMC\ methods more advanced than M--H.
+However, they all share characteristics with M--H
+  (initialization, tuning, judging convergence),
+  such that it is very valuable to understand \MH\ well before using anything more advanced.
+Furthermore, a scientist new to \MCMC\ benefits enormously from
+  building, tuning, and using her or his own \MCMC\ software.
+One piece of advice we give then,
+  to the new user of \MCMC,
+  is to code up, tune, and use a \MH\ \MCMC\ sampler for a scientific project.
+A huge amount is learned in doing this,
+  and it is very often the case that the home-built \MH\ \MCMC\ does everything needed;
+  you often don't need any more advanced tool.
+Only after you have concluded that your home-built \MH\ \MCMC\ sampler
+  is not suited to your project
+  (or not developed properly into a properly versatile software package)
+  should you download and start to use any professionally developed alternative.
+That is, even if---in the end---%
+  you want to leave the sampling code to the experts and run an industrial-strength code,
+  it is still valuable to build your own,
+  given the simplicity of the algorithm,
+  and given the intuition you gain by doing it (at least once) yourself.\note{%
+    We also make this point in a previous \documentname\ (\citealt{fittingaline});
+    the ambitious reader will find in that \documentname\ a solid example project
+    for using \MCMC\ in a real context.}
+
+Because many problems involve a huge amount of dynamic range in the
+density function $f(\pars)$, and we like to avoid underflows and overflows
+(in, say, ratios computed for accept--reject steps), it is often advisable
+to work in the (natural) logarithm of the density rather than the density.
+When working in logarithmic density, 
+the accept--reject step would change from a comparison of a random
+deviate with a ratio of probabilities to comparison of the log of a
+random deviate with a difference of log probabilities.
+That is, the accept--reject step becomes:
+\begin{itemize}
+\item If $\ln f(\pars') - \ln f(\pars_k) > \ln r$ then $\pars_{k+1} \leftarrow \pars'$;
+      otherwise $\pars_{k+1} \leftarrow \pars_k$.
+\end{itemize}
+This protects you from underflow, but exposes you to some (negative)
+infinities, if you end up taking the logarithm of a zero.  It is
+important to write your code to be infinity-safe.\note{Most code will
+  be infinity safe if written in the form in this paragraph. However,
+  there can be issues if \emph{both} $f(\pars')$ \emph{and}
+  $f(\pars_k)$ are negative infinity; in (present-day) Python this
+  will return a \texttt{NaN} rather than an infinity or zero. That's a
+  problem and is a case that should get caught (probably way before
+  the accept--reject step!).}
+
+HOGG SAY HERE: ...You must describe your priors in the same parameterization that you ``step'' in.
+That is, if you change your parameters, you \emph{must} change your priors correspondingly.
+Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
+
+\begin{problem}\label{prob:MH}
+In your scientific programming language of choice, write a very simple
+M-H \MCMC\ sampler.
+Sample in a single parameter $x$ and give the sampler as its density
+function $p(x)$ a Gaussian density with mean 2 and variance 2.
+(Note that variance is the \emph{square} of the standard deviation.)
+Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
+pdf with mean zero and variance 1.
+Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
+Plot the results as a histogram, with the true density over-plotted sensibly.
+The resulting  plot should look something like \figurename~\ref{fig:MH}.
+\end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/p2}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH}:
+MCMC samples from a Gaussian (black) and the true distribution (blue).}
+\label{fig:MH}
+\end{figure}
+
+\begin{problem}\label{prob:MH2}
+Re-do \problemname~\ref{prob:MH} but now with an input density
+that is uniform on $3<x<7$ and zero everywhere else.
+The plot should look like \figurename~\ref{fig:MH2}.
+What change did you have to make to the initialization, and why?
+\end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/p3}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH2}:
+MCMC samples from a uniform distribution (black) and the true distribution
+(blue).}
+\label{fig:MH2}
+\end{figure}
+
+\begin{problem}\label{prob:twod}
+Re-do \problemname~\ref{prob:MH} but now with an input density
+that is a function of two variables $(x, y)$.
+For the density function use two different functions.
+\emph{(a)} The first density function is a covariant two-dimensional Gaussian
+density with variance tensor
+\begin{eqnarray}
+V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
+\quad.
+\end{eqnarray}
+\emph{(b)} The second density function is a rectangular top-hat function that
+is uniform on the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
+For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
+density with zero mean and variance tensor set to the two-dimensional
+identity matrix.
+Plot the two one-dimensional histograms and also a two-dimensional scatter
+plot for each sampling.
+\figurename~\ref{fig:twod-a} shows the expected results for the Gaussian.
+Make a similar plot for the top-hat.
+\end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{problems/p4a}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:twod}a:
+MCMC samples from a two-dimensional Gaussian (scatter plot) and the
+one-dimensional marginalized distributions (histograms).}
+\label{fig:twod-a}
+\end{figure}
+
+\begin{problem}\label{prob:MH-sigma}
+Re-do \problemname~\ref{prob:twod}a but with different values for
+the variance of the proposal distribution $q(x'\given x)$.
+What happens when you go to very extreme values (like for instance
+$10^{-1}$ or $10^2$)?
+\end{problem}
+
+\begin{problem}
+Why, in all the previous problems, did we give the proposal
+distributions $q(x'\given x)$ zero mean?
+What would be bad if we hadn't done that?
+Re-do \problemname~\ref{prob:twod}a with a proposal $q(x'\given x)$ with a
+non-zero mean and see what happens.
+Bonus points: Modify the acceptance--rejection criterion to deal with
+the messed-up $q(x'\given x)$ and show that everything works once
+again.
+\end{problem}
+
+\section{Likelihoods and priors}\label{sec:likelihood}
 
 \MCMC\ is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
-  or a badly normalized all-positive function $f(\pars)$
+  given a badly normalized all-positive function $f(\pars)$
   that is different from the pdf by an unknown factor $Z$.
@@
 That is, \MCMC\ is sampling a pdf \emph{for the parameters}.
-If it is returning parameter values, then it is not sampling a pdf for the data.
-That is, \MCMC\ can only sample a posterior (or prior) pdf for the parameters.
-\emph{It cannot sample a likelihood.}%
-\note{Well, technically, \MCMC\ \emph{can} be used to sample a likelihood function,
-  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
-  The purist might say that even this is not sampling a likelihood function,
-  because you should only call it a ``likelihood function'' in contexts in which you are
-  treating the data as fixed (at the true data values) and the parameters as variable.
-  In this context, the likelihood function is \emph{not} a pdf for the parameters,
-  so it can't be sampled.}
-
-Despite all this, in many cases,
-  data analysts believe they are sampling the likelihood.
-This is because (we presume) they have put a likelihood function
-  (or log-likelihood function)
-  in as the input to the \MCMC\ code, where the probability function
-  (or log-probability function) should go.
-Then is it the likelihood that is being sampled?
-No, not really;
-  it is a posterior probability that is directly proportional to the likelihood function.
-That is, it is a posterior probability for some implicit (and improper) ``flat'' priors.
 
@@
 product of a prior times a likelihood.
-In terms of implementation, it is usually advisable to work in the
+As we noted above---in terms of implementation---it is usually advisable to work in the
 logarithm of the density function, so the function input to the \MCMC\ code
@@
 \code{ln_likelihood()}.
-(When working in logarithmic probability, in M--H-like \MCMC\ methods,
-the accept--reject step would change from a comparison of a random
-deviate with a ratio of probabilities to comparison of the log of a
-random deviate with a difference of log probabilities.)
 
@@
 
+\MCMC\ \emph{cannot sample a likelihood} (which is a probability \emph{for the data} given
+parameters).\note{Well, technically, \MCMC\ \emph{can} be used to sample a likelihood function,
+  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
+  The purist might say that even this is not sampling a likelihood function,
+  because you should only call it a ``likelihood function'' in contexts in which you are
+  treating the data as fixed (at the true data values) and the parameters as variable.
+  In this context, the likelihood function is \emph{not} a pdf for the parameters,
+  so it can't be sampled.}
+Despite this, in many cases,
+  data analysts believe they are sampling the likelihood.
+This is because (we presume) they have put a likelihood function
+  (or log-likelihood function)
+  in as the input to the \MCMC\ code, where the probability function
+  (or log-probability function) should go.
+Then is it the likelihood that is being sampled?
+No, not really;
+  it is a posterior probability that is directly proportional to the likelihood function.
+That is, it is a posterior probability for some implicit (and improper) ``flat'' priors.
+
 It should be outside of the scope of this document to note here that
@@
 (that is, at any input value of $x$ it returns unity).  That is, an
-improper function.  Run it for longer and longer and plot the chain
-value $x$ as a function of timestep.  What happens?
+improper function (as discussed in \sectionname~\ref{sec:likelihood}).
+Run it for longer and longer and plot the chain value $x$ as a
+function of timestep.  What happens?
 \end{problem}
@@
 document.
-\end{problem}
-
-\section{Metropolis--Hastings MCMC}\label{sec:MH}
-
-The simplest algorithm for \MCMC\ is the Metropolis--Hastings algorithm (\MH\ \MCMC).\note{%
-  There are many claims that this algorithm is incorrectly named,
-  with claims that it should be credited to Enrico Fermi or Stan Ulam.
-  We don't have any opinions of this matter, but encourage the reader to follow this up.
-  The original paper is \citet{metropolis}, and there are a few sketchy historical notes
-  in \citet{geyer}.}
-It is so simple that we recommend that any reader of this document
-  who has not previously implemented the algorithm take a break at the end
-  of this \sectionname\ and implement it forthwith, in a short piece of computer code,
-  in the context of some simple problems.\note{%
-    We request this in a trivial case below in
-    \problemname~\ref{prob:MH} and the following problems in this
-    \sectionname.
-    We make the same request in a less trivial context in our
-    model-fitting \documentname\ (\citealt{fittingaline},
-    \problemname~6 and 7), where we ask the Reader to implement \MCMC\ for
-    a useful mixture model.}
-
-The \MH\ \MCMC\ algorithm requires two inputs.
-The first is a handle to the function $f(\pars)$ that is the function to be sampled,
-  such that the algorithm can evaluate $f(\pars)$ for any value of the parameters $\pars$.
-In data-analysis contexts,
-  this function would be the prior $p(\pars)$ times the likelihood $p(\data\given\pars)$
-  evaluated at the observed data $\data$.
-The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
-  such that the algorithm can draw a new position $\pars'$ in the parameter space
-  given an ``old'' position $\pars$.
-This second function must meet a symmetry requirement (detailed
-  balance) we discuss further below.
-It permits us to random-walk around the parameter space in a fair way.
-
-The algorithm\note{Technically this is the \emph{Metropolis} algorithm
-  rather than the Metropolis--Hastings algorithm.  The difference is
-  that in the Metropolis--Hastings algorithm, we permit the proposal
-  distribution to disobey detailed balance and then we correct the
-  acceptance/rejection step to recover detailed balance.} is the
-following: We have generated some set of samples, the most recent of
-which is $\pars_k$ To generate the next sample $\pars_{k+1}$ do the
-following:
-\begin{itemize}
-\item Draw a proposal $\pars'$ from the proposal pdf $q(\pars'\given\pars_k)$.
-\item Draw a random number $0<r<1$ from the uniform distribution.
-\item If $f(\pars') / f(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
-      otherwise $\pars_{k+1} \leftarrow \pars_k$.
-\end{itemize}
-That is, at each step, either a new proposed position in the parameter
-  space gets accepted into the list of samples or else the previous sample
-  in the parameter space \emph{gets repeated}.
-The algorithm can be iterated a large number $K$ of times to produce $K$ samples.
-
-Why does this algorithm work?  The answer is not absolutely
-trivial\note{There is a very mathematical discussion in \citet{geyer}
-  that we do not fully understand.  There is a more heuristic answer
-  on \project{Wikipedia} that we are happier with!},
-but there are two components to the argument:
-The first is that the Markov process delivers a unique stationary
-distribution.
-The second is that that stationary distribution is proportional to the
-density function $f(\pars)$.
-
-This algorithm---and indeed any \MCMC\ algorithm---produces a
-  \emph{biased random walk} through parameter space.
-It is a random walk for the same reason that it is ``Markov'':
-The step it makes to position $\pars_{k+1}$ depends only on the state
-  of the sampler at position $\pars_k$ (and no previous state).
-It is a biased random walk, biased by the acceptance algorithm
-  involving the ratios of function values; this acceptance rule biases
-  the random walk such that the amount of time spent in the neighborhood
-  of location $\pars$ is proportional to $f(\pars)$.
-Because of this local (Markov) property, the nearby samples are not
-  independent; the algorithm only produces fair samples in the limit of
-  arbitrary run time, and two samples are only independent when they are
-  sufficiently separated in the chain (more on this below in \sectionname~\ref{sec:convergence}).
-
-The principal user-settable knob in \MH\ \MCMC\ is the proposal
-  pdf $q(\pars'\given\pars)$.
-A typical choice is a multi-variate Gaussian distribution for $\pars'$
-  centered on $\pars$ with some simple (diagonal, perhaps) variance
-  tensor.
-We will discuss the choice and tuning of this proposal distribution
-  below (\sectionname~\ref{sec:tuning} and \sectionname~\ref{sec:trouble}).
-
-Importantly---for the algorithm given above to work correctly---the
-  proposal pdf must satisfy a ``detailed balance'' condition; it
-  must have the property that
-\begin{eqnarray}
-q(\pars'\given\pars) &=& q(\pars\given\pars')
-\quad ;
-\end{eqnarray}
-that is, it must be just as easy to go one way in the parameter space
-  as the other.
-You can break this property, if you like, and then adjust the
-  acceptance condition accordingly (which is truly the Metropolis--Hastings algorithm), but we \emph{do not recommend this}
-  except under the supervision of a trained professional.\note{Actually,
-  our own favorite method, \project{emcee}, has a proposal pdf that does
-  violate detailed balance in this way and has a compensated
-  acceptance probability.  Read the paper \cite{emcee} for more details.}
-The reason is:  It is one thing to draw samples from $q(x'\given x)$.
-It is another thing to correctly write down $q(x'\given x)$.
-  If you violate detailed balance in $q(x'\given x)$ then you have to be
-  able to \emph{both} draw from \emph{and} write down $q(x'\given x)$
-  and you are subject to a set of new possible bugs for your code.
-If the detailed balance condition frightens you (as it frightens us),
-  then just stick with pdfs that are symmetric in $\pars'$ and $\pars$,
-  like the Gaussian, or a centered uniform distribution and forget about
-  it.
-
-In what follows, when we discuss tuning of the proposal pdf, we will
-  often cycle through the parameter blob components and propose changes
-  in only one dimension or element or component at a time.
-That is, at step $k+1$ you might only propose a one-dimensional move
-  in the $i$th dimension, not in all $D$ dimensions of the parameter
-  space.
-That won't change anything significant; but if you are implementing
-  this right now you might want to do that.
-It will help us tune the proposal pdf (\sectionname~\ref{sec:tuning})
-and diagnose problems (\sectionname~\ref{sec:trouble}).
-
-One note to make here
-  is that you must \emph{propose}
-  in the same coordinates (parameterization or transformation of parameters)
-  as that in which your priors are \emph{specified},
-  or else multiply in a (possibly nasty) Jacobian.
-That is, if your prior is ``flat in $\pars$''
-  but you find it easier to run your sampler by proposing steps in $\ln \pars$,
-  if you don't modify your acceptance probability by the Jacobian,
-  your \emph{real} prior won't be flat in $\pars$.
-These think-os can get subtle;
-  the best way to avoid them
-  is to write your code and your likelihood function
-  and your prior pdf function and your proposal distribution
-  all in precisely the same parameterization.
-We will return to this point below in our comments on testing (\sectionname~\ref{sec:trouble}).
-
-As we discuss briefly below (\sectionname~\ref{sec:methods}),
-  there are many \MCMC\ methods more advanced than M--H.
-However, they all share characteristics with M--H
-  (initialization, tuning, judging convergence),
-  such that it is very valuable to understand \MH\ well before using anything more advanced.
-Furthermore, a scientist new to \MCMC\ benefits enormously from
-  building, tuning, and using her or his own \MCMC\ software.
-One piece of advice we give then,
-  to the new user of \MCMC,
-  is to code up, tune, and use a \MH\ \MCMC\ sampler for a scientific project.
-A huge amount is learned in doing this,
-  and it is very often the case that the home-built \MH\ \MCMC\ does everything needed;
-  you often don't need any more advanced tool.
-Only after you have concluded that your home-built \MH\ \MCMC\ sampler
-  is not suited to your project
-  (or not developed properly into a properly versatile software package)
-  should you download and start to use any professionally developed alternative.
-That is, even if---in the end---%
-  you want to leave the sampling code to the experts and run an industrial-strength code,
-  it is still valuable to build your own,
-  given the simplicity of the algorithm,
-  and given the intuition you gain by doing it (at least once) yourself.\note{%
-    We also make this point in a previous \documentname\ (\citealt{fittingaline});
-    the ambitious reader will find in that \documentname\ a solid example project
-    for using \MCMC\ in a real context.}
-
-DFM says: should we say something about using the log density here?
-
-HOGG: ...You must describe your priors in the same parameterization that you ``step'' in.
-That is, if you change your parameters, you \emph{must} change your priors correspondingly.
-Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
-
-\begin{problem}\label{prob:MH}
-In your scientific programming language of choice, write a very simple
-M-H \MCMC\ sampler.
-Sample in a single parameter $x$ and give the sampler as its density
-function $p(x)$ a Gaussian density with mean 2 and variance 2.
-(Note that variance is the \emph{square} of the standard deviation.)
-Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
-pdf with mean zero and variance 1.
-Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
-Plot the results as a histogram, with the true density over-plotted sensibly.
-The resulting  plot should look something like \figurename~\ref{fig:MH}.
-\end{problem}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{problems/p2}
-\end{center}
-\caption{Solution to \problemname~\ref{prob:MH}:
-MCMC samples from a Gaussian (black) and the true distribution (blue).}
-\label{fig:MH}
-\end{figure}
-
-\begin{problem}\label{prob:MH2}
-Re-do \problemname~\ref{prob:MH} but now with an input density
-that is uniform on $3<x<7$ and zero everywhere else.
-The plot should look like \figurename~\ref{fig:MH2}.
-What change did you have to make to the initialization, and why?
-\end{problem}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{problems/p3}
-\end{center}
-\caption{Solution to \problemname~\ref{prob:MH2}:
-MCMC samples from a uniform distribution (black) and the true distribution
-(blue).}
-\label{fig:MH2}
-\end{figure}
-
-\begin{problem}\label{prob:twod}
-Re-do \problemname~\ref{prob:MH} but now with an input density
-that is a function of two variables $(x, y)$.
-For the density function use two different functions.
-\emph{(a)} The first density function is a covariant two-dimensional Gaussian
-density with variance tensor
-\begin{eqnarray}
-V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
-\quad.
-\end{eqnarray}
-\emph{(b)} The second density function is a rectangular top-hat function that
-is uniform on the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
-For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
-density with zero mean and variance tensor set to the two-dimensional
-identity matrix.
-Plot the two one-dimensional histograms and also a two-dimensional scatter
-plot for each sampling.
-\figurename~\ref{fig:twod-a} shows the expected results for the Gaussian.
-Make a similar plot for the top-hat.
-\end{problem}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{problems/p4a}
-\end{center}
-\caption{Solution to \problemname~\ref{prob:twod}a:
-MCMC samples from a two-dimensional Gaussian (scatter plot) and the
-one-dimensional marginalized distributions (histograms).}
-\label{fig:twod-a}
-\end{figure}
-
-\begin{problem}\label{prob:MH-sigma}
-Re-do \problemname~\ref{prob:twod}a but with different values for
-the variance of the proposal distribution $q(x'\given x)$.
-What happens when you go to very extreme values (like for instance
-$10^{-1}$ or $10^2$)?
-\end{problem}
-
-\begin{problem}
-Why, in all the previous problems, did we give the proposal
-distributions $q(x'\given x)$ zero mean?
-What would be bad if we hadn't done that?
-Re-do \problemname~\ref{prob:twod}a with a proposal $q(x'\given x)$ with a
-non-zero mean and see what happens.
-Bonus points: Modify the acceptance--rejection criterion to deal with
-the messed-up $q(x'\given x)$ and show that everything works once
-again.
 \end{problem}

### 2016-07-19 mcmc.tex
renaming problems

--- 2016-07-19 mcmc.tex
@@
 
-\begin{problem}
+\begin{problem}\label{prob:simple-stats}
 Look up (or choose) definitions for the mean, variance, skewness,
@@
 
-\section{Likelihoods and priors}
+\section{Metropolis--Hastings MCMC}\label{sec:MH}
+
+The simplest algorithm for \MCMC\ is the Metropolis--Hastings algorithm (\MH\ \MCMC).\note{%
+  There are many claims that this algorithm is incorrectly named,
+  with claims that it should be credited to Enrico Fermi or Stan Ulam.
+  We don't have any opinions of this matter, but encourage the reader to follow this up.
+  The original paper is \citet{metropolis}, and there are a few sketchy historical notes
+  in \citet{geyer}.}
+It is so simple that we recommend that any reader of this document
+  who has not previously implemented the algorithm take a break at the end
+  of this \sectionname\ and implement it forthwith, in a short piece of computer code,
+  in the context of some simple problems.\note{%
+    We request this in a trivial case below in
+    \problemname~\ref{prob:MH} and the following problems in this
+    \sectionname.
+    We make the same request in a less trivial context in our
+    model-fitting \documentname\ (\citealt{fittingaline},
+    \problemname~6 and 7), where we ask the Reader to implement \MCMC\ for
+    a useful mixture model.}
+
+The \MH\ \MCMC\ algorithm requires two inputs.
+The first is a handle to the function $f(\pars)$ that is the function to be sampled,
+  such that the algorithm can evaluate $f(\pars)$ for any value of the parameters $\pars$.
+In data-analysis contexts,
+  this function would be the prior $p(\pars)$ times the likelihood $p(\data\given\pars)$
+  evaluated at the observed data $\data$.
+The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
+  such that the algorithm can draw a new position $\pars'$ in the parameter space
+  given an ``old'' position $\pars$.
+This second function must meet a symmetry requirement (detailed
+  balance) we discuss further below.
+It permits us to random-walk around the parameter space in a fair way.
+
+The algorithm\note{Technically this is the \emph{Metropolis} algorithm
+  rather than the Metropolis--Hastings algorithm.  The difference is
+  that in the Metropolis--Hastings algorithm, we permit the proposal
+  distribution to disobey detailed balance and then we correct the
+  acceptance/rejection step to recover detailed balance.} is the
+following: We have generated some set of samples, the most recent of
+which is $\pars_k$ To generate the next sample $\pars_{k+1}$ do the
+following:
+\begin{itemize}
+\item Draw a proposal $\pars'$ from the proposal pdf $q(\pars'\given\pars_k)$.
+\item Draw a random number $0<r<1$ from the uniform distribution.
+\item If $f(\pars') / f(\pars_k) > r$ then $\pars_{k+1} \leftarrow \pars'$;
+      otherwise $\pars_{k+1} \leftarrow \pars_k$.
+\end{itemize}
+That is, at each step, either a new proposed position in the parameter
+  space gets accepted into the list of samples or else the previous sample
+  in the parameter space \emph{gets repeated}.
+The algorithm can be iterated a large number $K$ of times to produce $K$ samples.
+
+Why does this algorithm work?  The answer is not absolutely
+trivial\note{There is a very mathematical discussion in \citet{geyer}
+  that we do not fully understand.  There is a more heuristic answer
+  on \project{Wikipedia} that we are happier with!},
+but there are two components to the argument:
+The first is that the Markov process delivers a unique stationary
+distribution.
+The second is that that stationary distribution is proportional to the
+density function $f(\pars)$.
+
+This algorithm---and indeed any \MCMC\ algorithm---produces a
+  \emph{biased random walk} through parameter space.
+It is a random walk for the same reason that it is ``Markov'':
+The step it makes to position $\pars_{k+1}$ depends only on the state
+  of the sampler at position $\pars_k$ (and no previous state).
+It is a biased random walk, biased by the acceptance algorithm
+  involving the ratios of function values; this acceptance rule biases
+  the random walk such that the amount of time spent in the neighborhood
+  of location $\pars$ is proportional to $f(\pars)$.
+Because of this local (Markov) property, the nearby samples are not
+  independent; the algorithm only produces fair samples in the limit of
+  arbitrary run time, and two samples are only independent when they are
+  sufficiently separated in the chain (more on this below in \sectionname~\ref{sec:convergence}).
+
+The principal user-settable knob in \MH\ \MCMC\ is the proposal
+  pdf $q(\pars'\given\pars)$.
+A typical choice is a multi-variate Gaussian distribution for $\pars'$
+  centered on $\pars$ with some simple (diagonal, perhaps) variance
+  tensor.
+We will discuss the choice and tuning of this proposal distribution
+  below (\sectionname~\ref{sec:tuning} and \sectionname~\ref{sec:trouble}).
+
+Importantly---for the algorithm given above to work correctly---the
+  proposal pdf must satisfy a ``detailed balance'' condition; it
+  must have the property that
+\begin{eqnarray}
+q(\pars'\given\pars) &=& q(\pars\given\pars')
+\quad ;
+\end{eqnarray}
+that is, it must be just as easy to go one way in the parameter space
+  as the other.
+You can break this property, if you like, and then adjust the
+  acceptance condition accordingly (which is truly the Metropolis--Hastings algorithm), but we \emph{do not recommend this}
+  except under the supervision of a trained professional.\note{Actually,
+  our own favorite method, \project{emcee}, has a proposal pdf that does
+  violate detailed balance in this way and has a compensated
+  acceptance probability.  Read the paper \cite{emcee} for more details.}
+The reason is:  It is one thing to draw samples from $q(x'\given x)$.
+It is another thing to correctly write down $q(x'\given x)$.
+  If you violate detailed balance in $q(x'\given x)$ then you have to be
+  able to \emph{both} draw from \emph{and} write down $q(x'\given x)$
+  and you are subject to a set of new possible bugs for your code.
+If the detailed balance condition frightens you (as it frightens us),
+  then just stick with pdfs that are symmetric in $\pars'$ and $\pars$,
+  like the Gaussian, or a centered uniform distribution and forget about
+  it.
+
+In what follows, when we discuss tuning of the proposal pdf, we will
+  often cycle through the parameter blob components and propose changes
+  in only one dimension or element or component at a time.
+That is, at step $k+1$ you might only propose a one-dimensional move
+  in the $i$th dimension, not in all $D$ dimensions of the parameter
+  space.
+That won't change anything significant; but if you are implementing
+  this right now you might want to do that.
+It will help us tune the proposal pdf (\sectionname~\ref{sec:tuning})
+and diagnose problems (\sectionname~\ref{sec:trouble}).
+
+One note to make here
+  is that you must \emph{propose}
+  in the same coordinates (parameterization or transformation of parameters)
+  as that in which your priors are \emph{specified},
+  or else multiply in a (possibly nasty) Jacobian.
+That is, if your prior is ``flat in $\pars$''
+  but you find it easier to run your sampler by proposing steps in $\ln \pars$,
+  if you don't modify your acceptance probability by the Jacobian,
+  your \emph{real} prior won't be flat in $\pars$.
+These think-os can get subtle;
+  the best way to avoid them
+  is to write your code and your likelihood function
+  and your prior pdf function and your proposal distribution
+  all in precisely the same parameterization.
+We will return to this point below in our comments on testing (\sectionname~\ref{sec:trouble}).
+
+As we discuss briefly below (\sectionname~\ref{sec:methods}),
+  there are many \MCMC\ methods more advanced than M--H.
+However, they all share characteristics with M--H
+  (initialization, tuning, judging convergence),
+  such that it is very valuable to understand \MH\ well before using anything more advanced.
+Furthermore, a scientist new to \MCMC\ benefits enormously from
+  building, tuning, and using her or his own \MCMC\ software.
+One piece of advice we give then,
+  to the new user of \MCMC,
+  is to code up, tune, and use a \MH\ \MCMC\ sampler for a scientific project.
+A huge amount is learned in doing this,
+  and it is very often the case that the home-built \MH\ \MCMC\ does everything needed;
+  you often don't need any more advanced tool.
+Only after you have concluded that your home-built \MH\ \MCMC\ sampler
+  is not suited to your project
+  (or not developed properly into a properly versatile software package)
+  should you download and start to use any professionally developed alternative.
+That is, even if---in the end---%
+  you want to leave the sampling code to the experts and run an industrial-strength code,
+  it is still valuable to build your own,
+  given the simplicity of the algorithm,
+  and given the intuition you gain by doing it (at least once) yourself.\note{%
+    We also make this point in a previous \documentname\ (\citealt{fittingaline});
+    the ambitious reader will find in that \documentname\ a solid example project
+    for using \MCMC\ in a real context.}
+
+DFM says: should we say something about using the log density here?
+
+HOGG: ...You must describe your priors in the same parameterization that you ``step'' in.
+That is, if you change your parameters, you \emph{must} change your priors correspondingly.
+Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
+
+\begin{problem}\label{prob:MH}
+In your scientific programming language of choice, write a very simple
+M-H \MCMC\ sampler.
+Sample in a single parameter $x$ and give the sampler as its density
+function $p(x)$ a Gaussian density with mean 2 and variance 2.
+(Note that variance is the \emph{square} of the standard deviation.)
+Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
+pdf with mean zero and variance 1.
+Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
+Plot the results as a histogram, with the true density over-plotted sensibly.
+The resulting  plot should look something like \figurename~\ref{fig:MH}.
+\end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/MH}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH}:
+MCMC samples from a Gaussian (black) and the true distribution (blue).}
+\label{fig:MH}
+\end{figure}
+
+\begin{problem}\label{prob:MH2}
+Re-do \problemname~\ref{prob:MH} but now with an input density
+that is uniform on $3<x<7$ and zero everywhere else.
+The plot should look like \figurename~\ref{fig:MH2}.
+What change did you have to make to the initialization, and why?
+\end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/MH2}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:MH2}:
+MCMC samples from a uniform distribution (black) and the true distribution
+(blue).}
+\label{fig:MH2}
+\end{figure}
+
+\begin{problem}\label{prob:twod}
+Re-do \problemname~\ref{prob:MH} but now with an input density
+that is a function of two variables $(x, y)$.
+For the density function use two different functions.
+\emph{(a)} The first density function is a covariant two-dimensional Gaussian
+density with variance tensor
+\begin{eqnarray}
+V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
+\quad.
+\end{eqnarray}
+\emph{(b)} The second density function is a rectangular top-hat function that
+is uniform on the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
+For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
+density with zero mean and variance tensor set to the two-dimensional
+identity matrix.
+Plot the two one-dimensional histograms and also a two-dimensional scatter
+plot for each sampling.
+\figurename~\ref{fig:twod-a} shows the expected results for the Gaussian.
+Make a similar plot for the top-hat.
+\end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{problems/twod_a}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:twod}a:
+MCMC samples from a two-dimensional Gaussian (scatter plot) and the
+one-dimensional marginalized distributions (histograms).}
+\label{fig:twod-a}
+\end{figure}
+
+\begin{problem}\label{prob:MH-sigma}
+Re-do \problemname~\ref{prob:twod}a but with different values for
+the variance of the proposal distribution $q(x'\given x)$.
+What happens when you go to very extreme values (like for instance
+$10^{-1}$ or $10^2$)?
+\end{problem}
+
+\begin{problem}\label{prob:prop-mean}
+Why, in all the previous problems, did we give the proposal
+distributions $q(x'\given x)$ zero mean?
+What would be bad if we hadn't done that?
+Re-do \problemname~\ref{prob:twod}a with a proposal $q(x'\given x)$ with a
+non-zero mean and see what happens.
+Bonus points: Modify the acceptance--rejection criterion to deal with
+the messed-up $q(x'\given x)$ and show that everything works once
+again.
+\end{problem}
+
+\section{Likelihoods and priors}\label{sec:likelihood}
 

### 2016-07-19 problems/MH.py

@@ -12,13 +12,11 @@ setup()  # initialize the plotting styles
 np.random.seed(42)
 
 
-def log_p(x, a=3.0, b=7.0):
-    if a <= x <= b:
-        return -np.log(b-a)
-    return -np.inf
+def log_p(x, mean=2.0, variance=2.0):
+    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
 
 
-x = 5.0
+x = 0.0
 lp = log_p(x)
 chain = np.empty(2e4)
 for step in range(len(chain)):
@@ -31,14 +29,12 @@ for step in range(len(chain)):
 
 
 fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
-ax.axhline(0.0, color="k")
-ax.hist(chain, 50, histtype="step", color=COLORS["DATA"], normed=True)
-x = np.linspace(2.5, 7.5, 5000)
-y = np.exp([log_p(_) for _ in x])
-ax.plot(x, y, color=COLORS["MODEL_1"])
+ax.hist(chain, 100, histtype="step", color=COLORS["DATA"], normed=True)
+x = 2 + np.linspace(-4.5, 4.5, 5000)
+ax.plot(x, np.exp(log_p(x)), color=COLORS["MODEL_1"])
 ax.set_xlim(x.min(), x.max())
-ax.set_ylim(-0.01, y.max() + 0.05)
+ax.yaxis.set_major_locator(pl.MaxNLocator(4))
 ax.set_xlabel("$x$")
 ax.set_ylabel("$p(x)$")
 
-savefig(fig, "p3.pdf")
+savefig(fig, "MH.pdf")

### 2016-07-19 problems/MH2.py

@@ -12,11 +12,13 @@ setup()  # initialize the plotting styles
 np.random.seed(42)
 
 
-def log_p(x, mean=2.0, variance=2.0):
-    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
+def log_p(x, a=3.0, b=7.0):
+    if a <= x <= b:
+        return -np.log(b-a)
+    return -np.inf
 
 
-x = 0.0
+x = 5.0
 lp = log_p(x)
 chain = np.empty(2e4)
 for step in range(len(chain)):
@@ -29,12 +31,14 @@ for step in range(len(chain)):
 
 
 fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
-ax.hist(chain, 100, histtype="step", color=COLORS["DATA"], normed=True)
-x = 2 + np.linspace(-4.5, 4.5, 5000)
-ax.plot(x, np.exp(log_p(x)), color=COLORS["MODEL_1"])
+ax.axhline(0.0, color="k")
+ax.hist(chain, 50, histtype="step", color=COLORS["DATA"], normed=True)
+x = np.linspace(2.5, 7.5, 5000)
+y = np.exp([log_p(_) for _ in x])
+ax.plot(x, y, color=COLORS["MODEL_1"])
 ax.set_xlim(x.min(), x.max())
-ax.yaxis.set_major_locator(pl.MaxNLocator(4))
+ax.set_ylim(-0.01, y.max() + 0.05)
 ax.set_xlabel("$x$")
 ax.set_ylabel("$p(x)$")
 
-savefig(fig, "p2.pdf")
+savefig(fig, "MH2.pdf")

### 2016-07-19 problems/MH_sigma.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from p4a import run_mcmc, log_p_gauss
+from twod_a import run_mcmc, log_p_gauss
 from plot_setup import setup, COLORS, SQUARE_FIGSIZE, savefig
 
 setup()  # initialize the plotting styles

### 2016-07-19 problems/simple_stats.py

@@ -56,4 +56,4 @@ for i, (ax, title) in enumerate(zip(
 
 [ax.set_xlabel(r"$\log_2 K$") for ax in axes[1]]
 
-savefig(fig, "p1.pdf")
+savefig(fig, "simple_stats.pdf")

### 2016-07-19 problems/twod_a.py

@@ -41,4 +41,4 @@ if __name__ == "__main__":
     fig = corner.corner(chain, labels=["$x$", "$y$"],
                         range=[(-4.5, 4.5), (-4.5, 4.5)],
                         plot_density=False, plot_contours=False)
-    savefig(fig, "p4a.pdf", dpi=300)
+    savefig(fig, "twod_a.pdf", dpi=300)

### 2016-07-19 problems/twod_b.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import corner
 import numpy as np
 
-from p4a import run_mcmc
+from twod_a import run_mcmc
 from plot_setup import setup, savefig
 
 setup()  # initialize the plotting styles
@@ -23,4 +23,4 @@ chain, _ = run_mcmc(log_p_uniform, np.array([5.0, 5.0]), nsteps=1e5)
 fig = corner.corner(chain, labels=["$x$", "$y$"],
                     range=[(2.5, 7.5), (0.5, 9.5)],
                     plot_density=False, plot_contours=False)
-savefig(fig, "p4b.pdf", dpi=300)
+savefig(fig, "twod_b.pdf", dpi=300)

### 2016-07-19 mcmc.tex

--- 2016-07-19 mcmc.tex
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{problems/p2}
+\includegraphics[width=0.5\textwidth]{problems/MH}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{problems/p3}
+\includegraphics[width=0.5\textwidth]{problems/MH2}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{problems/p4a}
+\includegraphics[width=0.7\textwidth]{problems/twod_a}
 \end{center}
@@
 
-\begin{problem}
+\begin{problem}\label{prob:prop-mean}
 Why, in all the previous problems, did we give the proposal

### 2016-07-19 problems/MH_sigma.py
renaming problems

@@ -36,4 +36,4 @@ for n, ax in zip([0.0, -1.0, 2.0], axes):
 
 axes[0].set_ylabel("$y$")
 
-savefig(fig, "p5.pdf")
+savefig(fig, "MH_sigma.pdf")

### 2016-07-19 problems/convergence.py

@@ -50,4 +50,4 @@ if __name__ == "__main__":
     ax2.set_ylabel("$p(x)$")
     ax2.set_xlabel("$x$")
 
-    savefig(fig, "p7.pdf", dpi=300)
+    savefig(fig, "convergence.pdf", dpi=300)

### 2016-07-19 problems/estimatetau.py

@@ -8,7 +8,7 @@ import matplotlib.pyplot as pl
 
 from plot_setup import SQUARE_FIGSIZE, savefig
 
-from p7 import chain
+from convergence import chain
 
 
 def autocorr_function(x):
@@ -28,4 +28,4 @@ if __name__ == "__main__":
     ax.set_ylabel("$C_x(\Delta)$")
     ax.set_xlabel("$\Delta$")
 
-    savefig(fig, "p8.pdf")
+    savefig(fig, "estimatetau.pdf")

### 2016-07-19 problems/itertau.py

@@ -8,7 +8,7 @@ import matplotlib.pyplot as pl
 
 from plot_setup import SQUARE_FIGSIZE, COLORS, savefig
 
-from p8 import autocorr_function
+from estimatetau import autocorr_function
 
 
 def autocorr_time_simple(acf, window):
@@ -25,7 +25,7 @@ def autocorr_time_iterative(acf, c=10, low=10):
 
 
 if __name__ == "__main__":
-    from p8 import chain
+    from estimatetau import chain
 
     m = len(chain)
     N = 2 ** np.arange(2, 19)
@@ -50,4 +50,4 @@ if __name__ == "__main__":
     ax.set_ylabel(r"$\tau_x$")
     ax.set_xlabel("window size")
 
-    savefig(fig, "p9.pdf")
+    savefig(fig, "itertau.pdf")

### 2016-07-19 problems/prop_mean.py

@@ -6,7 +6,7 @@ from __future__ import division, print_function
 import numpy as np
 import matplotlib.pyplot as pl
 
-from p4a import log_p_gauss
+from twod_a import log_p_gauss
 from plot_setup import setup, COLORS, SQUARE_FIGSIZE, savefig
 
 setup()  # initialize the plotting styles
@@ -57,4 +57,4 @@ for broken, ax in zip([True, False], axes):
 
 axes[0].set_ylabel("$y$")
 
-savefig(fig, "p6.pdf")
+savefig(fig, "prop_mean.pdf")

### 2016-07-19 problems/tuning.py

@@ -7,7 +7,7 @@ import numpy as np
 import matplotlib.pyplot as pl
 from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
-from p4a import log_p_gauss, run_mcmc
+from twod_a import log_p_gauss, run_mcmc
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -28,4 +28,4 @@ if __name__ == "__main__":
     ax.set_xlabel(r"$\sigma_q$")
     ax.set_ylabel("acceptance fraction")
     ax.set_xlim(0.2, 40)
-    savefig(fig, "p10.pdf")
+    savefig(fig, "tuning.pdf")

### 2016-07-19 problems/tuningtau.py

@@ -7,8 +7,8 @@ import numpy as np
 import matplotlib.pyplot as pl
 from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
 
-from p9 import autocorr_function, autocorr_time_iterative
-from p10 import log_p_gauss, run_mcmc, sigs
+from itertau import autocorr_function, autocorr_time_iterative
+from tuning import log_p_gauss, run_mcmc, sigs
 
 setup()  # initialize the plotting styles
 np.random.seed(42)
@@ -32,4 +32,4 @@ ax.set_ylabel(r"$\tau_x$")
 ax.set_xlim(0.2, 40)
 ax.set_ylim(0, 450)
 ax.yaxis.set_major_locator(pl.MaxNLocator(5))
-savefig(fig, "p11.pdf")
+savefig(fig, "tuningtau.pdf")

### 2016-07-27 mcmc.tex
renaming problems
boldface names
Adding problems for likelihood section

@@
 
-\begin{problem}
+\begin{problem}\label{prob:improper}
 Run your M-H \MCMC\ sampler from \problemname~\ref{prob:MH}, but now
@@
 
-\begin{problem}
+\begin{problem}\label{prob:fittingaline}
+DFM: do this problem.
 For a real-world inference problem, read enough of
@@
 
-\begin{problem}
+\begin{problem}\label{prob:logprior}
 Modify the sampler you wrote in \problemname~\ref{prob:MH} to take
@@
 proposal distribution $q(x'\given x)$ with a Gaussian distribution in
-$\ln x$ $q(\ln x'\given \ln x)$, but---incorrectly---make no other
-changes.  What about your answers change?  How wrong is your answer?
-Bonus points: Compute the analytic function that you have sampled from
-(that is, not $p(x)$) and over-plot it on your histogram.
+$\ln x$ $q(\ln x'\given \ln x)$, but make no other changes.
+By doing this, you are no longer sampling the Gaussian $p(x)$ that you were in
+\problemname~\ref{prob:MH}.
+What about your answers change?  What distribution are you sampling now?
+Compute the analytic function that you have sampled from~--~this will no longer
+be the same $p(x)$~--~and over-plot it on your histogram.
 \end{problem}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{problems/logprior}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:logprior}:
+The black histogram shows the MCMC samples and the blue curve is the analytic
+target density.}
+\label{fig:estimatetau}
+\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{problems/p8}
+\includegraphics[width=0.5\textwidth]{problems/estimatetau}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{problems/p9}
+\includegraphics[width=0.5\textwidth]{problems/itertau}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.48\textwidth]{problems/p10}
-\includegraphics[width=0.48\textwidth]{problems/p11}
+\includegraphics[width=0.48\textwidth]{problems/tuning}
+\includegraphics[width=0.48\textwidth]{problems/tuningtau}
 \end{center}

### 2016-07-27 problems/improper.py

@@ -1,0 +1,34 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+x = 0.0
+chain = np.empty(1e6)
+for step in range(len(chain)):
+    x += np.random.randn()
+    chain[step] = x
+
+# Thin the chain for plotting purposes.
+chain = chain[::1000]
+mn, mx = np.min(chain), np.max(chain)
+mid = 0.5 * (mn + mx)
+rng = 0.5 * (mx - mn) * 1.1
+
+fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+ax.plot(np.arange(len(chain)) * 1e-2, chain, color=COLORS["DATA"])
+ax.yaxis.set_major_locator(pl.MaxNLocator(5))
+ax.set_xlabel("mcmc step [$\\times 10^5$]")
+ax.set_ylabel("$x$")
+ax.set_ylim(mid-rng, mid+rng)
+
+savefig(fig, "improper.pdf")

### 2016-07-27 problems/logprior.py

@@ -1,0 +1,40 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p(x, mean=2.0, variance=2.0):
+    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
+
+
+log_x = 0.0
+lp = log_p(log_x)
+chain = np.empty(2e6)
+for step in range(len(chain)):
+    log_x_prime = log_x + np.random.randn()
+    lp_prime = log_p(log_x_prime)
+    if np.random.rand() <= np.exp(lp_prime - lp):
+        log_x = log_x_prime
+        lp = lp_prime
+    chain[step] = np.exp(log_x)
+
+
+fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+x = np.linspace(0, 20, 5000)
+ax.plot(x, np.exp(log_p(np.log(x))) / x, color=COLORS["MODEL_1"])
+ax.hist(chain, 10000, histtype="step", color=COLORS["DATA"], normed=True)
+ax.set_xlim(x.min(), x.max())
+ax.yaxis.set_major_locator(pl.MaxNLocator(4))
+ax.set_xlabel("$x$")
+ax.set_ylabel("$p(x)$")
+
+savefig(fig, "logprior.pdf")

### 2016-07-29 mcmc.tex
nope

@@
 \noindent
-David~W.~Hogg\\
+\textbf{David~W.~Hogg}\\
 \affil{Simons~Center~for~Data~Analysis,
@@
 \affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}\\[1ex]
-Daniel~Foreman-Mackey\\
+\textbf{Daniel~Foreman-Mackey}\\
 \affil{Sagan Fellow}\\
@@
 \begin{problem}\label{prob:fittingaline}
-DFM: do this problem.
 For a real-world inference problem, read enough of

### 2016-08-29 mcmc.tex
autocorrelation function calculation

@@
 function.
-You can speed this computation up considerably by using a fast Fourier
-transform.
+You will probably want to speed this computation up by using a fast Fourier
+transform\note{%
+The calculation of the autocorrelation function can be seen as a convolution
+and it can, therefore, be computed using the fast Fourier transform in
+$\mathcal{O}(N\,\log N)$ operations instead of $\mathcal{O}(N^2)$ for a naive
+implementation.}.
 Run this on the chain you obtained from \problemname~\ref{prob:MH}.

### 2016-08-29 problems/estimatetau.py

@@ -11,6 +11,20 @@ from plot_setup import SQUARE_FIGSIZE, savefig
 from convergence import chain
 
 
+# This method is really very slow! Probably best to just avoid calling it.
+def autocorr_function_naive(x):
+    print("Warning: running naive autocorrelation function method. "
+          "This will be slow!")
+    mu = np.mean(x)
+    r = x - mu
+    C = np.empty(len(r) // 2)
+    for i in range(1, len(C)):
+        C[i] = np.mean(r[i:] * r[:-i])
+    C /= np.mean(r ** 2)
+    C[0] = 1.0
+    return C
+
+
 def autocorr_function(x):
     n = len(x)
     f = np.fft.fft(x - np.mean(x), n=2*n)

### 2016-09-01 problems/initialization.py
adding initialization problem

@@ -1,0 +1,38 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p(x, mean=2.0, variance=2.0):
+    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
+
+
+x = 1e3
+lp = log_p(x)
+chain = np.empty(6000)
+for step in range(len(chain)):
+    x_prime = x + np.random.randn()
+    lp_prime = log_p(x_prime)
+    if np.random.rand() <= np.exp(lp_prime - lp):
+        x = x_prime
+        lp = lp_prime
+    chain[step] = x
+
+
+fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+ax.plot(chain, color=COLORS["DATA"])
+ax.xaxis.set_major_locator(pl.MaxNLocator(4))
+ax.yaxis.set_major_locator(pl.MaxNLocator(4))
+ax.set_xlabel("step")
+ax.set_ylabel("$x$")
+
+savefig(fig, "initialization.pdf")

### 2016-09-01 mcmc.tex
Adding optimization problem

@@
 origin?
-Did you write the algorithm in terms of $\ln p(x)$ or $p(x)$?
-This makes a difference for extreme initializations!
 What is the scaling:  As you move the initialization further away,
@@
 
-\begin{problem}
+\begin{problem}\label{prob:optimization}
 Import (or write) an optimizer, and upgrade the code you wrote for

### 2016-09-01 problems/optimization.py

@@ -1,0 +1,42 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as pl
+from scipy.optimize import minimize
+
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p(x, mean=2.0, variance=2.0):
+    return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
+
+
+neg_log_p = lambda x: -log_p(x)
+r = minimize(neg_log_p, [1e3])
+
+x = r.x[0]
+lp = log_p(x)
+chain = np.empty(6000)
+for step in range(len(chain)):
+    x_prime = x + np.random.randn()
+    lp_prime = log_p(x_prime)
+    if np.random.rand() <= np.exp(lp_prime - lp):
+        x = x_prime
+        lp = lp_prime
+    chain[step] = x
+
+
+fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+ax.plot(chain, color=COLORS["DATA"])
+ax.xaxis.set_major_locator(pl.MaxNLocator(4))
+ax.yaxis.set_major_locator(pl.MaxNLocator(4))
+ax.set_xlabel("step")
+ax.set_ylabel("$x$")
+
+savefig(fig, "optimization.pdf")

### 2016-09-01 mcmc.tex
removed two dfms

@@
 Markov Chain Monte Carlo (\MCMC) methods for sampling probability
-distribution functions (combined with abundant computational resources)
+density functions (combined with abundant computational resources)
 have transformed the sciences.
@@
 
-As we noted above (\sectionname~\ref{sec:when}), in the context of \MCMC, we are often sampling or
+As we noted above (\sectionname~\ref{sec:when}), in the context of \MCMC, we are often
 using some \emph{badly normalized} function $f(\pars)$.
@@
 \end{eqnarray}
-
-DFM says: I don't agree with this section. Using MCMC you \emph{are} sampling
-from $p(\pars)$\ldots you \emph{can't} sample from $f(\pars)$.
 
@@
 like this:
-% DFM: Do you have a preferred environment for code other than verbatim?
 \begin{verbatim}
@@
 \begin{problem}\label{prob:rosenbrock}
-Look up the Rosenbrock density used as the demonstration case for the
-\project{emcee} sampler (see DFM: should cite G\&W here).
+Look up the Rosenbrock density used as a demonstration case for many samplers (see, for example, \citealt{gw}).
 Tune the Gaussian proposal distribution in your \MH\ \MCMC\ sampler to sample this density efficiently.
@@
   in \citet{mcmchandbook}, 3--48.
+\bibitem[Goodman \& Weare(2010)]{gw}
+  Goodman, J. \& Weare, J., 2010,
+  Ensemble samplers with affine invariance,
+  \textit{Comm.\.Appl.\,Math.\,Comp.\,Sci.}, \textbf{5}(1), 65–-80.
 \bibitem[Hogg \etal(2010a)]{fittingaline}

### 2016-09-02 mcmc.tex
typo
tweaks

@@
 parameter space.
-In many use cases for \MCMC, the investigator wants to use it to obtain
+In many use cases for \MCMC, the investigator wants to use \MCMC\ to obtain
 uncertainty information about or propagate uncertainty into parameter
@@
 make a representative sampling (dozens or hundreds of times, ideally),
-or you can use a high-end sampler (like nested sampling) that is
+or you can use a high-end sampler (like nested sampling; see \sectionname~\ref{sec:methods}) that is
 designed for such problems.
@@
 
-DFM: IS THIS PARAGRAPH USEFUL OR CLEAR?
-Of course there are some kinds of multi-modalities that aren't a
+Of course there are some kinds of multi-modalities that might not be a
 problem.
@@
 
-The parameter space $\pars$ is usually multi-dimensional, and more
-than one of those parameters is of interest.
+The parameter space $\pars$ is usually multi-dimensional, and usually it is more
+than one of those parameters that is of interest.
 In this case, the mean or median of sampling can be produced for each

### 2016-09-02 mcmc.tex
rosenbrock example

--- 2016-09-01 mcmc.tex
@@
 \begin{eqnarray}\label{eq:tauest}
-\tau &=& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
+\tau &=& 1 + 2\,\sum_{\Delta = 1}^{M} C(\Delta)
 \end{eqnarray}
@@
 \begin{problem}\label{prob:rosenbrock}
-Look up the Rosenbrock density used as a demonstration case for many samplers (see, for example, \citealt{gw}).
+The Rosenbrock density used as a demonstration case for many samplers (see, for example, \citealt{gw}).
+Test your sampler on this density:
+\begin{eqnarray}
+p(\theta_1,\,\theta_2) &\propto&
+    \exp\left(-\frac{100\,(\theta_2-{\theta_1}^2)^2+(1-\theta_1)^2}{20}\right)
+\quad.
+\end{eqnarray}
 Tune the Gaussian proposal distribution in your \MH\ \MCMC\ sampler to sample this density efficiently.
@@
 
-\begin{problem}
+\begin{problem}\label{prob:rosenbrock2}
 Execute \problemname~\ref{prob:rosenbrock}, or---if you are lazy---%

### 2016-09-04 mcmc.tex
building on travis

--- 2016-09-02 mcmc.tex
@@
 target density.}
-\label{fig:estimatetau}
+\label{fig:logprior}
 \end{figure}

### 2016-09-04 mcmc.tex
refs and dois

@@
 \project{emcee}\note{\citet{emcee}}, which is popular in the
-astrophysical sciences.
+astrophysical sciences, but there are others\note{One that we are
+  interested in is \project{Kombine} (Farr \& Farr, in preparation).}.
 It is worty of note that ensemble methods have particular
@@
 Sometimes it is possible to sample efficiently from a distribution
-that is close to the target function $f(\pars)$ you care about but
+that is close to the target function $f(\pars)$ you care about, even when the target $f(\pars)$ itself is
 hard to sample efficiently with \MCMC.
@@
 
+% DFM: Is the last sentence of the above paragraph correct??
+
 \paragraph{Nested:}
 Similar to tempering-like methods are a class of methods called ``nested''
-samplers, which also smoothly change the target distribution away from $f(\pars)$.
+samplers\note{For example, see \citet{skilling} and \citet{dnest}},
+which also smoothly change the target distribution away from $f(\pars)$.
 In the nested case, instead of increasing the temperature, the
@@
   \isbn{978-0471810339}.
+\bibitem[Brewer \etal(2011)]{dnest}
+  Brewer,~B.~J., P\'artay,~L.~B. & Cs\'anyi,~G., 2011,
+  {Diffusive nested sampling},
+  \textit{Stat.\,Comput.}\ \textbf{21} 649
+  \doi{10.1007/s11222-010-9198-8}.
 \bibitem[Brewer \etal(2013)]{brewer}
@@
   {Probabilistic catalogs for crowded stellar fields},
-  \textit{Astron.\,J.} \textbf{146} 7
+  \textit{Astron.\,J.}\ \textbf{146} 7
   \doi{10.1088/0004-6256/146/1/7}.
@@
   Ensemble samplers with affine invariance,
-  \textit{Comm.\.Appl.\,Math.\,Comp.\,Sci.}, \textbf{5}(1), 65–-80.
+  \textit{Comm.\.Appl.\,Math.\,Comp.\,Sci.}, \textbf{5}(1), 65–-80
+  \doi{10.2140/camcos.2010.5.65}.
 \bibitem[Hogg \etal(2010a)]{fittingaline}

### 2016-09-21 mcmc.tex
skilling
frenchspacing
Added comments

@@
 \newcommand{\pars}{\theta}
+\newcommand{\bl}[1]{{\color{red}{[BL:] #1}}}
 \newcommand{\best}{{\mathrm{(best)}}}
@@
 \end{abstract}
+
+\bl{
+Overall comments:
+\begin{itemize}
+\item Would it be useful to mention that MCMC is useful when you cannot or don't want to evaluate the posterior on a grid?
+\item $p(\theta)$ is used in place of $p(\theta|D)$ in various places.
+This is mildly perturbing - I would either change it to $p(\theta|D)$ everywhere or explicitly say that the shorter convention $p(\theta)$ is used.
+\item How much do you want to go into non Metropolis-H algorithms?
+At the moment this is either not sufficiently or too detailed.
+I would either reduce it to a brief enumeration or to a discussion of features, advantages and drawbacks of various methods.
+I am happy to help with that.
+However, it might be hard to avoid discussing specific technologies and codes.
+Worse: I am not convinced by the practical usefulness of a generic summary of all sampling methods.
+\end{itemize}
+}
 
@@
 It is so simple that we recommend that any reader of this document
-  who has not previously implemented the algorithm take a break at the end
+  who has not previously implemented the algorithm takes a break at the end
   of this \sectionname\ and implement it forthwith, in a short piece of computer code,
@@
 That is, if you change your parameters, you \emph{must} change your priors correspondingly.
-Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
+Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$. \bl{Isn't that discussed somewhere below?}
 
@@
 \begin{problem}\label{prob:itertau}
-Write a peice of code that estiamtes the integrated autocorrelation time for a
+Write a piece of code that estimates the integrated autocorrelation time for a
 chain of samples using an estimate of the autocorrelation function and a given
@@
 Some are global parameters, which touch or have an effect on every
-sub-part of the data blob $\data$, while others are local parameteres,
+sub-part of the data blob $\data$, while others are local parameters,
 which only touch a small subset of the data.
@@
 \bibitem[Brewer \etal(2011)]{dnest}
-  Brewer,~B.~J., P\'artay,~L.~B. & Cs\'anyi,~G., 2011,
+  Brewer,~B.~J., P\'artay,~L.~B. \& Cs\'anyi,~G., 2011,
   {Diffusive nested sampling},
@@
   \textit{Statistica Sinica}, \textbf{20}(1) 343--364.
+\bibitem[Skilling(2006)]{skilling}
+  Skilling, J., 2006,
+  Nested sampling for general bayesian computation,
+  \textit{Bayes.\,Anal.}\ \textbf{1}(4) 833-–860
+  \doi{10.1214/06-BA127}.
 \bibitem[van~der~Walt \etal(2011)]{numpy}

### 2016-10-14 mcmc.tex
swap section order
mcmc

--- 2016-09-04 mcmc.tex
@@
 % - Go through github issues, if any.
-% - Systematically cite chapters of Brooks \etal in each section.
-% - Skim MacKay and make sure we cite him in any relevant place.
-% - Mention & cite STAN in methods section
 % - Need to mention case where there is an MC integral INSIDE the likelihood function
 % - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
-% - Make sure we say ``don't use MCMC for 1-d problems or if you have alternatives to directly integrate!''
-% - Would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
 % - Send to friendlies for comments
@@
 
-\begin{document}\sloppy\sloppypar\raggedbottom\thispagestyle{plain}%
+\begin{document}\sloppy\sloppypar\raggedbottom\frenchspacing\thispagestyle{plain}%
 %
@@
 \caption{Solution to \problemname~\ref{prob:MH}:
-MCMC samples from a Gaussian (black) and the true distribution (blue).}
+\MCMC\ samples from a Gaussian (black) and the true distribution (blue).}
 \label{fig:MH}
@@
 \caption{Solution to \problemname~\ref{prob:MH2}:
-MCMC samples from a uniform distribution (black) and the true distribution
+\MCMC\ samples from a uniform distribution (black) and the true distribution
 (blue).}
@@
 \caption{Solution to \problemname~\ref{prob:twod}a:
-MCMC samples from a two-dimensional Gaussian (scatter plot) and the
+\MCMC\ samples from a two-dimensional Gaussian (scatter plot) and the
 one-dimensional marginalized distributions (histograms).}
@@
 \caption{Solution to \problemname~\ref{prob:logprior}:
-The black histogram shows the MCMC samples and the blue curve is the analytic
+The black histogram shows the \MCMC\ samples and the blue curve is the analytic
 target density.}
@@
 It is also worth noting that there is not just one autocorrelation function or
-autocorrelation time for an MCMC chain.
+autocorrelation time for an \MCMC\ chain.
 Instead, there is an autocorrelation time $\tau_f$ for every function
@@
 Estimates of the integrated autocorrelation time of different segments of the
-MCMC chain from \problemname~\ref{prob:MH} (black lines) and for the full
+\MCMC\ chain from \problemname~\ref{prob:MH} (black lines) and for the full
 chain (orange line) as a function of window size $M$.
@@
 
+\section{Troubleshooting and advice}\label{sec:trouble}
+
+When issues arise with \MCMC\ sampling,
+  they are sometimes difficult to diagnose:
+Is it the \MCMC\ code, the priors, the likelihood function,
+  the initialization, the tuning, or something else?
+
+\paragraph{functional testing:}
+In addition to a full set of unit tests for every part of your code
+  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
+  we recommend building some end-to-end functional tests that permit you to perform \MCMC\
+  runs with output that must meet certain expectations.
+The simplest kind of functional test is to sample a known distribution,
+  and check that the sampling has the moments you expect (to within tolerances).
+For example, you can sample a $D$-dimensional Gaussian distribution with known non-trivial covariance,
+  and check that the empirical covariance comes out as expected.
+This is an easy test to fail, so a success builds confidence in any \MCMC\ code.
+
+In the most extreme form of this test,
+  for a Gaussian in $D$ dimensions,
+  build the covariance matrix input to the Gaussian distribution
+  by taking the sum of the outer product of $D+1$ randomly generated vectors.
+Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
+Set up the \MCMC\ sampler with the Gaussian function with this covariance matrix
+  as the probability function (or the log of it as the log-probability function).
+Sample for many autocorrelation times,
+  or until you are confident you are getting good samples.
+Check that the empirical $D$-dimensional covariance of the samples
+  has the same (or similar) eigenvectors and eigenvlues to the input matrix.
+Also check that the individual one-dimensional samplings
+  (along each axis or any projection in the $D$-space)
+  look Gaussian and have nearly vanishing skew and kurtosis.
+It takes some thinking and some calculating to see ``how well'' these things all ought to match
+  expectations, but one thing that can be said with confidence is that agreement should improve
+  as the sampling proceeds.
+If it doesn't, it is likely that there is something broken about the sampler itself.
+
+Another important functional test is to \emph{sample the prior pdf}.
+This tests the sampler,
+  tests that your priors really are what you think they are
+  (and there are so many ways for them \emph{not} to be\note{It is
+    easy to think you have flat priors when in practice your sampler
+    has flat priors in the logarithms of the parameters or inverses of
+    the parameters.  Also, it is easy to have improper priors when you
+    think they are proper, because limits are not working, or you were
+    wrong about the convergence of some integral.  Sometimes the prior pdf in practice has
+    different edges than you think because there is some censoring
+    happening that you haven't considered.  And then, of course, it is
+    possible for your sampler itself to be buggy.}), and tests that
+  your prior pdfs are proper (functions that can be normalized).
+This test is most effective as a true end-to-end test of your model
+  if you structure your code such that the log probability function \code{lnprob}
+  input to the \MCMC\ scheme
+  is set up as a sum of a log likelihood function \code{lnlikelihood}
+  and a prior function \code{lnprior}.
+The sampler can be switched to sampling the prior by a simple replacement
+  of the \code{lnlikelihood} function
+  with a trivial function that always returns zero.
+
+\MCMC\ should run well in this flat-likelihood case (priors are usually easy to sample)
+  and the output sample histograms should look like what you expect
+  given the prior pdfs you coded.
+If the walkers run off to positive or negative infinity,
+  and nothing seems to converge,
+  your priors are probably not proper.
+If the priors don't look as you expect,
+  you either have a bug in your code,
+  or else a think-o about how your priors ought to look.
+Either way, diagnosis in order.
+
+\paragraph{visualization:}
+HOGG: What plots do you make to diagnose problems?
+
+HOGG: Plot parameters vs ``time'' in the chain.
+
+HOGG: Plot a cut through the density function and the samples near that cut.
+
+HOGG:...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?
+
+\paragraph{likelihood issues:}
+Sometimes there can be problems with the likelihood function itself.
+For example,
+  when you re-call the likelihood function with the same parameters,
+  do you get \emph{exactly} the same answer?
+If you don't, your likelihood function effectively depends on some random numbers;
+  maybe it includes within it an integral performed by Monte Carlo method?
+In general, you want to change your likelihood function such that,
+  even if it includes a numerical or stochastic integral,
+  it is designed such that it produces \emph{precisely} the same
+  value when it returns to the same point in parameter space.
+One way to make this happen is to choose the random numbers
+  (those used for any internal integration inside the likelihood function)
+  in advance and re-use them on every likelihood call.
+This provides the same integral according to the same approximation,
+  but both speeds up the code (no re-generation of random numbers)
+  and also makes the likelihood function single valued.
+Even better is to replace internal Monte Carlo integrals with
+  deterministic numerical (or even better analytic) integrals.
+
+If things are giving trouble and you suspect the likelihood function,
+  another important test is to visualize slices through parameter space.
+When you call the likelihood function on a grid in each parameter
+  (with the others fixed, say, to reasonable values),
+  you should see a likelihood value that is a smooth function of each parameter,
+  or at least as smooth as you expect.
+Things like numerical integration,
+  truncated expansions,
+  or adaptive approximations
+  inside likelihood functions
+  can make the function noisy or jagged at small scales in parameter space.
+These problems hurt optimizers more than samplers,
+  but they usually point to code issues.
+In general, plotting the likelihood function as a function of parameter along
+  slices in parameter space often reveals bugs or think-os.
+
+DFM:...SHOW EXAMPLES?
+
+\paragraph{low acceptance fraction:}
+If you find that you are getting low acceptance fraction
+  (in standard M-H \MCMC\ or other varieties for which there is such a concept)
+  you can simply reduce the sizes of the steps you consider to increase the fraction that are accepted.
+This means decreasing the variance of the proposal distribution,
+  or whatever parameter or parameters control the width of that distribution.
+If reducing the variance of the proposal distribution does \emph{not} increase the acceptance fraction,
+  then \emph{you have a bug}.
+You must find that bug by auditing your code or else returning to the functional tests listed above.
+
+Similarly for high acceptance fraction:
+If you are finding a high acceptance fraction,
+  then increasing the variance of the proposal distributon \emph{must} decrease the acceptance fraction.
+If it does \emph{not}, then \emph{you have a bug}.
+
+\paragraph{sticky chains:}
+If you are concerned about acceptance fraction or convergence,
+  it makes sense to plot parameter values as a function of ``step number'' or iteration number.
+That is, plot the ordered chain in each parameter dimension.
+These plots will have long horizontal patches if the chain is getting stuck;
+  a converged chain will show the walker traversing the parameter space in every dimension fully many times
+  over the length of the run.
+Not a few times; many times.
+If you are using an ensemble sampler\note{Such as \project{emcee} \citep{emcee};
+  one nice thing about the ensemble sampler is that it gives you great diagnostic information automatically.}
+  make plots showing all $M$ walkers, each a different color.
+A converged run for an ensemble sampler will show every single walker traversing the parameter space
+  in every dimension many times.
+
+DFM: SHOW EXAMPLES OF GOOD AND BAD?
+
+\paragraph{initialization-dependence:}
+If you initialize your \MCMC\ runs in different places in parameter space,
+  do you get the same (or very similar) final parameter samplings,
+  in mean and variance?
+If you don't,
+  then your posterior pdf is badly multi-modal;
+  trivial \MCMC\ methods probably won't fix your problem.
+Not to put too fine a point on it:
+If your results are initialization-dependent,
+  then your \MCMC\ runs \emph{are not converged}.
+
+You either have to go to a method that tries to explore the posterior
+  more liberally, such as nested sampling or simulated tempering,
+  or else split your ``model'' up into several sub-models,
+  each of which contains different posterior modes.
+In general, this problem is hard to fix;
+  as we mentioned above, it is provably impossible
+  to explore all of parameter space if the posterior pdf is
+  complex in morphology.
+We don't have very useful advice here, except to
+  spend time learning about the multiple modes in the posterior pdf
+  and what each ``means'' or corresponds to,
+  and to do your best to express in your results the multiple optima.
+One of the miracles of \MCMC\ is that if you \emph{can} fairly sample
+  as complex posterior pdf,
+  the fraction of samples in each mode
+  tends to the relative total integrated posterior probability inside each mode,
+  as the sampling converges.
+
+Ensemble samplers react to multiple modes
+  by obtaining very low acceptance ratio,
+  because the samplers in the different modes are not easily able to ``help one another''
+  move efficiently.
+Again, if the acceptance ratio is low and there appear to be multiple modes,
+  it is likely that the sampling will also be initialization dependent,
+  and it is very unlikely that your sampling is going to be converged.
+
+\paragraph{bad initialization:}
+If you initialize your sampler in a disallowed region of parameter
+space---that is, a part of the space with $f(\pars)=0$ (so your
+\code{ln_f()} function returns \code{-Inf})---it might have a very
+hard time random-walking out of that location.
+At initialization time, it is important to test that the
+initialization is permitted.
+For ensemble methods, it is important to test that \emph{all} the
+walkers in the ensemble are initialized to permitted values.
+
+\paragraph{parameterization problems:}
+It helps to think carefully about how to parameterize.
+For example, if there is an angle $\phi$ in your problem,
+  and your prior requires it to be between $0$ and $2\pi$,
+  and the posterior mode is very near $0$,
+  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
+  makes an intrinsically unimodal problem multimodal!
+In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
+  to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
+Of course this re-parameterization changes significantly the \emph{form} of the prior pdf
+  (the transformation brings in a Jacobian);
+  it has to be adjusted with care.
+
+DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?
+Choose a $p(A)$ and a $p(\phi) = [2\pi]^{-1}$,
+   and then transform to $p(a, b)$.
+
+%% ...Priors: Once again:
+%% If you describe your priors in one parameterization but
+%% then perform sampling in a different space, it is easy to make
+%% mistakes.  For example: Flat prior in amplitude and phase, sample in
+%% $A\,\cos\phi$ and $A\,\sin\phi$.  There are many ways to do this
+%% wrong.
+
+For another example, sometimes there is almost an exact degeneracy between two parameters,
+  say $a$ and $b$.  Then it makes sense to switch to parameters $a+b$ and $a-b$,
+  or some other linear combinations where correlations or near-degeneracies are likely to be reduced.
+Again, such transformations require also prior transformations, which must be made with care.
+There are some affine invariant samplers\note{Note how much we like to cite \citet{emcee}!} that are invariant to such transformations,
+  but if you aren't using such a sampler,
+  it is worth getting out ahead of these problems.
+They can be found by performing an exploratory sampling,
+  making a triangle plot as described above,
+  and looking for narrow diagonal lines in the two-dimentional scatter plots.
+
+HOGG: ...How to deal with discrete or integer or binary parameters?
+
+\paragraph{model checking:}
+\emph{All models are wrong, but some are useful.}
+That's not a quote from Box, but it's close.\note{The real quotation
+  is a parenthetical sentence ``Remember that all models are wrong;
+  the practical question is how wrong do they have to be to not be
+  useful.'' (\citealt{box}, p.~74).  It is amazing how rarely this is
+  ever quoted correctly; or maybe there is another source for this
+  quotation?}
+The point (for our purposes) is that if you check your model hard
+enough---that is, you take enough data---you will certainly rule it
+out.\note{Physicists sometimes forget this, and believe that various
+  physical theories, such as quantum electrodynamics or the
+  cold-dark-matter cosmological model, are strictly true.  First of
+  all, there is no way to know that for sure, and second of all, even
+  if they \emph{are} true, they don't precisely explain any
+  observation, which is also affected by various kinds of auxilliary
+  effects and noises for which there is (and can be, for deep reasons,
+  to be discussed in other \documentname s in this series) no extremely precise model.  So in
+  detail, any model of any specific observation must be wrong, even in
+  the (exceedingly unlikely) event that the fundamental model is
+  correct.}
+``Model checking'' is an enormous subject that goes way beyond the
+scope of this \documentname, for which reason we won't say much about
+it here except to make two points:
+
+The first---and most important---point about model checking is that
+you always learn a huge amount by checking the model.
+What this entails is looking at the distribution of residuals or the
+quality of the predictions of the model when compared to the data.
+There are methods (like the chi-squared statistic and the Bayesian
+evidence integral) that check the absolute quality of the model in a
+scalar way.
+These are usually uninformative, because (as we noted), all models are
+wrong, and eventually the data will be good enough to return a bad
+statistic here.\note{Sometimes Bayesians like to complain that the
+  chi-squared statistic is a measure of the size of your data!  That's
+  true when your model is wrong (that is, always).}
+
+Much more informative are methods (like plotting residuals of the data
+away from the model in the space of the data) that look at what parts
+of the data the model ``explains well'' and what parts of the data the
+model ``explains badly''.
+These kinds of experiments usually reveal inadequacies of the model in
+a rich way, or confirm that certain assumptions are bad.
+Cross-validation\note{Cross-validation is an incredibly valuable set
+  of techniques that all astronomers should know.
+  it is not clear what to cite here, but one possibility is \citet{crossval}.}
+is a framework for constructing
+valuable tests of this form, as is posterior predictive checks\note{%
+  The posterior predictive check is one of the Bayesian equivalents of
+  a goodness-of-fit test \citep{ppcheck}.}
+Importantly, you want to find ways to compare your model to your data
+\emph{in the space of the data}.
+Don't just look at the parameter space of your model!
+It is the data that really exist, especially when all models are wrong!
+
+There are model checking methods (like varying the prior and
+re-running \MCMC) that test the sensitivity of results to assumptions.
+Not only are these very valuable for testing the model, they are
+essential for writing any paper about your results: The reader always
+wants to know the sensitivity of results to assumptions.\note{And be
+  sure not to forget that there are (at least) as many assumptions
+  encoded in your likelihood function as there are in your prior
+  pdfs.}
+Once again, it is important to find ways to display the results of
+such tests in the space of the data.
+All we will say here about these kinds of rich model tests is that you
+should do them, and report the results.
+
+The second---and less important---point about model checking is that
+the ur-Bayesian thing to be doing is computing a Bayesian evidence
+integral.
+However, it is usually inadvisable:
+The integral is expensive to compute.
+Most \MCMC\ methods are cleverly designed to \emph{avoid} computing this
+integral (as we note above in \sectionname~\ref{sec:when}); so it usually adds
+expense to the project.
+At the end of the day, it returns only a single, scalar number, with
+no guidance about how to use it (and indeed there are only heuristics
+to reach for).
+If your model is wrong, the richer tests of visualization of residuals
+or sensitivity to assumptions are much more likely to lead to insight
+about what changes to make or what new directions to explore.
+
+%% HOGG DELETED / COMMENTED OUT THE BELOW
+%% - make sure everything below is discussed somewhere above and then delete it below.
+
+%% ...breaking Markov-ness.
+%% % ***** BJB
+%% % "Diminishing adaptations" is an idea that is quite easy
+%% % to explain. There are other ways to break Markov-ness but
+%% % this probably covers enough of them for an intro document
+%% % ***** Hogg
+%% % I was going to say you should NEVER break Markov-ness
+%% % unless you REALLY understand what you are doing.
+
+%% ...Posteriors:  You can do anything you like (within
+%% reason) to the output samples and be sampling the posterior in that
+%% transformed quantity.
+
+%% ...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.
+
+%% ...Finally, to repeat: You can't sample the likelihood!  You can only
+%% sample a posterior pdf!
+
 \section{Gibbs, ensemble, Hamiltonian, and so on}\label{sec:methods}
@@
 
-\section{Troubleshooting and advice}\label{sec:trouble}
-
-When issues arise with \MCMC\ sampling,
-  they are sometimes difficult to diagnose:
-Is it the \MCMC\ code, the priors, the likelihood function,
-  the initialization, the tuning, or something else?
-
-\paragraph{functional testing:}
-In addition to a full set of unit tests for every part of your code
-  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
-  we recommend building some end-to-end functional tests that permit you to perform \MCMC\
-  runs with output that must meet certain expectations.
-The simplest kind of functional test is to sample a known distribution,
-  and check that the sampling has the moments you expect (to within tolerances).
-For example, you can sample a $D$-dimensional Gaussian distribution with known non-trivial covariance,
-  and check that the empirical covariance comes out as expected.
-This is an easy test to fail, so a success builds confidence in any \MCMC\ code.
-
-In the most extreme form of this test,
-  for a Gaussian in $D$ dimensions,
-  build the covariance matrix input to the Gaussian distribution
-  by taking the sum of the outer product of $D+1$ randomly generated vectors.
-Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
-Set up the \MCMC\ sampler with the Gaussian function with this covariance matrix
-  as the probability function (or the log of it as the log-probability function).
-Sample for many autocorrelation times,
-  or until you are confident you are getting good samples.
-Check that the empirical $D$-dimensional covariance of the samples
-  has the same (or similar) eigenvectors and eigenvlues to the input matrix.
-Also check that the individual one-dimensional samplings
-  (along each axis or any projection in the $D$-space)
-  look Gaussian and have nearly vanishing skew and kurtosis.
-It takes some thinking and some calculating to see ``how well'' these things all ought to match
-  expectations, but one thing that can be said with confidence is that agreement should improve
-  as the sampling proceeds.
-If it doesn't, it is likely that there is something broken about the sampler itself.
-
-Another important functional test is to \emph{sample the prior pdf}.
-This tests the sampler,
-  tests that your priors really are what you think they are
-  (and there are so many ways for them \emph{not} to be\note{It is
-    easy to think you have flat priors when in practice your sampler
-    has flat priors in the logarithms of the parameters or inverses of
-    the parameters.  Also, it is easy to have improper priors when you
-    think they are proper, because limits are not working, or you were
-    wrong about the convergence of some integral.  Sometimes the prior pdf in practice has
-    different edges than you think because there is some censoring
-    happening that you haven't considered.  And then, of course, it is
-    possible for your sampler itself to be buggy.}), and tests that
-  your prior pdfs are proper (functions that can be normalized).
-This test is most effective as a true end-to-end test of your model
-  if you structure your code such that the log probability function \code{lnprob}
-  input to the \MCMC\ scheme
-  is set up as a sum of a log likelihood function \code{lnlikelihood}
-  and a prior function \code{lnprior}.
-The sampler can be switched to sampling the prior by a simple replacement
-  of the \code{lnlikelihood} function
-  with a trivial function that always returns zero.
-
-\MCMC\ should run well in this flat-likelihood case (priors are usually easy to sample)
-  and the output sample histograms should look like what you expect
-  given the prior pdfs you coded.
-If the walkers run off to positive or negative infinity,
-  and nothing seems to converge,
-  your priors are probably not proper.
-If the priors don't look as you expect,
-  you either have a bug in your code,
-  or else a think-o about how your priors ought to look.
-Either way, diagnosis in order.
-
-\paragraph{visualization:}
-HOGG: What plots do you make to diagnose problems?
-
-HOGG: Plot parameters vs ``time'' in the chain.
-
-HOGG: Plot a cut through the density function and the samples near that cut.
-
-HOGG:...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?
-
-\paragraph{likelihood issues:}
-Sometimes there can be problems with the likelihood function itself.
-For example,
-  when you re-call the likelihood function with the same parameters,
-  do you get \emph{exactly} the same answer?
-If you don't, your likelihood function effectively depends on some random numbers;
-  maybe it includes within it an integral performed by Monte Carlo method?
-In general, you want to change your likelihood function such that,
-  even if it includes a numerical or stochastic integral,
-  it is designed such that it produces \emph{precisely} the same
-  value when it returns to the same point in parameter space.
-One way to make this happen is to choose the random numbers
-  (those used for any internal integration inside the likelihood function)
-  in advance and re-use them on every likelihood call.
-This provides the same integral according to the same approximation,
-  but both speeds up the code (no re-generation of random numbers)
-  and also makes the likelihood function single valued.
-Even better is to replace internal Monte Carlo integrals with
-  deterministic numerical (or even better analytic) integrals.
-
-If things are giving trouble and you suspect the likelihood function,
-  another important test is to visualize slices through parameter space.
-When you call the likelihood function on a grid in each parameter
-  (with the others fixed, say, to reasonable values),
-  you should see a likelihood value that is a smooth function of each parameter,
-  or at least as smooth as you expect.
-Things like numerical integration,
-  truncated expansions,
-  or adaptive approximations
-  inside likelihood functions
-  can make the function noisy or jagged at small scales in parameter space.
-These problems hurt optimizers more than samplers,
-  but they usually point to code issues.
-In general, plotting the likelihood function as a function of parameter along
-  slices in parameter space often reveals bugs or think-os.
-
-DFM:...SHOW EXAMPLES?
-
-\paragraph{low acceptance fraction:}
-If you find that you are getting low acceptance fraction
-  (in standard M-H \MCMC\ or other varieties for which there is such a concept)
-  you can simply reduce the sizes of the steps you consider to increase the fraction that are accepted.
-This means decreasing the variance of the proposal distribution,
-  or whatever parameter or parameters control the width of that distribution.
-If reducing the variance of the proposal distribution does \emph{not} increase the acceptance fraction,
-  then \emph{you have a bug}.
-You must find that bug by auditing your code or else returning to the functional tests listed above.
-
-Similarly for high acceptance fraction:
-If you are finding a high acceptance fraction,
-  then increasing the variance of the proposal distributon \emph{must} decrease the acceptance fraction.
-If it does \emph{not}, then \emph{you have a bug}.
-
-\paragraph{sticky chains:}
-If you are concerned about acceptance fraction or convergence,
-  it makes sense to plot parameter values as a function of ``step number'' or iteration number.
-That is, plot the ordered chain in each parameter dimension.
-These plots will have long horizontal patches if the chain is getting stuck;
-  a converged chain will show the walker traversing the parameter space in every dimension fully many times
-  over the length of the run.
-Not a few times; many times.
-If you are using an ensemble sampler\note{Such as \project{emcee} \citep{emcee};
-  one nice thing about the ensemble sampler is that it gives you great diagnostic information automatically.}
-  make plots showing all $M$ walkers, each a different color.
-A converged run for an ensemble sampler will show every single walker traversing the parameter space
-  in every dimension many times.
-
-DFM: SHOW EXAMPLES OF GOOD AND BAD?
-
-\paragraph{initialization-dependence:}
-If you initialize your \MCMC\ runs in different places in parameter space,
-  do you get the same (or very similar) final parameter samplings,
-  in mean and variance?
-If you don't,
-  then your posterior pdf is badly multi-modal;
-  trivial \MCMC\ methods probably won't fix your problem.
-Not to put too fine a point on it:
-If your results are initialization-dependent,
-  then your \MCMC\ runs \emph{are not converged}.
-
-You either have to go to a method that tries to explore the posterior
-  more liberally, such as nested sampling or simulated tempering,
-  or else split your ``model'' up into several sub-models,
-  each of which contains different posterior modes.
-In general, this problem is hard to fix;
-  as we mentioned above, it is provably impossible
-  to explore all of parameter space if the posterior pdf is
-  complex in morphology.
-We don't have very useful advice here, except to
-  spend time learning about the multiple modes in the posterior pdf
-  and what each ``means'' or corresponds to,
-  and to do your best to express in your results the multiple optima.
-One of the miracles of \MCMC\ is that if you \emph{can} fairly sample
-  as complex posterior pdf,
-  the fraction of samples in each mode
-  tends to the relative total integrated posterior probability inside each mode,
-  as the sampling converges.
-
-Ensemble samplers react to multiple modes
-  by obtaining very low acceptance ratio,
-  because the samplers in the different modes are not easily able to ``help one another''
-  move efficiently.
-Again, if the acceptance ratio is low and there appear to be multiple modes,
-  it is likely that the sampling will also be initialization dependent,
-  and it is very unlikely that your sampling is going to be converged.
-
-\paragraph{bad initialization:}
-If you initialize your sampler in a disallowed region of parameter
-space---that is, a part of the space with $f(\pars)=0$ (so your
-\code{ln_f()} function returns \code{-Inf})---it might have a very
-hard time random-walking out of that location.
-At initialization time, it is important to test that the
-initialization is permitted.
-For ensemble methods, it is important to test that \emph{all} the
-walkers in the ensemble are initialized to permitted values.
-
-\paragraph{parameterization problems:}
-It helps to think carefully about how to parameterize.
-For example, if there is an angle $\phi$ in your problem,
-  and your prior requires it to be between $0$ and $2\pi$,
-  and the posterior mode is very near $0$,
-  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
-  makes an intrinsically unimodal problem multimodal!
-In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
-  to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
-Of course this re-parameterization changes significantly the \emph{form} of the prior pdf
-  (the transformation brings in a Jacobian);
-  it has to be adjusted with care.
-
-DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?
-Choose a $p(A)$ and a $p(\phi) = [2\pi]^{-1}$,
-   and then transform to $p(a, b)$.
-
-%% ...Priors: Once again:
-%% If you describe your priors in one parameterization but
-%% then perform sampling in a different space, it is easy to make
-%% mistakes.  For example: Flat prior in amplitude and phase, sample in
-%% $A\,\cos\phi$ and $A\,\sin\phi$.  There are many ways to do this
-%% wrong.
-
-For another example, sometimes there is almost an exact degeneracy between two parameters,
-  say $a$ and $b$.  Then it makes sense to switch to parameters $a+b$ and $a-b$,
-  or some other linear combinations where correlations or near-degeneracies are likely to be reduced.
-Again, such transformations require also prior transformations, which must be made with care.
-There are some affine invariant samplers\note{Note how much we like to cite \citet{emcee}!} that are invariant to such transformations,
-  but if you aren't using such a sampler,
-  it is worth getting out ahead of these problems.
-They can be found by performing an exploratory sampling,
-  making a triangle plot as described above,
-  and looking for narrow diagonal lines in the two-dimentional scatter plots.
-
-HOGG: ...How to deal with discrete or integer or binary parameters?
-
-\paragraph{model checking:}
-\emph{All models are wrong, but some are useful.}
-That's not a quote from Box, but it's close.\note{The real quotation
-  is a parenthetical sentence ``Remember that all models are wrong;
-  the practical question is how wrong do they have to be to not be
-  useful.'' (\citealt{box}, p.~74).  It is amazing how rarely this is
-  ever quoted correctly; or maybe there is another source for this
-  quotation?}
-The point (for our purposes) is that if you check your model hard
-enough---that is, you take enough data---you will certainly rule it
-out.\note{Physicists sometimes forget this, and believe that various
-  physical theories, such as quantum electrodynamics or the
-  cold-dark-matter cosmological model, are strictly true.  First of
-  all, there is no way to know that for sure, and second of all, even
-  if they \emph{are} true, they don't precisely explain any
-  observation, which is also affected by various kinds of auxilliary
-  effects and noises for which there is (and can be, for deep reasons,
-  to be discussed in other \documentname s in this series) no extremely precise model.  So in
-  detail, any model of any specific observation must be wrong, even in
-  the (exceedingly unlikely) event that the fundamental model is
-  correct.}
-``Model checking'' is an enormous subject that goes way beyond the
-scope of this \documentname, for which reason we won't say much about
-it here except to make two points:
-
-The first---and most important---point about model checking is that
-you always learn a huge amount by checking the model.
-What this entails is looking at the distribution of residuals or the
-quality of the predictions of the model when compared to the data.
-There are methods (like the chi-squared statistic and the Bayesian
-evidence integral) that check the absolute quality of the model in a
-scalar way.
-These are usually uninformative, because (as we noted), all models are
-wrong, and eventually the data will be good enough to return a bad
-statistic here.\note{Sometimes Bayesians like to complain that the
-  chi-squared statistic is a measure of the size of your data!  That's
-  true when your model is wrong (that is, always).}
-
-Much more informative are methods (like plotting residuals of the data
-away from the model in the space of the data) that look at what parts
-of the data the model ``explains well'' and what parts of the data the
-model ``explains badly''.
-These kinds of experiments usually reveal inadequacies of the model in
-a rich way, or confirm that certain assumptions are bad.
-Cross-validation\note{Cross-validation is an incredibly valuable set
-  of techniques that all astronomers should know.
-  it is not clear what to cite here, but one possibility is \citet{crossval}.}
-is a framework for constructing
-valuable tests of this form, as is posterior predictive checks\note{%
-  The posterior predictive check is one of the Bayesian equivalents of
-  a goodness-of-fit test \citep{ppcheck}.}
-Importantly, you want to find ways to compare your model to your data
-\emph{in the space of the data}.
-Don't just look at the parameter space of your model!
-It is the data that really exist, especially when all models are wrong!
-
-There are model checking methods (like varying the prior and
-re-running \MCMC) that test the sensitivity of results to assumptions.
-Not only are these very valuable for testing the model, they are
-essential for writing any paper about your results: The reader always
-wants to know the sensitivity of results to assumptions.\note{And be
-  sure not to forget that there are (at least) as many assumptions
-  encoded in your likelihood function as there are in your prior
-  pdfs.}
-Once again, it is important to find ways to display the results of
-such tests in the space of the data.
-All we will say here about these kinds of rich model tests is that you
-should do them, and report the results.
-
-The second---and less important---point about model checking is that
-the ur-Bayesian thing to be doing is computing a Bayesian evidence
-integral.
-However, it is usually inadvisable:
-The integral is expensive to compute.
-Most \MCMC\ methods are cleverly designed to \emph{avoid} computing this
-integral (as we note above in \sectionname~\ref{sec:when}); so it usually adds
-expense to the project.
-At the end of the day, it returns only a single, scalar number, with
-no guidance about how to use it (and indeed there are only heuristics
-to reach for).
-If your model is wrong, the richer tests of visualization of residuals
-or sensitivity to assumptions are much more likely to lead to insight
-about what changes to make or what new directions to explore.
-
-%% HOGG DELETED / COMMENTED OUT THE BELOW
-%% - make sure everything below is discussed somewhere above and then delete it below.
-
-%% ...breaking Markov-ness.
-%% % ***** BJB
-%% % "Diminishing adaptations" is an idea that is quite easy
-%% % to explain. There are other ways to break Markov-ness but
-%% % this probably covers enough of them for an intro document
-%% % ***** Hogg
-%% % I was going to say you should NEVER break Markov-ness
-%% % unless you REALLY understand what you are doing.
-
-%% ...Posteriors:  You can do anything you like (within
-%% reason) to the output samples and be sampling the posterior in that
-%% transformed quantity.
-
-%% ...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.
-
-%% ...Finally, to repeat: You can't sample the likelihood!  You can only
-%% sample a posterior pdf!
-
 \clearpage

### 2016-10-14 mcmc.tex
got rid of a DFM

@@
 We will return to this point below in our comments on testing (\sectionname~\ref{sec:trouble}).
+The point is that if you \emph{do} change coordinates between
+  the statement of your priors and the specific variables you
+  step in, you have to introduce Jacobian factors.
 
@@
   the accept--reject step!).}
-
-HOGG SAY HERE: ...You must describe your priors in the same parameterization that you ``step'' in.
-That is, if you change your parameters, you \emph{must} change your priors correspondingly.
-Give the example of going from $A$, $\phi$ to $A\,\cos\phi$, $A\,\sin\phi$.
 
@@
 
-\paragraph{visualization:}
-HOGG: What plots do you make to diagnose problems?
-
-HOGG: Plot parameters vs ``time'' in the chain.
-
-HOGG: Plot a cut through the density function and the samples near that cut.
-
-HOGG:...Make triangle plots.  Great value!...  Show example here... Mentioned above; drop here or reiterate?
-
 \paragraph{likelihood issues:}
@@
 
-DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?
-Choose a $p(A)$ and a $p(\phi) = [2\pi]^{-1}$,
-   and then transform to $p(a, b)$.
-
-%% ...Priors: Once again:
-%% If you describe your priors in one parameterization but
-%% then perform sampling in a different space, it is easy to make
-%% mistakes.  For example: Flat prior in amplitude and phase, sample in
-%% $A\,\cos\phi$ and $A\,\sin\phi$.  There are many ways to do this
-%% wrong.
+To be explicit, imagine that you have parameters $A$ and $\phi$ and
+priors $p(A)$ and $p(\phi)$.
+Now imagine that instead (and very sensibly) you want to sample in
+parameters $x\equiv A\,\cos(\phi)$ and $y\equiv A\,\sin(\phi)$.
+What is the proper prior to apply to $x, y$?
+It is not just $p(A)\,p(\phi)$ with $A=\sqrt{x^2 + y^2}$ and $\phi=\arctan(y,x)$!
+That's close, though! There is a correction factor which is the
+absolute value of the determinant of the Jacobian matrix
+$||\frac{\dd(A,\phi}{\dd(x,y)}||$.
+That is, it is the determinant of the full derivative of one set of
+parameters with respect to the other.
+Keep your units straight!\note{See \citet{hoggcalculus} for advice on
+  how to check and test these transformations using dimensional
+  analysis.}
+In this particular case, the Jacobian is equal to $A$, such that 
+\begin{eqnarray}
+  p(x,y) &=& \frac{1}{A}\,p(A)\,p(\phi)
+  \quad ,
+\end{eqnarray}
+where $A=\sqrt{x^2 + y^2}$ and $\phi=\arctan(y,x)$.\note{Exercise to
+  the reader: Do the same but with $x\equiv \sqrt{A}\,\cos(\phi)$ and
+  $y\equiv\sqrt{A}\,\sin(\phi)$.}
 
@@
 Again, such transformations require also prior transformations, which must be made with care.
-There are some affine invariant samplers\note{Note how much we like to cite \citet{emcee}!} that are invariant to such transformations,
-  but if you aren't using such a sampler,
-  it is worth getting out ahead of these problems.
+There are some affine invariant samplers\note{Note how much we like to
+  cite \citet{emcee}!} that are invariant to such transformations, but
+if you aren't using such a sampler, it is worth getting out ahead of
+these problems.
 They can be found by performing an exploratory sampling,
@@
 
-HOGG: ...How to deal with discrete or integer or binary parameters?
+For another example, sometimes some parameters are continuous, and some
+are integer (or discrete).
+In this case, it is sufficient, in the context of M-H \MCMC, to make
+custom proposal distributions for the integer parameters that make sure
+only integers are proposed (provided care is taken to make sure that detailed
+balance is preserved).
+An example is to have a proposal that has some probability of
+incrementing the parameter up, and an equal (ie, the same) probability
+of incrementing it down.
+Some samplers implicitly or explicitly assume that all parameters are
+continuous\note{Once again, \citet{emcee}.}; these should be avoided
+when some parameters are integer.
+If you \emph{really} want to use such samplers when some parameters
+are discrete, you have to put a \code{round()} or \code{int()}
+operation into the likelihood and prior functions.
+This makes for a stepwise-constant posterior pdf. It is a hack, but it
+works in most circumstances.
 

### 2016-10-14 mcmc.tex
some words about plotting

--- 2016-10-14 mcmc.tex
@@
 % - Go through github issues, if any.
+% - Systematically cite chapters of Brooks \etal in each section.
+% - Skim MacKay and make sure we cite him in any relevant place.
+% - Mention & cite STAN in methods section
 % - Need to mention case where there is an MC integral INSIDE the likelihood function
 % - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
+% - Make sure we say ``don't use MCMC for 1-d problems or if you have alternatives to directly integrate!''
+% - Would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
 % - Send to friendlies for comments
@@
 \caption{Solution to \problemname~\ref{prob:MH}:
-\MCMC\ samples from a Gaussian (black) and the true distribution (blue).}
+MCMC samples from a Gaussian (black) and the true distribution (blue).}
 \label{fig:MH}
@@
 \caption{Solution to \problemname~\ref{prob:MH2}:
-\MCMC\ samples from a uniform distribution (black) and the true distribution
+MCMC samples from a uniform distribution (black) and the true distribution
 (blue).}
@@
 \caption{Solution to \problemname~\ref{prob:twod}a:
-\MCMC\ samples from a two-dimensional Gaussian (scatter plot) and the
+MCMC samples from a two-dimensional Gaussian (scatter plot) and the
 one-dimensional marginalized distributions (histograms).}
@@
 \caption{Solution to \problemname~\ref{prob:logprior}:
-The black histogram shows the \MCMC\ samples and the blue curve is the analytic
+The black histogram shows the MCMC samples and the blue curve is the analytic
 target density.}
@@
 It is also worth noting that there is not just one autocorrelation function or
-autocorrelation time for an \MCMC\ chain.
+autocorrelation time for an MCMC chain.
 Instead, there is an autocorrelation time $\tau_f$ for every function
@@
 Estimates of the integrated autocorrelation time of different segments of the
-\MCMC\ chain from \problemname~\ref{prob:MH} (black lines) and for the full
+MCMC chain from \problemname~\ref{prob:MH} (black lines) and for the full
 chain (orange line) as a function of window size $M$.
@@
 
+There are several figures that are useful and should be made every time you
+run MCMC and we'll describe a few here.
+
+\begin{itemize}
+
+\item \emph{Trace plots}~---~The first set of plots are the ``trace
+plots''~--~the parameter values as a function of step number.
+These plots can be used to select burn-in lengths, indicate problems with the
+model or sampler, and qualitatively judge convergence.
+This being said, it is important to remember that \emph{good looking trace
+plots do not guarantee converged sampling} and, even if the traces look okay,
+the convergence diagnostics discussed in \sectionname~\ref{sec:convergence}
+might identify problems.
+
+\item \emph{Posterior predictive plots}~---~Another useful plot is the
+predictive distribution for the model in data space.
+For this plot, you take some $K$ random samples from your chain, plot the
+prediction that each sample makes for the data and over-plot the observed
+data.
+This plot gives a qualitative sense of the ``quality'' of the fit and it can
+identify problems with sampling or convergence.
+
+\end{itemize}
+
 Of course figures are always better than tables, at least in the
@@
 If you are clever, all these plots can be arranged into a lovely and
-informative triangle.\note{Our favorite package these days is our own \project{corner.py}.  Find it on the internets.}
+informative triangle.\note{Our favorite package these days is our own
+\project{corner.py}.  Find it on the internets.}
 These two-dimensional plots do not in any sense contain all the
-information in the sampling!  However, they are remarkable for
+information in the sampling.  However, they are remarkable for
 locating expected and unexpected parameter relationships, and often

### 2016-10-14 mcmc.tex
self citation
more words about plots

@@
 
+Of course figures are always better than tables, at least in the
+printed form of scientific publications.
 There are several figures that are useful and should be made every time you
@@
 data.
-This plot gives a qualitative sense of the ``quality'' of the fit and it can
-identify problems with sampling or convergence.
-
-\end{itemize}
-
-Of course figures are always better than tables, at least in the
-printed form of scientific publications.
-In a $D$-dimensional parameter space, we recommend plotting all $D$
-one-dimensional sample histograms, on a pretty fine binning, and all
-$D$-choose-2 two-dimensional histograms (scatter plots) to show the
-low-level covariances and non-linearities.
+This plot gives a qualitative sense of how well the model fits the data and it
+can identify problems with sampling or convergence.
+
+\item \emph{Corner plots or scatterplot matrices}~---~In a $D$-dimensional
+parameter space, we recommend plotting all $D$ one-dimensional sample
+histograms and all $D$-choose-2 two-dimensional histograms (scatter plots) to
+show the low-level covariances and non-linearities.
 If you are clever, all these plots can be arranged into a lovely and
 informative triangle.\note{Our favorite package these days is our own
-\project{corner.py}.  Find it on the internets.}
-These two-dimensional plots do not in any sense contain all the
-information in the sampling.  However, they are remarkable for
-locating expected and unexpected parameter relationships, and often
-invaluable for suggesting re-parameterizations and transformations
-that simplify your problem.
+\project{corner.py}.  Find it on the internets.} These two-dimensional plots
+do not in any sense contain all the information in the sampling but they are
+remarkable for locating expected and unexpected parameter relationships, and
+often invaluable for suggesting re-parameterizations and transformations that
+simplify your problem.
+
+\end{itemize}
+
+DFM: Add an example of each of these plots.
 

### 2016-10-14 mcmc.tex
less is more

--- 2016-10-14 mcmc.tex
@@
 In addition to a full set of unit tests for every part of your code
-  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
+  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.
+It makes sense to put in unit tests for every part of your code, especially the parts of your code that are relied upon for correctness in multiple locations, like your \MCMC\ sampler.}),
   we recommend building some end-to-end functional tests that permit you to perform \MCMC\
@@
 This is an easy test to fail, so a success builds confidence in any \MCMC\ code.
-
-In the most extreme form of this test,
-  for a Gaussian in $D$ dimensions,
-  build the covariance matrix input to the Gaussian distribution
-  by taking the sum of the outer product of $D+1$ randomly generated vectors.
-Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
-Set up the \MCMC\ sampler with the Gaussian function with this covariance matrix
-  as the probability function (or the log of it as the log-probability function).
-Sample for many autocorrelation times,
-  or until you are confident you are getting good samples.
-Check that the empirical $D$-dimensional covariance of the samples
-  has the same (or similar) eigenvectors and eigenvlues to the input matrix.
-Also check that the individual one-dimensional samplings
-  (along each axis or any projection in the $D$-space)
-  look Gaussian and have nearly vanishing skew and kurtosis.
-It takes some thinking and some calculating to see ``how well'' these things all ought to match
-  expectations, but one thing that can be said with confidence is that agreement should improve
-  as the sampling proceeds.
-If it doesn't, it is likely that there is something broken about the sampler itself.
 
@@
   maybe it includes within it an integral performed by Monte Carlo method?
-In general, you want to change your likelihood function such that,
+In general, you want to write your likelihood function such that,
   even if it includes a numerical or stochastic integral,
@@
 
-DFM:...SHOW EXAMPLES?
-
 \paragraph{low acceptance fraction:}
@@
   or whatever parameter or parameters control the width of that distribution.
-If reducing the variance of the proposal distribution does \emph{not} increase the acceptance fraction,
-  then \emph{you have a bug}.
-You must find that bug by auditing your code or else returning to the functional tests listed above.
+If reducing the variance of the proposal distribution does \emph{not}
+increase the acceptance fraction, then \emph{you have a bug}.
+You must find that bug by auditing your code or else returning to the
+functional tests listed above.
 
 Similarly for high acceptance fraction:
-If you are finding a high acceptance fraction,
-  then increasing the variance of the proposal distributon \emph{must} decrease the acceptance fraction.
+If you are finding a high acceptance fraction, then increasing the
+variance of the proposal distributon \emph{must} decrease the
+acceptance fraction.
 If it does \emph{not}, then \emph{you have a bug}.
@@
   in every dimension many times.
-
-DFM: SHOW EXAMPLES OF GOOD AND BAD?
 
@@
 There are some affine invariant samplers\note{Note how much we like to
-  cite \citet{emcee}!} that are invariant to such transformations, but
-if you aren't using such a sampler, it is worth getting out ahead of
-these problems.
+  cite \project{emcee} \citep{emcee}!} that are invariant to such
+transformations, but if you aren't using such a sampler, it is worth
+getting out ahead of these problems.
 They can be found by performing an exploratory sampling,
@@
   Data analysis recipes: Fitting a model to data,
-  \arxiv{1008.4686v1}.
+  \arxiv{1008.4686}.
 \bibitem[Hogg \etal(2010b)]{eccentricity}
@@
   \textit{Astrophys.\ J.}\ \textbf{725} 2166--2175.
+\bibitem[Hogg(2012)]{hoggcalculus}
+  Hogg,~D.~W., 2012,
+  Data analysis recipes: Probability calculus for inference,
+  \arxiv{1205.4446}.
 \bibitem[Hou \etal(2014)]{fengji}

### 2016-10-14 mcmc.tex
constraints

@@
 
+For yet another example, there might be constraints such that all
+acceptable models lie on a non-trivial, non-linear subspace of your
+parameter space.
+Here almost no arbitrary steps in parameter space could lead to an
+acceptance, since if you move arbitrarily, you are very unlikely to
+hit the subspace!
+Your best move here is to re-parameterize so there are only models on
+that subspace.
+We're not claiming that that re-parameterization is always easy to
+find; the mathematics of Lagrange multipliers can be useful here.
+
 \paragraph{model checking:}
@@
 about what changes to make or what new directions to explore.
-
-%% HOGG DELETED / COMMENTED OUT THE BELOW
-%% - make sure everything below is discussed somewhere above and then delete it below.
-
-%% ...breaking Markov-ness.
-%% % ***** BJB
-%% % "Diminishing adaptations" is an idea that is quite easy
-%% % to explain. There are other ways to break Markov-ness but
-%% % this probably covers enough of them for an intro document
-%% % ***** Hogg
-%% % I was going to say you should NEVER break Markov-ness
-%% % unless you REALLY understand what you are doing.
-
-%% ...Posteriors:  You can do anything you like (within
-%% reason) to the output samples and be sampling the posterior in that
-%% transformed quantity.
-
-%% ...How to deal with constraints that pin you to a subspace of the parameter space?  Method-dependent.
 

### 2016-10-14 mcmc.tex
tweaks

@@
 absolute value of the determinant of the Jacobian matrix
-$||\frac{\dd(A,\phi}{\dd(x,y)}||$.
+$||{\dd(A,\phi)}/{\dd(x,y)}||$.
 That is, it is the determinant of the full derivative of one set of
@@
   analysis.}
-In this particular case, the Jacobian is equal to $A$, such that
+In this particular case, the Jacobian is equal to $1/A$, such that
 \begin{eqnarray}
@@
 An example is to have a proposal that has some probability of
-incrementing the parameter up, and an equal (ie, the same) probability
+incrementing the parameter up, and an equal (that is, the same) probability
 of incrementing it down.

### 2016-10-14 mcmc.tex
some ensemble refs

--- 2016-10-14 mcmc.tex
@@
 In addition to a full set of unit tests for every part of your code
-  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.
-It makes sense to put in unit tests for every part of your code, especially the parts of your code that are relied upon for correctness in multiple locations, like your \MCMC\ sampler.}),
+  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
   we recommend building some end-to-end functional tests that permit you to perform \MCMC\
@@
 This is an easy test to fail, so a success builds confidence in any \MCMC\ code.
+
+In the most extreme form of this test,
+  for a Gaussian in $D$ dimensions,
+  build the covariance matrix input to the Gaussian distribution
+  by taking the sum of the outer product of $D+1$ randomly generated vectors.
+Then, before sampling, get the eigenvalues and eigenvectors of the matrix.
+Set up the \MCMC\ sampler with the Gaussian function with this covariance matrix
+  as the probability function (or the log of it as the log-probability function).
+Sample for many autocorrelation times,
+  or until you are confident you are getting good samples.
+Check that the empirical $D$-dimensional covariance of the samples
+  has the same (or similar) eigenvectors and eigenvlues to the input matrix.
+Also check that the individual one-dimensional samplings
+  (along each axis or any projection in the $D$-space)
+  look Gaussian and have nearly vanishing skew and kurtosis.
+It takes some thinking and some calculating to see ``how well'' these things all ought to match
+  expectations, but one thing that can be said with confidence is that agreement should improve
+  as the sampling proceeds.
+If it doesn't, it is likely that there is something broken about the sampler itself.
 
@@
   maybe it includes within it an integral performed by Monte Carlo method?
-In general, you want to write your likelihood function such that,
+In general, you want to change your likelihood function such that,
   even if it includes a numerical or stochastic integral,
@@
 
+DFM:...SHOW EXAMPLES?
+
 \paragraph{low acceptance fraction:}
@@
   or whatever parameter or parameters control the width of that distribution.
-If reducing the variance of the proposal distribution does \emph{not}
-increase the acceptance fraction, then \emph{you have a bug}.
-You must find that bug by auditing your code or else returning to the
-functional tests listed above.
+If reducing the variance of the proposal distribution does \emph{not} increase the acceptance fraction,
+  then \emph{you have a bug}.
+You must find that bug by auditing your code or else returning to the functional tests listed above.
 
 Similarly for high acceptance fraction:
-If you are finding a high acceptance fraction, then increasing the
-variance of the proposal distributon \emph{must} decrease the
-acceptance fraction.
+If you are finding a high acceptance fraction,
+  then increasing the variance of the proposal distributon \emph{must} decrease the acceptance fraction.
 If it does \emph{not}, then \emph{you have a bug}.
@@
   in every dimension many times.
+
+DFM: SHOW EXAMPLES OF GOOD AND BAD?
 
@@
 There are some affine invariant samplers\note{Note how much we like to
-  cite \project{emcee} \citep{emcee}!} that are invariant to such
-transformations, but if you aren't using such a sampler, it is worth
-getting out ahead of these problems.
+  cite \citet{emcee}!} that are invariant to such transformations, but
+if you aren't using such a sampler, it is worth getting out ahead of
+these problems.
 They can be found by performing an exploratory sampling,
@@
 
-\section{Gibbs, ensemble, Hamiltonian, and so on}\label{sec:methods}
+\section{More sophisticated sampling methods}\label{sec:methods}
 
@@
 \project{emcee}\note{\citet{emcee}}, which is popular in the
-astrophysical sciences, but there are others\note{One that we are
-  interested in is \project{Kombine} (Farr \& Farr, in preparation).}.
+astrophysical sciences, but there are others\note{A few ensemble methods that
+are being used in astronomy: Differential Evolution MCMC (\citealt{demcmc,
+run-dmc}) and \project{kombine} (Farr \& Farr, in preparation).}.
 It is worty of note that ensemble methods have particular
@@
   in \citet{mcmchandbook}, 113--162.
+\bibitem[Nelson \etal(2014)]{run-dmc}
+  Nelson, B., Ford, E.~B., \& Payne, M.~J., 2014,
+  \href{http://adsabs.harvard.edu/abs/2014ApJS..210...11N}{RUN DMC: An
+Efficient, Parallel Code for Analyzing Radial Velocity Observations Using
+N-body Integrations and Differential Evolution Markov Chain Monte Carlo}
+  \textbf{210}, 11.
+  \doi{10.1088/0067-0049/210/1/11}.
 \bibitem[Pasarica \& Gelman(2010)]{esjd}
@@
   \doi{10.1214/06-BA127}.
+\bibitem[Ter Braak(2006)]{demcmc}
+  Ter Braak, C.~J.~F., 2006,
+  A Markov Chain Monte Carlo version of the genetic algorithm
+  Differential Evolution: easy Bayesian computing for real
+  parameter spaces
+  \textit{Stat.\,Comp.}\ \textbf{16} 239--249.
+  \doi{10.1007/s11222-006-8769-1}.
 \bibitem[van~der~Walt \etal(2011)]{numpy}

### 2017-10-09 mcmc.tex
start reformatting for arXiv and ApJ

@@
 %   which is (in turn) part of the Data Analysis Recipes project.
-% Copyright 2013, 2014, 2015, 2016 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).
+% Copyright 2013, 2014, 2015, 2016, 2017 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).
 
@@
 
-% low priority
-% ------------
-% - Make endnote numbers clickable links to endnotes and back.  Relevant ideas here:
-%   http://tex.stackexchange.com/questions/8452/making-endnotes-clickable-links-with-hyperref
-
 % style notes
@@
 
-\documentclass[12pt,twoside,pdftex]{article}
-\usepackage{styles/dar_endnotes}
+\documentclass[12pt,pdftex]{article}
 
@@
 %
-\section*{Data analysis recipes:\\ \chaptertitle\footnotemark}
-%
-\footnotetext{%
-    The notes begin on page~\pageref{note:first}, including the
-    license\note{\label{note:first}
-        Copyright 2013, 2014, 2015, 2016 the authors (OMG this took us a long time).
+\section*{Data analysis recipes:\\ \chaptertitle\note{%
+\label{note:first}%
+Copyright 2013, 2014, 2015, 2016, 2017 the authors (OMG this took us a long time).
         This work is licensed under a
@@
             Creative Commons Attribution-NonCommercial-NoDerivs 3.0 Unported
-            License}.}
-    and the acknowledgements\note{%
-        We would like to thank
-          Jo Bovy (IAS),
-          Brendon Brewer (Auckland),
-          Will Farr (Birmingham),
-          Jonathan Goodman (NYU),
-          Fengji Hou (NYU),
-          Dustin Lang (CMU),
-          Phil Marshall (KIPAC), and
-          Iain Murray (Edinburgh)
-        for valuable advice and comments over the years, and
-          Alessandro Gentilini
-        for a very close read.
-        This project benefitted from various research grants, especially
-          NSF grants AST-0908357, IIS-1124794, and AST-1517237,
-          NASA grant NNX12AI50G,
-          and the The Moore-Sloan Data Science Environment at NYU.}
-}
-
+            License}.}}%
 \noindent
 \textbf{David~W.~Hogg}\\
-\affil{Simons~Center~for~Data~Analysis,
-       Simons~Foundation}\\
+\affil{Flatiron Institute, a division of the Simons~Foundation}\\
 \affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
@@
 \textbf{Daniel~Foreman-Mackey}\\
-\affil{Sagan Fellow}\\
+\affil{Flatiron Institute, a division of the Simons~Foundation}\\
 \affil{Department~of~Astronomy, University~of~Washington}
@@
 methods on real problems.
-We will leave parenthetical and philosophical matters to the end-notes,
+We will leave parenthetical and philosophical matters to the footnotes,
 which appear after the main text.
@@
 \clearpage
-\markright{Notes}\theendnotes
-
-\clearpage
+\paragraph{Acknowledgements:}
+We would like to thank
+          Alex Barnett (Flatiron),
+          Jo Bovy (Toronto),
+          Brendon Brewer (Auckland),
+          Will Farr (Flatiron)
+          Jonathan Goodman (NYU),
+          Fengji Hou (NYU),
+          Dustin Lang (Toronto),
+          Phil Marshall (KIPAC), and
+          Iain Murray (Edinburgh)
+        for valuable advice and comments over the years, and
+          Alessandro Gentilini
+        for a very close read.
+        This project benefitted from various research grants, especially
+          NSF grants AST-0908357, IIS-1124794, and AST-1517237,
+          NASA grant NNX12AI50G,
+          and the The Moore-Sloan Data Science Environment at NYU.
+
 \raggedright

### 2017-10-09 problems/MH.py
fixing problem bugs

@@ -18,7 +18,7 @@ def log_p(x, mean=2.0, variance=2.0):
 
 x = 0.0
 lp = log_p(x)
-chain = np.empty(2e4)
+chain = np.empty(int(2e4))
 for step in range(len(chain)):
     x_prime = x + np.random.randn()
     lp_prime = log_p(x_prime)

### 2017-10-09 problems/MH2.py

@@ -20,7 +20,7 @@ def log_p(x, a=3.0, b=7.0):
 
 x = 5.0
 lp = log_p(x)
-chain = np.empty(2e4)
+chain = np.empty(int(2e4))
 for step in range(len(chain)):
     x_prime = x + np.random.randn()
     lp_prime = log_p(x_prime)

### 2017-10-09 problems/convergence.py

@@ -18,7 +18,7 @@ def log_p(x, mean=2.0, variance=2.0):
 
 x = 0.0
 lp = log_p(x)
-chain = np.empty(1e6)
+chain = np.empty(int(1e6))
 for step in range(len(chain)):
     x_prime = x + np.random.randn()
     lp_prime = log_p(x_prime)
@@ -41,8 +41,8 @@ if __name__ == "__main__":
     ax1.set_xlabel("thousands of steps")
 
     for i in range(4):
-        a = i*len(chain)/4
-        b = (i+1)*len(chain)/4
+        a = int(i*len(chain)/4)
+        b = int((i+1)*len(chain)/4)
         print(i+1, np.mean(chain[a:b]), np.var(chain[a:b]))
         ax2.hist(chain[a:b], 50, histtype="step")
     ax2.set_xlim(2-5.5, 2+5.5)

### 2017-10-09 problems/prop_mean.py

@@ -16,7 +16,7 @@ np.random.seed(42)
 def run_broken_mcmc(log_p, x, prop_sigma=1.0, prop_mean=0.5, nsteps=2e3,
                     broken=True):
     lp = log_p(x)
-    chain = np.empty((nsteps, len(x)))
+    chain = np.empty((int(nsteps), len(x)))
     for step in range(len(chain)):
         x_prime = np.array(x)
         ind = np.random.randint(len(x))

### 2017-10-09 problems/twod_a.py

@@ -20,7 +20,7 @@ def log_p_gauss(x):
 
 def run_mcmc(log_p, x, prop_sigma=1.0, nsteps=2e4, thin=1):
     lp = log_p(x)
-    chain = np.empty((nsteps // thin, len(x)))
+    chain = np.empty((int(nsteps) // int(thin), len(x)))
     acc = 0
     for step in range(len(chain)):
         for i in range(thin):

### 2017-10-10 mcmc.tex
works in aastex 6.1, baby

@@
 
-\documentclass[12pt,pdftex]{article}
-
-\newcommand{\chaptertitle}{Using Markov Chain Monte Carlo}
+\documentclass[modern]{aastex61}
 \include{styles/dar}
+
+\shorttitle{using markov chain monte carlo}
+\shortauthors{hogg \& foreman-mackey}
 
@@
 %
-\section*{Data analysis recipes:\\ \chaptertitle\note{%
-\label{note:first}%
+\title{Data analysis recipes:\\
+       Using Markov Chain Monte Carlo%
+\note{\label{note:first}%
 Copyright 2013, 2014, 2015, 2016, 2017 the authors (OMG this took us a long time).
@@
             License}.}}%
-\noindent
-\textbf{David~W.~Hogg}\\
-\affil{Flatiron Institute, a division of the Simons~Foundation}\\
-\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
-       New~York~University}\\
-\affil{Center~for~Data~Science,
-       New~York~University}\\
-\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}\\[1ex]
-\textbf{Daniel~Foreman-Mackey}\\
-\affil{Flatiron Institute, a division of the Simons~Foundation}\\
+
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
+\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
+\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
+\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}
+
+\author[0000-0002-9328-5652]{Dan Foreman-Mackey}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
 \affil{Department~of~Astronomy, University~of~Washington}
@@
 full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
-the posterior pdf $p(\pars\given\data$).
-% ************* BJB
-% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
-% refer the interested reader to the paper "Monte Carlo is Fundamentally
-% Unsound" by Tony O'Hagan.
+the posterior pdf $p(\pars\given\data)$. We will give some sense of what ``fair'' means
+in this context, below.
 

### 2017-10-10 mcmc.tex
abstract

@@
 density functions (combined with abundant computational resources)
-have transformed the sciences.
-Here we give a brief overview of the most basic \MCMC\ method and then turn to
+have transformed the sciences, especially in performing probabilistic
+inferences, or fitting models to data.
+In this primarily pedagogical contribution,
+we give a brief overview of the most basic \MCMC\ method and some
 practical advice for the use of \MCMC\ in real inference problems.
@@
 output to produce or report parameter estimates with associated uncertainties.
+We emphasize that sampling is a method for doing integrals; this guides our thinking
+about how \MCMC\ output is best used.
 \end{abstract}

### 2017-10-10 mcmc.tex
first few pages

@@
 % -----
+% - deal with \MCMC\ macro in footnotes.
 % - Deal with all mentions of ``DFM'' or ``HOGG''
@@
 % - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
-% - Send to friendlies for comments
-%   - Gelman
-%   - Goodman
-%   - Lang
-%   - Rix
-%   - Bovy
-%   - Brewer
 % - Kill all issues in this list, above, and then:
 %   - Post on arXiv
+%   - Submit to ApJ special issue
 %   - Blog about it
@@
 \author[0000-0003-2866-9403]{David W. Hogg}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 \affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
@@
 \affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
-\affil{Department~of~Astronomy, University~of~Washington}
+\affil{Department~of~Astronomy, University~of~Washington, Box 351580, Seattle, WA 98195, USA}
 
@@
 ``Bayes integral'', and ``prior predictive probability'', and is usually
-\emph{extremely hard to calculate}.\note{Some would call
-  the factor $Z$ in \equationname~(\ref{eq:bayes}) something like
-  ``$p(\data)$'' or ``the fully marginalized likelihood'' or the
-  probability of the data.
-  Of course it is only the probability of the data within the model
-  space under consideration!
-  We (intend to) discuss this object and these issues elsewhere
-  (in various things in preparation).
-  The factor $Z$ is often hugely difficult to compute, because the
+\emph{extremely hard to calculate}.\note{%
+  The factor $Z$ is often difficult to compute, because the
   likelihood (or the prior) can have extremely complex structure,
@@
   and so have many others before us.
-  Perhaps more importantly, we also have views about why you probably never really
-  \emph{want} to compute it.
-  These views relate to the fact that the value of this integral
-  very rarely delivers the answer to a question you care about (see future \documentname s in this series).
-  These philosophical matters and the computational issues are both
-  outside the scope of this \documentname,
-  but suffice it to say:
-  If the computation of $Z$ is in your travel plans, you will need to
-  think about implementing some non-trivial integrating or sampling
-  technology; we will give some brief description of options in
-  \sectionname~\ref{sec:methods}.}
+  We also have (unpublished) philosophical arguments against calculating
+  this $Z$ if you can possibly avoid it, but these are
+  outside the scope of this \documentname.
+  The point is that \MCMC\ methods will not require that you know $Z$.}
 That is, you often know the function $p(\pars\given\data)$ up to a
@@
 You have a huge blob of data $\data$ (think of this as a vector or
-list or heterogeneous collection of observations\note{In general in
-  this \documentname\ we will treat any variable as being capable of
-  containing not just a value but some huge arbitrarily structured set
-  of values in matrix, vector, or an ordered list of any set of such objects,
-  or perhaps some even more complex form.}).
+list or heterogeneous-but-ordered collection of observations).
 You also have a model sophisticated enough---a probabilistic,
-generative model, if you will\note{For more discussion
-  of our definition of a \emph{model}, see [some other as-yet-unwritten \documentname] in
-  this series.
+generative model, if you will\note{%
   Briefly, a ``model'' for us is a likelihood function (a pdf for the data given
@@
 }---that, given a
-setting of a huge blob of parameters\note{Again, the parameter
-  blob $\pars$ should also be thought of as an immense, arbitrarily
-  structured set or list of parameters, possibly even with no fixed
-  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We will
-  bash the terminology ``likelihood'' in another, as-yet-unwritten, \documentname\ in this
-  series.  Techically, this thing $p(\data\given\pars)$
+setting of a huge blob of parameters (again, think of this as a vector or list or heterogeneous-but-ordered collection of values) $\pars$,
+  you can compute a pdf for data (or likelihood\note{%
+  Techically, $p(\data\given\pars)$
   is only properly a likelihood function when
@@
 posterior pdf in all of parameter space, many investigators use \MCMC\
-because it will sample \emph{all} of the parameter space
+because (they believe) it will sample \emph{all} of the parameter space
 ($\pars$-space).
@@
 search algorithm, and that it is a simple-to-code effective
-optimizer---are \emph{not good reasons}.  \MCMC\ is first and foremost a
+optimizer---are \emph{not good reasons}.  \MCMC\ is a
 \emph{sampler}.  If you are trying to find the optimum of the
@@
 Heuristically, a sampling $\setof{\pars_k}_{k=1}^K$ from some pdf $p(\pars)$
-is a set of $K$ values $\pars_k$ that are independent draws from the pdf.
+is a set of $K$ values $\pars_k$ that are draws from the pdf.
 Heuristically, if an enormous (large $K$) sampling could be displayed
@@
 where, implicitly, the integral is over the full domain of $\pars$.
-Importantly, although $p(\pars)$ is single-valued, $\pars$ can be a scalar or
+Importantly, although $p(\pars)$ is single-valued, $\pars$ can be
 a multi-element vector or list or blob; it can be arbitrarily large
@@
 then along those dimensions the integral becomes a discrete sum.
-This latter is a detail to which we return below.
+This latter is a detail to which we return below. HOGG DO WE?
 
@@
 \end{eqnarray}
+And recall that we don't actually know the value of $Z$, but we do know
+that it is finite.
 

### 2017-10-15 mcmc.tex
symbols

@@
 $E_{p(\pars)}[\pars]$
-for $\pars$ or for any quantity that can be expressed as a function $q(\pars)$
+for $\pars$ or for any quantity that can be expressed as a function $g(\pars)$
 of $\pars$:
@@
 \\
-E_{p(\pars)}[q(\pars)] &\equiv& \int q(\pars)\,p(\pars)\,\dd \pars
+E_{p(\pars)}[g(\pars)] &\equiv& \int g(\pars)\,p(\pars)\,\dd \pars
 \quad ,
@@
 integrals are multi-dimensional if $\pars$ is multi-dimensional.
-These expectation values are the mean values of $\pars$ and $q(\pars)$ under
+These expectation values are the mean values of $\pars$ and $g(\pars)$ under
 the pdf.  A good sampling---and really this is the definition of a good
@@
 \\
-E_{p(\pars)}[q(\pars)] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
+E_{p(\pars)}[g(\pars)] &\approx& \frac{1}{K}\,\sum_{k=1}^K g(\pars_k)
 \quad .
@@
 \begin{eqnarray}
-E_{p(\pars)}[q(\pars)] &\equiv& \frac{\int q(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
+E_{p(\pars)}[g(\pars)] &\equiv& \frac{\int g(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
 \\
-E_{p(\pars)}[q(\pars)] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
+E_{p(\pars)}[g(\pars)] &\approx& \frac{1}{K}\,\sum_{k=1}^K g(\pars_k)
 \label{eq:expectationvalue}
@@
 
-That is, if the expectation value $E[q]$ is computed by the sampling
+That is, if the expectation value $E[g]$ is computed by the sampling
 approximation given in \equationname~(\ref{eq:expectationvalue}),
@@
 \begin{eqnarray}
-  \Var[E_{p(\pars)}[q(\pars)]] &=& \frac{\tau_q}{K}\,\Var_{p(\pars)}[q(\pars)]
+  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_q}{K}\,\Var_{p(\pars)}[g(\pars)]
   \\
-  &\approx& \frac{\tau_q}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) -
-    E_{p(\pars)}[q(\pars)]\right)^2
+  &\approx& \frac{\tau_q}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
+    E_{p(\pars)}[g(\pars)]\right)^2
   \quad,
@@
 where we are being a bit loose with terminology because this is the
-variance not on $E_{p(\pars)}[q(\pars)]$ but on our sampling-based
+variance not on $E_{p(\pars)}[g(\pars)]$ but on our sampling-based
 \emph{estimator} for
-$E_{p(\pars)}[q(\pars)]$, and the $E_{p(\pars)}[q(\pars)]$ inside the sum is
+$E_{p(\pars)}[g(\pars)]$, and the $E_{p(\pars)}[g(\pars)]$ inside the sum is
 that estimator too.
@@
 autocorrelation time for an \MCMC\ chain.
-Instead, there is an autocorrelation time $\tau_f$ for every function
-$f(\pars)$ and these will, in general, be different.
+Instead, there is an autocorrelation time $\tau_g$ for every function
+$g(\pars)$ and these will, in general, be different.
 Therefore, the autocorrelation time used in \equationname~(\ref{eq:tauvar}) is
-the time computed on the chain of $q(\pars)$ values instead of the
+the time computed on the chain of $g(\pars)$ values instead of the
 autocorrelation time for $\pars$.
@@
   $K$ samples $\pars_k$ of a parameter vector $\pars$.
-Now you have a scalar function $q(\pars)$ which takes the parameter vector
+Now you have a scalar function $g(\pars)$ which takes the parameter vector
   and returns a scalar value%
-  \note{For the median-of-sampling value, $q(\pars)$ needs to return a scalar,
-  but technically, for the mean-of-sampling, $q(\pars)$ can be a vector or something high-dimensional.}
+  \note{For the median-of-sampling value, $g(\pars)$ needs to return a scalar,
+  but technically, for the mean-of-sampling, $g(\pars)$ can be a vector or something high-dimensional.}
   which could be as simple as a single component of the parameter vector
@@
 \begin{eqnarray}
-\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
+\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K g(\pars_k)
 \quad ,
 \end{eqnarray}
-  and the median-of-sampling value is just the $[K/2]$th value of $q(\pars_k)$
-  when the $q(\pars_k)$ have been ordered (sorted).
+  and the median-of-sampling value is just the $[K/2]$th value of $g(\pars_k)$
+  when the $g(\pars_k)$ have been ordered (sorted).
 These are both produced by integrals;
@@
   A frequentist estimator of a Bayesian quantity, to be sure, but a frequentist estimator nonetheless.}
-  by ordering the $q(\pars_k)$ values,
+  by ordering the $g(\pars_k)$ values,
   and finding the $\pars$ values such that some fraction (0.68 or 0.95)
   of the samples lie between them.
-Again, to form these limits, $q(\pars)$ must be a scalar function (so sorting is well defined).
+Again, to form these limits, $g(\pars)$ must be a scalar function (so sorting is well defined).
 Because they are so afraid of being confused with frequentists,

### 2017-10-15 mcmc.tex
typo; less math

@@
   which the samples are drawn, the histogram (thought of as a step
-  function) must integrate to unity.}  looks very much like the pdf
-from which the samples were drawn.
+  function) must integrate to unity.})
+looks very much like the pdf from which the samples were drawn.
 This is a way to ``reconstruct'' the pdf from the sampling:

### 2017-10-15 mcmc.tex
nopagebreak doesn't work; other tweaks

@@
 \section{What is a sampling?}\label{sec:sampling}
-
+\nopagebreak
 Heuristically, a sampling $\setof{\pars_k}_{k=1}^K$ from some pdf $p(\pars)$
@@
 $x$ from a uniform distribution in the interval $0<x<1$.
-Now use those $K$ numbers to compute a the sampling estimate of the
-mean, variance, skewness, and kurtosis (four estimates)
+Now use those $K$ numbers to compute a sampling estimate of the
+mean, variance, skewness, and kurtosis (four estimates; look up
+definitions as needed).
 Make four plot of these four estimates as a function of $1/K$ or
@@
 \section{Metropolis--Hastings MCMC}\label{sec:MH}
-
+\nopagebreak
 The simplest algorithm for \MCMC\ is the Metropolis--Hastings algorithm (\MH\ \MCMC).\note{%
@@
   evaluated at the observed data $\data$.
-The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
+The second input is a handle to a proposal pdf function $q(\pars'\given\pars)$ that can deliver samples,
   such that the algorithm can draw a new position $\pars'$ in the parameter space
@@
 Importantly---for the algorithm given above to work correctly---the
-  proposal pdf must satisfy a ``detailed balance'' condition; it
-  must have the property that
+  proposal pdf must satisfy a ``detailed-balance'' condition\footnote{This
+  detailed-balance condition in \equationname~(\ref{eq:detailed-balance})
+  requires implicitly that the function $q(\pars'\given\pars)$ is also
+  properly normalized. That is, that it integrates (over $\pars'$) to unity
+  for any setting of $\pars$. That is an extremely technical point, but stands
+  as a reminder that you don't want to mess with detailed balance casually!};
+  it must have the property that
 \begin{eqnarray}
 q(\pars'\given\pars) &=& q(\pars\given\pars')
-\quad ;
+\label{eq:detailed-balance}\quad ;
 \end{eqnarray}
@@
   then just stick with pdfs that are symmetric in $\pars'$ and $\pars$,
-  like the Gaussian, or a centered uniform distribution and forget about
+  like the Gaussian, or a centered uniform distribution, and forget about
   it.
@@
 Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
-pdf with mean zero and variance 1.
+pdf for $x'$ with mean $x$ and variance 1.
 Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
@@
 For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
-density with zero mean and variance tensor set to the two-dimensional
+density with mean at $[x, y]$ and variance tensor set to the two-dimensional
 identity matrix.
@@
 Why, in all the previous problems, did we give the proposal
-distributions $q(x'\given x)$ zero mean?
+distributions $q(x'\given x)$ a mean of $x$?
 What would be bad if we hadn't done that?
 Re-do \problemname~\ref{prob:twod}a with a proposal $q(x'\given x)$ with a
-non-zero mean and see what happens.
-Bonus points: Modify the acceptance--rejection criterion to deal with
+stupidly shifted mean of $x + 2$ and see what happens.
+\emph{Bonus points:} Modify the acceptance--rejection criterion to deal with
 the messed-up $q(x'\given x)$ and show that everything works once
@@
 \section{Likelihoods and priors}\label{sec:likelihood}
-
+\nopagebreak
 \MCMC\ is used to obtain samples $\pars_k$ from a pdf $p(\pars)$,
@@
 If it is flat with bounds (that is, proper), the \code{ln_prior()}
-should check the bounds and return \code{-Inf} values if the parameter
+should check the bounds and return \code{-Inf} values when the parameter
 vector is out of bounds.
@@
 accept--reject step of the \MCMC\ method, the programming language handles
-properly the \code{-Inf} values.\footnote{If you are working in a language
+properly the \code{-Inf} values.\note{If you are working in a language
 that doesn't have an \code{Inf} or doesn't evaluate comparisons correctly
@@
 \section{Convergence}\label{sec:convergence}
-
+\nopagebreak
 A key question for an \MCMC\ operator---\emph{the} key question in some
@@
 \section{Tuning}\label{sec:tuning}
-
+\nopagebreak
 Most \MCMC\ methods make use of something like a ``proposal distribution''
@@
 \section{Initialization and burn-in}\label{sec:initialization}
-
+\nopagebreak
 Just as most (though not all) \MCMC\ methods require a choice of
@@
 \section{Results, error bars, and figures}\label{sec:results}
-
+\nopagebreak
 For a committed probabilistic scientist---frequentist or
@@
 \section{Troubleshooting and advice}\label{sec:trouble}
-
+\nopagebreak
 When issues arise with \MCMC\ sampling,
@@
 \section{More sophisticated sampling methods}\label{sec:methods}
-
+\nopagebreak
 As we have discussed above (\sectionname~\ref{sec:MH}),
@@
 \clearpage
-\paragraph{Acknowledgements:}
+\acknowledgements
 We would like to thank

### 2017-10-15 mcmc.tex
tweaks

@@
 That is, the projection of the sampling to the subspace of interest produces
-  a sampling of the marginalized pdf, marginalizing (or projecting) out the nuisance parameters.%
-  \note{This point about marginalization is, once again, a use of the sampling to perform an \emph{integral};
+  a sampling of the marginalized pdf, marginalizing (or projecting) out the
+  nuisance parameters.\note{This point about marginalization is, once again,
+        a use of the sampling to perform an \emph{integral};
         the marginalized pdf is obtained from the full pdf by an integration.
@@
   where there are always parameters with very different levels of importance to the scientific conclusions.
-This point will return again below (\sectionname~\ref{sec:results}).
+This point generalizes from a subspace to any function of the parameters;
+and it will return again below (\sectionname~\ref{sec:results}).
 

### 2017-10-15 mcmc.tex
g functions

@@
 
-\section{Convergence}\label{sec:convergence}
-\nopagebreak
+\section{Convergence}\label{sec:convergence}%
+\nopagebreak%
 A key question for an \MCMC\ operator---\emph{the} key question in some
@@
 The question of whether a sampling is converged to a representative sampling
-  of the posterior pdf is actually outside the domain of science,
+  of the posterior pdf is actually \emph{outside the domain of science},
   because the domain of science is the domain of questions answerable by scientists.
@@
 definitely converged, or that you have the posterior pdf correct to
-some given level of accuracy.  The reason is related to the ``no free
-lunch'' theorem of discrete optimization: In most problems there is no
+some given level of accuracy.  The reason is related to the above-mentioned ``no free
+lunch'' theorem of discrete optimization \citep{nofreelunch}: In most problems there is no
 way you could have searched your parameter space finely enough to know
@@
 second half of the chain, or indeed as any substantial subset.
+Or, relatedly, you have sampled long enough when different walkers initialized
+differently and run with different random number seeds create posterior inferences
+that are substantially the same.
 
@@
 
-The autocorrelation function is defined by
+The autocorrelation function $C_g$ is defined by
 \begin{eqnarray}
-C_f(\Delta) &\equiv& \frac{E[(f(\pars_k)-\mu_f)\,(f(\pars_{k+\Delta})-\mu_f)]}
-                          {E[(f(\pars)-\mu_f)^2]}
+C_g(\Delta) &\equiv& \frac{E[(g(\pars_k)-\mu_g)\,(g(\pars_{k+\Delta})-\mu_g)]}
+                          {E[(g(\pars)-\mu_g)^2]}
 \label{eq:autocorrelationfn}\\
-\mu_f &=& E[f(\pars)]
+\mu_g &=& E[g(\pars)]
 \quad ;
 \end{eqnarray}
-where $f(\pars)$ is a scalar valued function of the parameters.
+where $g(\pars)$ is a scalar valued function of the parameters, and $\Delta$
+is a lag (difference in index number) between points in the chain.
 In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
@@
 \begin{eqnarray}
-C_f(0) &=& 1
+C_g(0) &=& 1
 \\
-\lim_{\Delta\rightarrow\infty} C_f(\Delta) &=& 0
+\lim_{\Delta\rightarrow\infty} C_g(\Delta) &=& 0
 \quad ,
@@
 \begin{eqnarray}
-\tau_f = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_f(\Delta)
+\tau_g = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_g(\Delta)
 \label{eq:autocorrelationtime}\quad.
@@
 simplify your problem.
-
 \end{itemize}
-
-DFM: Add an example of each of these plots.
 

### 2017-10-15 mcmc.tex
ref

@@
   it proposes steps too small---%
-  almost all steps will be accepted (REFERENCE ALGORITHM HERE),
+  almost all steps will be accepted (recall the acceptance-rejection step from \sectionname~\ref{sec:MH})
   but it will take a long time to move anywhere because of timidity.

### 2017-10-15 mcmc.tex
autocorrelation advice

@@
 then along those dimensions the integral becomes a discrete sum.
-This latter is a detail to which we return below. HOGG DO WE?
+This latter is a detail to which we return below. \textbf{HOGG DO WE?}
 
@@
 \begin{eqnarray}
-  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_q}{K}\,\Var_{p(\pars)}[g(\pars)]
+  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_g}{K}\,\Var_{p(\pars)}[g(\pars)]
   \\
-  &\approx& \frac{\tau_q}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
+  &\approx& \frac{\tau_g}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
     E_{p(\pars)}[g(\pars)]\right)^2
@@
 (\ref{eq:autocorrelationtime}) does not by itself provide an
-\emph{estimator} for the time.
-We might first try the na\"ive estimator
+\emph{estimator} for the time. Estimating the auto-correlation time is
+notoriously difficult: It is a two-point statistic, and two-point statistics
+are much harder to estimate than one-point statistics.\note{See: All of cosmology!}
+If you are not a hard-core user of MCMC, and if all you want is heuristic indicators
+of convergence, then what you should take from this section is that the auto-correlation
+time is involved in variance estimates through \equationname~(\ref{eq:tauvar}) but that
+it is hard to estimate.
+
+If you want to think about autocorrelation estimation:\note{One of us (DFM) has written
+  much more about autocorrelation estimation in the \project{emcee} documentation
+  at \url{http://emcee.readthedocs.io/en/develop/tutorials/autocorr/}.}
+You might first try the na\"ive estimator
 \begin{eqnarray}\label{eq:tauestnaive}
@@
 An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
-trusted if the chain is many times ($\sim 10$) longer than $\tau$.
+trusted if the chain is many times ($\sim 100$) longer than $\tau$.\footnote{This
+  point is also emphasized at the aforementioned tutorial
+  \url{http://emcee.readthedocs.io/en/develop/tutorials/autocorr/}, where
+  a number like 50 is recommended.}
 
@@
 deeply.
-
-% *********** BJB talking:
-% I suggest partitioning the discussion of different methods into the following
-% two categories:
-%
-%  i) Methods for getting around the target distribution
-% ii) Methods that change the target distribution to something
-% other than the posterior pdf, to help the sampling somehow
-%
-% i) Would include things like M-H, the stretch move,
-% Hamiltonian, slice sampling (maybe skip the latter)
-% ii) would include things like annealing and Nested Sampling
-
-%%%% Hogg talking:
-% DFM: In this section, I think we should just summarize the other
-% methods, not describe them in detail.  Then we can cite the relevant
-% papers for the enthusiastic.
 

### 2017-10-15 mcmc.tex
tuning

@@
 \nopagebreak
-Most \MCMC\ methods make use of something like a ``proposal distribution''
+Most \MCMC\ methods make use of something like a ``proposal distribution'',
   which determines what kinds of steps the walker can take
   as it random-walks through the parameter space.
+This is the function we called $q(\pars'\given\pars)$ in \sectionname~\ref{sec:MH}.
 The user generally has a lot of control over what this proposal distribution might be,
@@
   much less put it into some kind of optimization loop.
-We usually use proxies for this of various kinds.
+When tuning, we usually use proxies for this.
 
@@
   taking a random step in just one parameter at a time.\note{This has a lot to do with Gibbs sampling, which is discussed in \sectionname~\ref{sec:methods}.}
-Aside from involving some persistence of state,
-  this kind of proposal distribution can be valuable,
+This kind of proposal distribution can be valuable,
   in part because it reduces the (almost impossible) $D$-dimensional tuning problem

### 2017-10-15 mcmc.tex
problems and punctuation

@@
 \begin{eqnarray}
-p(\theta_1,\,\theta_2) &\propto&
+f(\theta_1,\,\theta_2) &=&
     \exp\left(-\frac{100\,(\theta_2-{\theta_1}^2)^2+(1-\theta_1)^2}{20}\right)
@@
 What autocorrelation time do you get?
-Compare to what \project{emcee} gets.
+Compare to what \project{emcee}\note{Available at \url{http://dfm.io/emcee}.} gets.
 \end{problem}
@@
 If this might be an issue, then the best diagnosis is to start many
-chains in parallel and check that they lead to identical (or very
+chains in parallel with different initializations,
+  and check that they lead to identical (or very
 similar) posterior inferences (same pdf location and shape and
@@
 
-But it is worth remembering that just as \MCMC\ is not a good optimizer
-(generically, samples will not lie close to the maximum of the posterior pdf),
+But it is worth remembering that---just as \MCMC\ is not a good optimizer
+(generically, samples will not lie close to the maximum of the posterior pdf)---%
 it is also not a good \emph{search algorithm}.
@@
 Bayesian---there is no ``answer'', there are only probability density
-functions (pdfs).  Any time in a paper or in an email or in a conversation
-we say what ``the answer is''---even if we say it with an error
-bar---we have departed the probabilistic program.  There are many
-principled ways to depart the program; in another \documentname, we
-discuss the economic explanation of how we can make hard decisions in
-the context of probabilistic reasoning\note{We intend to write a subsequent
-  \documentname\ in this series about these issues.  The idea is that you
+functions (pdfs).  In a paper or in an email or in a conversation
+we might say what ``the answer is'', but even if we say it with an error
+bar, we have departed the probabilistic program.  There are many
+principled ways to depart the program; in another forum, we hope to say more
+about the economic explanation of how we can make hard decisions in
+the context of probabilistic reasoning\note{The idea is that you
   can make principled decisions by optimizing the expected utility under

### 2017-10-15 mcmc.tex
tweaks

@@
 
-One thing we can keep in mind to guide us in this is what we said above (\sectionname~\ref{sec:sampling}),
+One thing we can keep in mind to guide us in this is what we said above
+  (in \sectionname~\ref{sec:sampling}),
   which is that samplings are good for doing \emph{integrals}.
@@
   and the median-of-sampling value is just the $[K/2]$th value of $g(\pars_k)$
-  when the $g(\pars_k)$ have been ordered (sorted).
+  when the $g(\pars_k)$ have been ordered (sorted) from lowest to highest.
 These are both produced by integrals;
@@
 As we mentioned above, scientists often like to report the ``best sample'';
-  that is, the sample with the highest posterior pdf value.
+  that is, the sample with the highest posterior pdf or $f(\pars)$ value.
 This is not usually a good idea.
@@
 in general the need for samples grows exponentially (or perhaps even
-factorially!)\ with the number of dimensions.
+factorially!) with the number of dimensions.
 So if you have a ten-dimensional space, expect to be publishing your
@@
 \begin{itemize}
-
 \item \emph{Trace plots}~---~The first set of plots are the ``trace

### 2017-10-16 mcmc.tex
troubleshooting

@@
 then along those dimensions the integral becomes a discrete sum.
-This latter is a detail to which we return below. \textbf{HOGG DO WE?}
+This latter is a detail to which we return below (briefly, in
+  \sectionname~\ref{sec:trouble}).
 
@@
 If you really need to check all of parameter space, you should put
-\MCMC\ aside, and do an exhaustive search.\note{You might want to open a
-  bottle of wine or two while the code runs.}
+\MCMC\ aside, and do an exhaustive search (which will in general take an enormous
+  amount of time).
 
@@
 This test is most effective as a true end-to-end test of your model
-  if you structure your code such that the log probability function \code{lnprob}
+  if you structure your code such that the log probability function \code{ln_f}
   input to the \MCMC\ scheme
-  is set up as a sum of a log likelihood function \code{lnlikelihood}
-  and a prior function \code{lnprior}.
+  is set up as a sum of a log-likelihood function \code{ln_likelihood}
+  and a log-prior function \code{ln_prior}.
 The sampler can be switched to sampling the prior by a simple replacement
-  of the \code{lnlikelihood} function
+  of the \code{ln_likelihood} function
   with a trivial function that always returns zero.
@@
   or else a think-o about how your priors ought to look.
-Either way, diagnosis in order.
+Either way, diagnosis is in order.
 
@@
   but they usually point to code issues.
-In general, plotting the likelihood function as a function of parameter along
+In general, plotting the likelihood function as a function of parameters along
   slices in parameter space often reveals bugs or think-os.

### 2017-10-16 mcmc.tex
fewer other methods

@@
 \author[0000-0002-9328-5652]{Dan Foreman-Mackey}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 \affil{Department~of~Astronomy, University~of~Washington, Box 351580, Seattle, WA 98195, USA}
@@
 \paragraph{Gibbs:}
-In many problems, the parameters---the components or entries of the
+Neither \MH\ \MCMC\ nor most ensemble methods good at going to very
+  large numbers of parameters.
+Even dozens of parameters can be enough to drive autocorrelation times
+  extremely large.
+With larger numbers of parameters, new methods need to be considered.
+Gibbs sampling is worth considering when the number of parameters is
+  large, and the different parameters have different ``scopes'' in the problem.
+
+In many problems that have large numbers of parameters,
+  the parameters---the components or entries of the
   blob $\pars$---are not equivalent at all.
@@
 For problems with these structures, Gibbs samplers are ideal, and
-there are good implementations.\note{For example, \project{JAGS}; find it on the internets.}
+there are good implementations.\note{For example, \project{STAN} and \project{JAGS}; find them on the internets. DFM INSERT URLs AND REFERENCES.}
 
@@
 
-Gibbs sampling can be very good for exploiting multiprocessing:
+Gibbs sampling can also be very good for exploiting multiprocessing:
 Global parameter updates might require a serial re-calculation of the
@@
 
-\paragraph{Hamiltonian or Hybrid:}
+\paragraph{Hamiltonian:}
+At time of writing, the premier approaches for problems with large numbers
+  of parameters are Hamiltonian methods.\note{For a review of the theory
+  behind these methods, see \cite{hamiltonian}.}
+These methods perform very well at large dimensions, but require (for speed)
+  analytic derivatives of the function $f(\pars)$ with respect to the parameters.
+
 In many cases of interest, the function $f(\pars)$ can be analytically
@@
 
-DFM: more here!  For example, does this work better in large numbers
-of dimensions?  If so, we should say so.  And cite \citep{hamiltonian}.
+DFM: more here! The reason these methods are ``Hamiltonian'' is that... And cite \citep{hamiltonian}.
 
@@
 must be greater than or equal to the approximate function everywhere!
-
-\paragraph{Slice:}
-DFM ... When is slice sampling a good idea?  DFM, can you help here?
-
-Slice sampling makes use of the point that (in one dimension) a fair
-sampling of $f(\pars)$ is a projection onto the $\pars$ axis of a
-uniform sampling of the two-dimensional region under the
-two-dimensional plot of $f(\pars)$ \vs\ $\pars$.
-The $D$-dimensional generalization is that a fair sampling of
-$f(\pars)$ is a projection onto the $\pars$ $D$-dimensional
-hypersurface of $\pars$ of a uniform sampling of the $D+1$-dimensional
-hypervolume under the hypersurface $f(\pars)$ in a hypothetical plot
-of $f(\pars)$ over the whole parameter space.
-Thinking in terms of the two-dimensional case (for simplicity), slice
-samplers make both ``horizontal'' moves in the $\pars$ direction at
-fixed $f(\pars)$ and also ``vertical'' moves in the $f(\pars)$
-direction.
-Because the sampling in the higher-dimensional space is uniform, it is
-easy to make moves to very different parts of parameter space, or
-different peaks or modes in the function $f(\pars)$.
-In our own research, we use a variant of slice sampling for problems
-involving samplings governed by Gaussian Processes.\note{DFM: Cite us
-  and Murray here.}
 
@@
 (the Bayes factor or fully marginalized likelihood).
-
-\paragraph{Data augmentation:}
-...HOGG Look at the relevant chapter in Brooks.
 

### 2017-10-16 mcmc.tex
brutal whipping into shape

@@
 % - \MCMC not MCMC
-% - These documents are ``Chapters'', no?  Use \documentname.
 % - [LaTeX] newline after every full stop for proper git diffing
@@
 All that said, and adhering to the traditions of the \project{Data
-  Analysis Recipes} project\note{Every \documentname\ in the \project{Data
+  Analysis Recipes} project\note{Every entry in the \project{Data
     Analysis Recipes} series begins with a rant in which we argue that
@@
     We make the same request in a less trivial context in our
-    model-fitting \documentname\ (\citealt{fittingaline},
+    model-fitting screed (\citealt{fittingaline},
     \problemname~6 and 7), where we ask the Reader to implement \MCMC\ for
@@
   that in the Metropolis--Hastings algorithm, we permit the proposal
-  distribution to disobey detailed balance and then we correct the
-  acceptance/rejection step to recover detailed balance.} is the
+  distribution to disobey detailed balance but correct the
+  accept--reject step to recover detailed balance.} is the
 following: We have generated some set of samples, the most recent of
@@
   and given the intuition you gain by doing it (at least once) yourself.\note{%
-    We also make this point in a previous \documentname\ (\citealt{fittingaline});
-    the ambitious reader will find in that \documentname\ a solid example project
+    We also make this point in a previous piece (\citealt{fittingaline});
+    the ambitious reader will find in that piece a solid example project
     for using \MCMC\ in a real context.}
@@
   effects and noises for which there is (and can be, for deep reasons,
-  to be discussed in other \documentname s in this series) no extremely precise model.  So in
+  to be discussed in other documents in this series) no extremely precise model.  So in
   detail, any model of any specific observation must be wrong, even in
@@
 deeply.
+There is no sense in which this list is comprehensive; these are just
+methods (or method classes) that are likely to be useful to projects similar
+to projects we ourselves have executed.
 
@@
 \project{emcee}\note{\citet{emcee}}, which is popular in the
-astrophysical sciences, but there are others\note{A few ensemble methods that
+physical sciences, but there are others\note{A few ensemble methods that
 are being used in astronomy: Differential Evolution MCMC (\citealt{demcmc,
@@
 
-DFM: more here! The reason these methods are ``Hamiltonian'' is that... And cite \citep{hamiltonian}.
+The reason these methods are ``Hamiltonian'' is that they use a kind of
+  Hamiltonian dynamics to improve the sampling: They augment the position
+  of the walker in parameter space with a fictitious momentum in
+  parameter space, and use the derivatives in the accept-reject step.
+The momentum helps the sampler move through the space on a more efficient
+  variant of the random walk.
+These samplers are essentially the only good options when the number of
+  parameters gets very large.
 
@@
 approximate function as a weight or probability.
-We use this quite a bit in our hierarchical inferences.\note{%
+We use this quite a bit in our hierarchical inferences,\note{%
   For example, we use importance sampling to convert individual-planet
@@
   consistent hierarchical model in \citet{eccentricity}.}
+and in problems where the dimensionality is low.\note{see,
+  for example, \citet{joker}}
 
@@
 distribution being sampled is not exactly $f(\pars)$, but the results
-of the sampling are only kept for those samples taken at unit
-temperature; these in the end make a fair sampling.
-
-% DFM: Is the last sentence of the above paragraph correct??
+can be transformed into a fair sampling.
 
@@
           Brendon Brewer (Auckland),
-          Will Farr (Flatiron)
+          Will Farr (Flatiron),
+          Marla Geha (Yale),
           Jonathan Goodman (NYU),
@@
   \textit{Statistica Sinica}, \textbf{20}(1) 343--364.
+\bibitem[Price-Whelan \etal(2017)]{joker}
+  Price-Whelan, A.~M., Hogg, D.~W., Foreman-Mackey, D. \& Rix, H.-W., 2017,
+  \href{http://adsabs.harvard.edu/abs/2017ApJ...837...20P}{\project{The Joker}: A custom Monte Carlo sampler for binary-star and
+  exoplanet radial-velocity data},
+  \textit{Astrophys.\ J.} \textbf{837} 20
+  \doi{10.3847/1538-4357/aa5e50}.
 \bibitem[Skilling(2006)]{skilling}

### 2017-10-16 .ci/travis.sh
trying to fix travis build

@@ -1,0 +1,25 @@
+#!/bin/bash -x
+
+wget https://repo.continuum.io/miniconda/Miniconda3-latest-Linux-x86_64.sh -O miniconda.sh;
+
+bash miniconda.sh -b -p $HOME/miniconda
+export PATH="$HOME/miniconda/bin:$PATH"
+hash -r
+conda config --set always_yes yes --set changeps1 no
+conda update -q conda
+conda info -a
+
+# Python problems
+conda create --yes -n test python=$PYTHON_VERSION numpy scipy matplotlib
+source activate test
+pip install corner
+
+# Set up tectonic
+conda install -c conda-forge -c pkgw-forge tectonic
+
+# Build the paper
+tectonic --help
+cd problems
+make
+cd ..
+tectonic ms.tex

### 2017-10-16 .ci/upload-paper.sh

@@ -1,0 +1,15 @@
+#!/bin/bash -x
+
+if [[ $MCMC_BUILDING_PAPER == false ]]; then
+  return
+fi
+
+# Push to GitHub
+if [ -n "$GITHUB_API_KEY" ]; then
+  cd $TRAVIS_BUILD_DIR
+  git checkout --orphan pdf
+  git rm -rf .
+  git add -f ms.pdf
+  git -c user.name='travis' -c user.email='travis' commit -m "building the paper"
+  git push -q -f https://dfm:$GITHUB_API_KEY@github.com/davidwhogg/mcmc pdf
+fi

### 2017-10-16 .ci/build.sh
headless

@@ -1,0 +1,6 @@
+#!/bin/bash -x
+
+cd problems
+make
+cd ..
+tectonic ms.tex

### 2017-10-16 .ci/install.sh

@@ -16,10 +16,4 @@ pip install corner
 
 # Set up tectonic
 conda install -c conda-forge -c pkgw-forge tectonic
-
-# Build the paper
 tectonic --help
-cd problems
-make
-cd ..
-tectonic ms.tex

### 2017-10-16 .ci/build.sh
tex file name

@@ -3,4 +3,4 @@
 cd problems
 make
 cd ..
-tectonic ms.tex
+tectonic mcmc.tex

### 2017-10-16 .ci/build.sh
TRAVIS

@@ -1,6 +1,11 @@
 #!/bin/bash -x
 
+# Do the problems
 cd problems
 make
 cd ..
+
+# Set up tectonic & build
+conda install -c conda-forge -c pkgw-forge tectonic
+tectonic --help
 tectonic mcmc.tex

### 2017-10-16 .ci/install.sh

@@ -13,7 +13,3 @@ conda info -a
 conda create --yes -n test python=$PYTHON_VERSION numpy scipy matplotlib
 source activate test
 pip install corner
-
-# Set up tectonic
-conda install -c conda-forge -c pkgw-forge tectonic
-tectonic --help

### 2017-10-16 .ci/upload.sh
paper name

@@ -9,7 +9,7 @@ if [ -n "$GITHUB_API_KEY" ]; then
   cd $TRAVIS_BUILD_DIR
   git checkout --orphan pdf
   git rm -rf .
-  git add -f ms.pdf
+  git add -f mcmc.pdf
   git -c user.name='travis' -c user.email='travis' commit -m "building the paper"
   git push -q -f https://dfm:$GITHUB_API_KEY@github.com/davidwhogg/mcmc pdf
 fi

### 2017-10-16 mcmc.tex
cite rant

@@
 Neither \MH\ \MCMC\ nor most ensemble methods good at going to very
-  large numbers of parameters.
+  large numbers of parameters.\footnote{See the excellent rant ``Ensemble Methods are Doomed to Fail in High Dimensions'' at \url{http://andrewgelman.com/2017/03/15/ensemble-methods-doomed-fail-high-dimensions/}.}
 Even dozens of parameters can be enough to drive autocorrelation times

### 2017-10-16 mcmc.tex

@@
 We usually use quantiles.
-That is, the ``one-sigma'' error bar can be taken to be the half-size of the central (or smallest\note{%
-  We usually create a 68-percent (or 95-percent) interval or region by
-  excluding the top and bottom 16 percent (or 2.5 percent) of the
+That is, the ``one-sigma'' error bar can be taken to be the half-size of the central (or smallest) interval of the parameter that contains 68 percent of the posterior
+  samples.
+In our own work, We usually create a 68-percent interval or region by
+  excluding the top and bottom 16 percent of the
   posterior samples.  Some more aggressive investigators find the
@@
   Both are legitimate, and they are near-identical for distributions
-  that are near-symmetric around the mean.})
-  interval of the parameter that contains 68 percent of the posterior
-  samples.
-The ``two-sigma'' error bar would be the same but for the 95-percent interval.
+  that are near-symmetric around the mean.
+The ``two-sigma'' error bar would be the same but for the 95-percent interval
+  (excluding the top and bottom 2.5 percent).
 Again, these limits can be estimated%

### 2017-10-16 mcmc.tex
formatting

@@
 
+\usepackage{needspace}
+
 \shorttitle{using markov chain monte carlo}
@@
 
-\author[0000-0002-9328-5652]{Dan Foreman-Mackey}
+\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
 \affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affil{Department~of~Astronomy, University~of~Washington, Box 351580, Seattle, WA 98195, USA}
+\affil{NASA Sagan Fellow; Department~of~Astronomy, University~of~Washington, Box 351580, Seattle, WA 98195, USA}
 
@@
 
+\Needspace{7\baselineskip}
 \section{Convergence}\label{sec:convergence}%
@@
 the autocorrelation function measures this.
+
+More formally,
 
@@
   much more about autocorrelation estimation in the \project{emcee} documentation
-  at \url{http://emcee.readthedocs.io/en/develop/tutorials/autocorr/}.}
+  at \url{http://dfm.io/posts/autocorr}.}
 You might first try the na\"ive estimator
@@
   point is also emphasized at the aforementioned tutorial
-  \url{http://emcee.readthedocs.io/en/develop/tutorials/autocorr/}, where
+  \url{http://dfm.io/posts/autocorr}, where
   a number like 50 is recommended.}
@@
 
-\clearpage
+%\clearpage
 \acknowledgements

### 2017-10-16 mcmc.tex
edits to autocorr section

@@
 \quad ,
+    \label{eq:the-real-integral}
 \end{eqnarray}
@@
 \quad .
+    \label{eq:the-real-samples}
 \end{eqnarray}
@@
 \Needspace{7\baselineskip}
-\section{Convergence}\label{sec:convergence}%
+\section{Autocorrelation \& convergence}\label{sec:convergence}%
 \nopagebreak%
@@
   This is very closely related---somehow---to the difficulty of sampling.}
-Another way to put it is that you might have two modes---two separated
-regions of substantial total probability---in the posterior pdf that
+Qualitatively, you can consider the scenario where you have two modes---two
+separated regions of substantial total probability---in the posterior pdf that
 are both important but separated by a large region of low probability.
@@
 the other mode.
-Any test of convergence would be passed, but in fact the sampling
-would be incomplete and not representative.
-That is, knowing that you have a complete and representative sampling
+In most real problems, the situation is much worse, with an unknown number of
+modes, and knowing that you have a complete and representative sampling
 is effectively impossible.
@@
 
-The above heuristics can be made more precise (in terms of the amount
+The above heuristics can be made more precise in terms of the amount
 of deviation one expects between the means and variances, say, of two
@@
 The premier tool for making this heuristic precise, however, is to
-look at the \emph{autocorrelation time} of the chain.
+look at the \emph{``integrated autocorrelation time''} of the chain.
 In general, when one has a sequence $(\pars_1, \pars_2, \theta_3, \cdots)$
@@
 
-More formally,
-
-The autocorrelation function $C_g$ is defined by
-\begin{eqnarray}
-C_g(\Delta) &\equiv& \frac{E[(g(\pars_k)-\mu_g)\,(g(\pars_{k+\Delta})-\mu_g)]}
-                          {E[(g(\pars)-\mu_g)^2]}
-\label{eq:autocorrelationfn}\\
-\mu_g &=& E[g(\pars)]
-\quad ;
-\end{eqnarray}
-where $g(\pars)$ is a scalar valued function of the parameters, and $\Delta$
-is a lag (difference in index number) between points in the chain.
-In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
-of points in the chain separated by $\Delta$ steps, taken relative to the
-variance.
-In the above equation, the assumption is that the samples $\pars_k$ are
-given in the order that they were generated by the \MCMC\ algorithm.
-By construction
-\begin{eqnarray}
-C_g(0) &=& 1
-\\
-\lim_{\Delta\rightarrow\infty} C_g(\Delta) &=& 0
-\quad ,
-\end{eqnarray}
-and by definition, the autocorrelation ``time'' $\tau$ is
-\begin{eqnarray}
-\tau_g = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_g(\Delta)
-\label{eq:autocorrelationtime}\quad.
-\end{eqnarray}
-Heuristically, the autocorrelation time is the number of samples you
-need to take before you get a new independent sample.
-More precisely, the autocorrelation time is the factor by which you
-have to reduce the number of samples when you compute the variance on
-a sampling-based expectation value estimate.
-
-That is, if the expectation value $E[g]$ is computed by the sampling
-approximation given in \equationname~(\ref{eq:expectationvalue}),
-there is an uncertainty on that expectation given by
-\begin{eqnarray}
-  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_g}{K}\,\Var_{p(\pars)}[g(\pars)]
-  \\
-  &\approx& \frac{\tau_g}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
-    E_{p(\pars)}[g(\pars)]\right)^2
-  \quad,
-\label{eq:tauvar}
-\end{eqnarray}
-where we are being a bit loose with terminology because this is the
-variance not on $E_{p(\pars)}[g(\pars)]$ but on our sampling-based
-\emph{estimator} for
-$E_{p(\pars)}[g(\pars)]$, and the $E_{p(\pars)}[g(\pars)]$ inside the sum is
-that estimator too.
-It is slightly off-topic here, but related to this, if you have a long
-chain of length $K$ and with autocorrelation time $\tau$ and you thin
-the chain down to every $\tau$th sample (instead of using all $K$
-samples), you will get just as good sampling-based estimates as if you
-had used all $K$ samples.
-
-It is also worth noting that there is not just one autocorrelation function or
-autocorrelation time for an \MCMC\ chain.
-Instead, there is an autocorrelation time $\tau_g$ for every function
-$g(\pars)$ and these will, in general, be different.
-Therefore, the autocorrelation time used in \equationname~(\ref{eq:tauvar}) is
-the time computed on the chain of $g(\pars)$ values instead of the
-autocorrelation time for $\pars$.
-
-A sampler with a smaller autocorrelation time is just better; you have
+More formally, as discussed in \sectionname~\ref{sec:sampling}, the goal of
+MCMC sampling is to compute integrals of the form given in
+\equationname~\ref{eq:the-real-integral} using the Monte Carlo approximation
+in \equationname~\ref{eq:the-real-samples}.
+The Monte Carlo error introduced by this approximation is proportional to
+$\sqrt{\tau_\mathrm{int}/N}$ where $\tau_\mathrm{int}$ is the
+integrated autocorrelation time and $N$ is the total number of
+samples of $p(\pars)$.
+In other words, $\tau_\mathrm{int}$ is the number of steps required for the
+chain to produce an \emph{independent} sample.
+
+%The autocorrelation function $C_g$ is defined by
+%\begin{eqnarray}
+%C_g(\Delta) &\equiv& \frac{E[(g(\pars_k)-\mu_g)\,(g(\pars_{k+\Delta})-\mu_g)]}
+%                          {E[(g(\pars)-\mu_g)^2]}
+%\label{eq:autocorrelationfn}\\
+%\mu_g &=& E[g(\pars)]
+%\quad ;
+%\end{eqnarray}
+%where $g(\pars)$ is a scalar valued function of the parameters, and $\Delta$
+%is a lag (difference in index number) between points in the chain.
+%In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
+%of points in the chain separated by $\Delta$ steps, taken relative to the
+%variance.
+%In the above equation, the assumption is that the samples $\pars_k$ are
+%given in the order that they were generated by the \MCMC\ algorithm.
+%By construction
+%\begin{eqnarray}
+%C_g(0) &=& 1
+%\\
+%\lim_{\Delta\rightarrow\infty} C_g(\Delta) &=& 0
+%\quad ,
+%\end{eqnarray}
+%and by definition, the autocorrelation ``time'' $\tau$ is
+%\begin{eqnarray}
+%\tau_g = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_g(\Delta)
+%\label{eq:autocorrelationtime}\quad.
+%\end{eqnarray}
+%Heuristically, the autocorrelation time is the number of samples you
+%need to take before you get a new independent sample.
+%More precisely, the autocorrelation time is the factor by which you
+%have to reduce the number of samples when you compute the variance on
+%a sampling-based expectation value estimate.
+
+%That is, if the expectation value $E[g]$ is computed by the sampling
+%approximation given in \equationname~(\ref{eq:expectationvalue}),
+%there is an uncertainty on that expectation given by
+%\begin{eqnarray}
+%  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_g}{K}\,\Var_{p(\pars)}[g(\pars)]
+%  \\
+%  &\approx& \frac{\tau_g}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
+%    E_{p(\pars)}[g(\pars)]\right)^2
+%  \quad,
+%\label{eq:tauvar}
+%\end{eqnarray}
+%where we are being a bit loose with terminology because this is the
+%variance not on $E_{p(\pars)}[g(\pars)]$ but on our sampling-based
+%\emph{estimator} for
+%$E_{p(\pars)}[g(\pars)]$, and the $E_{p(\pars)}[g(\pars)]$ inside the sum is
+%that estimator too.
+%It is slightly off-topic here, but related to this, if you have a long
+%chain of length $K$ and with autocorrelation time $\tau$ and you thin
+%the chain down to every $\tau$th sample (instead of using all $K$
+%samples), you will get just as good sampling-based estimates as if you
+%had used all $K$ samples.
+
+%It is also worth noting that there is not just one autocorrelation function or
+%autocorrelation time for an \MCMC\ chain.
+%Instead, there is an autocorrelation time $\tau_g$ for every function
+%$g(\pars)$ and these will, in general, be different.
+%Therefore, the autocorrelation time used in \equationname~(\ref{eq:tauvar}) is
+%the time computed on the chain of $g(\pars)$ values instead of the
+%autocorrelation time for $\pars$.
+
+A sampler with a smaller integrated autocorrelation time is better; you have
 to do fewer $f(\pars)$ calls per independent sample, and you have to
@@
 
-Of course, the autocorrelation-time definition
-(\ref{eq:autocorrelationtime}) does not by itself provide an
-\emph{estimator} for the time. Estimating the auto-correlation time is
+We won't go into details about how to estimate $\tau_\mathrm{int}$, but it is
 notoriously difficult: It is a two-point statistic, and two-point statistics
-are much harder to estimate than one-point statistics.\note{See: All of cosmology!}
-If you are not a hard-core user of MCMC, and if all you want is heuristic indicators
-of convergence, then what you should take from this section is that the auto-correlation
-time is involved in variance estimates through \equationname~(\ref{eq:tauvar}) but that
-it is hard to estimate.
-
-If you want to think about autocorrelation estimation:\note{One of us (DFM) has written
-  much more about autocorrelation estimation in the \project{emcee} documentation
+are much harder to estimate than one-point statistics.\note{See: All of
+cosmology!}
+If you are not a hard-core user of MCMC, and if all you want is
+heuristic indicators of convergence, then what you should take from this
+section is that the autocorrelation time is involved in variance estimates,
+but that it is hard to estimate.
+If you want to think about autocorrelation estimation, more detailed
+references can be found elsewhere.\note{%
+    A canonical reference is a set of lecture notes by Alan Sokal
+    \citep{sokal}.
+    Another good reference is a blog post by one of us (DFM) that can be found
     at \url{http://dfm.io/posts/autocorr}.}
-You might first try the na\"ive estimator
-\begin{eqnarray}\label{eq:tauestnaive}
-\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
-\end{eqnarray}
-where the sum is over the autocorrelation function estimated based the chain
-using \equationname~(\ref{eq:autocorrelationfn}).
-In practice, this is a high variance estimator of the integrated
-autocorrelation time and this variance doesn't go to zero for an infinitely
-long chain.\note{%
-  A.~Sokal has a clear discussion of this topic and several proposals for
-  robust autocorrelation time estimation in a set of lecture notes
-  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
-Instead, we can get a much more robust estimate of the autocorrelation time by
-restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
-where
-$\tau \le M \ll N$
-\begin{eqnarray}\label{eq:tauest}
-\tau &=& 1 + 2\,\sum_{\Delta = 1}^{M} C(\Delta)
-\end{eqnarray}
-In practice, $M$ can be chosen iteratively to satisfy these constraints but it
-will only be as reliable as the chain itself.
-An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
-trusted if the chain is many times ($\sim 100$) longer than $\tau$.\footnote{This
-  point is also emphasized at the aforementioned tutorial
-  \url{http://dfm.io/posts/autocorr}, where
-  a number like 50 is recommended.}
-
-One simple and sensible test of convergence is the Gelman--Rubin
-diagnostic \note{\citet{gelmanrubin}.}, which compares the variance
-(in one parameter, or your most important parameter, or all
-parameters) \emph{within} a chain to the variance \emph{across}
-chains.
+
+%Of course, the autocorrelation-time definition
+%(\ref{eq:autocorrelationtime}) does not by itself provide an
+%\emph{estimator} for the time. Estimating the auto-correlation time is
+%notoriously difficult: It is a two-point statistic, and two-point statistics
+%are much harder to estimate than one-point statistics.\note{See: All of cosmology!}
+%If you are not a hard-core user of MCMC, and if all you want is heuristic indicators
+%of convergence, then what you should take from this section is that the auto-correlation
+%time is involved in variance estimates through \equationname~(\ref{eq:tauvar}) but that
+%it is hard to estimate.
+
+%If you want to think about autocorrelation estimation:\note{One of us (DFM) has written
+%  much more about autocorrelation estimation in the \project{emcee} documentation
+%  at \url{http://dfm.io/posts/autocorr}.}
+%You might first try the na\"ive estimator
+%\begin{eqnarray}\label{eq:tauestnaive}
+%\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
+%\end{eqnarray}
+%where the sum is over the autocorrelation function estimated based the chain
+%using \equationname~(\ref{eq:autocorrelationfn}).
+%In practice, this is a high variance estimator of the integrated
+%autocorrelation time and this variance doesn't go to zero for an infinitely
+%long chain.\note{%
+%  A.~Sokal has a clear discussion of this topic and several proposals for
+%  robust autocorrelation time estimation in a set of lecture notes
+%  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
+%Instead, we can get a much more robust estimate of the autocorrelation time by
+%restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
+%where
+%$\tau \le M \ll N$
+%\begin{eqnarray}\label{eq:tauest}
+%\tau &=& 1 + 2\,\sum_{\Delta = 1}^{M} C(\Delta)
+%\end{eqnarray}
+%In practice, $M$ can be chosen iteratively to satisfy these constraints but it
+%will only be as reliable as the chain itself.
+%An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
+%trusted if the chain is many times ($\sim 100$) longer than $\tau$.\footnote{This
+%  point is also emphasized at the aforementioned tutorial
+%  \url{http://dfm.io/posts/autocorr}, where
+%  a number like 50 is recommended.}
+
+Besides estimating the integrated autocorrelation time, another simple and
+sensible test of convergence is the Gelman--Rubin
+diagnostic,\note{\citet{gelmanrubin}.} which compares the variance (in one
+parameter, or your most important parameter, or all parameters) \emph{within}
+a chain to the variance \emph{across} chains.
 This requires running multiple chains, and looking at the empirical
@@
 than an autocorrelation time.
-It can also be used to assess the accuracy of the chain for performing
-integrals---by replacing the mean of the parameter with whatever
-integral is of most interest.
+%It can also be used to assess the accuracy of the chain for performing
+%integrals---by replacing the mean of the parameter with whatever
+%integral is of most interest.
 
@@
 chain of samples using an estimate of the autocorrelation function and a given
-``window'' size $M$ (see \equationname~\ref{eq:tauest}).
+``window'' size $M$ (see \citealt{sokal}).
 Plot the estimated $\tau$ as a function of $M$ for several contiguous segments
@@
 Implement an iterative procedure for automatically choosing $M$.\note{%
-The recipe given on page 16 of A.~Sokal's notes
-(\url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}) might be helpful.
-Note that the definition of $\tau$ in that document is half of the definition
-that we use.}
+The recipe given on page 16 of A.~Sokal's notes \citep{sokal} might be helpful.
+Note that the definition of $\tau$ that we adopt is twice the value used by
+Sokal.}
 Overplot this estimate on the plot of $\tau(M)$ and the result should look
@@
   \doi{10.1214/06-BA127}.
+\bibitem[Sokal(1997)]{sokal}
+    Sokal, A., 1997,
+    Monte Carlo methods in statistical mechanics: foundations and new algorithms.
+    \textit{Functional integration.}\ 131--192
 \bibitem[Ter Braak(2006)]{demcmc}

### 2017-10-16 mcmc.tex
acronym improvements; abstract tweak

@@
 % -----
-% - deal with \MCMC\ macro in footnotes.
-% - Deal with all mentions of ``DFM'' or ``HOGG''
 % - Go through github issues, if any.
-% - Need to mention case where there is an MC integral INSIDE the likelihood function
-% - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
-% - Kill all issues in this list, above, and then:
 % - Post on arXiv
@@
 output to produce or report parameter estimates with associated uncertainties.
+We argue that autocorrelation time is the most important test for convergence,
+as it directly connects to the variances of sampling estimates of quantities of interest.
 We emphasize that sampling is a method for doing integrals; this guides our thinking
@@
 More formally, as discussed in \sectionname~\ref{sec:sampling}, the goal of
-MCMC sampling is to compute integrals of the form given in
+\MCMC\ sampling is to compute integrals of the form given in
 \equationname~\ref{eq:the-real-integral} using the Monte Carlo approximation
@@
 cosmology!}
-If you are not a hard-core user of MCMC, and if all you want is
+If you are not a hard-core user of \MCMC, and if all you want is
 heuristic indicators of convergence, then what you should take from this
@@
 %are much harder to estimate than one-point statistics.\note{See: All of cosmology!}
-%If you are not a hard-core user of MCMC, and if all you want is heuristic indicators
+%If you are not a hard-core user of \MCMC, and if all you want is heuristic indicators
 %of convergence, then what you should take from this section is that the auto-correlation
@@
 This optimal-posterior parameter value is called the
-  ``maximum \foreign{a posteriori}'' or ``MAP'' value for the parameters;
+  ``maximum \foreign{a posteriori}'' or ``\acronym{MAP}'' value for the parameters;
   it is like a maximum-likelihood value but regularized by the prior.
@@
 There are several figures that are useful and should be made every time you
-run MCMC and we'll describe a few here.
+run \MCMC\ and we'll describe a few here.
 
@@
 physical sciences, but there are others\note{A few ensemble methods that
-are being used in astronomy: Differential Evolution MCMC (\citealt{demcmc,
+are being used in astronomy: Differential Evolution \MCMC\ (\citealt{demcmc,
 run-dmc}) and \project{kombine} (Farr \& Farr, in preparation).}.

### 2017-10-16 mcmc.tex
typesetting

@@
 
-\begin{abstract}
+\begin{abstract}\noindent
 Markov Chain Monte Carlo (\MCMC) methods for sampling probability

### 2017-10-16 mcmc.tex
closing #29

--- 2017-10-16 mcmc.tex
@@
 For problems with these structures, Gibbs samplers are ideal, and
-there are good implementations.\note{For example, \project{STAN} and \project{JAGS}; find them on the internets. DFM INSERT URLs AND REFERENCES.}
+there are good implementations.\note{For example, \project{Stan}
+(\url{http://mc-stan.org}).}
 

### 2017-10-16 mcmc.tex
fixing problem numbering

@@
 
-\begin{problem}
+\begin{problem}\label{prob:initialization2}
 Check the scaling you found in \problemname~\ref{prob:initialization}

### 2017-10-16 problems/initialization.py

@@ -15,24 +15,28 @@ np.random.seed(42)
 def log_p(x, mean=2.0, variance=2.0):
     return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
 
-
-x = 1e3
-lp = log_p(x)
-chain = np.empty(6000)
-for step in range(len(chain)):
-    x_prime = x + np.random.randn()
-    lp_prime = log_p(x_prime)
-    if np.random.rand() <= np.exp(lp_prime - lp):
-        x = x_prime
-        lp = lp_prime
-    chain[step] = x
+inits = 10**np.linspace(-5.0, 3.0, 5)
+times = np.empty_like(inits)
+for ix, x in enumerate(10**np.linspace(-5.0, 3.0, 5)):
+    x += 4.0
+    lp = log_p(x)
+    for step in range(int(1e5)):
+        x_prime = x + np.random.randn()
+        lp_prime = log_p(x_prime)
+        if np.random.rand() <= np.exp(lp_prime - lp):
+            x = x_prime
+            lp = lp_prime
+        if np.abs(x - 2.0) < np.sqrt(2.0):
+            times[ix] = step
+            print(ix, inits[ix], times[ix])
+            break
 
 
 fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
-ax.plot(chain, color=COLORS["DATA"])
-ax.xaxis.set_major_locator(pl.MaxNLocator(4))
-ax.yaxis.set_major_locator(pl.MaxNLocator(4))
-ax.set_xlabel("step")
-ax.set_ylabel("$x$")
+ax.plot(inits, times, color=COLORS["DATA"])
+# ax.xaxis.set_major_locator(pl.MaxNLocator(4))
+# ax.yaxis.set_major_locator(pl.MaxNLocator(4))
+# ax.set_xlabel("step")
+# ax.set_ylabel("$x$")
 
-savefig(fig, "initialization.pdf")
+# savefig(fig, "initialization.pdf")

### 2017-10-16 .ci/install.sh
adding solution to problem 20

@@ -10,6 +10,6 @@ conda update -q conda
 conda info -a
 
 # Python problems
-conda create --yes -n test python=$PYTHON_VERSION numpy scipy matplotlib
+conda create --yes -n test python=$PYTHON_VERSION numpy scipy matplotlib tqdm
 source activate test
-pip install corner
+pip install corner https://github.com/dfm/emcee/archive/develop.zip

### 2017-10-16 problems/initialization.py

@@ -15,10 +15,10 @@ np.random.seed(42)
 def log_p(x, mean=2.0, variance=2.0):
     return -0.5 * (x - mean)**2 / variance - 0.5*np.log(2*np.pi*variance)
 
-inits = 10**np.linspace(-5.0, 3.0, 5)
+inits = np.linspace(0.0, 3.5, 8)
 times = np.empty_like(inits)
-for ix, x in enumerate(10**np.linspace(-5.0, 3.0, 5)):
-    x += 4.0
+for ix, x in enumerate(10**inits):
+    x = 2.0 + x * np.sqrt(2)
     lp = log_p(x)
     for step in range(int(1e5)):
         x_prime = x + np.random.randn()
@@ -26,17 +26,15 @@ for ix, x in enumerate(10**np.linspace(-5.0, 3.0, 5)):
         if np.random.rand() <= np.exp(lp_prime - lp):
             x = x_prime
             lp = lp_prime
-        if np.abs(x - 2.0) < np.sqrt(2.0):
+        if np.abs(x - 2.0) < 0.5 * np.sqrt(2.0):
             times[ix] = step
             print(ix, inits[ix], times[ix])
             break
 
 
 fig, ax = pl.subplots(1, 1, figsize=SQUARE_FIGSIZE)
-ax.plot(inits, times, color=COLORS["DATA"])
-# ax.xaxis.set_major_locator(pl.MaxNLocator(4))
-# ax.yaxis.set_major_locator(pl.MaxNLocator(4))
-# ax.set_xlabel("step")
-# ax.set_ylabel("$x$")
+ax.loglog(inits, times, ".-", color=COLORS["DATA"])
+ax.set_xlabel("$\Delta$ initialization; [$\\times \sigma$]")
+ax.set_ylabel("approx.\ number of burn-in steps")
 
-# savefig(fig, "initialization.pdf")
+savefig(fig, "initialization.pdf")

### 2017-10-16 problems/rosenbrock2.py

@@ -1,0 +1,35 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import emcee
+import corner
+import numpy as np
+import matplotlib.pyplot as pl
+
+from plot_setup import setup, SQUARE_FIGSIZE, COLORS, savefig
+
+setup()  # initialize the plotting styles
+np.random.seed(42)
+
+
+def log_p(x):
+    return -(100.0*(x[1]-x[0]**2)**2 + (1-x[0])**2) / 20.0
+
+
+coords = 1.0 + 1e-5*np.random.randn(32, 2)
+nwalkers, ndim = coords.shape
+sampler = emcee.EnsembleSampler(nwalkers, ndim, log_p)
+coords, _, _ = sampler.run_mcmc(coords, 10000, progress=True)
+sampler.reset()
+sampler.run_mcmc(coords, 250000, progress=True)
+
+samples = sampler.get_chain(thin=25*32, flat=True)
+
+fig = corner.corner(samples, 50,
+                    range=[(-10.2, 10.2), (-5.0, 55.0)],
+                    labels=[r"$\theta_1$", r"$\theta_2$"],
+                    truths=np.mean(samples, axis=0),
+                    quantiles=[0.16, 0.5, 0.84])
+savefig(fig, "rosenbrock2.pdf")

### 2017-10-16 mcmc.tex
adding solution to problem 20
adding Sagan ack

@@
 Test your sampler on this density:
-\begin{eqnarray}
+\begin{eqnarray}\label{eq:rosenbrock}
 f(\theta_1,\,\theta_2) &=&
@@
 
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{problems/rosenbrock2}
+\end{center}
+\caption{Solution to \problemname~\ref{prob:rosenbrock2}:
+\MCMC\ samples from the Rosenbrock density
+(\equationname~\ref{eq:rosenbrock}) with the mean shown with blue lines and the
+0.16-, 0.5-, and 0.84th quantiles shown as vertical dashed lines.
+    \label{fig:rosenbrock2}}
+\end{figure}
+
 \section{Troubleshooting and advice}\label{sec:trouble}
@@
           and the The Moore-Sloan Data Science Environment at NYU.
+This work was performed in part under contract with the Jet Propulsion
+Laboratory (JPL) funded by NASA through the Sagan Fellowship Program executed
+by the NASA Exoplanet Science Institute.
 

### 2017-10-17 mcmc.tex
shorten footnote

@@
   \note{If you have a $K$-point sampling in a $D$-dimensional parameter space,
-    there is some best sample $\pars_k^{\best}$.  Now imagine that you want to make
-    that best sample 10 times better---that is, you want to find a point $\pars^{\better}$
-    that is ten times closer to the true optimum.  How many more samples do you need to take?
-    In the limit that $K$ is large, on average, you have to take a factor of
+    there is some best sample $\pars_k^{\best}$.  Now imagine that you want to
+    find a point $\pars^{\better}$
+    that is ten times closer to the true optimum.
+    In the limit that $K$ is large you have to take (on average) a \emph{factor} of
     $10^D$ more samples!

### 2017-10-17 mcmc.tex
footnote improvements

@@
   treating the data as fixed (at the true data values) and the parameters as variable.
-  In this context, the likelihood function is \emph{not} a pdf for the parameters,
+  In this context, the likelihood function is \emph{not} a pdf for anything,
   so it can't be sampled.}
@@
   optimization problem without exhaustive search of all possibilities.
-  This is very closely related---somehow---to the difficulty of sampling.}
+  This is very closely related---somehow---to the difficulty of sampling.\label{note:nofreelunch}}
 Qualitatively, you can consider the scenario where you have two modes---two
@@
 some given level of accuracy.  The reason is related to the above-mentioned ``no free
-lunch'' theorem of discrete optimization \citep{nofreelunch}: In most problems there is no
+lunch'' theorem of discrete optimization (see \notename~\ref{note:nofreelunch}): In most problems there is no
 way you could have searched your parameter space finely enough to know
@@
 sensible test of convergence is the Gelman--Rubin
-diagnostic,\note{\citet{gelmanrubin}.} which compares the variance (in one
+diagnostic\note{\citet{gelmanrubin}.}, which compares the variance (in one
 parameter, or your most important parameter, or all parameters) \emph{within}
@@
 Implement an iterative procedure for automatically choosing $M$.\note{%
-The recipe given on page 16 of A.~Sokal's notes \citep{sokal} might be helpful.
+The recipe given on page 16 of Sokal's notes \citep{sokal} might be helpful.
 Note that the definition of $\tau$ that we adopt is twice the value used by
@@
   even with an \emph{enormous} sampling, if you are in large
-  dimensions, not a single sample will be very near the optimum.}, but
+  dimensions, not a single sample will be very near the optimum.\label{note:highD}}, but
 in low dimensions (few to tens) this is often sensible.
@@
   smallest interval that contains 68 (or 95) percent of the samples.
-  Both are legitimate, and they are near-identical for distributions
-  that are near-symmetric around the mean.
+  Both are legitimate, and they are nearly identical for distributions
+  that are nearly symmetric around the mean.
 The ``two-sigma'' error bar would be the same but for the 95-percent interval
@@
     many normal intuitions one might have (if any) about what is reasonably likely in a pdf
-    are regularly violated.}
+    are regularly violated. See also \notename~\ref{note:highD}.}
 For two, the closeness of the best sample to the posterior mode is a strong function of sample size,
@@
   is rarely correct or specific enough for another investigator to
-  duplicate exactly.  So just augment your samples!}
+  duplicate exactly.  So just augment your samples!\label{note:importance}}
 
@@
 Ensemble samplers react to multiple modes
-  by obtaining very low acceptance ratio,
+  by obtaining very low acceptance ratio\note{Although maybe this isn't
+    fundamental; see \project{kombine} at \url{https://github.com/bfarr/kombine}.},
   because the samplers in the different modes are not easily able to ``help one another''
@@
   physical theories, such as quantum electrodynamics or the
-  cold-dark-matter cosmological model, are strictly true.  First of
+  cold-dark-matter cosmological model, are strictly True with a capital ``T''.  First of
   all, there is no way to know that for sure, and second of all, even
@@
   the (exceedingly unlikely) event that the fundamental model is
-  correct.}
+  correct.\label{note:truth}}
 ``Model checking'' is an enormous subject that goes way beyond the
@@
   chi-squared statistic is a measure of the size of your data!  That's
-  true when your model is wrong (that is, always).}
+  true when your model is wrong (that is, always); see \notename~\ref{note:truth}.}
 
@@
   encoded in your likelihood function as there are in your prior
-  pdfs.}
+  pdfs, probably far more. No data analysis is protected from \emph{subjectivity}.}
 Once again, it is important to find ways to display the results of
@@
   is that any empirical measure of autocorrelation time will give a
-  finite answer, but the density won't be properly sampled, ever.  In
-  principle you can find these with multiple re-starts of the chain
+  finite answer, but the density won't be properly sampled, ever. This
+  problem is related to a problem that any simple empirical measure of
+  the autocorrelation time will be an under-estimate. In
+  principle you can find these badly multimodal cases
+  with multiple re-starts of the chain
   and (things like) the Gelman--Rubin diagnostic \citep{gelmanrubin}.}).
@@
 reducing the tuning requirements.
+This tuning improvement comes at the cost of a burn-in phase at the beginning
+of any run---a phase that lasts a few autocorrelation times---in which the
+ensemble expands or shrinks to fill the posterior volume.
 We are co-authors on one of these methods,
-\project{emcee}\note{\citet{emcee}}, which is popular in the
+\project{emcee}\note{\citet{emcee}.}, which is popular in the
 physical sciences, but there are others\note{A few ensemble methods that
@@
 run-dmc}) and \project{kombine} (Farr \& Farr, in preparation).}.
-It is worty of note that ensemble methods have particular
+It is worthy of note that ensemble methods have particular
 peculiarities about initialization and convergence diagnostics that
@@
 Parallelization is built in to some Gibbs implementations.\note{For
-  example, \project{JAGS} automatically
+  example, \project{JAGS} (\url{http://mcmc-jags.sourceforge.net/}) automatically
   parallelizes Gibbs problems.}
@@
   functions!  The idea is \emph{not} to take numerical (small
-  difference) derivatives of the code; there is no way that
+  difference) derivatives of the code; there is almost no way that
   finite-difference differentiation could help your sampling in the
@@
 approximate function as a weight or probability.
-We use this quite a bit in our hierarchical inferences,\note{%
+We use this quite a bit in our hierarchical inferences\note{%
   For example, we use importance sampling to convert individual-planet
   exoplanet inferences (all performed with dumb priors) into a full-population
-  consistent hierarchical model in \citet{eccentricity}.}
-and in problems where the dimensionality is low.\note{see,
-  for example, \citet{joker}}
+  consistent hierarchical model in \citet{eccentricity}. See also \notename~\ref{note:importance}.},
+and in problems where the dimensionality is low.\note{see, for example, \citet{joker}.}
 
@@
 case the number of stars itself is a parameter, so the number of
-parameters is itself a function of the parameters.\note{Our attempt to sample this very problem is \citet{brewer}.}
+parameters is itself a function of the parameters.\note{Our attempt to sample in this very problem is \citet{brewer}.}
 In these cases you can only use certain kinds of samplers (though M--H

### 2017-10-17 mcmc.tex
barnett

@@
 
-Not being mathematicians, we don't know the precise definition of a
-pdf\note{Seriously.  It has something to do with measure theory.}, but for our purposes here a pdf is
+We don't know all the relevant mathematics (measure theory),
+but for our purposes here a pdf is
 any single-valued (scalar) function that is non-negative everywhere
@@
   at least not in any provable sense.
-The question of whether a sampling is converged to a representative sampling
+The question of whether a sampling is definitely converged to a representative sampling
   of the posterior pdf is actually \emph{outside the domain of science},
   because the domain of science is the domain of questions answerable by scientists.
-That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
+That is, unless you have some bounds on the support and smoothness of
+  the posterior pdf,
+  you can \emph{never know} that you have correctly sampled the posterior pdf,
   despite many statements in the literature to the contrary.\note{%
@@
 chain to produce an \emph{independent} sample.
-
-%The autocorrelation function $C_g$ is defined by
-%\begin{eqnarray}
-%C_g(\Delta) &\equiv& \frac{E[(g(\pars_k)-\mu_g)\,(g(\pars_{k+\Delta})-\mu_g)]}
-%                          {E[(g(\pars)-\mu_g)^2]}
-%\label{eq:autocorrelationfn}\\
-%\mu_g &=& E[g(\pars)]
-%\quad ;
-%\end{eqnarray}
-%where $g(\pars)$ is a scalar valued function of the parameters, and $\Delta$
-%is a lag (difference in index number) between points in the chain.
-%In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
-%of points in the chain separated by $\Delta$ steps, taken relative to the
-%variance.
-%In the above equation, the assumption is that the samples $\pars_k$ are
-%given in the order that they were generated by the \MCMC\ algorithm.
-%By construction
-%\begin{eqnarray}
-%C_g(0) &=& 1
-%\\
-%\lim_{\Delta\rightarrow\infty} C_g(\Delta) &=& 0
-%\quad ,
-%\end{eqnarray}
-%and by definition, the autocorrelation ``time'' $\tau$ is
-%\begin{eqnarray}
-%\tau_g = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_g(\Delta)
-%\label{eq:autocorrelationtime}\quad.
-%\end{eqnarray}
-%Heuristically, the autocorrelation time is the number of samples you
-%need to take before you get a new independent sample.
-%More precisely, the autocorrelation time is the factor by which you
-%have to reduce the number of samples when you compute the variance on
-%a sampling-based expectation value estimate.
-
-%That is, if the expectation value $E[g]$ is computed by the sampling
-%approximation given in \equationname~(\ref{eq:expectationvalue}),
-%there is an uncertainty on that expectation given by
-%\begin{eqnarray}
-%  \Var[E_{p(\pars)}[g(\pars)]] &=& \frac{\tau_g}{K}\,\Var_{p(\pars)}[g(\pars)]
-%  \\
-%  &\approx& \frac{\tau_g}{K^2}\,\sum_{k=1}^K \left(g(\pars_k) -
-%    E_{p(\pars)}[g(\pars)]\right)^2
-%  \quad,
-%\label{eq:tauvar}
-%\end{eqnarray}
-%where we are being a bit loose with terminology because this is the
-%variance not on $E_{p(\pars)}[g(\pars)]$ but on our sampling-based
-%\emph{estimator} for
-%$E_{p(\pars)}[g(\pars)]$, and the $E_{p(\pars)}[g(\pars)]$ inside the sum is
-%that estimator too.
-%It is slightly off-topic here, but related to this, if you have a long
-%chain of length $K$ and with autocorrelation time $\tau$ and you thin
-%the chain down to every $\tau$th sample (instead of using all $K$
-%samples), you will get just as good sampling-based estimates as if you
-%had used all $K$ samples.
-
-%It is also worth noting that there is not just one autocorrelation function or
-%autocorrelation time for an \MCMC\ chain.
-%Instead, there is an autocorrelation time $\tau_g$ for every function
-%$g(\pars)$ and these will, in general, be different.
-%Therefore, the autocorrelation time used in \equationname~(\ref{eq:tauvar}) is
-%the time computed on the chain of $g(\pars)$ values instead of the
-%autocorrelation time for $\pars$.
 
@@
 
-% BJB
-% Please suggest mixtures of scales for M-H proposals.
-% You can mimic the properties of slice sampling but with
-% much easier code.
-
-In a deep sense, the only scalar that makes sense to optimize,
+In terms of long-term computational efficiency,
+  the only scalar that makes sense to optimize,
   when choosing proposal distribution (or, loosely speaking, step size),
@@
   that each step depends \emph{only} on the state at the previous step.
-Violating the Markov property can be very bad;
+Violating the Markov property can be very bad:
   when you violate it,
@@
 Like the acceptance fraction, it can also be used to tune
-  in the case that the proposal distribution loops over parameters;
-  once again, the user gets (in this case) $D$ one-dimensional tunings.
+  in the case that the proposal distribution loops over parameters.
+Once again, the user gets (in this case) $D$ one-dimensional tunings.
 
@@
   there are various tricks and tips to work through.
-% One is to see if any of the parameters has a conditional posterior
-%   probability that is guaranteed to be Gaussian
-%   (this is true for some linear parameters in least-square problems
-%   when the priors are Gaussian, for example).
-% These dimensions can often be sampled \emph{exactly},
-%   outside of the \MCMC\ framework;
-%   the algorithm looks like:
-% Take an \MCMC\ step in the hard parameters,
-%   then take an exact step in the easy parameters,
-%   repeat.
-% Another related trick is to analytically marginalize,
-%   rather than sample,
-%   easy nuisance parameters.
 The huge bag of possible tricks is so large,
@@
 sample, you are fine!
-Ensemble methods (like \project{emcee}\note{\citet{emcee}.}; and
+Ensemble methods (like \project{emcee}\note{\citet{emcee}.}, and
 mentioned below in \sectionname~\ref{sec:methods}) require a burn-in
@@
 make a representative sampling (dozens or hundreds of times, ideally),
-or you can use a high-end sampler (like nested sampling; see \sectionname~\ref{sec:methods}) that is
+or you can use a high-end sampler (like nested sampling, see \sectionname~\ref{sec:methods}) that is
 designed for such problems.
@@
 For a committed probabilistic scientist---frequentist or
-Bayesian---there is no ``answer'', there are only probability density
+Bayesian---there is no single ``answer'', there are only probability density
 functions (pdfs).  In a paper or in an email or in a conversation
@@
   but use as the ``measurement'' the posterior mean,
-  you can get pathological situations in which the posterior mean measurement
+  you can get pathological situations (from large skewness)
+  in which the posterior mean measurement
   is actually \emph{outside} the one-sigma confidence interval.
@@
 This is not usually a good idea.
-For one, \MCMC\ is a sampling algorithm, not an optimization algorithm;
+For one, \MCMC\ is a sampling algorithm, not an optimization algorithm:
   there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.%
@@
 This optimal-posterior parameter value is called the
-  ``maximum \foreign{a posteriori}'' or ``\acronym{MAP}'' value for the parameters;
-  it is like a maximum-likelihood value but regularized by the prior.
+  ``maximum \foreign{a posteriori}'' or ``\acronym{MAP}'' value for the parameters:
+  It is like a maximum-likelihood value but regularized by the prior.
 It is an estimator with some good (and some bad) properties, but it is not the point of \MCMC,
@@
 
-\paragraph{functional testing:}
+\paragraph{Functional testing:}
 In addition to a full set of unit tests for every part of your code
@@
 
-\paragraph{likelihood issues:}
+\paragraph{Likelihood issues:}
 Sometimes there can be problems with the likelihood function itself.
@@
   can make the function noisy or jagged at small scales in parameter space.
-These problems hurt optimizers more than samplers,
+These problems hurt optimizers more than samplers (although Hamiltonian
+  methods depend strongly on good derivatives, see \sectionname~\ref{sec:methods}),
   but they usually point to code issues.
@@
 
-\paragraph{low acceptance fraction:}
+\paragraph{Low acceptance fraction:}
 If you find that you are getting low acceptance fraction
@@
 
-\paragraph{sticky chains:}
+\paragraph{Sticky chains:}
 If you are concerned about acceptance fraction or convergence,
@@
 
-\paragraph{initialization-dependence:}
+\paragraph{Initialization-dependence:}
 If you initialize your \MCMC\ runs in different places in parameter space,
@@
 
-\paragraph{bad initialization:}
+\paragraph{Bad initialization:}
 If you initialize your sampler in a disallowed region of parameter
@@
 
-\paragraph{parameterization problems:}
+\paragraph{Parameterization problems:}
 It helps to think carefully about how to parameterize.
@@
 
-\paragraph{model checking:}
+\paragraph{Model checking:}
 \emph{All models are wrong, but some are useful.}
@@
   variant of the random walk.
-These samplers are essentially the only good options when the number of
+The ergodicity property of (relevant) dynamical systems ensure good sampling (in
+  the infinite-time limit, of course); the connection to dynamics is that
+  the negative logarithm of $f(\pars)$ becomes the potential for the
+  dynamical system.
+At the present day,
+  these samplers are essentially the only good options when the number of
   parameters gets very large.
@@
   consistent hierarchical model in \citet{eccentricity}. See also \notename~\ref{note:importance}.},
-and in problems where the dimensionality is low.\note{see, for example, \citet{joker}.}
+and in problems where the dimensionality is low.\note{See, for example, \citet{joker}.}
 
@@
 easier.
+As with Hamiltonian methods, the sense in which this system makes use
+  of a ``temperature''
+  is that the logarithm of $f(\pars)$ can be related to a potential,
+  in a physical analogy.
 At each step, when the temperature has moved away from unity, the

### 2017-10-17 mcmc.tex
more tweaks

@@
 We argue that autocorrelation time is the most important test for convergence,
-as it directly connects to the variances of sampling estimates of quantities of interest.
+as it directly connects to the uncertainty on the sampling estimate of any
+quantity of interest.
 We emphasize that sampling is a method for doing integrals; this guides our thinking
@@
 
-\paragraph{Functional testing:}
+\paragraph{Functional testing}
 In addition to a full set of unit tests for every part of your code
@@
 
-\paragraph{Likelihood issues:}
+\paragraph{Likelihood issues}
 Sometimes there can be problems with the likelihood function itself.
@@
 
-\paragraph{Low acceptance fraction:}
+\paragraph{Low acceptance fraction}
 If you find that you are getting low acceptance fraction
@@
 
-\paragraph{Sticky chains:}
+\paragraph{Sticky chains}
 If you are concerned about acceptance fraction or convergence,
@@
 
-\paragraph{Initialization-dependence:}
+\paragraph{Initialization-dependence}
 If you initialize your \MCMC\ runs in different places in parameter space,
@@
 
-\paragraph{Bad initialization:}
+\paragraph{Bad initialization}
 If you initialize your sampler in a disallowed region of parameter
@@
 
-\paragraph{Parameterization problems:}
+\paragraph{Parameterization problems}
 It helps to think carefully about how to parameterize.
@@
 
-\paragraph{Model checking:}
+\paragraph{Model checking}
 \emph{All models are wrong, but some are useful.}
@@
 
-\paragraph{Ensembles:}
+\paragraph{Ensembles}
 As we discussed (\sectionname~\ref{sec:tuning}), one of the critical challenges in the use of
@@
 
-\paragraph{Gibbs:}
+\paragraph{Gibbs}
 Neither \MH\ \MCMC\ nor most ensemble methods good at going to very
@@
 
-\paragraph{Hamiltonian:}
+\paragraph{Hamiltonian}
 At time of writing, the premier approaches for problems with large numbers
@@
 
-\paragraph{Importance:}
+\paragraph{Importance}
 Sometimes it is possible to sample efficiently from a distribution
@@
 
-\paragraph{Tempering:}
+\paragraph{Tempering}
 A huge problem for \MCMC\ samplers---which make local moves based on
@@
 
-\paragraph{Nested:}
+\paragraph{Nested}
 Similar to tempering-like methods are a class of methods called ``nested''
@@
 
-\paragraph{Reversible-jump:}
+\paragraph{Reversible-jump}
 Finally, there are some extreme problems in which it is not just hard

### 2018-01-08 response_to_referee.txt
start referee response

Dear Frank:

Please find enclosed our revised version of MS#AAS07769, Data analysis
recipes: Using Markov Chain Monte Carlo by Hogg & Foreman-Mackey.

In what follows, the referee comments are indented and our responses
are not.

Thank you,

David

     ---------------------------------------------------------------------- 
     Referee Report

     The authors provide a simple, and somehow anecdotal discussion
     about MCMC methods (for astronomers, although the title don't
     explicitly says it) in an informal and pragmatic style.

We figured that because this is in the AAS journals, that would supply
the relevant context, though we wouldn't object to modifying the title
slightly.

     The concept of conveying an intuitive description of statistical
     methods for the broad astronomical community is certainly
     something I support, however I would like to ask the authors and
     editor to consider some caveats below.

     Novelty and references

     The draft lacks of some important literature contextualization. A
     critical one is a recent Annual Review of Astronomy and
     Astrophysics entitled Markov Chain Monte Carlo Methods for
     Bayesian Data Analysis in Astronomy
     https://arxiv.org/abs/1706.01629

Yes, we know this paper. We now cite it [WHERE].

     I would like the authors to please explain how their review
     brings something new to the literature, considering the amount of
     good and available material about MCMC.

Good point. We are more explicit about this [WHERE]. The idea is to
delve deeper here, but on MCMC specifically, but not so much about new
developments; more like we are trying to specifically emphasize the
importance of basic good pratices; we say that more clearly now.

     Such as e.g.

     Doing Bayesian Data Analysis, Second Edition: A Tutorial with R,
     JAGS, and Stan, by John Kruschke, and many others

We also now cite some of this literature [WHERE; DO THIS].

     A potential reason is to update the readers about recent
     developments in the field.  But again, the authors seems to
     ignore the current astronomical literature. For example, the
     authors are aware of Statistical Bayesian packages such as JAGS
     and Stan, but do not cite two recent books in Astronomy that make
     use of them, such as:

     Bayesian Methods for the Physical Sciences: Learning from
     Examples in Astronomy and Physics, Springer Series in
     Astrostatistics, 2015 Andreon, and Weaver

     And

     Bayesian Models for Astrophysical Data: Using R, JAGS, Python,
     and Stan, Cambridge University press, Hilbe, de Souza and Ishida,
     2017

Good point; these are all now cited [WHERE].

     2. Confidence versus credible intervals

     I am sure the authors are aware of the difference between both,
     and should know they don't even agree beyond simple cases: such
     as for normally distributed data. Hence the quote "Because they
     are so afraid of being confused with frequentists, Bayesians
     often call these regions "credible" rather than "confidence""

     Not only seems inappropriate for a scientific paper, but it is
     actually misleading. See for instance The fallacy of placing
     confidence in confidence intervals, Morey et al. 2015
     https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4742505/

     Moreover, such discussion -- Bayesian vs Frequentist -- has far
     deeper roots. See e.g. Probability theory, the logic of science,
     E.T. Jaynes, Cambridge University Press 2003.

Actually, our problem is with the *names*. We agree with the referee
that these two things are importantly and substantially different! We
have fixed that all up in the text now [DO THIS].

     3. Lack of discussion about prediction vs credible intervals.

[UHHH SUP?]

     4. Model check

     Lack of contemporary discussion such as e.g. Gabry et
     al. Visualization in Bayesian workflow,
     https://arxiv.org/abs/1709.01449

Ooh that's a great reference! We cite this and were pleased to be
pointed at this.

     5. MORE SOPHISTICATED SAMPLING METHODS

     Please consider adding some material about

     Approximate Bayesian Computation, since it has gained space in astronomy e.g.

     https://arxiv.org/abs/1608.07606
     https://arxiv.org/abs/1607.01782 
     https://arxiv.org/abs/1504.06129 
     https://arxiv.org/abs/1202.1426

Good point -- we are even working on this ourselves. We added a small
bit of discussion about this [WHERE].

     ---------------------------------------------------------------------- 
     Statistics Editor Comments:

     As a didactic presentation, I suggest two additions:

     (a) Add a table listing MCMC approaches so the reader appreciates
     the variety of options available
     (e.g. https://www.rdocumentation.org/packages/LaplacesDemon/versions/16.0.1/topics/LaplacesDemon)

     (b) Discuss how to choose when MCMC sampling and when
     optimization (e.g. EM Algorithm or simplex) is appropriate. The
     former is commonly used for characterizing Bayesian posteriors
     while the latter are commonly used for maximizing likelihoods.

### 2018-01-21 mcmc.tex
no todos
commentstrip
flatten for submission
one change done

@@
 % Copyright 2013, 2014, 2015, 2016, 2017 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).
-
-% to-do
-% -----
-% - Go through github issues, if any.
-% - Post on arXiv
-% - Submit to ApJ special issue
-% - Blog about it
 
@@
 \documentclass[modern]{aastex61}
-\include{styles/dar}
+\include{dar}
 
@@
 be about the mathematical properties or structure of the method or
-methods.
+methods.\note{For a description of various different sampling methods,
+  given from the point of view of an astrophysicist, see the review by
+  \citet{sharma}. We say a bit about different \MCMC\ methods in
+  \sectionname~\ref{sec:methods}, but our discussion is limited
+  relative to Sharma's review.}
 It will be about how to use \MCMC, including diagnosis and
@@
 
-%Of course, the autocorrelation-time definition
-%(\ref{eq:autocorrelationtime}) does not by itself provide an
-%\emph{estimator} for the time. Estimating the auto-correlation time is
-%notoriously difficult: It is a two-point statistic, and two-point statistics
-%are much harder to estimate than one-point statistics.\note{See: All of cosmology!}
-%If you are not a hard-core user of \MCMC, and if all you want is heuristic indicators
-%of convergence, then what you should take from this section is that the auto-correlation
-%time is involved in variance estimates through \equationname~(\ref{eq:tauvar}) but that
-%it is hard to estimate.
-
-%If you want to think about autocorrelation estimation:\note{One of us (DFM) has written
-%  much more about autocorrelation estimation in the \project{emcee} documentation
-%  at \url{http://dfm.io/posts/autocorr}.}
-%You might first try the na\"ive estimator
-%\begin{eqnarray}\label{eq:tauestnaive}
-%\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
-%\end{eqnarray}
-%where the sum is over the autocorrelation function estimated based the chain
-%using \equationname~(\ref{eq:autocorrelationfn}).
-%In practice, this is a high variance estimator of the integrated
-%autocorrelation time and this variance doesn't go to zero for an infinitely
-%long chain.\note{%
-%  A.~Sokal has a clear discussion of this topic and several proposals for
-%  robust autocorrelation time estimation in a set of lecture notes
-%  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
-%Instead, we can get a much more robust estimate of the autocorrelation time by
-%restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
-%where
-%$\tau \le M \ll N$
-%\begin{eqnarray}\label{eq:tauest}
-%\tau &=& 1 + 2\,\sum_{\Delta = 1}^{M} C(\Delta)
-%\end{eqnarray}
-%In practice, $M$ can be chosen iteratively to satisfy these constraints but it
-%will only be as reliable as the chain itself.
-%An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
-%trusted if the chain is many times ($\sim 100$) longer than $\tau$.\footnote{This
-%  point is also emphasized at the aforementioned tutorial
-%  \url{http://dfm.io/posts/autocorr}, where
-%  a number like 50 is recommended.}
-
 Besides estimating the integrated autocorrelation time, another simple and
@@
 than an autocorrelation time.
-%It can also be used to assess the accuracy of the chain for performing
-%integrals---by replacing the mean of the parameter with whatever
-%integral is of most interest.
 
@@
   or something more complex.
-The mean-of-sampling value $\mean{q}$ is just
+The mean-of-sampling value $\mean{g}$ is just
 \begin{eqnarray}
-\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K g(\pars_k)
+\mean{g} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K g(\pars_k)
 \quad ,
@@
 
-%% ...Finally, to repeat: You can't sample the likelihood!  You can only
-%% sample a posterior pdf!
-
 \section{More sophisticated sampling methods}\label{sec:methods}
@@
 There is no sense in which this list is comprehensive; these are just
-methods (or method classes) that are likely to be useful to projects similar
-to projects we ourselves have executed.
+methods (or method classes) that are likely to be useful to projects
+similar to projects we ourselves have executed.\note{There is a
+  more in-depth and more comprehensive list of astrophyically relevant
+  \MCMC\ methods in the review by \citet{sharma}.}
 
@@
 
-%\clearpage
 \acknowledgements
@@
   \doi{10.3847/1538-4357/aa5e50}.
+\bibitem[Sharma(2017)]{sharma}
+  Sharma, S., 2017,
+  \href{http://adsabs.harvard.edu/abs/2017ARA&A..55..213S}{Markov Chain Monte Carlo methods for Bayesian data analysis in astronomy},
+  \textit{Ann.\ Rev.\ Astron.\ Astrophys.} \textbf{55} 213
+  \doi{10.1146/annurev-astro-082214-122339}.
 \bibitem[Skilling(2006)]{skilling}

### 2018-01-21 response_to_referee.txt
one change done
tweak

@@
 the relevant context, though we wouldn't object to modifying the title
-slightly.
+slightly. We have not made a change here.
 
@@
 
-Yes, we know this paper. We now cite it [WHERE].
+Yes, we know this paper. We now cite it near the end of the first
+Section, and near the beginning of Section 10. We note in both places
+that the Sharma review goes into much more depth on methods than we
+do; our goals here are far more introductory.
 
@@
 
-We also now cite some of this literature [WHERE; DO THIS].
+We also now cite this too [WHERE; DO THIS].
 

### 2018-01-21 response_to_referee.txt
tweak
credible issue

@@
 Actually, our problem is with the *names*. We agree with the referee
-that these two things are importantly and substantially different! We
-have fixed that all up in the text now [DO THIS].
+that these two things are importantly and substantially different!
+They have totally different meanings, of course. We have fixed that
+all up in the text now, making it clear that these are not the same
+thing at ALL. We changed both the footnote (and extended it to make
+the point) and also the text that is so footnoted.
 

### 2018-01-21 mcmc.tex
credible issue

@@
 Again, to form these limits, $g(\pars)$ must be a scalar function (so sorting is well defined).
-Because they are so afraid of being confused with frequentists,
-Bayesians often call these regions ``credible'' rather than
-``confidence''\note{We don't like this ``credible'' terminology,
-  though we sometimes adopt it to avoid confusion.}
+Because of the danger of having them confused with frequentist uncertainties,
+Bayesians usually call these regions ``credible'' rather than
+``confidence'' regions.\note{We don't like this terminology, because there is
+  nothing about the words ``confidence'' and ``credible'' that properly
+  distinguish the difference between these very different things.
+  Confidence regions say what regions of parameter space make data like the
+  observed data; credible regions say what regions of parameter space are
+  probable given the data. The latter require priors for their construction
+  and are the intervals produced by \MCMC.}
 

### 2018-01-21 response_to_referee.txt
item 3 done

@@
 
-[UHHH SUP?]
+We interpreted this telegraphic comment to mean: Shouldn't the paper
+say something about the fact that you might want to put a
+probabilistic interval on a prediction for new data (like a new
+observation of a binary star's velocity, given past observations).  We
+considered this out of scope, because we don't have much or any
+content on decision theory, experimental design, or model
+checking. There is lots there, though, and it is deserving of more
+attention.
 
@@
 Ooh that's a great reference! We cite this and were pleased to be
-pointed at this.
+pointed at this. [DO THIS]
 

### 2018-01-22 response_to_referee.txt
issues 4 and 5

@@
 observation of a binary star's velocity, given past observations).  We
-considered this out of scope, because we don't have much or any
-content on decision theory, experimental design, or model
-checking. There is lots there, though, and it is deserving of more
-attention.
+put in a mention in the Model Checking subsection. It is certainly
+deserving of more attention than it gets here.
 
@@
 
-Ooh that's a great reference! We cite this and were pleased to be
-pointed at this. [DO THIS]
+Ooh that's a great reference! We cite this in the relevant place now,
+and were pleased to be pointed at this. Indeed, Gelman has been a
+great public voice in support of taking this broader view of what
+Bayesian inference is.
 
@@
 
-Good point -- we are even working on this ourselves. We added a small
-bit of discussion about this [WHERE].
+While we agree that the ABC developments are exciting -- and some of
+them are being worked on by us!! -- we do consider this out of
+scope. As we now better explain in the paper, we aren't trying to
+cover the whole ground of inference, we are just trying to introduce
+the uninitiated to the method and help them get over the initial
+stumbling blocks. So we have not made a change here.
 
@@
 
+We explicitly consider this out of scope, as we have said both early in
+the paper, and in the Advanced-Methods section. We call out to much more
+comprehensive reviews of methods now, which have these tables and full
+descriptions of the methods.
+
      (b) Discuss how to choose when MCMC sampling and when
@@
      while the latter are commonly used for maximizing likelihoods.
+
+We have addes more detail to our optimization comments. [DO THIS]

### 2018-01-22 mcmc.tex
issues 4 and 5

@@
 is a framework for constructing
-valuable tests of this form, as is posterior predictive checks\note{%
+valuable tests of this form, as is posterior predictive checking.\note{%
   The posterior predictive check is one of the Bayesian equivalents of
   a goodness-of-fit test \citep{ppcheck}.}
+Indeed, making posterior predictions of new data, and testing the
+model by taking completely new data and comparing to those predictions,
+is the strongest test of any model. Each sample from the \MCMC\ chain
+provides a prediction for future data; the distribution over these sample
+predictions is the posterior prediction for new data.
 Importantly, you want to find ways to compare your model to your data
 \emph{in the space of the data}.
-Don't just look at the parameter space of your model!
+Don't just look at the parameter space of your model. Look at the data.
 It is the data that really exist, especially when all models are wrong!
 
-There are model checking methods (like varying the prior and
+There are model checking methods (like varying the prior or assumptions
+in the likelihood function and
 re-running \MCMC) that test the sensitivity of results to assumptions.
@@
 or sensitivity to assumptions are much more likely to lead to insight
-about what changes to make or what new directions to explore.
+about what changes to make or what new directions to explore.\footnote{%
+  There is a growing idea that Bayesian inference---and inference in general---is
+  about much more than inferring parameter values: It is about understanding and
+  elaborating models (for example, \citealt{gabry}).}
 
@@
   \doi{10.1088/0004-637X/795/1/64}.
+\bibitem[Gabry \etal(2017)]{gabry}
+  Gabry, J., Simpson, D., Vehtari, A., Betancourt, M., \& Gelman, A., 2017,
+  {Visualization in Bayesian workflow},
+  \arxiv{1709.01449}.
 \bibitem[Geisser(1993)]{crossval}

### 2018-01-22 mcmc.tex
typo

@@
 \paragraph{Gibbs}
-Neither \MH\ \MCMC\ nor most ensemble methods good at going to very
-  large numbers of parameters.\footnote{See the excellent rant ``Ensemble Methods are Doomed to Fail in High Dimensions'' at \url{http://andrewgelman.com/2017/03/15/ensemble-methods-doomed-fail-high-dimensions/}.}
+Neither \MH\ \MCMC\ nor most ensemble methods are good at going to very
+  large numbers of parameters.\note{See the excellent rant ``Ensemble Methods are Doomed to Fail in High Dimensions'' at \url{http://andrewgelman.com/2017/03/15/ensemble-methods-doomed-fail-high-dimensions/}.}
 Even dozens of parameters can be enough to drive autocorrelation times

### 2018-01-22 mcmc.tex
typos

@@
 Importantly---for the algorithm given above to work correctly---the
-  proposal pdf must satisfy a ``detailed-balance'' condition\footnote{This
+  proposal pdf must satisfy a ``detailed-balance'' condition\note{This
   detailed-balance condition in \equationname~(\ref{eq:detailed-balance})
@@
 or sensitivity to assumptions are much more likely to lead to insight
-about what changes to make or what new directions to explore.\footnote{%
+about what changes to make or what new directions to explore.\note{%
   There is a growing idea that Bayesian inference---and inference in general---is

### 2018-01-22 response_to_referee.txt
tweak
item 1

@@
 
-Good point. We are more explicit about this [WHERE]. The idea is to
-delve deeper here, but on MCMC specifically, but not so much about new
-developments; more like we are trying to specifically emphasize the
-importance of basic good pratices; we say that more clearly now.
+Good point. We are slightly more explicit about this now in the
+introductory section, in the paragraph starting "In what
+follows...". The idea is to provide a getting-started user-manual for
+MCMC conceptually or specifically, but not really review new
+developments or sophisticated methods. It is more like we are trying
+to specifically emphasize the importance of basic good pratices; we
+say that more clearly now.
 
@@
      JAGS, and Stan, by John Kruschke, and many others
-
-We also now cite this too [WHERE; DO THIS].
 
@@
 
-Good point; these are all now cited [WHERE].
+Good point; these are all now cited in the same "In what follows..."
+paragraph, where we explain our difference from the rest of the
+literature.
 
@@
 comprehensive reviews of methods now, which have these tables and full
-descriptions of the methods.
+descriptions of the methods. [MAKE THIS TRUE]
 

### 2018-01-22 mcmc.tex
item 1

@@
   \sectionname~\ref{sec:methods}, but our discussion is limited
-  relative to Sharma's review.}
-It will be about how to use \MCMC, including diagnosis and
-trouble-shooting.
+  relative to Sharma's review. For even more depth, there are relevant
+  book-length treatments of \MCMC; our favorite is \citet{mcmchandbook}.
+  There are also
+  good books on Bayesian data analysis more generally that
+  include discussion of \MCMC\ as a part of that;
+  for example, \citet{kruschke}, and \citet{andreon} and \citet{hilbe}.}
+Think of this \documentname\ as a getting-started document for a new user
+of \MCMC\ methods.
+It concentrates on conceptual points, and questions we get asked
+frequently by users, with a concentration on how an \MCMC\ project is
+diagnosed and tested, and how its outputs are used responsibly.
+Most of what we say will
+be applicable to any \MCMC\ method; in that sense the discussion is general.
+And all in the language---more-or-less---of an astrophysicist.
 
@@
   \citet{mcmchandbook}, but much has been developed and changed even
-  since that was written.}
+  since that was written. For less in-depth but more current---and
+  more astrophysically relevant---coverage, see also the review by \citet{sharma}.}
 The spaces of \MCMC\ methods and \MCMC\ method implementations are both
@@
 methods (or method classes) that are likely to be useful to projects
-similar to projects we ourselves have executed.\note{There is a
-  more in-depth and more comprehensive list of astrophyically relevant
-  \MCMC\ methods in the review by \citet{sharma}.}
+similar to projects we ourselves have executed.
 
@@
 \begin{thebibliography}{}\markright{References}
+\bibitem[Andreon \& Weaver(2015)]{andreon}
+  Andreon, S. \& Weaver, B., 2015,
+  \textit{Bayesian Methods for the Physical Sciences: Learning from Examples in Astronomy and Physics},
+  Springer
+  \isbn{9783319152875}.
 \bibitem[Box \& Draper(1987)]{box}
@@
   \doi{10.2140/camcos.2010.5.65}.
+\bibitem[Hilbe \etal(2017)]{hilbe}
+  Hilbe, J. M., de~Souza, R. S., \& Ishida, E. E. O., 2017,
+  \textit{Bayesian Models for Astrophysical Data: Using R, JAGS, Python, and Stan},
+  Cambridge University Press
+  \isbn{9781107133082}.
 \bibitem[Hogg \etal(2010a)]{fittingaline}
@@
   \arxiv{1401.6128}.
+\bibitem[Kruschke(2015)]{kruschke}
+  Kruschke, J. K., 2015,
+  \textit{Doing Bayesian Data Analysis: A Tutorial with R, JAGS, and Stan}, 2ed,
+  Academic Press
+  \isbn{9780124058880}.
 \bibitem[Lang \& Hogg(2012)]{cometholmes}

### 2018-01-22 response_to_referee.txt
stats editor

@@
 
-Actually, our problem is with the *names*. We agree with the referee
-that these two things are importantly and substantially different!
-They have totally different meanings, of course. We have fixed that
-all up in the text now, making it clear that these are not the same
-thing at ALL. We changed both the footnote (and extended it to make
-the point) and also the text that is so footnoted.
+Actually, our problem is not with the differences -- they are huge and
+real -- it is with the *names*. We agree with the referee that these
+two things are importantly and substantially different!  They have
+totally different meanings, of course. We have fixed that all up in
+the text now, making it clear that these are not the same thing at
+ALL. We changed both the footnote (and extended it to make the point)
+and also the text that is so footnoted.
 
@@
 the paper, and in the Advanced-Methods section. We call out to much more
-comprehensive reviews of methods now, which have these tables and full
-descriptions of the methods. [MAKE THIS TRUE]
+comprehensive reviews of methods, which have these tables and full
+descriptions of the methods.
 
@@
 
-We have addes more detail to our optimization comments. [DO THIS]
+We have added more detail to our optimization comment in the introduction.

### 2018-01-22 mcmc.tex
stats editor

@@
   choose the ``wrong'' tool, because that ``wrong'' tool gives them
-  back something far more useful than the output of any optimizer!}
+  back something far more useful than the output of any optimizer!
+  However it is true that when the data are extremely informative
+  (that is, the posterior is very narrow), there is not much difference
+  between optimizing the posterior and sampling it, nor between optimizing
+  the likelihood and optimizing the posterior. That is, all inference
+  methods sort-of converge as the data become very informative. That can be
+  an important point when you are working on a problem where the posterior
+  is very narrow.}
 This is another bad reason!

### 2018-01-23 mcmc.tex
tweak

@@
   There are also
-  good books on Bayesian data analysis more generally that
+  good books on Bayesian data analysis from a computing perspective more generally that
   include discussion of \MCMC\ as a part of that;
-  for example, \citet{kruschke}, and \citet{andreon} and \citet{hilbe}.}
+  for example, \citet{kruschke} and \citet{andreon} and \citet{hilbe}.}
 Think of this \documentname\ as a getting-started document for a new user
